### 2014-03-17 paper_enahpc14/Introduction.tex
adds .bib file for bibliography entries; starts drafting introduction and COSMO-ART section

The  interaction  between aerosols  and  clouds  presents  one of  the
biggest  uncertainties  in   present  day  climate  simulations.   The
chemistry involved is complex,  but there is a concerted international
effort to model the underlying processes through models. These models,
such as  the Aerosol  Reactive Transport (ART),  developed at  the KIT
(Karlsruhe  Institute of  Technology)  , offer  a  key opportunity  to
reduce   the  climate  uncertainty,   particularly  on   the  regional
scale. These  models can be coupled  with climate models,  such as the
regional  weather  and   climate  COSMO  (Consortium  for  Small-scale
Modelling) model  developed by the German Weather  Service (DWD).  The
COSMO model  is a weather forecast  model which has  become a standard
all  over Europe.   Beside the  federal weather  forecast  stations in
Germany  (Deutscher  Wetterdienst,  DWD),  Switzerland  (MeteoSchweiz,
MCH),  Italy (Ufficio  Generale  Spazio Aereo  e Meteorologia,  USAM),
Greece  (Hellenic   National  Meteorological  Service,   HNMS)  Poland
(Institute  of   Meteorology  and  Water   Management,  IMGW)  Romania
(National  Meteorological  Administration,  NMA) and  Russia  (Federal
Service for Hydrometeorology  and Environmental Monitoring, RHM), also
a   large  number   of  agencies   including  military   and  research
institutions  base their  forecasts on  COSMO. The  combined COSMO-ART
model is capable of simulating  aerosol distributions as well as their
interactions over Europe as well as other regional domains.

It has  been a major  scientific achievement to effectively  model the
underlying  chemical  processes   through  these  applications.   This
achievement  now   needs  to  be  fostered  by   reducing  its  energy
expense. Currently, resource allocation  is one of the key limitations
to scientists  in running the simulation periods  and resolutions they
would like.

### 2014-03-17 paper_enahpc14/Sec1:COSMO-ART.tex
adds .bib file for bibliography entries; starts drafting introduction and COSMO-ART section

This   study   uses   the   regional   atmospheric   model   COSMO-ART
\citep{Vogel-2009} (ART stands for Aerosols and Reactive Trace gases),
in which atmospheric chemistry is online-coupled to the COSMO regional
numerical      weather      prediction      and     climate      model
\citep{Baldauf-2011}. The online coupling  is realized in a consistent
way using the same spatial and temporal resolution for all scalars, as
for    example   advection,   the    same   numerical    schemes   are
applied. Detailed model description can be found in the aforementioned
publications as well as in \citep{Stanelle-2010}, \citep{Bangert-2012}
and \citep{Knote-2011}.

The gaseous chemistry in COSMO-ART  is simulated by a modified version
of  the Regional Acid  Deposition Model,  Version 2  (RADM2) mechanism
\citep{Stockwell-1990},  which  is   extended  to  describe  secondary
organic  aerosol formation  \citep{Schell-2001}  and hydroxyl  radical
recycling  due  to  isoprene chemistry  \citep{Geiger-2003}.   Aerosol
dynamics   are  simulated   with   the  modal   aerosol  module   MADE
\citep{Ackermann-1998}, improved  by explicit treatment  of soot aging
through  condensation  of  inorganic \citep{Riemer-2003}  and  organic
substances.   The five  modes  that represent  the aerosol  population
contain:  pure   soot,  secondary  mixtures   of  sulfates,  nitrates,
ammonium, organics  and water (nucleation and  accumulation mode), and
the  internal mixtures  of all  these species  in both  modes. Aerosol
species are transferred among  these mixtures through condensation and
coagulation.   A  separate  coarse  mode contains  additional  primary
emitted  particles.    Meteorologically-affected  emissions  are  also
online-coupled to the model  system. Sea-salt and desert dust emission
algorithms are described in \citep{Vogel-2006} and \citep{Vogel-2009}.
The biogenic VOC emissions are calculated as functions of the land use
type    based    on   the    Global    Land    Cover   2000    dataset
\citep{Bartholome-2005}  and the  modeled  temperatures and  radiative
fluxes  \citep{Vogel-1995}.   Sedimentation,  advection,  washout  and
turbulent diffusion also modify the aerosol distributions.

### 2014-03-17 paper_enahpc14/draft.tex
adds .bib file for bibliography entries; starts drafting introduction and COSMO-ART section

%%%%%%%%%%%%%%%%%%%%%%% file draft.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is a general template file for the LaTeX package SVJour3
% for Springer journals.          Springer Heidelberg 2010/09/16
%
% Copy it to a new file with a new name and use it as the basis
% for your article. Delete % signs as needed.
%
% This template includes a few options for different layouts and
% content for various journals. Please consult a previous issue of
% your journal as needed.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% First comes an example EPS file -- just ignore it and proceed on the
% \documentclass line your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
%\documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
%\documentclass[smallextended]{svjour3}      % onecolumn (second format)
\documentclass[twocolumn]{svjour3}           % twocolumn
%
\smartqed % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage[round, sort, numbers, authoryear]{natbib}
%
% please place your own definitions here and don't use \def but
% \newcommand{}{}
%
% Insert the name of "your journal" with
% \journalname{myjournal}
%
\begin{document}

\title{Evaluating the performance  and energy efficiency of COSMO-ART,
  a fully online coupled model  system composed of a numerical weather
  forecast model  and a  chemical transport model.   \thanks{Grants or
    other notes  about the  article that should  go on the  front page
    should be placed here.}}
%\subtitle{Write here subtitle}

%\titlerunning{Short form of title} % if too long for running head

\author{J.Charles \and M.Dolz}

%\authorrunning{Short form of author list} % if too long for running
%head

\institute{Dr. Joseph Charles \at Swiss National Supercomputing Centre
  (CSCS)\\ CH-6900  Lugano, Switzerland \\  Tel.: +41 (0) 91  610 8216
  \\  \email{joseph.charles@cscs.ch}  \\  \and  Dr.  Manuel  Dolz  \at
  University of Hamburg \\ 20148  Hamburg, Germany \\ Tel.: +49 (0) 40
  460094-404 \\ \email{manuel.dolz@informatik.uni-hamburg.de}}

\date{Received: date / Accepted: date} % The correct dates will be
                                       % entered by the editor

\maketitle

\begin{abstract}
  In this paper we present  COSMO-ART, an extension of the operational
  weather  forecast  model  of   the  German  Weather  Service  (DWD),
  developed for  the evaluation of the interactions  of reactive gases
  and aerosol particles  with the state of atmosphere  at the regional
  scale. It  includes secondary aerosols,  directly emitted components
  like  soot,  mineral  dust,  sea  salt and  biological  material  as
  pollen. Processes such  as emissions, coagulation, condensation, dry
  deposition,  wet removal,  and sedimentation  of aerosols  are taken
  into account.   The overall performance  of this application  on HPC
  systems is  analysed by a  profiling and tracing study  to determine
  hotspots, i.e.  the  parts of the program that  require the dominant
  fraction of runtime, and to identify critical paths, i.e.  sequences
  of functions that shows dominant inclusive contributions for all its
  elements. Moreover, we describe measurement devices and energy-aware
  techniques  employed  to  evaluate   the  energy  footprint  of  the
  considered  application and  to  get detailed  insights about  power
  bottlenecks.  Our approach is to improve corresponding code sections
  to  sustain high  performance  while minimizing  energy-to-solution.
  This preliminary  study sets the basis of  broader considerations to
  tackle challenges  related to energy efficient HPC  in the framework
  of the Exa2Green project (\url{http://exa2green.eu/}).

\keywords{High performance computing  \and Energy-aware computing \and
  Numerical Weather Prediction}
% \PACS{PACS code1 \and PACS code2 \and more}
% \subclass{MSC code1 \and MSC code2 \and more}
\end{abstract}

\section{Introduction}
\label{intro}
\input{Introduction.tex}

\section{COSMO-ART model description}
\label{sec:1}
\input{Sec1:COSMO-ART.tex}

\subsection{Subsection title}
\label{subsec:1}
Don't forget to  give each section and subsection  a unique label (see
Sect.~\ref{sec:1}).
\paragraph{Paragraph headings} Use paragraph headings as needed.
\begin{equation}
a^2+b^2=c^2
\end{equation}

% For one-column wide figures use
\begin{figure}
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics{example.eps}
% figure caption is below the figure
\caption{Please write your figure caption here}
\label{fig:1}       % Give a unique label
\end{figure}
%
% For two-column wide figures use
\begin{figure*}
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[width=0.75\textwidth]{example.eps}
% figure caption is below the figure
\caption{Please write your figure caption here}
\label{fig:2}       % Give a unique label
\end{figure*}
%
% For tables use
\begin{table}
% table caption is above the table
\caption{Please write your table caption here}
\label{tab:1}       % Give a unique label
% For LaTeX tables use
\begin{tabular}{lll}
\hline\noalign{\smallskip}
first & second & third  \\
\noalign{\smallskip}\hline\noalign{\smallskip}
number & number & number \\
number & number & number \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

\begin{acknowledgements}
The  authors  would  like  to  thank the  High  Performance  and  High
Productivity  Computing  (\url{www.hp2c.ch})  Initiative and  the  FP7
funding this project.)
\end{acknowledgements}

%\bibliographystyle{spbasic}  % basic style, author-year citations
%\bibliographystyle{spphys}   % APS-like style for physics
%\bibliographystyle{spmpsci}  % mathematics and physical sciences
\bibliographystyle{unsrtnat}

\bibliography{draft}  % name your BibTeX data base

\end{document}

### 2014-03-17 paper_enahpc14/compile.sh

@@ -1,6 +1,9 @@
 #/bin/bash -l
 ./clean.sh
-latex template.tex
-dvips -o template.ps template.dvi
-ps2pdf -sPAPERSIZE=a4 template.ps
-acroread template.pdf
+latex draft.tex
+bibtex draft.aux
+latex draft.tex
+latex draft.tex
+dvips -o draft.ps draft.dvi
+ps2pdf -sPAPERSIZE=a4 draft.ps
+acroread draft.pdf

### 2014-03-17 trunk/paper_enahpc14/clean.sh
adds .bib file for bibliography entries; starts drafting introduction and COSMO-ART section

@@ -1,2 +1,2 @@
 #/bin/bash -l
-rm -vf *~ *.aux *.eps *.pdf *.log *.ps *.dvi *.out
+rm -vf *~ *.aux *.eps *.pdf *.log *.ps *.dvi *.out *.bbl *.blg

### 2014-03-25 trunk/paper_enahpc14/draft.tex
updates bibliography and introduction

--- 2014-03-17 paper_enahpc14/draft.tex
@@
-%%%%%%%%%%%%%%%%%%%%%%% file draft.tex %%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %
@@
 %
+\DeclareRobustCommand\IPCClongname{}
+%
 % please place your own definitions here and don't use \def but
@@
 
-\author{J.Charles \and M.Dolz}
+\author{J.Charles \and W.Sawyer \and M.Dolz \and C.Malossi}
 
@@
 \institute{Dr. Joseph Charles \at Swiss National Supercomputing Centre
-  (CSCS)\\ CH-6900  Lugano, Switzerland \\  Tel.: +41 (0) 91  610 8216
-  \\  \email{joseph.charles@cscs.ch}  \\  \and  Dr.  Manuel  Dolz  \at
-  University of Hamburg \\ 20148  Hamburg, Germany \\ Tel.: +49 (0) 40
-  460094-404 \\ \email{manuel.dolz@informatik.uni-hamburg.de}}
+  (CSCS)\\ CH-6900  Lugano, Switzerland  \\ Tel: +41  (0) 91  610 8216
+  \\ \email{joseph.charles@cscs.ch} \and Dr.  William Sawyer \at Swiss
+  National Supercomputing Centre  (CSCS)\\ CH-6900 Lugano, Switzerland
+  \\ Tel:  +41 (0) 91 610 8229  \\ \email{william.sawyer@cscs.ch} \and
+  Dr.  Manuel Dolz \at University of Hamburg \\ 20148 Hamburg, Germany
+  \\         Tel:         +49         (0)        40         460094-404
+  \\ \email{manuel.dolz@informatik.uni-hamburg.de}  \and Dr. Cristiano
+  Malossi  \at  IBM  Zurich  Research Laboratory  \\  CH-8000  Zurich,
+  Switzerland     \\     Tel:      +41     (0)     44     724     8616
+  \\ \email{ACM@zurich.ibm.com}}
 
@@
   considered  application and  to  get detailed  insights about  power
-  bottlenecks.  Our approach is to improve corresponding code sections
-  to  sustain high  performance  while minimizing  energy-to-solution.
-  This preliminary  study sets the basis of  broader considerations to
-  tackle challenges  related to energy efficient HPC  in the framework
-  of the Exa2Green project (\url{http://exa2green.eu/}).
+  bottlenecks.   Our  motivation  is  to  improve  corresponding  code
+  sections    to   sustain    high   performance    while   minimizing
+  energy-to-solution.   This  preliminary  study  sets  the  basis  of
+  broader  considerations  to  tackle  challenges  related  to  energy
+  efficient   HPC  in   the   framework  of   the  Exa2Green   project
+  (\url{http://exa2green.eu/}).
 
 \keywords{High performance computing  \and Energy-aware computing \and
-  Numerical Weather Prediction}
+  Green computing  \and Numerical Weather  Prediction \and Atmospheric
+  Chemistry \and Aerosol Modelling \and COSMO-ART}
 % \PACS{PACS code1 \and PACS code2 \and more}
@@
 \begin{acknowledgements}
-The  authors  would  like  to  thank the  High  Performance  and  High
-Productivity  Computing  (\url{www.hp2c.ch})  Initiative and  the  FP7
-funding this project.)
+This  research  was  supported   by  the  Exa2Green  research  project
+co-funded  under the  EU  7th Framework  Program  Future and  Emerging
+Technologies (FET) Proactive Initiative: Minimising Energy Consumption
+of Computing to  the Limit (MINECC).  The authors  would like to thank
+the    High    Performance    and    High    Productivity    Computing
+(\url{www.hp2c.ch}) Initiative.)
 \end{acknowledgements}
 
-%\bibliographystyle{spbasic}  % basic style, author-year citations
-%\bibliographystyle{spphys}   % APS-like style for physics
-%\bibliographystyle{spmpsci}  % mathematics and physical sciences
-\bibliographystyle{unsrtnat}
-
-\bibliography{draft}  % name your BibTeX data base
+\DeclareRobustCommand\IPCClongname{ Intergovernmental Panel on Climate Change}
+\bibliographystyle{plainnat}
+\bibliography{\jobname}
 

### 2014-03-25 paper_enahpc14/Introduction.tex
updates bibliography and introduction

@@
-The  interaction  between aerosols  and  clouds  presents  one of  the
-biggest  uncertainties  in   present  day  climate  simulations.   The
-chemistry involved is complex,  but there is a concerted international
-effort to model the underlying processes through models. These models,
-such as  the Aerosol  Reactive Transport (ART),  developed at  the KIT
-(Karlsruhe  Institute of  Technology)  , offer  a  key opportunity  to
-reduce   the  climate  uncertainty,   particularly  on   the  regional
-scale. These  models can be coupled  with climate models,  such as the
-regional  weather  and   climate  COSMO  (Consortium  for  Small-scale
-Modelling) model  developed by the German Weather  Service (DWD).  The
-COSMO model  is a weather forecast  model which has  become a standard
-all  over Europe.   Beside the  federal weather  forecast  stations in
-Germany  (Deutscher  Wetterdienst,  DWD),  Switzerland  (MeteoSchweiz,
-MCH),  Italy (Ufficio  Generale  Spazio Aereo  e Meteorologia,  USAM),
-Greece  (Hellenic   National  Meteorological  Service,   HNMS)  Poland
-(Institute  of   Meteorology  and  Water   Management,  IMGW)  Romania
-(National  Meteorological  Administration,  NMA) and  Russia  (Federal
-Service for Hydrometeorology  and Environmental Monitoring, RHM), also
-a   large  number   of  agencies   including  military   and  research
-institutions  base their  forecasts on  COSMO. The  combined COSMO-ART
-model is capable of simulating  aerosol distributions as well as their
-interactions over Europe as well as other regional domains.
+Comprehensive  scientific assessments  of climate-warming  trends over
+the  past 50  years have  shown  that the  state and  dynamics of  the
+Earth's  climate system  have undergone  unprecedented  major changes.
+The  last report  of the  Intergovernmental Panel  on  Climate Change,
+released  in  September  2013  \citep{IPCC}, states  with  95  percent
+confidence that the Earth  climate system is unequivocally warming, as
+evidenced  by  increases  in   global  average  atmosphere  and  ocean
+temperatures,  widespread melting  of  snow and  ice,  and rising  sea
+levels.   Besides it  confirms the  overwhelming  scientific consensus
+that  it's extremely  likely that  increasing emissions  of greenhouse
+gases  from   anthropogenic  activities,  such   as  land-use  change,
+transportation, burning fossil fuels, are the primary driver.
 
-It has  been a major  scientific achievement to effectively  model the
-underlying  chemical  processes   through  these  applications.   This
-achievement  now   needs  to  be  fostered  by   reducing  its  energy
-expense. Currently, resource allocation  is one of the key limitations
-to scientists  in running the simulation periods  and resolutions they
-would like.
+Greenhouse gases  have long been  studied because of their  ability to
+trap heat  in the atmosphere  and their significant warming  effect on
+surface  and  lower  atmosphere  temperatures.   Some  of  them  occur
+naturally  in the  atmosphere, such  as water  vapor,  carbon dioxide,
+methane,  nitrous  oxide  and  ozone,  while  others  are  exclusively
+human-induced, such as fluorinated  gases with a global warming effect
+up to  23000 times greater than  carbon dioxide. Some,  such as carbon
+dioxode  or nitrous  oxide,  are  long-lived as  they  persist in  the
+atmosphere over time scales of  a decade to thousands of years, become
+homogeneously well-mixed in the troposphere, and thus have a long-term
+influence on climate change.
+
+The Earth's average surface  temperature is regulated by the transfert
+of incoming  solar shortwave  absorbed radiation and  outgoing thermal
+longwave infrared radiation emitted  back into Space. Many factors can
+alter  Earth's radiation  balance  such  as the  type  of surface  the
+sunlight encounters  (i.e.  white glacier, dark  desert soil, forests,
+grasslands, ocean surfaces, ice  caps, cities), the intensity of solar
+energy,  the cloud  cover, the  reflectivity of  clouds or  gases, the
+absorption by various greenhouse gases or surfaces and the emission of
+heat by various materials.
+
+Numerous  studies over  the last  decades have  demonstrated  that the
+enhancement  of aerosol  concentration since  industrialization  has a
+profound impact  on the climate  system.  Aerosols are  tiny particles
+suspended in  the air,  including sea salt,  mineral dust,  ash, soot,
+sulphate,   nitrate,  organic  carbon   and  black   carbon.  Although
+atmospheric aerosols are not considered a heat-trapping greenhouse gas
+and  have shorter  atmospheric  lifetimes (around  10  days), they  do
+affect the  Earth's radiation balance  and have short-term  warming or
+cooling  influences  of the  earth-atmosphere  system.  Aerosols  that
+mainly scatter solar radiation have a cooling effect, by enhancing the
+total  reflected  solar  radiation  from  the  Earth,  while  strongly
+absorbing  aerosols  have  a  warming  effect.   Indirectly,  enhanced
+aerosol  concentrations  can boost  cloud  formation  and  lead to  an
+increase in their albedo, providing a net cooling force.

### 2014-03-26 trunk/paper_enahpc14/clean.sh
renames source files; creates Makefiles for main directory and the newly created Figs/ subdirectory; deletes bash scripts

@@ -1,2 +1,0 @@
-#/bin/bash -l
-rm -vf *~ *.aux *.eps *.pdf *.log *.ps *.dvi *.out *.bbl *.blg

### 2014-03-26 paper_enahpc14/compile.sh
renames source files; creates Makefiles for main directory and the newly created Figs/ subdirectory; deletes bash scripts

@@ -1,9 +1,0 @@
-#/bin/bash -l
-./clean.sh
-latex draft.tex
-bibtex draft.aux
-latex draft.tex
-latex draft.tex
-dvips -o draft.ps draft.dvi
-ps2pdf -sPAPERSIZE=a4 draft.ps
-acroread draft.pdf

### 2014-03-30 trunk/paper_enahpc14/paper.tex
updates introduction and bibliography

--- 2014-03-25 trunk/paper_enahpc14/draft.tex
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-% First comes an example EPS file -- just ignore it and proceed on the
-% \documentclass line your LaTeX will extract the file if required
-\begin{filecontents*}{example.eps}
-%!PS-Adobe-3.0 EPSF-3.0
-%%BoundingBox: 19 19 221 221
-%%CreationDate: Mon Sep 29 1997
-%%Creator: programmed by hand (JK)
-%%EndComments
-gsave
-newpath
-  20 20 moveto
-  20 220 lineto
-  220 220 lineto
-  220 20 lineto
-closepath
-2 setlinewidth
-gsave
-  .4 setgray fill
-grestore
-stroke
-grestore
-\end{filecontents*}
 %
@@
 %
-%\documentclass{svjour3}                     % onecolumn (standard format)
-%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
-%\documentclass[smallextended]{svjour3}      % onecolumn (second format)
 \documentclass[twocolumn]{svjour3}           % twocolumn
@@
 \usepackage{hyperref}
-\usepackage[round, sort, numbers, authoryear]{natbib}
+\usepackage[round, sort&compress, numbers]{natbib}
 %
@@
   systems is  analysed by a  profiling and tracing study  to determine
-  hotspots, i.e.  the  parts of the program that  require the dominant
-  fraction of runtime, and to identify critical paths, i.e.  sequences
-  of functions that shows dominant inclusive contributions for all its
-  elements. Moreover, we describe measurement devices and energy-aware
-  techniques  employed  to  evaluate   the  energy  footprint  of  the
-  considered  application and  to  get detailed  insights about  power
-  bottlenecks.   Our  motivation  is  to  improve  corresponding  code
-  sections    to   sustain    high   performance    while   minimizing
-  energy-to-solution.   This  preliminary  study  sets  the  basis  of
-  broader  considerations  to  tackle  challenges  related  to  energy
-  efficient   HPC  in   the   framework  of   the  Exa2Green   project
-  (\url{http://exa2green.eu/}).
+  hotspots  and  identify   critical  paths.   Moreover,  we  describe
+  measurement devices and energy-aware techniques employed to evaluate
+  the  energy  footprint of  the  considered  application  and to  get
+  detailed  insights about  power bottlenecks.   Our motivation  is to
+  improve  corresponding  code sections  to  sustain high  performance
+  while  minimizing energy-to-solution.   This preliminary  study sets
+  the basis of broader  considerations to tackle challenges related to
+  energy efficient high performance  computing in the framework of the
+  Exa2Green project (\url{http://exa2green.eu/}).
 
 \keywords{High performance computing  \and Energy-aware computing \and
-  Green computing  \and Numerical Weather  Prediction \and Atmospheric
-  Chemistry \and Aerosol Modelling \and COSMO-ART}
+  Green computing  \and Numerical weather  prediction \and Atmospheric
+  chemistry  \and  Aerosols  modelling  \and  Profiling  methods  \and
+  Benchmark analysis \and COSMO-ART coupled model}
 % \PACS{PACS code1 \and PACS code2 \and more}
@@
 
-\section{COSMO-ART model description}
+\section{COSMO coupled with ART extension}
 \label{sec:1}
+\subsection{Model description and set-up}
+\label{subsec:1}
 \input{Sec1:COSMO-ART.tex}
 
-\subsection{Subsection title}
-\label{subsec:1}
-Don't forget to  give each section and subsection  a unique label (see
-Sect.~\ref{sec:1}).
-\paragraph{Paragraph headings} Use paragraph headings as needed.
-\begin{equation}
-a^2+b^2=c^2
-\end{equation}
+\section{Hardware description}
+\label{sec:2}
 
-% For one-column wide figures use
-\begin{figure}
-% Use the relevant command to insert your figure file.
-% For example, with the graphicx package use
-  \includegraphics{example.eps}
-% figure caption is below the figure
-\caption{Please write your figure caption here}
-\label{fig:1}       % Give a unique label
-\end{figure}
-%
-% For two-column wide figures use
-\begin{figure*}
-% Use the relevant command to insert your figure file.
-% For example, with the graphicx package use
-  \includegraphics[width=0.75\textwidth]{example.eps}
-% figure caption is below the figure
-\caption{Please write your figure caption here}
-\label{fig:2}       % Give a unique label
-\end{figure*}
-%
-% For tables use
-\begin{table}
-% table caption is above the table
-\caption{Please write your table caption here}
-\label{tab:1}       % Give a unique label
-% For LaTeX tables use
-\begin{tabular}{lll}
-\hline\noalign{\smallskip}
-first & second & third  \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-number & number & number \\
-number & number & number \\
-\noalign{\smallskip}\hline
-\end{tabular}
-\end{table}
+\section{Power measurement systems}
+\label{sec:3}
+
+\section{Benchmark results}
+\label{sec:4}
+
+%\paragraph{Paragraph headings} Use paragraph headings as needed.
+
+%\begin{figure*}
+%  \includegraphics[width=0.4\textwidth]{Figs/earth.eps}
+%  \caption{Please write your figure caption here}
+%  \label{fig:1}
+%\end{figure*}
+
+%\begin{table}
+%  \caption{Please write your table caption here}
+%  \label{tab:1}
+%  \begin{tabular}{lll}
+%    \hline\noalign{\smallskip}
+%    first & second & third  \\
+%    \noalign{\smallskip}\hline\noalign{\smallskip}
+%    number & number & number \\
+%    number & number & number \\
+%    \noalign{\smallskip}\hline
+%  \end{tabular}
+%\end{table}
 
@@
 of Computing to  the Limit (MINECC).  The authors  would like to thank
-the    High    Performance    and    High    Productivity    Computing
-(\url{www.hp2c.ch}) Initiative.)
+the  High  Performance  and  High  Productivity  Computing  Initiative
+(\url{www.hp2c.ch}).)
 \end{acknowledgements}
 
-\DeclareRobustCommand\IPCClongname{ Intergovernmental Panel on Climate Change}
+\DeclareRobustCommand\IPCClongname{ - Intergovernmental Panel on Climate Change}
+
 \bibliographystyle{plainnat}

### 2014-03-30 paper_enahpc14/Introduction.tex
updates introduction and bibliography

--- 2014-03-17 paper_enahpc14/Introduction.tex
@@
-The  interaction  between aerosols  and  clouds  presents  one of  the
-biggest  uncertainties  in   present  day  climate  simulations.   The
-chemistry involved is complex,  but there is a concerted international
-effort to model the underlying processes through models. These models,
-such as  the Aerosol  Reactive Transport (ART),  developed at  the KIT
-(Karlsruhe  Institute of  Technology)  , offer  a  key opportunity  to
+Comprehensive  scientific assessments  of climate-warming  trends over
+the  past 50  years have  shown  that the  state and  dynamics of  the
+Earth's climate system have  undergone unprecedented major changes, as
+evidenced  by  increases  in   global  average  atmosphere  and  ocean
+temperatures,  widespread melting  of  snow and  ice,  and rising  sea
+levels.   The last report  of the  Intergovernmental Panel  on Climate
+Change, released  in September 2013  \citep{IPCC-2013}, concluded that
+most of the observed temperature  increases since the mid 20th Century
+have  been caused  by  increasing concentrations  of greenhouse  gases
+resulting from  anthropogenic activities  such as fossil  fuel burning
+and  deforestation.  Long-lived greenhouse  gases, for  example carbon
+dioxide, methane  and nitrous oxide, are chemically  stable and linger
+homogeneously well-mixed and over time scales of a decade to thousands
+of years in  the atmosphere; whereas short-lived gases  such as water,
+suphur  dioxide  and  carbon  monoxide  which  respond  physically  or
+chemically  to changes  in temperature,  act primarily  as  a feedback
+mechanism.   Some   extremly  powerful  greenhouse   gases,  such  as
+fluorinated gases can even have  a warming effect on the atmosphere up
+to  23000  times  greater  than  carbon  dioxide  \citep{Zehner-2012}.
+Fortunately  the  European  Parliament,  which  has  gained  worldwide
+recognition as a  leader in climate policy, has  passed legislation in
+March 2014  to phase  out fluorinated gas  emissions by  two-thirds by
+2030,  towards  the  internationally  agreed goal  of  keeping  global
+warming  below  2  degrees  Celsius  compared to  the  temperature  in
+pre-industrial times.\\
+
+\noindent
+Current  cumulative releases of  human-induced gas  emissions enhances
+the greenhouse  effect to  the Earth's atmosphere  and thus  creates a
+large imbalance between incoming solar absorbed radiation and outgoing
+longwave radiation emitted back  to space, through radiative trapping.
+While the  Earth's temperature  is dependent upon  the greenhouse-like
+action of  the atmosphere, the  Earth's radiation balance  is strongly
+influenced by several  other factors such as the  type of surface that
+sunlight first  encounters.  Forests, grasslands,  ocean surfaces, ice
+caps, deserts,  and cities all absorb, reflect,  and radiate radiation
+differently.   Sunlight falling  on a  white glacier  surface strongly
+reflects back into space, resulting  in minimal heating of the surface
+and  lower atmosphere.   Sunlight falling  on  a dark  desert soil  is
+strongly absorbed,  on the other hand, and  contributes to significant
+heating of the surface and lower atmosphere.  Cloud cover also affects
+greenhouse  warming by  both reducing  the amount  of  solar radiation
+reaching the earth's  surface and by reducing the  amount of radiation
+energy emitted into space.\\
+
+\noindent 
+Aerosols are  suspensions of liquid,  solid or mixed particles  in the
+air  (including sea  salt,  mineral dust,  sulphate, nitrate,  organic
+carbon and  black carbon),  with highly variable  chemical composition
+and  size  distribution \citep{Putaud-2010}.   Although  they are  not
+considered a heat-trapping greenhouse gas and have shorter atmospheric
+lifetimes, they significantly affect  the Earth's radiation budget and
+are acknowledged as one of  the most significant and uncertain aspects
+in  anthropogenic forcing  over the  last 150  years \citep{Koch-2009,
+IPCC-2013}.  Enhanced  aerosols concentrations can  impact the climate
+system by  reflecting (e.g.  pure sulfates and  nitrates) or absorbing
+(e.g.  black  carbon) solar radiation  and thereby exert a  cooling or
+warming  effect on  the Earth-atmosphere  system, causing  a so-called
+direct    radiative   forcing    \citep{Charlson-1991,   Haywood-2000,
+Ramanathan-2001,  Liao-2005, Bangert-2012,  Lundgren-2013}.  Depending
+on their  size and  chemical composition, they  can also act  as cloud
+condensation  and   ice  nuclei   and  profoundly  impact   the  cloud
+microphysical   processes  and   optical   properties  (cloud   albedo
+effect,   \citealp{Twomey-1977}),  leading   to  changes   in  droplet
+concentrations                \citep{Albrecht-1989}                and
+precipitation   \citep{Rosenfeld-2000,   Khain-2008,  Pruppacher-2010,
+Seifert-2012, Tao-2012, Lee-2012},  and therefore producing a negative
+indirect   radiative    forcing   \citep{Haywood-2000,   Lohmann-2005,
+VandenHeever-2011,   Rosenfeld-2013}.    However,   the   effects   of
+anthropogenic aerosols on clouds and the hydrological cycle as well as
+the  cloud   lifetime  effect  are  especially  hard   to  assess  and
+quantify   \citep{IPCC-2013},   and   remain   one  of   the   largest
+uncertainties in climate modeling and in climate change prediction due
+to the lack of  understanding of cloud feedbacks \citep{Sherwood-2013,
+Rosenfeld-2013, Lee-2013}. \\
+
+\noindent 
+Hence to  enhance our understanding of  aerosol-cloud interactions and
+reduce  uncertainties  of aerosol  effects  in  climate, the  research
+community is making a  concerted international effort to represent the
+underlying chemical  processes through models.  These  models, such as
+ART (Aerosols  and Reactive Trace  gases) extension, developed  at the
+Karlsruhe Institute  of Technology (KIT),  offer a key  opportunity to
 reduce   the  climate  uncertainty,   particularly  on   the  regional
-scale. These  models can be coupled  with climate models,  such as the
-regional  weather  and   climate  COSMO  (Consortium  for  Small-scale
-Modelling) model  developed by the German Weather  Service (DWD).  The
-COSMO model  is a weather forecast  model which has  become a standard
-all  over Europe.   Beside the  federal weather  forecast  stations in
-Germany  (Deutscher  Wetterdienst,  DWD),  Switzerland  (MeteoSchweiz,
-MCH),  Italy (Ufficio  Generale  Spazio Aereo  e Meteorologia,  USAM),
-Greece  (Hellenic   National  Meteorological  Service,   HNMS)  Poland
-(Institute  of   Meteorology  and  Water   Management,  IMGW)  Romania
-(National  Meteorological  Administration,  NMA) and  Russia  (Federal
-Service for Hydrometeorology  and Environmental Monitoring, RHM), also
-a   large  number   of  agencies   including  military   and  research
-institutions  base their  forecasts on  COSMO. The  combined COSMO-ART
-model is capable of simulating  aerosol distributions as well as their
-interactions over Europe as well as other regional domains.
+scale \citep{Knote-2011, Bangert-2011,  Knote-2013}.  These models can
+be coupled with climate models,  such as the regional weather forecast
+model COSMO (Consortium  for Small-scale Modelling), jointly developed
+by  a consortium  of  European weather  centers  including the  German
+weather service  DWD and MeteoSwiss,  and used in the  climate version
+COSMO-CLM by a wide  research community.  The extended COSMO-ART model
+provides a detailed description of air pollution chemistry and aerosol
+processes, and  is mainly  designed to study  air quality  and aerosol
+meteorology feedbacks on short, episodic to annual time scales.  It is
+capable  of   simulating  aerosol  distributions  as   well  as  their
+interactions over Europe as well as other regional domains.\\
 
-It has  been a major  scientific achievement to effectively  model the
-underlying  chemical  processes   through  these  applications.   This
-achievement  now   needs  to  be  fostered  by   reducing  its  energy
-expense. Currently, resource allocation  is one of the key limitations
-to scientists  in running the simulation periods  and resolutions they
-would like.
+\noindent 
+COSMO-ART is  computationally much more demanding than  the COSMO core
+since a  large number of additional  tracers and processes  have to be
+considered.  Thus this model is currently severely-limited in terms of
+applicability and expensive in terms of energy consumption.  COSMO has
+recently  been  ported  to  GPUs  within the  framework  of  the  High
+Performance  and  High  Productivity  Computing (HP2C)  Initiative  to
+optimize it  for computational and energy  efficiency.  Although these
+developments will  facilitate the  application of COSMO  for numerical
+weather prediction and climate  simulations, they do little to address
+the  coupling with  the  ART model  extension,  for which  significant
+investments  are still required  to take  it to  a similar  level. The
+efficiency of ART is being  addressed in the EU Exa2Green project.  In
+Sec.~\ref{sec:1},  we   introduce  the  aerosol  model   and  its  its
+integration  into  COSMO.    Sec.~\ref{sec:2}  presents  the  hardware
+considerations.  In  Sec.~\ref{sec:3} we describe  measurement devices
+and energy-aware techniques employed  to evaluate the energy footprint
+of the  considered application.  Finally,  Sec.~\ref{sec:4} presents a
+benchmark  analysis  of  COSMO-ART   from  the  Exa2Green  project  to
+highlight the critical components and an outline of the next steps for
+model optimization.  The ultimate goal  of the project is to deliver a
+prototype code, which  provides an energy efficiency of  at least five
+times of the  baseline value.  Such an implementation  would allow the
+community to  investigate critical questions at  higher resolution and
+over longer periods, at reduced cost to the environment.

### 2014-03-31 paper_enahpc14/Introduction.tex
updates paper; adds two empty files for the redaction of COSMO-ART model description and setup.

@@
 considered a heat-trapping greenhouse gas and have shorter atmospheric
-lifetimes, they significantly affect  the Earth's radiation budget and
-are acknowledged as one of  the most significant and uncertain aspects
-in  anthropogenic forcing  over the  last 150  years \citep{Koch-2009,
-IPCC-2013}.  Enhanced  aerosols concentrations can  impact the climate
-system by  reflecting (e.g.  pure sulfates and  nitrates) or absorbing
-(e.g.  black  carbon) solar radiation  and thereby exert a  cooling or
-warming  effect on  the Earth-atmosphere  system, causing  a so-called
-direct    radiative   forcing    \citep{Charlson-1991,   Haywood-2000,
-Ramanathan-2001,  Liao-2005, Bangert-2012,  Lundgren-2013}.  Depending
-on their  size and  chemical composition, they  can also act  as cloud
-condensation  and   ice  nuclei   and  profoundly  impact   the  cloud
-microphysical   processes  and   optical   properties  (cloud   albedo
-effect,   \citealp{Twomey-1977}),  leading   to  changes   in  droplet
-concentrations                \citep{Albrecht-1989}                and
-precipitation   \citep{Rosenfeld-2000,   Khain-2008,  Pruppacher-2010,
-Seifert-2012, Tao-2012, Lee-2012},  and therefore producing a negative
-indirect   radiative    forcing   \citep{Haywood-2000,   Lohmann-2005,
-VandenHeever-2011,   Rosenfeld-2013}.    However,   the   effects   of
-anthropogenic aerosols on clouds and the hydrological cycle as well as
-the  cloud   lifetime  effect  are  especially  hard   to  assess  and
+lifetimes, they significantly  affect the atmospheric radiative fluxes
+and  are acknowledged  as one  of the  most significant  and uncertain
+aspects    in    anthropogenic    forcing    over   the    last    150
+years \citep{Koch-2009,  IPCC-2013}.  Enhanced aerosols concentrations
+can impact the  climate system by reflecting (e.g.   pure sulfates and
+nitrates)  or  absorbing  (e.g.   black carbon)  solar  radiation  and
+thereby  exert a  cooling or  warming effect  on  the Earth-atmosphere
+system,      causing       a      so-called      direct      radiative
+forcing     \citep{Charlson-1991,    Haywood-2000,    Ramanathan-2001,
+Liao-2005, Bangert-2012, Lundgren-2013}.   Depending on their size and
+chemical composition, they can also  act as cloud condensation and ice
+nuclei, and  thereby influence  the cloud microphysical  processes and
+optical   properties  (cloud  albedo   effect,  \citealp{Twomey-1977})
+through the contribution to  cloud formation.  This results in changes
+in      droplet      concentrations     \citep{Albrecht-1989}      and
+precipitation     \citep{Rosenfeld-2000,    Khain-2008,    Khain-2009,
+Pruppacher-2010,   Seifert-2012,  Tao-2012,  Lee-2012},   producing  a
+so-called  negative  indirect  radiative forcing  \citep{Haywood-2000,
+Lohmann-2005,  VandenHeever-2011, Rosenfeld-2013}.  Wet  scavenging of
+aerosol  particles  represents  a  major  removal  mechanism  for  air
+pollutants from  the atmosphere, however the  effects of anthropogenic
+aerosols on  clouds and  the hydrological cycle  as well as  the cloud
+lifetime    effect    are    especially    hard    to    assess    and
 quantify   \citep{IPCC-2013},   and   remain   one  of   the   largest
 uncertainties in climate modeling and in climate change prediction due
-to the lack of  understanding of cloud feedbacks \citep{Sherwood-2013,
+to  the lack  of  knowledge in  cloud feedbacks  \citep{Sherwood-2013,
 Rosenfeld-2013, Lee-2013}. \\
@@
 \noindent 
-Hence to  enhance our understanding of  aerosol-cloud interactions and
+Hence to  improve our understanding of  aerosol-cloud interactions and
 reduce  uncertainties  of aerosol  effects  in  climate, the  research
@@
 investments  are still required  to take  it to  a similar  level. The
-efficiency of ART is being  addressed in the EU Exa2Green project.  In
-Sec.~\ref{sec:1},  we   introduce  the  aerosol  model   and  its  its
-integration  into  COSMO.    Sec.~\ref{sec:2}  presents  the  hardware
-considerations.  In  Sec.~\ref{sec:3} we describe  measurement devices
-and energy-aware techniques employed  to evaluate the energy footprint
-of the  considered application.  Finally,  Sec.~\ref{sec:4} presents a
-benchmark  analysis  of  COSMO-ART   from  the  Exa2Green  project  to
-highlight the critical components and an outline of the next steps for
-model optimization.  The ultimate goal  of the project is to deliver a
-prototype code, which  provides an energy efficiency of  at least five
-times of the  baseline value.  Such an implementation  would allow the
-community to  investigate critical questions at  higher resolution and
-over longer periods, at reduced cost to the environment.
+efficiency of ART is being  addressed in the EU Exa2Green project. The
+ultimate goal  of the  project is to  deliver a prototype  code, which
+provides an energy  efficiency of at least five  times of the baseline
+value.   Such   an  implementation   would  allow  the   community  to
+investigate critical  questions at  higher resolution and  over longer
+periods, at reduced cost  to the environment.  In Sec.~\ref{sec:1}, we
+introduce the online coupled  mesoscale atmosphere and chemistry model
+COSMO-ART  \citep{Vogel-2009}  that  accounts  for  feedbacks  between
+chemistry,  aerosols,  radiation,  and clouds.   Sec.~\ref{sec:2}  and
+Sec.~\ref{sec:3}  successively describe  the hardware  and measurement
+devices used in evaluation as well as energy-aware techniques employed
+to  determine  the energy  footprint  of  the considered  application.
+Sec.~\ref{sec:4}  shows benchmark results  from the  Exa2Green project
+and highlight the critical  components of the application.  Finally, a
+summary   and  directions   for  model   optimization  are   given  in
+Sec.~\ref{concl}.

### 2014-03-31 trunk/paper_enahpc14/paper.tex
updates paper; adds two empty files for the redaction of COSMO-ART model description and setup.

@@
   while  minimizing energy-to-solution.   This preliminary  study sets
-  the basis of broader  considerations to tackle challenges related to
-  energy efficient high performance  computing in the framework of the
-  Exa2Green project (\url{http://exa2green.eu/}).
+  the basis of subsequent  considerations to tackle challenges related
+  to energy  efficient high performance computing in  the framework of
+  the Exa2Green project (\url{http://exa2green.eu/}).
 
@@
 
-\section{COSMO coupled with ART extension}
+\section{The operational COSMO-ART}
 \label{sec:1}
-\subsection{Model description and set-up}
-\label{subsec:1}
-\input{Sec1:COSMO-ART.tex}
+\subsection{Model description}
+\label{subsec:1.1}
+\input{Subsec_1.1:Model_Description.tex}
+
+\subsection{Model setup}
+\label{subsec:1.2}
+\input{Subsec_1.2:Model_Setup.tex}
 
@@
 \label{sec:4}
+
+\section{Conclusion}
+\label{concl}
 
@@
 the  High  Performance  and  High  Productivity  Computing  Initiative
-(\url{www.hp2c.ch}).)
+(\url{www.hp2c.ch}) for  results that will be  leveraged in subsequent
+code refactoring.
 \end{acknowledgements}

### 2014-03-31 trunk/paper_enahpc14/Introduction.tex
updates paper and tidy files up; starts redaction of Section 1.

--- 2014-03-31 paper_enahpc14/Introduction.tex
@@
 warming  below  2  degrees  Celsius  compared to  the  temperature  in
-pre-industrial times.\\
+pre-industrial times.
 
-\noindent
 Current  cumulative releases of  human-induced gas  emissions enhances
@@
 reaching the earth's  surface and by reducing the  amount of radiation
-energy emitted into space.\\
+energy emitted into space.
 
-\noindent 
 Aerosols are  suspensions of liquid,  solid or mixed particles  in the
@@
 to  the lack  of  knowledge in  cloud feedbacks  \citep{Sherwood-2013,
-Rosenfeld-2013, Lee-2013}. \\
+Rosenfeld-2013, Lee-2013}.
 
-\noindent 
 Hence to  improve our understanding of  aerosol-cloud interactions and
@@
 weather service  DWD and MeteoSwiss,  and used in the  climate version
-COSMO-CLM by a wide  research community.  The extended COSMO-ART model
-provides a detailed description of air pollution chemistry and aerosol
-processes, and  is mainly  designed to study  air quality  and aerosol
-meteorology feedbacks on short, episodic to annual time scales.  It is
-capable  of   simulating  aerosol  distributions  as   well  as  their
-interactions over Europe as well as other regional domains.\\
+(COSMO-CLM)  by a  wide  research community.   The extended  COSMO-ART
+model provides  a detailed description of air  pollution chemistry and
+aerosol processes,  and is  mainly designed to  study air  quality and
+aerosol  meteorology  feedbacks  on  short, episodic  to  annual  time
+scales.  It is capable of  simulating aerosol distributions as well as
+their interactions over Europe as well as other regional domains.
 
-\noindent 
 COSMO-ART is  computationally much more demanding than  the COSMO core
@@
 periods, at reduced cost  to the environment.  In Sec.~\ref{sec:1}, we
-introduce the online coupled  mesoscale atmosphere and chemistry model
-COSMO-ART  \citep{Vogel-2009}  that  accounts  for  feedbacks  between
-chemistry,  aerosols,  radiation,  and clouds.   Sec.~\ref{sec:2}  and
-Sec.~\ref{sec:3}  successively describe  the hardware  and measurement
-devices used in evaluation as well as energy-aware techniques employed
-to  determine  the energy  footprint  of  the considered  application.
-Sec.~\ref{sec:4}  shows benchmark results  from the  Exa2Green project
-and highlight the critical  components of the application.  Finally, a
-summary   and  directions   for  model   optimization  are   given  in
-Sec.~\ref{concl}.
+introduce the regional  atmospheric model COSMO-ART \citep{Vogel-2009}
+that  accounts for feedbacks  between chemistry,  aerosols, radiation,
+and   clouds.   Sec.~\ref{sec:2}  and   Sec.~\ref{sec:3}  successively
+describe the  hardware and measurement  devices used in  evaluation as
+well  as  energy-aware techniques  employed  to  determine the  energy
+footprint  of  the  considered  application.   Sec.~\ref{sec:4}  shows
+benchmark  results  from  the  Exa2Green  project  and  highlight  the
+critical  components  of  the  application.  Finally,  a  summary  and
+directions for model optimization are given in Sec.~\ref{concl}.

### 2014-03-31 paper_enahpc14/Sec1:COSMO-ART.tex
updates paper and tidy files up; starts redaction of Section 1.

@@
-This   study   uses   the   regional   atmospheric   model   COSMO-ART
-\citep{Vogel-2009} (ART stands for Aerosols and Reactive Trace gases),
-in which atmospheric chemistry is online-coupled to the COSMO regional
-numerical      weather      prediction      and     climate      model
-\citep{Baldauf-2011}. The online coupling  is realized in a consistent
-way using the same spatial and temporal resolution for all scalars, as
-for    example   advection,   the    same   numerical    schemes   are
-applied. Detailed model description can be found in the aforementioned
-publications as well as in \citep{Stanelle-2010}, \citep{Bangert-2012}
-and \citep{Knote-2011}.
+\subsection{Model description}
+\label{subsec:1.1}
 
-The gaseous chemistry in COSMO-ART  is simulated by a modified version
-of  the Regional Acid  Deposition Model,  Version 2  (RADM2) mechanism
-\citep{Stockwell-1990},  which  is   extended  to  describe  secondary
-organic  aerosol formation  \citep{Schell-2001}  and hydroxyl  radical
-recycling  due  to  isoprene chemistry  \citep{Geiger-2003}.   Aerosol
-dynamics   are  simulated   with   the  modal   aerosol  module   MADE
-\citep{Ackermann-1998}, improved  by explicit treatment  of soot aging
-through  condensation  of  inorganic \citep{Riemer-2003}  and  organic
-substances.   The five  modes  that represent  the aerosol  population
-contain:  pure   soot,  secondary  mixtures   of  sulfates,  nitrates,
-ammonium, organics  and water (nucleation and  accumulation mode), and
-the  internal mixtures  of all  these species  in both  modes. Aerosol
-species are transferred among  these mixtures through condensation and
-coagulation.   A  separate  coarse  mode contains  additional  primary
-emitted  particles.    Meteorologically-affected  emissions  are  also
-online-coupled to the model  system. Sea-salt and desert dust emission
-algorithms are described in \citep{Vogel-2006} and \citep{Vogel-2009}.
-The biogenic VOC emissions are calculated as functions of the land use
-type    based    on   the    Global    Land    Cover   2000    dataset
-\citep{Bartholome-2005}  and the  modeled  temperatures and  radiative
-fluxes  \citep{Vogel-1995}.   Sedimentation,  advection,  washout  and
-turbulent diffusion also modify the aerosol distributions.
+COSMO-ART  (ART   stands  for  Aerosols  and   Reactive  Trace  gases,
+\citealp{Vogel-2009}),  developed  at KIT  Karlsruhe  (Germany), is  a
+regional  to  continental scale  model  coupled  online  to the  COSMO
+numerical    weather    prediction    (NWP)    and    climate    model
+\citep{Baldauf-2011}. COSMO is used  by several European countries for
+operational weather  forecast. The  gaseous chemistry in  COSMO-ART is
+solved by  a modified version  of the Regional Acid  Deposition Model,
+Version  2 (RADM2)  mechanism \citep{Stockwell-1990},  which  has been
+extended   to    describe   secondary   organic    aerosol   formation
+\citep{Schell-2001, Athanasopoulou-2013} and hydroxyl recycling due to
+isoprene chemistry \citep{Geiger-2003}  and heterogeneous reactions as
+hydrolysis  of N2O5.  Aerosols are  represented by  the  modal aerosol
+module MADE \citep{Ackermann-1998},  improved by explicit treatment of
+soot aging  through condensation of  inorganic \citep{Riemer-2003} and
+organic substances (MADEsoot).
+
+The  five modes that  represent the  aerosol population  contain: pure
+soot, secondary mixtures of sulfates, nitrates, ammonium, organics and
+water (nucleation  and accumulation) and the internal  mixtures of all
+these species in  both modes. Separate fine and  coarse emission modes
+for  sea-salt,  dust  \citep{Stanelle-2010},  and  rest  anthropogenic
+species  are treated by  six additional  modes.  Specific  modules are
+included    to   simulate    the   dispersion    of    pollen   grains
+\citep{Vogel-2008}   and  other  biological   particles.   Meteorology
+affected emissions are also online coupled within the model system.
+
+\subsection{Model setup}
+\label{subsec:1.2}

### 2014-04-01 trunk/paper_enahpc14/paper.tex
updates paper and tidy files up; starts redaction of Section 1.
Manuel's changes

@@
 \usepackage[round, sort&compress, numbers]{natbib}
+
+\usepackage{multirow}
+\usepackage{graphicx}
+\usepackage{units}
+\usepackage{color}
+
+\usepackage{xspace}
+\newcommand{\pmlib}{\texttt{pmlib}\xspace}
+
 %
@@
 
-\title{Evaluating the performance  and energy efficiency of COSMO-ART,
-  a fully online coupled model  system composed of a numerical weather
-  forecast model  and a  chemical transport model.   \thanks{Grants or
-    other notes  about the  article that should  go on the  front page
-    should be placed here.}}
+\title{Evaluating the Performance  and Energy Efficiency of COSMO-ART,
+  a Fully Online Coupled Model System with Numerical Weather
+    Forecast and Chemical Transport Models }
+ % \thanks{This work was supported 
+ %   by the EU Project FP7 318793 ``EXA2GREEN''}}
 %\subtitle{Write here subtitle}
@@
 
-\author{J.Charles \and W.Sawyer \and M.Dolz \and C.Malossi}
+\author{Joseph~Charles \and William~Sawyer \and Manuel~F.~Dolz \and Cristiano~Malossi}
 
@@
 
-\institute{Dr. Joseph Charles \at Swiss National Supercomputing Centre
-  (CSCS)\\ CH-6900  Lugano, Switzerland  \\ Tel: +41  (0) 91  610 8216
-  \\ \email{joseph.charles@cscs.ch} \and Dr.  William Sawyer \at Swiss
-  National Supercomputing Centre  (CSCS)\\ CH-6900 Lugano, Switzerland
-  \\ Tel:  +41 (0) 91 610 8229  \\ \email{william.sawyer@cscs.ch} \and
-  Dr.  Manuel Dolz \at University of Hamburg \\ 20148 Hamburg, Germany
-  \\         Tel:         +49         (0)        40         460094-404
-  \\ \email{manuel.dolz@informatik.uni-hamburg.de}  \and Dr. Cristiano
-  Malossi  \at  IBM  Zurich  Research Laboratory  \\  CH-8000  Zurich,
-  Switzerland     \\     Tel:      +41     (0)     44     724     8616
+\institute{J.~Charles, W.~Sawyer 
+	\at Swiss National Supercomputing Centre (CSCS)
+	\\ CH-6900 Lugano, Switzerland  
+	%\\ Tel: +41  (0) 91  610 8216
+ 	\\ E-mails:~{\{joseph.charles,william.sawyer\}@cscs.ch} 
+        %\and Dr.  William Sawyer 
+        %\at Swiss National Supercomputing Centre  (CSCS)
+        %\\ CH-6900 Lugano, Switzerland
+        %\\ Tel:  +41 (0) 91 610 8229  
+        %\\ \email{william.sawyer@cscs.ch} 
+        \and M.~F.~Dolz 
+        \at Dept. of Informatics, University of Hamburg
+	\\ D-22.527 Hamburg, Germany
+        %\\ Tel: +49 (0) 40 460094-404
+        \\ \email{manuel.dolz@informatik.uni-hamburg.de}  
+        \and C.~Malossi  
+        \at IBM Zurich Research Laboratory 
+        \\  CH-8000 Zurich, Switzerland
+        %\\  Tel: +41 (0) 44 724 8616
         \\ \email{ACM@zurich.ibm.com}}
@@
 
-\section{The operational COSMO-ART}
+\section{The COSMO-ART model system}
 \label{sec:1}
-\subsection{Model description}
-\label{subsec:1.1}
-\input{Subsec_1.1:Model_Description.tex}
-
-\subsection{Model setup}
-\label{subsec:1.2}
-\input{Subsec_1.2:Model_Setup.tex}
+\input{Sec1:COSMO-ART.tex}
 
@@
 \label{sec:3}
+\input{power-measurement.tex}
 

### 2014-04-01 trunk/paper_enahpc14/power-measurement.tex
Manuel's changes

\subsection{Power-Performance measurement framework}

To assess the performance and the energy efficiency of COSMO-ART,
we employ a version of the integrated framework presented in~\cite{energy13}
that works in combination with VampirTrace and Vampir, which are profiling/tracing
and visualization tools, respectively. 
%The left part of the \vref{fig:Lustre} offers a graphical representation of
%the Lustre architecture; the right depicts the tracing and profiling framework.
To use our approach, COSMO-ART is compiled using the VampirTrace compiler wrappers, 
which automatically instrument the Fortran code of the model. Next, COSMO-ART 
is run on the nodes, thus dissipating certain amount of power.
The server nodes are connected to power measurement devices that account for
the dissipated power/consumed energy and send the power data to the tracing server.
The attached VampirTrace \pmlib plugin employs the client API that
sends start/stop primitives in order to gather captured data by the wattmeters onto
the tracing server, where an instance of the \pmlib server is running.
Once COSMO-ART run is finished, the VampirTrace \pmlib plugin
receives the power data from the tracing server.
The instrumentation post-process generates the performance trace files and the
\pmlib plugin inserts the power data into them.

In addition to the power measurements, we also account for the resource utilization
values of the nodes: CPU load, memory usage and storage device utilization. % and network utilization.
We run special \pmlib server instances on the server nodes
that retrieve these values from the \texttt{proc} file system (leveraging the \texttt{psutil} Python library).
Thus, \pmlib plugin instances running with the instrumented application
connect with the \pmlib servers. Finally, using the Vampir visualization tool, the power-performance traces
can be easily analyzed through a series of plots and statistics.

### 2014-04-01 trunk/paper_enahpc14/Sec1:COSMO-ART.tex
renames file for "Power Measurement" section; adds a few updates

\subsection{Model description}
\label{subsec:1.1}
COSMO-ART  (ART  stands  for   Aerosols  and  Reactive  Trace  gases),
developed  at  KIT  Karlsruhe  \citep{Vogel-2009}, is  a  regional  to
continental scale model coupled  online to the COSMO numerical weather
prediction  (NWP) and  climate  model \citep{Baldauf-2011}.   Physical
processes  like  transport,  turbulent  diffusion,  and  dry  and  wet
deposition  are  treated  together  with  photochemistry  and  aerosol
dynamics using  the modal  approach.  Aerosols dynamics  are simulated
with the modal aerosol module MADE \citep{Ackermann-1998}, improved by
explicit  treatment of  soot aging  through condensation  of inorganic
\citep{Riemer-2003}   and   organic   substances   (MADEsoot),   which
represents the  aerosol population using  eleven overlapping lognormal
modes.   Five  modes  represent  sub-micron  particles  consisting  of
sulphate,  ammonium,  nitrate,   organic  compounds,  water  and  soot
\citep{Riemer-2004}  in a  range  of mixing  state.   These modes  are
coupled with  the gas  phase by condensation  and nucleation,  and are
strongly influenced by anthropogenic emissions of gases and particles.
Sea    salt     \citep{Lundgren-2013},    mineral    dust    particles
\citep{Vogel-2006, Stanelle-2010}  and rest anthropogenic  species are
treated by  seven additional modes. For each  mode, mass contributions
and total  number concentration  are prognostic quantities,  while the
standard deviation is fixed.

Specific  modules are included  to simulate  the dispersion  of pollen
grains    \citep{Vogel-2008}   and    other    biological   particles.
Meteorologically-influenced emissions  are also online  coupled within
the  model  system.  The  biogenic  VOC  (volatile organic  compounds)
emissions are  calculated as functions of  the land use  type based on
the Global  Land Cover 2000  dataset and the modeled  temperatures and
radiative fluxes \citep{Vogel-1995}.

The gaseous chemistry in COSMO-ART  is solved by a modified version of
the  Regional  Acid  Deposition  Model, Version  2  (RADM2)  mechanism
\citep{Stockwell-1990},  which  is   extended  to  describe  secondary
organic aerosol formation based on a volatile basis set (VBS) approach
\citep{Athanasopoulou-2013}  and  hydroxyl  radical recycling  due  to
isoprene chemistry \citep{Geiger-2003}.  The thermodynamic equilibrium
between the  gas and particulate  phases of the inorganic  material is
achieved through the  ISORROPIA II module \citep{Fountoukis-2007}.

COSMO-ART  is  fully  online-coupled,  and  allows  for  feedbacks  of
aerosols  on  temperature,  radiation  and cloud  condensation  nuclei
(CCN).   Analytical  description  of  modules  incorporated  for  this
purpose  exists in  \citep{Vogel-2009,  Bangert-2011}.  The  radiation
scheme used  within the  model to calculate  the vertical  profiles of
shortwave and longwave radiative fluxes is GRAALS \citep{Ritter-1992}.
In order to account for  the interaction of aerosol particles with the
cloud microphysics and radiation,  COSMO-ART uses the two-moment cloud
microphysics  scheme of  Seifert and  Beheng  \citep{Seifert-2006} and
comprehensive parameterizations for cloud condensation and ice crystal
nucleation  \citep{Bangert-2011, Bangert-2012}.   The system  of stiff
ordinary differential equations described by chemical reactions in the
aqueous-phase together with the transfer reactions is solved using the
Kinetic PreProcessor (KPP, \citealp{Damian-2002}).

Detailed  model  description  can   be  found  in  the  aforementioned
publications  as   well  as  in   \citep{Stanelle-2010,  Bangert-2012,
  Knote-2011, Knote-2013}.

\subsection{Model setup}
\label{subsec:1.2}
To define a baseline within a code under development, it was necessary
to  find  a  run-configuration  capable  of  being  recreated  in  all
subsequent versions. The  energy-to-solution benchmarking of COSMO-ART
concerns  one-day  long simulations  without  spin-up  stage, using  a
222x216x40  discretization of  Europe and  a  time step  of 120s.  The
baseline code incorporates 34 2-d and  45 3-d fields to be written out
every hour for a simulation starting  on April 13th, which is close to
the equinox and thus brings  benefits of having approximately half day
of night and  half day of sun exposure, to  ensure a proper activation
of  the chemistry  cycle. Ultimately  new  input files  closer to  the
equinox (on March 20th) will be  created, but the current set of input
files is fine for the time being.

This COSMO-ART  version is configured  to deal with a  semi Lagrangian
horizontal advection scheme  with tricubic interpolation and selective
filling diffusion option in combination with the Runge-Kutta dynamical
core.   Concerning the modelling  of wet  deposition in  aerosols, the
baseline  has  only  indirect  cloud  feedbacks  but  doesn't  include
in-cloud  scavenging (rainout)  and  below-cloud scavenging  (washout)
yet.  Amongst  physical parameterizations, precipitation  formation is
performed  by a  two-moment  cloud microphysics  scheme  instead of  a
classical  bulk microphysics  scheme. Another  important point  is the
fact that  this version makes  use of the Kinetic  PreProcessor solver
for  the  resolution of  atmospheric  chemistry ordinary  differential
equations.

### 2014-04-01 paper_enahpc14/Introduction.tex
renames file for "Power Measurement" section; adds a few updates

@@
 periods, at reduced cost  to the environment.  In Sec.~\ref{sec:1}, we
-introduce the online coupled  mesoscale atmosphere and chemistry model
-COSMO-ART  \citep{Vogel-2009}  that  accounts  for  feedbacks  between
-chemistry,  aerosols,  radiation,  and clouds.   Sec.~\ref{sec:2}  and
-Sec.~\ref{sec:3}  successively describe  the hardware  and measurement
-devices used in evaluation as well as energy-aware techniques employed
-to  determine  the energy  footprint  of  the considered  application.
-Sec.~\ref{sec:4}  shows benchmark results  from the  Exa2Green project
-and highlight the critical  components of the application.  Finally, a
-summary   and  directions   for  model   optimization  are   given  in
-Sec.~\ref{concl}.
+introduce the regional  atmospheric model COSMO-ART \citep{Vogel-2009}
+that  accounts for feedbacks  between chemistry,  aerosols, radiation,
+and   clouds.   Sec.~\ref{sec:2}  and   Sec.~\ref{sec:3}  successively
+describe the  hardware and measurement  devices used in  evaluation as
+well  as  energy-aware techniques  employed  to  determine the  energy
+footprint  of  the  considered  application.   Sec.~\ref{sec:4}  shows
+benchmark  results from  the  Exa2Green project  and highlights  areas
+where improvements  will be  necessary for the  baseline optimisation.
+Finally, we  conclude and give directions for  further developments of
+the modeling system in Sec.~\ref{concl}.

### 2014-04-01 paper_enahpc14/paper.tex
renames source files; creates Makefiles for main directory and the newly created Figs/ subdirectory; deletes bash scripts
renames file for "Power Measurement" section; adds a few updates

--- 2014-03-25 trunk/paper_enahpc14/draft.tex
@@
 \usepackage{hyperref}
-\usepackage[round, sort, numbers, authoryear]{natbib}
+\usepackage[round, sort&compress, numbers]{natbib}
+
+\usepackage{multirow}
+\usepackage{graphicx}
+\usepackage{units}
+\usepackage{color}
+
+\usepackage{xspace}
+\newcommand{\pmlib}{\texttt{pmlib}\xspace}
+
 %
@@
   systems is  analysed by a  profiling and tracing study  to determine
-  hotspots, i.e.  the  parts of the program that  require the dominant
-  fraction of runtime, and to identify critical paths, i.e.  sequences
-  of functions that shows dominant inclusive contributions for all its
-  elements. Moreover, we describe measurement devices and energy-aware
-  techniques  employed  to  evaluate   the  energy  footprint  of  the
-  considered  application and  to  get detailed  insights about  power
-  bottlenecks.   Our  motivation  is  to  improve  corresponding  code
-  sections    to   sustain    high   performance    while   minimizing
-  energy-to-solution.   This  preliminary  study  sets  the  basis  of
-  broader  considerations  to  tackle  challenges  related  to  energy
-  efficient   HPC  in   the   framework  of   the  Exa2Green   project
-  (\url{http://exa2green.eu/}).
+  hotspots  and  identify   critical  paths.   Moreover,  we  describe
+  measurement devices and energy-aware techniques employed to evaluate
+  the  energy  footprint of  the  considered  application  and to  get
+  detailed  insights about  power bottlenecks.   Our motivation  is to
+  improve  corresponding  code sections  to  sustain high  performance
+  while minimizing energy-to-solution.  This preliminary work sets the
+  basis for subsequent studies  to tackle challenges related to energy
+  efficient  high  performance  computing  in  the  framework  of  the
+  Exa2Green project (\url{http://exa2green.eu/}).
 
 \keywords{High performance computing  \and Energy-aware computing \and
-  Green computing  \and Numerical Weather  Prediction \and Atmospheric
-  Chemistry \and Aerosol Modelling \and COSMO-ART}
+  Green computing  \and Numerical weather  prediction \and Atmospheric
+  chemistry  \and  Aerosols  modelling  \and  Profiling  methods  \and
+  Benchmark analysis \and COSMO-ART coupled model}
 % \PACS{PACS code1 \and PACS code2 \and more}
@@
 
-\section{COSMO-ART model description}
+\section{The COSMO-ART model system}
 \label{sec:1}
@@
 
-\subsection{Subsection title}
-\label{subsec:1}
-Don't forget to  give each section and subsection  a unique label (see
-Sect.~\ref{sec:1}).
-\paragraph{Paragraph headings} Use paragraph headings as needed.
-\begin{equation}
-a^2+b^2=c^2
-\end{equation}
+\section{Hardware description}
+\label{sec:2}
 
-% For one-column wide figures use
-\begin{figure}
-% Use the relevant command to insert your figure file.
-% For example, with the graphicx package use
-  \includegraphics{example.eps}
-% figure caption is below the figure
-\caption{Please write your figure caption here}
-\label{fig:1}       % Give a unique label
-\end{figure}
-%
-% For two-column wide figures use
-\begin{figure*}
-% Use the relevant command to insert your figure file.
-% For example, with the graphicx package use
-  \includegraphics[width=0.75\textwidth]{example.eps}
-% figure caption is below the figure
-\caption{Please write your figure caption here}
-\label{fig:2}       % Give a unique label
-\end{figure*}
-%
-% For tables use
-\begin{table}
-% table caption is above the table
-\caption{Please write your table caption here}
-\label{tab:1}       % Give a unique label
-% For LaTeX tables use
-\begin{tabular}{lll}
-\hline\noalign{\smallskip}
-first & second & third  \\
-\noalign{\smallskip}\hline\noalign{\smallskip}
-number & number & number \\
-number & number & number \\
-\noalign{\smallskip}\hline
-\end{tabular}
-\end{table}
+\section{Power measurement systems}
+\label{sec:3}
+\input{Sec3:Power-measurement.tex}
+
+\section{Benchmark results}
+\label{sec:4}
+
+\section{Conclusion}
+\label{concl}
+
+%\paragraph{Paragraph headings} Use paragraph headings as needed.
+
+%\begin{figure*}
+%  \includegraphics[width=0.4\textwidth]{Figs/earth.eps}
+%  \caption{Please write your figure caption here}
+%  \label{fig:1}
+%\end{figure*}
+
+%\begin{table}
+%  \caption{Please write your table caption here}
+%  \label{tab:1}
+%  \begin{tabular}{lll}
+%    \hline\noalign{\smallskip}
+%    first & second & third  \\
+%    \noalign{\smallskip}\hline\noalign{\smallskip}
+%    number & number & number \\
+%    number & number & number \\
+%    \noalign{\smallskip}\hline
+%  \end{tabular}
+%\end{table}
 
@@
 of Computing to  the Limit (MINECC).  The authors  would like to thank
-the    High    Performance    and    High    Productivity    Computing
-(\url{www.hp2c.ch}) Initiative.)
+the  High  Performance  and  High  Productivity  Computing  Initiative
+(\url{www.hp2c.ch}) for  results that will be  leveraged in subsequent
+code refactoring.
 \end{acknowledgements}
 
-\DeclareRobustCommand\IPCClongname{ Intergovernmental Panel on Climate Change}
+\DeclareRobustCommand\IPCClongname{ - Intergovernmental Panel on Climate Change}
+
 \bibliographystyle{plainnat}

### 2014-04-01 trunk/paper_enahpc14/Sec1:COSMO-ART.tex
adds a few updates

@@
 deposition  are  treated  together  with  photochemistry  and  aerosol
-dynamics using  the modal  approach.  Aerosols dynamics  are simulated
-with the modal aerosol module MADE \citep{Ackermann-1998}, improved by
+dynamics  using the  modal approach.
+
+Aerosols dynamics  are simulated  with the modal  aerosol microphysics
+module  MADE \citep{Ackermann-1998}, expanded  in MADEsoot  to feature
 explicit  treatment of  soot aging  through condensation  of inorganic
-\citep{Riemer-2003}   and   organic   substances   (MADEsoot),   which
-represents the  aerosol population using  eleven overlapping lognormal
-modes.   Five  modes  represent  sub-micron  particles  consisting  of
-sulphate,  ammonium,  nitrate,   organic  compounds,  water  and  soot
-\citep{Riemer-2004}  in a  range  of mixing  state.   These modes  are
-coupled with  the gas  phase by condensation  and nucleation,  and are
-strongly influenced by anthropogenic emissions of gases and particles.
-Sea    salt     \citep{Lundgren-2013},    mineral    dust    particles
-\citep{Vogel-2006, Stanelle-2010}  and rest anthropogenic  species are
-treated by  seven additional modes. For each  mode, mass contributions
-and total  number concentration  are prognostic quantities,  while the
-standard deviation is fixed.
+\citep{Riemer-2003} and organic substances. MADE describes the aerosol
+population  through  five   modes  representing  sub-micron  particles
+consisting of sulphate, ammonium, nitrate, particulate organic matter,
+water and soot  \citep{Riemer-2004} in a range of  mixing state. These
+modes are coupled  with the gas phase by  condensation and nucleation,
+and are  strongly influenced by  anthropogenic emissions of  gases and
+particles. MADEsoot  describes the sub-micrometer  aerosol population,
+composed  of sea  salt \citep{Lundgren-2013},  mineral  dust particles
+\citep{Vogel-2006,  Stanelle-2010},   by  means  of   six  interacting
+log-normal modes.  For each  mode, mass contributions and total number
+concentration are prognostic  quantities, while the standard deviation
+is fixed.
 

### 2014-04-01 trunk/paper_enahpc14/paper.tex
adds a few updates

@@
 %
-\documentclass[twocolumn]{svjour3}           % twocolumn
+\documentclass[twocolumn]{svjour3}
 %
-\smartqed % flush right qed marks, e.g. at end of proof
+\smartqed
 %
@@
 \usepackage[round, sort&compress, numbers]{natbib}
-
 \usepackage{multirow}
@@
 \usepackage{color}
-
 \usepackage{xspace}
-\newcommand{\pmlib}{\texttt{pmlib}\xspace}
-
 %
@@
 %
-% please place your own definitions here and don't use \def but
-% \newcommand{}{}
+% please place your own definitions here and don't use \def but \newcommand{}{}
+\newcommand{\pmlib}{\texttt{pmlib}\xspace}
 %
@@
     Forecast and Chemical Transport Models }
- % \thanks{This work was supported 
- %   by the EU Project FP7 318793 ``EXA2GREEN''}}
+% \thanks{This work was supported by the EU Project FP7 318793 ``EXA2GREEN''}}
 %\subtitle{Write here subtitle}
@@
  	\\ E-mails:~{\{joseph.charles,william.sawyer\}@cscs.ch} 
-        %\and Dr.  William Sawyer 
-        %\at Swiss National Supercomputing Centre  (CSCS)
-        %\\ CH-6900 Lugano, Switzerland
-        %\\ Tel:  +41 (0) 91 610 8229  
-        %\\ \email{william.sawyer@cscs.ch} 
         \and M.~F.~Dolz 
@@
   improve  corresponding  code sections  to  sustain high  performance
-  while  minimizing energy-to-solution.   This preliminary  study sets
-  the basis of subsequent  considerations to tackle challenges related
-  to energy  efficient high performance computing in  the framework of
-  the Exa2Green project (\url{http://exa2green.eu/}).
+  while minimizing energy-to-solution.  This preliminary work sets the
+  basis for subsequent studies  to tackle challenges related to energy
+  efficient  high  performance  computing  in  the  framework  of  the
+  Exa2Green project (\url{http://exa2green.eu/}).
 
@@
 \label{sec:3}
-\input{power-measurement.tex}
+\input{Sec3:Power-measurement.tex}
 

### 2014-04-01 trunk/paper_enahpc14/Sec2:Power-measurement.tex
adds missing files from previous commit

--- 2014-04-01 trunk/paper_enahpc14/power-measurement.tex
@@
 \subsection{Power-Performance measurement framework}
-
-To assess the performance and the energy efficiency of COSMO-ART,
-we employ a version of the integrated framework presented in~\cite{energy13}
-that works in combination with VampirTrace and Vampir, which are profiling/tracing
-and visualization tools, respectively. 
+\label{subsec:3.1}
+To assess the  performance and the energy efficiency  of COSMO-ART, we
+employ   a    version   of   the    integrated   framework   presented
+in~\cite{energy13}  that  works in  combination  with VampirTrace  and
+Vampir,   which  are   profiling/tracing   and  visualization   tools,
+respectively.
 %The left part of the \vref{fig:Lustre} offers a graphical representation of
 %the Lustre architecture; the right depicts the tracing and profiling framework.
-To use our approach, COSMO-ART is compiled using the VampirTrace compiler wrappers, 
-which automatically instrument the Fortran code of the model. Next, COSMO-ART 
-is run on the nodes, thus dissipating certain amount of power.
-The server nodes are connected to power measurement devices that account for
-the dissipated power/consumed energy and send the power data to the tracing server.
-The attached VampirTrace \pmlib plugin employs the client API that
-sends start/stop primitives in order to gather captured data by the wattmeters onto
-the tracing server, where an instance of the \pmlib server is running.
-Once COSMO-ART run is finished, the VampirTrace \pmlib plugin
-receives the power data from the tracing server.
-The instrumentation post-process generates the performance trace files and the
-\pmlib plugin inserts the power data into them.
+To  use our  approach,  COSMO-ART is  compiled  using the  VampirTrace
+compiler wrappers, which automatically  instrument the Fortran code of
+the  model. Next,  COSMO-ART is  run  on the  nodes, thus  dissipating
+certain  amount of  power.  The  server nodes  are connected  to power
+measurement  devices that  account for  the  dissipated power/consumed
+energy and  send the power data  to the tracing  server.  The attached
+VampirTrace \pmlib plugin employs the client API that sends start/stop
+primitives in order to gather captured data by the wattmeters onto the
+tracing server,  where an  instance of the  \pmlib server  is running.
+Once COSMO-ART run is finished, the VampirTrace \pmlib plugin receives
+the  power   data  from  the  tracing   server.   The  instrumentation
+post-process  generates the  performance  trace files  and the  \pmlib
+plugin inserts the power data into them.
 
-In addition to the power measurements, we also account for the resource utilization
-values of the nodes: CPU load, memory usage and storage device utilization. % and network utilization.
-We run special \pmlib server instances on the server nodes
-that retrieve these values from the \texttt{proc} file system (leveraging the \texttt{psutil} Python library).
-Thus, \pmlib plugin instances running with the instrumented application
-connect with the \pmlib servers. Finally, using the Vampir visualization tool, the power-performance traces
-can be easily analyzed through a series of plots and statistics.
+In  addition  to the  power  measurements,  we  also account  for  the
+resource utilization values  of the nodes: CPU load,  memory usage and
+storage device utilization. % and network utilization.
+
+We  run special  \pmlib  server  instances on  the  server nodes  that
+retrieve these  values from the \texttt{proc}  file system (leveraging
+the  \texttt{psutil} Python library).   Thus, \pmlib  plugin instances
+running  with the  instrumented  application connect  with the  \pmlib
+servers.   Finally,   using  the   Vampir   visualization  tool,   the
+power-performance traces  can be easily  analyzed through a  series of
+plots and statistics.

### 2014-04-02 paper_enahpc14/Introduction.tex
updates sections 1 & 2; adds file "Conclusion.tex"

@@
 that  accounts for feedbacks  between chemistry,  aerosols, radiation,
-and   clouds.   Sec.~\ref{sec:2}  and   Sec.~\ref{sec:3}  successively
-describe the  hardware and measurement  devices used in  evaluation as
-well  as  energy-aware techniques  employed  to  determine the  energy
-footprint  of  the  considered  application.   Sec.~\ref{sec:4}  shows
-benchmark  results from  the  Exa2Green project  and highlights  areas
-where improvements  will be  necessary for the  baseline optimisation.
-Finally, we  conclude and give directions for  further developments of
+and clouds and additionally specify its technical setup related to the
+investigated    performance    and     energy    evaluation    methods.
+Sec.~\ref{sec:2}  describes the different  HPC systems  and associated
+power  measurement  equipments used  for  conducting the  benchmarking
+study and  presents energy-aware techniques employed  to determine the
+energy  footprint  of  the considered  application.   Sec.~\ref{sec:3}
+gathers benchmark  results from  the Exa2Green project  and highlights
+areas where improvements will be necessary for the subsequent baseline
+refactoring.  Some related work  will be discussed in Sec.~\ref{sec:4}
+Finally, we  conclude with practical guidelines  to identifying energy
+saving opportunities  and give directions for  further developments of
 the modeling system in Sec.~\ref{concl}.

### 2014-04-02 paper_enahpc14/Sec1:COSMO-ART.tex
updates sections 1 & 2; adds file "Conclusion.tex"

@@
 \label{subsec:1.1}
+COSMO-ART  (ART  stands  for   Aerosols  and  Reactive  Trace  gases),
+developed  at  KIT  Karlsruhe  \citep{Vogel-2009}, is  a  regional  to
+continental scale model coupled  online to the COSMO numerical weather
+prediction  (NWP) and  climate  model \citep{Baldauf-2011}.   Physical
+processes  like  transport,  turbulent  diffusion,  and  dry  and  wet
+deposition  are  treated  together  with  photochemistry  and  aerosol
+dynamics  using the  modal approach.
 
-COSMO-ART  (ART   stands  for  Aerosols  and   Reactive  Trace  gases,
-\citealp{Vogel-2009}),  developed  at KIT  Karlsruhe  (Germany), is  a
-regional  to  continental scale  model  coupled  online  to the  COSMO
-numerical    weather    prediction    (NWP)    and    climate    model
-\citep{Baldauf-2011}. COSMO is used  by several European countries for
-operational weather  forecast. The  gaseous chemistry in  COSMO-ART is
-solved by  a modified version  of the Regional Acid  Deposition Model,
-Version  2 (RADM2)  mechanism \citep{Stockwell-1990},  which  has been
-extended   to    describe   secondary   organic    aerosol   formation
-\citep{Schell-2001, Athanasopoulou-2013} and hydroxyl recycling due to
-isoprene chemistry \citep{Geiger-2003}  and heterogeneous reactions as
-hydrolysis  of N2O5.  Aerosols are  represented by  the  modal aerosol
-module MADE \citep{Ackermann-1998},  improved by explicit treatment of
-soot aging  through condensation of  inorganic \citep{Riemer-2003} and
-organic substances (MADEsoot).
+Aerosols dynamics  are simulated  with the modal  aerosol microphysics
+module  MADE \citep{Ackermann-1998}, expanded  in MADEsoot  to feature
+explicit  treatment of  soot aging  through condensation  of inorganic
+\citep{Riemer-2003} and organic substances. MADE describes the aerosol
+population  through  five   modes  representing  sub-micron  particles
+consisting of sulphate, ammonium, nitrate, particulate organic matter,
+water and soot  \citep{Riemer-2004} in a range of  mixing state. These
+modes are coupled  with the gas phase by  condensation and nucleation,
+and are  strongly influenced by  anthropogenic emissions of  gases and
+particles. MADEsoot  describes the sub-micrometer  aerosol population,
+composed  of sea  salt \citep{Lundgren-2013},  mineral  dust particles
+\citep{Vogel-2006,  Stanelle-2010},   by  means  of   six  interacting
+log-normal modes.  For each  mode, mass contributions and total number
+concentration are prognostic  quantities, while the standard deviation
+is fixed.
 
-The  five modes that  represent the  aerosol population  contain: pure
-soot, secondary mixtures of sulfates, nitrates, ammonium, organics and
-water (nucleation  and accumulation) and the internal  mixtures of all
-these species in  both modes. Separate fine and  coarse emission modes
-for  sea-salt,  dust  \citep{Stanelle-2010},  and  rest  anthropogenic
-species  are treated by  six additional  modes.  Specific  modules are
-included    to   simulate    the   dispersion    of    pollen   grains
-\citep{Vogel-2008}   and  other  biological   particles.   Meteorology
-affected emissions are also online coupled within the model system.
+Specific  modules are included  to simulate  the dispersion  of pollen
+grains    \citep{Vogel-2008}   and    other    biological   particles.
+Meteorologically-influenced emissions  are also online  coupled within
+the  model  system.  The  biogenic  VOC  (volatile organic  compounds)
+emissions are  calculated as functions of  the land use  type based on
+the Global  Land Cover 2000  dataset and the modeled  temperatures and
+radiative fluxes \citep{Vogel-1995}.
+
+The gaseous chemistry in COSMO-ART  is solved by a modified version of
+the  Regional  Acid  Deposition  Model, Version  2  (RADM2)  mechanism
+\citep{Stockwell-1990},  which  is   extended  to  describe  secondary
+organic aerosol formation based on a volatile basis set (VBS) approach
+\citep{Athanasopoulou-2013}  and  hydroxyl  radical recycling  due  to
+isoprene chemistry \citep{Geiger-2003}.  The thermodynamic equilibrium
+between the  gas and particulate  phases of the inorganic  material is
+achieved through the  ISORROPIA II module \citep{Fountoukis-2007}.
+
+COSMO-ART  is  fully  online-coupled,  and  allows  for  feedbacks  of
+aerosols  on  temperature,  radiation  and cloud  condensation  nuclei
+(CCN).   Analytical  description  of  modules  incorporated  for  this
+purpose  exists in  \citep{Vogel-2009,  Bangert-2011}.  The  radiation
+scheme used  within the  model to calculate  the vertical  profiles of
+shortwave and longwave radiative fluxes is GRAALS \citep{Ritter-1992}.
+In order to account for  the interaction of aerosol particles with the
+cloud microphysics and radiation,  COSMO-ART uses the two-moment cloud
+microphysics  scheme of  Seifert and  Beheng  \citep{Seifert-2006} and
+comprehensive parameterizations for cloud condensation and ice crystal
+nucleation  \citep{Bangert-2011, Bangert-2012}.   The system  of stiff
+ordinary differential equations described by chemical reactions in the
+aqueous-phase together with the transfer reactions is solved using the
+Kinetic PreProcessor (KPP, \citealp{Damian-2002}).
+
+Detailed  model  description  can   be  found  in  the  aforementioned
+publications  as   well  as  in   \citep{Stanelle-2010,  Bangert-2012,
+  Knote-2011, Knote-2013}.
 
@@
 \label{subsec:1.2}
+Establishing effective energy performance benchmarking of a code under
+intense development such as COSMO-ART is a challenging task because of
+the  absolute necessity that  results must  be reproducible  within an
+expected  variance for the  duration of  the Exa2Green  project. After
+consultation  with  the  climate  community  to  properly  define  the
+baseline,  it was  necessary to  find a  run-configuration  capable of
+being recreated in all subsequent versions of the code.
+
+Thus, we give here a brief  overview of the main features of the model
+setup for  our benchmark's daily  runs.  Three-dimensional simulations
+are  performed over  Europe  for April  13th,  which is  close to  the
+equinox and thus  brings benefits of having approximately  half day of
+night  and half  day  of  sun exposure,  therefore  ensuring a  proper
+activation of the chemistry cycle. We consider daily 24-hour forecasts
+without  any  previous spin-up  simulations.   The calculation  domain
+corresponds to  the CORDEX-EU-44  domain and is  covered by a  grid of
+$222\times   216$    points   with   a    horizontal   resolution   of
+$0.22\,^{\circ}$,  i.e.  50  km in  both directions,  and  40 vertical
+layers.   Boundary  and initial  conditions,  emmissions and  external
+parameters  were obtained by  the INT2LM  pre-processing interpolation
+tool, which  performed the usual interpolation from  coarse grid model
+data, and  INT2LM\_trcr\_ART, which consists mostly  of the definition
+of a large number of  tracers.  The meteorological initial and bounday
+conditions are  obtained by  the the ECMWF  global spectral  model IFS
+with an update  frequency of 3h.  Boundary data  for gas-phase species
+are taken from IFS-MOZART  output at 6h temporal resolution.  Multiple
+runs of COSMO-ART were  performed to illustrate the reproducibility of
+the baseline, and quantify  the significant uncertainties in the power
+measurement,   as   dictated   by   the  available   technology   (see
+subsection~\ref{subsec:3.2}).   The run-configuration  incorporates 34
+2-d and 45 3-d fields to be written out every hour and doesn't include
+data assimilation methods.
+
+This COSMO-ART  version is configured  to deal with a  semi Lagrangian
+horizontal advection scheme  with tricubic interpolation and selective
+filling diffusion option in combination with the Runge-Kutta dynamical
+core. It  also makes  use of the  Kinetic PreProcessor solver  for the
+resolution of  atmospheric chemistry ordinary  differential equations.
+Concerning the  modelling of wet deposition in  aerosols, the baseline
+has  only  indirect  cloud  feedbacks  but  doesn't  include  in-cloud
+scavenging  (rainout)   and  below-cloud  scavenging   (washout)  yet.
+Amongst   physical  parameterizations,   precipitation   formation  is
+performed  by a  two-moment  cloud microphysics  scheme  instead of  a
+classical  bulk microphysics  scheme.

### 2014-04-02 paper_enahpc14/paper.tex
updates sections 1 & 2; adds file "Conclusion.tex"

@@
 \title{Evaluating the Performance  and Energy Efficiency of COSMO-ART,
-  a Fully Online Coupled Model System with Numerical Weather
-    Forecast and Chemical Transport Models }
- % \thanks{This work was supported 
- %   by the EU Project FP7 318793 ``EXA2GREEN''}}
+  a Fully Online Coupled  Model System with Numerical Weather Forecast
+  and Chemical Transport Models }
+% \thanks{This work was supported by the EU Project FP7 318793 ``EXA2GREEN''}}
 %\subtitle{Write here subtitle}
@@
 
-\author{Joseph~Charles \and William~Sawyer \and Manuel~F.~Dolz \and Cristiano~Malossi}
+\author{Joseph~Charles  \and William~Sawyer  \and  Manuel~F.~Dolz \and
+  Cristiano~Malossi}
 
@@
   In this paper we present  COSMO-ART, an extension of the operational
-  weather  forecast  model  of   the  German  Weather  Service  (DWD),
+  weather  forecast  model  of   the  German  weather  service  (DWD),
   developed for  the evaluation of the interactions  of reactive gases
@@
   into account.   The overall performance  of this application  on HPC
-  systems is  analysed by a  profiling and tracing study  to determine
-  hotspots  and  identify   critical  paths.   Moreover,  we  describe
-  measurement devices and energy-aware techniques employed to evaluate
-  the  energy  footprint of  the  considered  application  and to  get
-  detailed  insights about  power bottlenecks.   Our motivation  is to
-  improve  corresponding  code sections  to  sustain high  performance
-  while minimizing energy-to-solution.  This preliminary work sets the
-  basis for subsequent studies  to tackle challenges related to energy
-  efficient  high  performance  computing  in  the  framework  of  the
-  Exa2Green project (\url{http://exa2green.eu/}).
+  systems is analysed  by a profiling study to  determine hotspots and
+  identify critical paths.   Moreover, we describe measurement devices
+  and  energy-aware   techniques  employed  to   evaluate  the  energy
+  footprint of the considered application and to get detailed insights
+  about power bottlenecks.  Our motivation is to improve corresponding
+  code   sections  to  sustain   high  performance   while  minimizing
+  energy-to-solution.   This  preliminary  work  sets  the  basis  for
+  subsequent studies to tackle  challenges related to energy efficient
+  high performance computing in the framework of the Exa2Green project
+  (\url{http://exa2green.eu/}).
 
@@
 
-\section{Hardware description}
+\section{Power-performance measurement framework}
 \label{sec:2}
+\input{Sec2:Power-measurement.tex}
 
-\section{Power measurement systems}
+\section{Evaluation}
 \label{sec:3}
-\input{Sec3:Power-measurement.tex}
+\input{Sec3:Evaluation.tex}
 
-\section{Benchmark results}
+\section{Related work}
 \label{sec:4}
+\input{Sec4:Related.tex}
 
@@
 \label{concl}
+\input{Conclusion.tex}
 

### 2014-04-02 trunk/paper_enahpc14/Sec2:Power-measurement.tex
updates sections 1 & 2; adds file "Conclusion.tex"

@@
+\subsection{Hardware description}
+\label{subsec:3.1}
+
+\subsubsection{Monch (CSCS - ETH Zurich)}
+While hardware  platforms mature  and are replaced,  we have  chosen a
+state-of-the-art  dual-socket Intel  Ivybridge-based  platform (called
+``Monch'')  for our power  measurements, as  it is  slated to  stay in
+service without upgrade for the duration of the Exa2Green project.  In
+principle at least,  this architecture could be recreated  or found in
+an identical configuration beyond  the lifetime of the project.  Given
+that  the baseline  benchmark  can be  reproduced  within an  expected
+variance   (cf  previous   section),   and  that   the  baseline   run
+configuration can be  used in all future versions of  the code, a fair
+comparison  will  be  made  between  the baseline  and  the  milestone
+versions of COSMO-ART.
+
+Monch, which is installed  at the Swiss National Supercomputing Center
+(CSCS)  of  the  ETH  Zurich,  is a  10  rack  NEC-provided  (LX-2400)
+Intel-based cluster, utilised  from people that are part  of the Swiss
+Platform  for  Advanced Scientific  Computing  (PASC),  which aims  at
+positioning  Swiss  computational sciences  in  the emerging  exascale
+era. It is  composed of 16 administration nodes,  312 standard compute
+nodes, 24 large-memory compute nodes, 24 huge-memory compute nodes and
+8  servers to  provide  I/O  to Lustre.   Each  standard compute  node
+comprises 2 Intel 10-core Ivy Bridge EP E5-2660v2 processors operating
+at  2.2 GHz,  offering 32GB  of DDR3  1600MHz RAM,  with a  base clock
+frequency of 2.2 GHz, themselves  connected by a high speed Infiniband
+network,  based on Mellanox  SX6036 managed  FDR (Fourteen  Data Rate)
+switches, with a 56 Gb/s speed.
+
+A snapshot of the code, which includes, at least conceptually, all the
+information needed  to reproduce the  energy-to-solution benchmarks of
+COSMO-ART, was  produced and run  on a full  rack of 52  nodes without
+hyperthreading (hyperthreading was attempted, but always led to higher
+energy-to-solution), i.e. on a total  of 1040 cores using 20 tasks per
+node.
+
 \subsection{Power-Performance measurement framework}
-\label{subsec:3.1}
+\label{subsec:3.2}
+
+\subsubsection{CSCS - ETH Zurich}
+E3METER  Intelligent Power Strips  (IPS) and  Monitors (IPM)  are high
+quality electricity meters which enable to monitor and optimize energy
+consumption of datacenters or large facilities.  A central data logger
+- the  E3METER Data Concentrator  - aggregates  all measured  data and
+makes them available in various forms (SNMP, Web, etc). Data transfers
+between E3METER  IPS and E3METER Concentrator  use reliable narrowband
+powerline  communication (PLC)  technology which  avoids the  need for
+extra cabling.  It has  one internal temperature sensor. Two dedicated
+extension  ports  can be  used  to  measure  external temperature  and
+humidity  through E3METER  remote sensors.  It  features non-intrusive
+voltage, current and frequency measurement through current transformer
+with an accuracy of 1\%.
+
+\subsubsection{University of Hamburg}
 To assess the  performance and the energy efficiency  of COSMO-ART, we

### 2014-04-03 trunk/paper_enahpc14/Introduction.tex
updates introduction

@@
 resulting from  anthropogenic activities  such as fossil  fuel burning
-and  deforestation.  Long-lived greenhouse  gases, for  example carbon
-dioxide, methane  and nitrous oxide, are chemically  stable and linger
-homogeneously well-mixed and over time scales of a decade to thousands
-of years in  the atmosphere; whereas short-lived gases  such as water,
-suphur  dioxide  and  carbon  monoxide  which  respond  physically  or
-chemically  to changes  in temperature,  act primarily  as  a feedback
-mechanism.   Some   extremly  powerful  greenhouse   gases,  such  as
-fluorinated gases can even have  a warming effect on the atmosphere up
-to  23000  times  greater  than  carbon  dioxide  \citep{Zehner-2012}.
-Fortunately  the  European  Parliament,  which  has  gained  worldwide
-recognition as a  leader in climate policy, has  passed legislation in
-March 2014  to phase  out fluorinated gas  emissions by  two-thirds by
-2030,  towards  the  internationally  agreed goal  of  keeping  global
-warming  below  2  degrees  Celsius  compared to  the  temperature  in
-pre-industrial times.
+and  deforestation.
 
-Current  cumulative releases of  human-induced gas  emissions enhances
-the greenhouse  effect to  the Earth's atmosphere  and thus  creates a
-large imbalance between incoming solar absorbed radiation and outgoing
-longwave radiation emitted back  to space, through radiative trapping.
-While the  Earth's temperature  is dependent upon  the greenhouse-like
-action of  the atmosphere, the  Earth's radiation balance  is strongly
-influenced by several  other factors such as the  type of surface that
-sunlight first  encounters.  Forests, grasslands,  ocean surfaces, ice
-caps, deserts,  and cities all absorb, reflect,  and radiate radiation
-differently.   Sunlight falling  on a  white glacier  surface strongly
-reflects back into space, resulting  in minimal heating of the surface
-and  lower atmosphere.   Sunlight falling  on  a dark  desert soil  is
-strongly absorbed,  on the other hand, and  contributes to significant
-heating of the surface and lower atmosphere.  Cloud cover also affects
-greenhouse  warming by  both reducing  the amount  of  solar radiation
-reaching the earth's  surface and by reducing the  amount of radiation
-energy emitted into space.
+The Earth's  surface temperature is determined by  the balance between
+incoming  solar radiation and  outgoing infrared  radiation. Radiative
+forcing describes any imbalance  in Earth’s radiation budget caused by
+human  or natural interventions  in the  climate system.   It directly
+measures the capacity of a gas  or other forcing agents to affect that
+energy balance, thereby contributing to climate change.  The intensity
+of  solar energy,  reflectivity  or absorption  of  gases or  surfaces
+(forests, grasslands, oceans, deserts, cloud cover, ice caps, cities),
+emission of heat by various  materials, all differ in their warming or
+cooling contributions to climate forcing.
 
-Aerosols are  suspensions of liquid,  solid or mixed particles  in the
-air  (including sea  salt,  mineral dust,  sulphate, nitrate,  organic
-carbon and  black carbon),  with highly variable  chemical composition
-and  size  distribution \citep{Putaud-2010}.   Although  they are  not
-considered a heat-trapping greenhouse gas and have shorter atmospheric
-lifetimes, they significantly  affect the atmospheric radiative fluxes
-and  are acknowledged  as one  of the  most significant  and uncertain
-aspects    in    anthropogenic    forcing    over   the    last    150
-years \citep{Koch-2009,  IPCC-2013}.  Enhanced aerosols concentrations
-can impact the  climate system by reflecting (e.g.   pure sulfates and
-nitrates)  or  absorbing  (e.g.   black carbon)  solar  radiation  and
-thereby  exert a  cooling or  warming effect  on  the Earth-atmosphere
-system,      causing       a      so-called      direct      radiative
+The radiative forcing  potential of a greenhouse gas  is determined by
+its  atmospheric concentration, warming  capacity, residence  time and
+spatial  distribution.   Long-lived greenhouse  gases  such as  carbon
+dioxide, methane and nitrous oxide, which linger chemically stable and
+homogeneously well-mixed over time scales  of a decade to thousands of
+years  in the  atmosphere,  are major  human-induced climate  drivers;
+whereas  short-lived gases such  as water,  suphur dioxide  and carbon
+monoxide,  which are  abundant heat-trapping  gases but  physically or
+chemically  reactive  to changes  in  temperature,  cause positive  or
+negative  feedbacks effects which  may vary  with forcing  history and
+climate  state. On a  per molecule  basis, other  gases can  have more
+potent  heat-trapping ability  than carbon  dioxide  (e.g. fluorinated
+gases  are 23000  times more  effective \citep{Zehner-2012}),  but are
+fortunately far less abundant in the atmosphere.
+
+One  factor   overwhelmingly  affects  the   uncertainty  in  defining
+anthropogenic  radiative  forcing:  the  effects  of  aerosols  (small
+airborne liquid,  solid or mixed  particles, \citealp{Putaud-2010}) in
+the  atmosphere.  Although  they  are not  considered a  heat-trapping
+greenhouse   gas  and   have  shorter   atmospheric   lifetimes,  they
+significantly affected the atmospheric  radiative fluxes over the last
+150    years   \citep{Koch-2009,   IPCC-2013}.     Enhanced   aerosols
+concentrations can impact the climate system by reflecting (e.g.  pure
+sulfates  and  nitrates)  or  absorbing  (e.g.   black  carbon)  solar
+radiation  and  thereby exert  a  cooling  or  warming effect  on  the
+Earth-atmosphere   system,  causing   a  so-called   direct  radiative
 forcing     \citep{Charlson-1991,    Haywood-2000,    Ramanathan-2001,
-Liao-2005, Bangert-2012, Lundgren-2013}.   Depending on their size and
-chemical composition, they can also  act as cloud condensation and ice
-nuclei, and  thereby influence  the cloud microphysical  processes and
-optical   properties  (cloud  albedo   effect,  \citealp{Twomey-1977})
-through the contribution to  cloud formation.  This results in changes
-in      droplet      concentrations     \citep{Albrecht-1989}      and
+Liao-2005,  Bangert-2012,  Lundgren-2013}.   Depending on  their  size
+distribution  and chemical  composition, they  can also  act  as cloud
+condensation  and   ice  nuclei,  and  thereby   influence  the  cloud
+microphysical   processes  and   optical   properties  (cloud   albedo
+effect,  \citealp{Twomey-1977})  through  the  contribution  to  cloud
+formation.      This     results     in     changes     in     droplet
+concentrations                \citep{Albrecht-1989}                and
 precipitation     \citep{Rosenfeld-2000,    Khain-2008,    Khain-2009,
@@
 investigate critical  questions at  higher resolution and  over longer
-periods, at reduced cost  to the environment.  In Sec.~\ref{sec:1}, we
-introduce the regional  atmospheric model COSMO-ART \citep{Vogel-2009}
-that  accounts for feedbacks  between chemistry,  aerosols, radiation,
-and   clouds.   Sec.~\ref{sec:2}  and   Sec.~\ref{sec:3}  successively
-describe the  hardware and measurement  devices used in  evaluation as
-well  as  energy-aware techniques  employed  to  determine the  energy
-footprint  of  the  considered  application.   Sec.~\ref{sec:4}  shows
-benchmark  results  from  the  Exa2Green  project  and  highlight  the
-critical  components  of  the  application.  Finally,  a  summary  and
-directions for model optimization are given in Sec.~\ref{concl}.
+periods, at reduced cost  to the environment.
+
+In  Sec.~\ref{sec:1},  we  introduce  the regional  atmospheric  model
+COSMO-ART  \citep{Vogel-2009}  that  accounts  for  feedbacks  between
+chemistry,  aerosols, radiation, and  clouds and  additionally specify
+its technical setup related to the investigated performance and energy
+evaluation  methods.   Sec.~\ref{sec:2}  describes the  different  HPC
+systems   and  associated  power   measurement  equipments   used  for
+conducting the benchmarking study and presents energy-aware techniques
+employed  to   determine  the  energy  footprint   of  the  considered
+application.   Sec.~\ref{sec:3}  gathers  benchmark results  from  the
+Exa2Green  project and  highlights  areas where  improvements will  be
+necessary for the subsequent  baseline refactoring.  Some related work
+will  be  discussed  in  Sec.~\ref{sec:4} Finally,  we  conclude  with
+practical  guidelines to identifying  energy saving  opportunities and
+give  directions for further  developments of  the modeling  system in
+Sec.~\ref{concl}.

### 2014-04-04 paper_enahpc14/Introduction.tex
adds latest modifications to the document

@@
 resulting from  anthropogenic activities  such as fossil  fuel burning
-and  deforestation.  Long-lived greenhouse  gases, for  example carbon
-dioxide, methane  and nitrous oxide, are chemically  stable and linger
-homogeneously well-mixed and over time scales of a decade to thousands
-of years in  the atmosphere; whereas short-lived gases  such as water,
-suphur  dioxide  and  carbon  monoxide  which  respond  physically  or
-chemically  to changes  in temperature,  act primarily  as  a feedback
-mechanism.   Some   extremly  powerful  greenhouse   gases,  such  as
-fluorinated gases can even have  a warming effect on the atmosphere up
-to  23000  times  greater  than  carbon  dioxide  \citep{Zehner-2012}.
-Fortunately  the  European  Parliament,  which  has  gained  worldwide
-recognition as a  leader in climate policy, has  passed legislation in
-March 2014  to phase  out fluorinated gas  emissions by  two-thirds by
-2030,  towards  the  internationally  agreed goal  of  keeping  global
-warming  below  2  degrees  Celsius  compared to  the  temperature  in
-pre-industrial times.
+and  deforestation.
 
-Current  cumulative releases of  human-induced gas  emissions enhances
-the greenhouse  effect to  the Earth's atmosphere  and thus  creates a
-large imbalance between incoming solar absorbed radiation and outgoing
-longwave radiation emitted back  to space, through radiative trapping.
-While the  Earth's temperature  is dependent upon  the greenhouse-like
-action of  the atmosphere, the  Earth's radiation balance  is strongly
-influenced by several  other factors such as the  type of surface that
-sunlight first  encounters.  Forests, grasslands,  ocean surfaces, ice
-caps, deserts,  and cities all absorb, reflect,  and radiate radiation
-differently.   Sunlight falling  on a  white glacier  surface strongly
-reflects back into space, resulting  in minimal heating of the surface
-and  lower atmosphere.   Sunlight falling  on  a dark  desert soil  is
-strongly absorbed,  on the other hand, and  contributes to significant
-heating of the surface and lower atmosphere.  Cloud cover also affects
-greenhouse  warming by  both reducing  the amount  of  solar radiation
-reaching the earth's  surface and by reducing the  amount of radiation
-energy emitted into space.
+The Earth's  surface temperature is determined by  the balance between
+incoming  solar radiation and  outgoing infrared  radiation. Radiative
+forcing describes any imbalance  in Earth's radiation budget caused by
+human  or natural interventions  in the  climate system.   It directly
+measures the capacity of a gas  or other forcing agents to affect that
+energy balance, thereby contributing to climate change.  The intensity
+of  solar energy,  reflectivity  or absorption  of  gases or  surfaces
+(forests, grasslands, oceans, deserts, cloud cover, ice caps, cities),
+emission of heat by various  materials, all differ in their warming or
+cooling contributions to climate forcing.
 
-Aerosols are  suspensions of liquid,  solid or mixed particles  in the
-air  (including sea  salt,  mineral dust,  sulphate, nitrate,  organic
-carbon and  black carbon),  with highly variable  chemical composition
-and  size  distribution \citep{Putaud-2010}.   Although  they are  not
-considered a heat-trapping greenhouse gas and have shorter atmospheric
-lifetimes, they significantly  affect the atmospheric radiative fluxes
-and  are acknowledged  as one  of the  most significant  and uncertain
-aspects    in    anthropogenic    forcing    over   the    last    150
-years \citep{Koch-2009,  IPCC-2013}.  Enhanced aerosols concentrations
-can impact the  climate system by reflecting (e.g.   pure sulfates and
-nitrates)  or  absorbing  (e.g.   black carbon)  solar  radiation  and
-thereby  exert a  cooling or  warming effect  on  the Earth-atmosphere
-system,      causing       a      so-called      direct      radiative
+The radiative forcing  potential of a greenhouse gas  is determined by
+its  atmospheric concentration, warming  capacity, residence  time and
+spatial  distribution.   Long-lived greenhouse  gases  such as  carbon
+dioxide, methane and nitrous oxide, which linger chemically stable and
+homogeneously well-mixed over time scales  of a decade to thousands of
+years  in the  atmosphere,  are major  human-induced climate  drivers;
+whereas  short-lived gases such  as water,  suphur dioxide  and carbon
+monoxide,  which are  abundant heat-trapping  gases but  physically or
+chemically  reactive  to changes  in  temperature,  cause positive  or
+negative  feedbacks effects which  may vary  with forcing  history and
+climate  state. On a  per molecule  basis, other  gases can  have more
+potent  heat-trapping ability  than carbon  dioxide  (e.g. fluorinated
+gases  are 23000  times more  effective \citep{Zehner-2012}),  but are
+fortunately far less abundant in the atmosphere.
+
+One  factor   overwhelmingly  affects  the   uncertainty  in  defining
+anthropogenic  radiative  forcing:  the  effects  of  aerosols  (small
+airborne liquid,  solid or mixed  particles, \citealp{Putaud-2010}) in
+the  atmosphere.  Although  they  are not  considered a  heat-trapping
+greenhouse   gas  and   have  shorter   atmospheric   lifetimes,  they
+significantly affected the atmospheric  radiative fluxes over the last
+150    years   \citep{Koch-2009,   IPCC-2013}.     Enhanced   aerosols
+concentrations can impact the climate system by reflecting (e.g.  pure
+sulfates  and  nitrates)  or  absorbing  (e.g.   black  carbon)  solar
+radiation  and  thereby exert  a  cooling  or  warming effect  on  the
+Earth-atmosphere   system,  causing   a  so-called   direct  radiative
 forcing     \citep{Charlson-1991,    Haywood-2000,    Ramanathan-2001,
-Liao-2005, Bangert-2012, Lundgren-2013}.   Depending on their size and
-chemical composition, they can also  act as cloud condensation and ice
-nuclei, and  thereby influence  the cloud microphysical  processes and
-optical   properties  (cloud  albedo   effect,  \citealp{Twomey-1977})
-through the contribution to  cloud formation.  This results in changes
-in      droplet      concentrations     \citep{Albrecht-1989}      and
+Liao-2005,  Bangert-2012,  Lundgren-2013}.   Depending on  their  size
+distribution  and chemical  composition, they  can also  act  as cloud
+condensation  and   ice  nuclei,  and  thereby   influence  the  cloud
+microphysical   processes  and   optical   properties  (cloud   albedo
+effect,  \citealp{Twomey-1977})  through  the  contribution  to  cloud
+formation.      This     results     in     changes     in     droplet
+concentrations                \citep{Albrecht-1989}                and
 precipitation     \citep{Rosenfeld-2000,    Khain-2008,    Khain-2009,
@@
 investigate critical  questions at  higher resolution and  over longer
-periods, at reduced cost  to the environment.  In Sec.~\ref{sec:1}, we
-introduce the regional  atmospheric model COSMO-ART \citep{Vogel-2009}
-that  accounts for feedbacks  between chemistry,  aerosols, radiation,
-and clouds and additionally specify its technical setup related to the
-investigated    performance    and     energy    evaluation    methods.
-Sec.~\ref{sec:2}  describes the different  HPC systems  and associated
-power  measurement  equipments used  for  conducting the  benchmarking
-study and  presents energy-aware techniques employed  to determine the
-energy  footprint  of  the considered  application.   Sec.~\ref{sec:3}
-gathers benchmark  results from  the Exa2Green project  and highlights
-areas where improvements will be necessary for the subsequent baseline
-refactoring.  Some related work  will be discussed in Sec.~\ref{sec:4}
-Finally, we  conclude with practical guidelines  to identifying energy
-saving opportunities  and give directions for  further developments of
-the modeling system in Sec.~\ref{concl}.
+periods, at reduced cost  to the environment.
+
+In  Sec.~\ref{sec:1},  we  introduce  the regional  atmospheric  model
+COSMO-ART  \citep{Vogel-2009}  that  accounts  for  feedbacks  between
+chemistry,  aerosols, radiation, and  clouds and  additionally specify
+its technical setup related to the investigated performance and energy
+evaluation  methods.   Sec.~\ref{sec:2}  describes the  different  HPC
+systems   and  associated  power   measurement  equipments   used  for
+conducting the benchmarking study and presents energy-aware techniques
+employed  to   determine  the  energy  footprint   of  the  considered
+application.   Sec.~\ref{sec:3}  gathers  benchmark results  from  the
+Exa2Green  project and  highlights  areas where  improvements will  be
+necessary for the subsequent  baseline refactoring.  Some related work
+will  be  discussed  in  Sec.~\ref{sec:4} Finally,  we  conclude  with
+practical  guidelines to identifying  energy saving  opportunities and
+give  directions for further  developments of  the modeling  system in
+Sec.~\ref{concl}.

### 2014-04-04 paper_enahpc14/Sec2:Power-measurement.tex
adds latest modifications to the document

--- 2014-04-02 trunk/paper_enahpc14/Sec2:Power-measurement.tex
@@
 \subsection{Hardware description}
-\label{subsec:3.1}
+\label{subsec:2.1}
 
@@
 that  the baseline  benchmark  can be  reproduced  within an  expected
-variance   (cf  previous   section),   and  that   the  baseline   run
+variance (see subsection~\ref{subsec:1.2}),  and that the baseline run
 configuration can be  used in all future versions of  the code, a fair
@@
 
-A snapshot of the code, which includes, at least conceptually, all the
-information needed  to reproduce the  energy-to-solution benchmarks of
-COSMO-ART, was  produced and run  on a full  rack of 52  nodes without
-hyperthreading (hyperthreading was attempted, but always led to higher
-energy-to-solution), i.e. on a total  of 1040 cores using 20 tasks per
-node.
-
 \subsection{Power-Performance measurement framework}
-\label{subsec:3.2}
+\label{subsec:2.2}
 

### 2014-04-04 paper_enahpc14/Sec3:Evaluation.tex
adds latest modifications to the document

\subsection{Sofware environment}
\label{subsec:3.1}
The  software environment  on Monch  is controlled  using  the modules
framework which gives an easy  and flexible mechanism to access to all
of  the  CSCS  provided  compilers,  tools and  applications.   It  is
particularly  useful for  testing code  portability  between different
compilers  or for  changing  between different  versions  of the  same
compiler.  INT2LM  and the COSMO-ART model are  implemented in Fortran
90 for distributed memory parallel computers using the Message Passing
Interface (MPI).  For  our initial benchmarking, we opted  for the GNU
compiler (gcc/4.8.1) with the -O3  compiler flag as it generally gives
a  good  level  of optimization  and  the  code  runs faster  in  this
configuration  than  when  compiled  with  the  intel  compiler,  also
available on  Monch.  Besides, we installed  the MPICH2 implementation
of MPI (mvapich2/1.9) as well  as the commonly used HDF5 (hdf5/1.8.12)
and NetCDF  (netcdf/4.3.1) libraries in favor of  the traditional GRIB
library  for  the  management  of  extremely large  and  complex  data
collections.    Note   that   the  Linux   2.6.32-358.11.1.el6.x86\_64
operating system was used for all the Monch compute nodes.

\subsection{Run configuration}
\label{subsec:3.2}
The COSMO-ART model uses NAMELIST-input to specify runtime parameters
splitted into several groups:

\begin{itemize}
\item LMGRID: specifying the domain and the size of the grid,
\item RUNCTL: parameters for the model run,
\item TUNING: parameters for tuning physics and dynamics,
\item DYNCTL: parameters for the adiabatic model,
\item PHYCTL: parameters for the diabatic model,
\item COSMO\_ART: parameters for gases and aerosols model,
\item DIACTL: parameters for the diagnostic calculations,
\item SATCTL: controlling computation of synthetic satellite images,
\item IOCTL: controlling the I/O environment,
\item GRIBIN: controlling the grib input,
\item GRIBOUT: controlling the grib output,
\end{itemize}

\noindent
To run COSMO-ART the following input data are necessary:
\begin{itemize}
\item  Gas phase:  Anthropogenic emissions  for different  species and
  land use data for biogenic emissions and deposition,
\item Aerosol particles: Anthropogenic emissions,
\item Mineral dust: Soil specific land use data.
\end{itemize}

A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was  produced and run on a  full rack of 52  nodes on Monch
(monchc[029-080]),  which represents a  total of  1040 cores  using 20
tasks per node.  The calculated region was mapped to the participating
processors  using a 2D-partitioning  strategy. The  distribution along
the  x and  y coordinates  is defined  in the  namelist  INPUT\_ORG by
setting:  $nprocx=40$   and  $nprocy=26$,  as  the   total  number  of
processors has to be equal to $nprocx \times nprocy$.  Besides we take
$nprocio=0$.   Hyperthreading  is  not  considered in  this  study  as
previous attempts  of its  use revealed that  it always led  to higher
energy-to-solution.\\

Multiple production runs of COSMO-ART were performed to illustrate the
reproducibility  of   the  baseline,  and   quantify  the  significant
uncertainties in  the power measurement, as dictated  by the available
technology (see subsection~\ref{subsec:2.2}).

\subsection{Power-measurement results}
\label{subsec:3.3}
Measuring  the power  consumption on  one entire  Cray cabinet  is the
easiest way to calculate energy-to-solution for given applications.

\begin{figure*}
  \includegraphics[width=0.4\textwidth]{Figs/NRJ_benchmark_Monch.eps}
  \caption{Isola E1 Rack 2 Total Power and Isola E1 Total Power}
  \label{fig:1}
\end{figure*}

The run  was issued successfully  twice, the start time  and execution
time for both jobs are:
\begin{itemize}
\item start=12:51:03, end=13:20:46 $\Rightarrow$ T = 00:29:43 = 1783 s
\item start=14:28:40, end=14:58:02 $\Rightarrow$ T = 00:29:22 = 1762 s
\end{itemize}

The power  consumption for  both jobs is  calculated by  averaging the
corresponding power  measurements from the Excel  file.  Power results
account for the  Isola E1 Rack 2 Total Power. As  time resolution is 5
minutes  for the  output  results, the  average  power consumption  is
computed   by  considering  6   values  for   each  single   run  (see
Table~\ref{tab:1}).

\begin{table}
  \begin{center}
    \caption{}
    \label{tab:1}
    \begin{tabular}{lll}
      \hline\noalign{\smallskip}
      Time & Measured power (W) & Average power (W)  \\
      \noalign{\smallskip}\hline\noalign{\smallskip}
      12:55:00 & 1.2355833333e+04 & 1.265852278e+04 \\ 
      13:00:00 & 1.2820266667e+04 &  \\
      13:05:00 & 1.2600743333e+04 &  \\ 
      13:10:00 & 1.2811580000e+04 &  \\
      13:15:00 & 1.2609680000e+04 &  \\
      13:20:00 & 1.2753033333e+04 &  \\
      \noalign{\smallskip}\hline\noalign{\smallskip}
      14:30:00 & 1.2019266667e+04 & 1.258640833e+04 \\ 
      14:35:00 & 1.2632253333e+04 &  \\
      14:40:00 & 1.2779040000e+04 &  \\
      14:45:00 & 1.2753180000e+04 &  \\
      14:50:00 & 1.2610476667e+04 &  \\
      14:55:00 & 1.2724233333e+04 &  \\
      \noalign{\smallskip}\hline
    \end{tabular}
  \end{center}
\end{table}

The  total  energy to  solution  is  the straightforward  calculation:\\

Energy-to-solution  =  Integral  of  power consumption  over  the  job
duration

Often  the  average  power  consumption  times  the  job  duration  is
sufficient.

Thus, the total energy consumption corresponds to: 
\begin{itemize}
\item E  = 1783 x 12658.52278  = 22570146.11674 J $\sim$  22.57 MJ per
  day of simulation
\item E  = 1762 x 12586.40833  = 22177251.47746 J $\sim$  22.18 MJ per
  day of simulation
\end{itemize}

### 2014-04-04 trunk/paper_enahpc14/Sec1:COSMO-ART.tex
adds latest modifications to the document

@@
 continental scale model coupled  online to the COSMO numerical weather
-prediction  (NWP) and  climate  model \citep{Baldauf-2011}.   Physical
-processes  like  transport,  turbulent  diffusion,  and  dry  and  wet
-deposition  are  treated  together  with  photochemistry  and  aerosol
-dynamics  using the  modal approach.
+prediction (NWP) and climate model \citep{Baldauf-2011}. It allows the
+online calculation of reactive  trace substances and their interaction
+with  the atmosphere.   Physical processes  like  transport, turbulent
+diffusion,  and  dry and  wet  deposition  are  treated together  with
+photochemistry and aerosol dynamics using the modal approach.
 
@@
 \label{subsec:1.2}
-To define a baseline within a code under development, it was necessary
-to  find  a  run-configuration  capable  of  being  recreated  in  all
-subsequent versions. The  energy-to-solution benchmarking of COSMO-ART
-concerns  one-day  long simulations  without  spin-up  stage, using  a
-222x216x40  discretization of  Europe and  a  time step  of 120s.  The
-baseline code incorporates 34 2-d and  45 3-d fields to be written out
-every hour for a simulation starting  on April 13th, which is close to
-the equinox and thus brings  benefits of having approximately half day
-of night and  half day of sun exposure, to  ensure a proper activation
-of  the chemistry  cycle. Ultimately  new  input files  closer to  the
-equinox (on March 20th) will be  created, but the current set of input
-files is fine for the time being.
+Establishing effective energy performance benchmarking of a code under
+intense development such as COSMO-ART is a challenging task because of
+the  absolute necessity that  results must  be reproducible  within an
+expected  variance for the  duration of  the Exa2Green  project. After
+consultation  with  the  climate  community  to  properly  define  the
+baseline,  it was  necessary to  find a  run-configuration  capable of
+being recreated in all subsequent versions of the code.
+
+Thus, we give here a brief  overview of the main features of the model
+setup for  our benchmark's daily  runs.  Three-dimensional simulations
+are performed over  Europe for April 13th 2010, which  is close to the
+equinox and thus  brings benefits of having approximately  half day of
+night  and half  day  of  sun exposure,  therefore  ensuring a  proper
+activation of the chemistry cycle. We consider daily 24-hour forecasts
+without  any  previous spin-up  simulations.   The calculation  domain
+corresponds to  the CORDEX-EU-44  domain and is  covered by a  grid of
+$222\times   216$    points   with   a    horizontal   resolution   of
+$0.22\,^{\circ}$,  i.e.  50  km in  both directions,  and  40 vertical
+layers.   Boundary  and initial  conditions,  emmissions and  external
+parameters  were obtained by  the INT2LM  pre-processing interpolation
+tool, which interpolate  data from a coarse-grid driving  model to the
+rotated    latitude-longitude   grid    of   the    COSMO-Model,   and
+INT2LM\_trcr\_ART,  which  interpolates  aerosol  or  gaseous  species
+initial and boundary conditions  and consists mostly of the definition
+of a large number of  tracers.  The meteorological initial and bounday
+conditions are  obtained by  the the ECMWF  global spectral  model IFS
+with an update  frequency of 3h.  Boundary data  for gas-phase species
+are taken from IFS-MOZART output at 6h temporal resolution.  The model
+setup incorporates  34 2-d and 45  3-d fields to be  written out every
+hour and doesn't include data assimilation methods.
 
@@
 filling diffusion option in combination with the Runge-Kutta dynamical
-core.   Concerning the modelling  of wet  deposition in  aerosols, the
-baseline  has  only  indirect  cloud  feedbacks  but  doesn't  include
-in-cloud  scavenging (rainout)  and  below-cloud scavenging  (washout)
-yet.  Amongst  physical parameterizations, precipitation  formation is
+core. It  also makes  use of the  Kinetic PreProcessor solver  for the
+resolution of  atmospheric chemistry ordinary  differential equations.
+Concerning the  modelling of wet deposition in  aerosols, the baseline
+has  only  indirect  cloud  feedbacks  but  doesn't  include  in-cloud
+scavenging  (rainout)   and  below-cloud  scavenging   (washout)  yet.
+Amongst   physical  parameterizations,   precipitation   formation  is
 performed  by a  two-moment  cloud microphysics  scheme  instead of  a
-classical  bulk microphysics  scheme. Another  important point  is the
-fact that  this version makes  use of the Kinetic  PreProcessor solver
-for  the  resolution of  atmospheric  chemistry ordinary  differential
-equations.
+classical  bulk microphysics  scheme.

### 2014-04-07 trunk/paper_enahpc14/paper.tex
Manuel's changes
adds Pilatus results

@@
 \title{Evaluating the Performance  and Energy Efficiency of COSMO-ART,
-  a Fully Online Coupled Model System with Numerical Weather
-    Forecast and Chemical Transport Models }
+  a Coupled Numerical Weather Forecast and Chemical Transport Model}
 % \thanks{This work was supported by the EU Project FP7 318793 ``EXA2GREEN''}}
@@
 \label{concl}
+\input{Conclusion.tex}
 
 %\paragraph{Paragraph headings} Use paragraph headings as needed.
-
-%\begin{figure*}
-%  \includegraphics[width=0.4\textwidth]{Figs/earth.eps}
-%  \caption{Please write your figure caption here}
-%  \label{fig:1}
-%\end{figure*}
-
-%\begin{table}
-%  \caption{Please write your table caption here}
-%  \label{tab:1}
-%  \begin{tabular}{lll}
-%    \hline\noalign{\smallskip}
-%    first & second & third  \\
-%    \noalign{\smallskip}\hline\noalign{\smallskip}
-%    number & number & number \\
-%    number & number & number \\
-%    \noalign{\smallskip}\hline
-%  \end{tabular}
-%\end{table}
 

### 2014-04-07 paper_enahpc14/Sec2:Power-measurement.tex
adds Pilatus results

@@
 \label{subsec:2.1}
+While hardware  platforms mature  and are replaced,  we have  chosen a
+state-of-the-art  Intel's third-generation  Core (aka  ``Ivy Bridge'')
+processing platform (called ``Monch'')  for our power measurements, as
+it  is slated  to stay  in service  without hardware  upgrade  for the
+duration  of  the Exa2Green  project.   In  principle  at least,  this
+architecture could be recreated or found in an identical configuration
+beyond the lifetime of the project.  Given that the baseline benchmark
+can    be    reproduced    within    an   expected    variance    (see
+subsection~\ref{subsec:1.2}), and that  the baseline run configuration
+can be used in all future versions of the code, a fair comparison will
+be made between the baseline  and the milestone versions of COSMO-ART.
+In  this   study,  a  complementary   energy-to-solution  benchmarking
+comparison is carried out on  the ``Pilatus'' cluster based on Intel's
+previous-generation  ``Sandy  Bridge'' processors,  known  to be  more
+power consuming.
 
 \subsubsection{Monch (CSCS - ETH Zurich)}
-While hardware  platforms mature  and are replaced,  we have  chosen a
-state-of-the-art  dual-socket Intel  Ivybridge-based  platform (called
-``Monch'')  for our power  measurements, as  it is  slated to  stay in
-service without upgrade for the duration of the Exa2Green project.  In
-principle at least,  this architecture could be recreated  or found in
-an identical configuration beyond  the lifetime of the project.  Given
-that  the baseline  benchmark  can be  reproduced  within an  expected
-variance (see subsection~\ref{subsec:1.2}),  and that the baseline run
-configuration can be  used in all future versions of  the code, a fair
-comparison  will  be  made  between  the baseline  and  the  milestone
-versions of COSMO-ART.
-
 Monch, which is installed  at the Swiss National Supercomputing Center
-(CSCS)  of  the  ETH  Zurich,  is a  10  rack  NEC-provided  (LX-2400)
-Intel-based cluster, utilised  from people that are part  of the Swiss
-Platform  for  Advanced Scientific  Computing  (PASC),  which aims  at
-positioning  Swiss  computational sciences  in  the emerging  exascale
-era. It is  composed of 16 administration nodes,  312 standard compute
-nodes, 24 large-memory compute nodes, 24 huge-memory compute nodes and
-8  servers to  provide  I/O  to Lustre.   Each  standard compute  node
-comprises 2 Intel 10-core Ivy Bridge EP E5-2660v2 processors operating
-at  2.2 GHz,  offering 32GB  of DDR3  1600MHz RAM,  with a  base clock
-frequency of 2.2 GHz, themselves  connected by a high speed Infiniband
+(CSCS)  of the ETH  Zurich, is  a 10  rack NEC-provided  (LX-2400) and
+dual-socket Intel-based cluster, utilised from people that are part of
+the   Swiss  Platform   for  Advanced   Scientific   Computing  (PASC,
+\url{http://www.pasc-ch.org/}).   It  is   composed  of  312  standard
+compute  nodes,  24  large-memory  compute nodes  and  24  huge-memory
+compute nodes.   Each standard compute node comprises  2 Intel 10-core
+Ivy Bridge EP E5-2660v2 processors operating at 2.2 GHz, offering 32GB
+of DDR3 1600MHz  RAM, themselves connected by a  high speed Infiniband
 network,  based on Mellanox  SX6036 managed  FDR (Fourteen  Data Rate)
 switches, with a 56 Gb/s speed.
+
+\subsubsection{Pilatus (CSCS - ETH Zurich)}
+Pilatus is a dual-socket Intel  Sandy-Bridge based cluster used as Piz
+Daint pre-post processing cluster.  It is composed of 42 compute nodes
+with  a total  number of  1344 cores  in hyper-threading  and  2688 GB
+aggregate memory and has 2  high-speed interconnects based on FDR: the
+first is  dedicated to the MPI  traffic and the second  to the storage
+high  speed traffic.  The  2 login  nodes  and the  42 computes  nodes
+consists in  11 twin-pair Intel  E5-Series DALCO r2264i4t  2U scalable
+compute  modules.  Each  module contains  4 compute  nodes based  on 2
+Intel  8-core Intel  Xeon  E5-2670 processors  operating  at 2.6  GHz,
+offering  64GB of  DDR3 1600MHz  RAM, themselves  connected by  a high
+speed  Infiniband  network,  based  on  Mellanox  SX6036  managed  FDR
+(Fourteen Data Rate) switches, with a 56 Gb/s speed.
 
@@
 the  power   data  from  the  tracing   server.   The  instrumentation
-post-process  generates the  performance  trace files  and the  \pmlib
+post-processing generates  the performance trace files  and the \pmlib
 plugin inserts the power data into them.

### 2014-04-07 paper_enahpc14/Sec3:Evaluation.tex
adds Pilatus results

@@
 \label{subsec:3.1}
-The  software environment  on Monch  is controlled  using  the modules
-framework which gives an easy  and flexible mechanism to access to all
-of  the  CSCS  provided  compilers,  tools and  applications.   It  is
+The software environment on Monch  and Pilatus is controlled using the
+modules framework which gives an easy and flexible mechanism to access
+to all of the CSCS  provided compilers, tools and applications.  It is
 particularly  useful for  testing code  portability  between different
@@
 Interface (MPI).  For  our initial benchmarking, we opted  for the GNU
-compiler (gcc/4.8.1) with the -O3  compiler flag as it generally gives
-a  good  level  of optimization  and  the  code  runs faster  in  this
-configuration  than  when  compiled  with  the  intel  compiler,  also
-available on  Monch.  Besides, we installed  the MPICH2 implementation
-of MPI (mvapich2/1.9) as well  as the commonly used HDF5 (hdf5/1.8.12)
-and NetCDF  (netcdf/4.3.1) libraries in favor of  the traditional GRIB
-library  for  the  management  of  extremely large  and  complex  data
-collections.    Note   that   the  Linux   2.6.32-358.11.1.el6.x86\_64
-operating system was used for all the Monch compute nodes.
+compiler  (gcc/4.8.1 on  Monch,  gcc/4.8.2 on  Pilatus)  with the  -O3
+compiler flag as  it generally gives a good  level of optimization and
+the code runs faster in this configuration than when compiled with the
+intel compiler  (intel/14.0.1), also available on  Monch.  Besides, we
+installed the  MPICH2 implementation of MPI (mvapich2/1.9)  as well as
+the  commonly  used   HDF5  (hdf5/1.8.12)  and  NetCDF  (netcdf/4.3.1)
+libraries in favor of the  traditional GRIB library for the management
+of  extremely  large and  complex  data  collections.   Note that  all
+compute  nodes of  Monch  and Pilatus  used  respectively the  ``Linux
+2.6.32-358.11.1.el6.x86\_64''   and   ``Linux   3.0.101-0.15-default''
+operating systems.
 
@@
 COSMO-ART, was  produced and run on a  full rack of 52  nodes on Monch
-(monchc[029-080]),  which represents a  total of  1040 cores  using 20
-tasks per node.  The calculated region was mapped to the participating
-processors  using a 2D-partitioning  strategy. The  distribution along
-the  x and  y coordinates  is defined  in the  namelist  INPUT\_ORG by
-setting:  $nprocx=40$   and  $nprocy=26$,  as  the   total  number  of
-processors has to be equal to $nprocx \times nprocy$.  Besides we take
-$nprocio=0$.   Hyperthreading  is  not  considered in  this  study  as
-previous attempts  of its  use revealed that  it always led  to higher
-energy-to-solution.\\
+(monchc[029-080]), i.e.  a total of 1040 cores using 20 tasks per node
+and on  a full rack of  42 nodes on Pilatus  (pilatus[03-44]), i.e.  a
+total of  1344 cores using 16  tasks per node.   The calculated region
+was  mapped to  the participating  processors using  a 2D-partitioning
+strategy.  The distribution along the  x and y coordinates was defined
+by setting: $nprocx=40$ and  $nprocy=26$ for Monch and $nprocx=28$ and
+$nprocy=24$.  Besides as this version of COSMO-ART doesn't make use of
+the   GRIB   library,  we   specified   $nprocio=0$   for  GRIB   I/O.
+Hyperthreading is not considered in this study as previous attempts of
+its use revealed that it always led to higher energy-to-solution.\\
 
@@
 Measuring  the power  consumption on  one entire  Cray cabinet  is the
-easiest way to calculate energy-to-solution for given applications.
+easiest    way    to    calculate   energy-to-solution    for    given
+applications.  The total  energy  to solution  is the  straightforward
+calculation:\\
 
+Energy-to-solution  =  Integral  of  power consumption  over  the  job
+duration\\
+
+However, the average power consumption times the job duration is often
+sufficient.
+
+\subsubsection{Monch (CSCS - ETH Zurich)}
 \begin{figure*}
@@
 
-The run  was issued successfully  twice, the start time  and execution
-time for both jobs are:
+The 1-day simulation was issued successfully twice, the start time and
+execution time for both runs are:
 \begin{itemize}
@@
 
-The  total  energy to  solution  is  the straightforward  calculation:\\
-
-Energy-to-solution  =  Integral  of  power consumption  over  the  job
-duration
-
-Often  the  average  power  consumption  times  the  job  duration  is
-sufficient.
-
 Thus, the total energy consumption corresponds to: 
@@
 \end{itemize}
+
+\subsubsection{Pilatus (CSCS - ETH Zurich)}
+\begin{figure*}
+  \includegraphics[width=0.4\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
+  \caption{Isola HD Pilatus Total Power}
+  \label{fig:2}
+\end{figure*}
+
+The 1-day simulation was issued successfully four times and the 2-days
+run only once, the start time and execution time for all jobs are:
+\begin{itemize}
+\item start=13:07:25, end=13:29:47 $\Rightarrow$ T = 00:22:22 = 1342 s
+\item start=13:46:04, end=14:08:30 $\Rightarrow$ T = 00:22:26 = 1346 s
+\item start=14:27:51, end=14:50:01 $\Rightarrow$ T = 00:22:10 = 1330 s
+\item start=15:10:04, end=15:32:17 $\Rightarrow$ T = 00:22:13 = 1333 s
+\item start=15:41:48, end=16:26:13 $\Rightarrow$ T = 00:44:25 = 2665 s
+\end{itemize}
+
+Again, the power  consumption for all jobs is  calculated by averaging
+the  corresponding  power  measurements  from the  Excel  file.  Power
+results  account  for  the  Isola  HD Pilatus  Total  Power.  As  time
+resolution  is 5  minutes for  the output  results, the  average power
+consumption is computed by considering  4 values for each single 1-day
+run and 9 values for the 2-days run (see Table~\ref{tab:2}).
+
+\begin{table}
+  \begin{center}
+    \caption{}
+    \label{tab:2}
+    \begin{tabular}{lll}
+      \hline\noalign{\smallskip}
+      Time & Measured power (W) & Average power (W)  \\
+      \noalign{\smallskip}\hline\noalign{\smallskip}
+       13:10:00 & 1.7824080000e+04 & 1.812201417e+04 \\ 
+       13:15:00 & 1.8293650000e+04 &  \\ 
+       13:20:00 & 1.8125140000e+04 &  \\ 
+       13:25:00 & 1.8245186667e+04 &  \\ 
+      \noalign{\smallskip}\hline\noalign{\smallskip}
+       13:50:00 & 1.7849206667e+04 & 1.797961083e+04 \\ 
+       13:55:00 & 1.7977790000e+04 &  \\ 
+       14:00:00 & 1.8030646667e+04 &  \\ 
+       14:05:00 & 1.8060800000e+04 &  \\ 
+      \noalign{\smallskip}\hline\noalign{\smallskip}
+       14:30:00 & 1.7868826667e+04 & 1.806545167e+04 \\ 
+       14:35:00 & 1.8115786667e+04 &  \\ 
+       14:40:00 & 1.8188520000e+04 &  \\ 
+       14:45:00 & 1.8088673333e+04 &  \\ 
+      \noalign{\smallskip}\hline\noalign{\smallskip}
+       15:15:00 & 1.7857800000e+04 & 1.797302833e+04 \\ 
+       15:20:00 & 1.7962513333e+04 &  \\ 
+       15:25:00 & 1.8078226667e+04 &  \\ 
+       15:30:00 & 1.7993573333e+04 &  \\ 
+      \noalign{\smallskip}\hline\noalign{\smallskip}
+       15:45:00 & 1.7801980000e+04 & 1.799757815e+04 \\ 
+       15:50:00 & 1.7925386667e+04 &  \\ 
+       15:55:00 & 1.8076986667e+04 &  \\ 
+       16:00:00 & 1.8082966667e+04 &  \\ 
+       16:05:00 & 1.7903206667e+04 &  \\ 
+       16:10:00 & 1.8063913333e+04 &  \\ 
+       16:15:00 & 1.8355030000e+04 &  \\ 
+       16:20:00 & 1.7971580000e+04 &  \\ 
+       16:25:00 & 1.7797153333e+04 &  \\ 
+      \noalign{\smallskip}\hline
+    \end{tabular}
+  \end{center}
+\end{table}
+
+Thus, the total energy consumption corresponds to: 
+\begin{itemize}
+\item E = 1342 x 18122.01417 = 24319743.01614 J $\sim$ 24.32 MJ per day of simulation
+\item E = 1346 x 17979.61083 = 24200556.17718 J $\sim$ 24.20 MJ per day of simulation
+\item E = 1330 x 18065.45167 = 24027050.72110 J $\sim$ 24.03 MJ per day of simulation
+\item E = 1333 x 17973.02833 = 23958046.76389 J $\sim$ 23.96 MJ per day of simulation
+\item E = 2665 x 17997.57815 = 47963545.76975 J $\sim$ 47.96 MJ per day of simulation
+\end{itemize}

### 2014-04-07 paper_enahpc14/Sec3:Evaluation.tex
updates tables

@@
 The COSMO-ART model uses NAMELIST-input to specify runtime parameters
-splitted into several groups:
+splitted into several groups (see Table~\ref{tab:1}).
 
-\begin{itemize}
-\item LMGRID: specifying the domain and the size of the grid,
-\item RUNCTL: parameters for the model run,
-\item TUNING: parameters for tuning physics and dynamics,
-\item DYNCTL: parameters for the adiabatic model,
-\item PHYCTL: parameters for the diabatic model,
-\item COSMO\_ART: parameters for gases and aerosols model,
-\item DIACTL: parameters for the diagnostic calculations,
-\item SATCTL: controlling computation of synthetic satellite images,
-\item IOCTL: controlling the I/O environment,
-\item GRIBIN: controlling the grib input,
-\item GRIBOUT: controlling the grib output,
-\end{itemize}
+\begin{table}[htbf]
+  \begin{center}
+    \caption{}
+    \label{tab:1}
+    \begin{tabular}{ll}
+      \hline\noalign{\smallskip} 
+      \textbf{Group} & \textbf{Description} \\
+      \noalign{\smallskip}\hline\noalign{\smallskip}
+      LMGRID & Grid domain and size parameters \\
+      RUNCTL & Model run parameters \\
+      TUNING & Physics and dynamics parameters \\
+      DYNCTL & Adiabatic model parameters \\
+      PHYCTL & Diabatic model parameters \\
+      COSMO\_ART & Gases and aerosols model parameters \\
+      DIACTL & Diagnostic calculations parameters \\
+      SATCTL & Synthetic satellite images parameters \\
+      IOCTL & I/O environment parameters \\
+      GRIBIN & GRIB input parameters \\
+      GRIBOUT & GRIB output parameters \\
+     \noalign{\smallskip}\hline
+    \end{tabular}
+  \end{center}
+\end{table}
 
@@
 \subsubsection{Monch (CSCS - ETH Zurich)}
-\begin{figure*}
+\begin{figure}[htbf]
   \includegraphics[width=0.4\textwidth]{Figs/NRJ_benchmark_Monch.eps}
@@
   \label{fig:1}
-\end{figure*}
+\end{figure}
 
@@
 computed   by  considering  6   values  for   each  single   run  (see
-Table~\ref{tab:1}).
+Table~\ref{tab:2}).
 
@@
     \caption{}
-    \label{tab:1}
+    \label{tab:2}
     \begin{tabular}{lll}
       \hline\noalign{\smallskip}
-      Time & Measured power (W) & Average power (W)  \\
+      \textbf{Time} & \textbf{Measured power (W)} & \textbf{Average power (W)} \\
       \noalign{\smallskip}\hline\noalign{\smallskip}
@@
 \subsubsection{Pilatus (CSCS - ETH Zurich)}
-\begin{figure*}
+\begin{figure}[htbf]
   \includegraphics[width=0.4\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
@@
   \label{fig:2}
-\end{figure*}
+\end{figure}
 
@@
 consumption is computed by considering  4 values for each single 1-day
-run and 9 values for the 2-days run (see Table~\ref{tab:2}).
+run and 9 values for the 2-days run (see Table~\ref{tab:3}).
 
@@
     \caption{}
-    \label{tab:2}
+    \label{tab:3}
     \begin{tabular}{lll}
       \hline\noalign{\smallskip}
-      Time & Measured power (W) & Average power (W)  \\
+      \textbf{Time} & \textbf{Measured power (W)} & \textbf{Average power (W)}  \\
       \noalign{\smallskip}\hline\noalign{\smallskip}

### 2014-04-08 paper_enahpc14/Sec1:COSMO-ART.tex
adjusts structure of the document

@@
-\subsection{Model description}
+\subsection{Model Description}
 \label{subsec:1.1}
@@
 
-\subsection{Model setup}
+\subsection{Model Setup}
 \label{subsec:1.2}
@@
 setup for  our benchmark's daily  runs.  Three-dimensional simulations
-are  performed over  Europe  for April  13th,  which is  close to  the
+are performed over  Europe for April 13th 2010, which  is close to the
 equinox and thus  brings benefits of having approximately  half day of
@@
 parameters  were obtained by  the INT2LM  pre-processing interpolation
-tool, which  performed the usual interpolation from  coarse grid model
-data, and  INT2LM\_trcr\_ART, which consists mostly  of the definition
+tool, which interpolate  data from a coarse-grid driving  model to the
+rotated    latitude-longitude   grid    of   the    COSMO-Model,   and
+INT2LM\_trcr\_ART,  which  interpolates  aerosol  or  gaseous  species
+initial and boundary conditions  and consists mostly of the definition
 of a large number of  tracers.  The meteorological initial and bounday
@@
 with an update  frequency of 3h.  Boundary data  for gas-phase species
-are taken from IFS-MOZART  output at 6h temporal resolution.  Multiple
-runs of COSMO-ART were  performed to illustrate the reproducibility of
-the baseline, and quantify  the significant uncertainties in the power
-measurement,   as   dictated   by   the  available   technology   (see
-subsection~\ref{subsec:3.2}).   The run-configuration  incorporates 34
-2-d and 45 3-d fields to be written out every hour and doesn't include
-data assimilation methods.
+are taken from IFS-MOZART output at 6h temporal resolution.  The model
+setup incorporates  34 2-d and 45  3-d fields to be  written out every
+hour and doesn't include data assimilation methods.
 

### 2014-04-09 paper_enahpc14/Sec3:Evaluation.tex
modifies "Evaluation" section; adds figure on T2S and E2S

--- 2014-04-04 paper_enahpc14/Sec3:Evaluation.tex
@@
 \label{subsec:3.1}
-The  software environment  on Monch  is controlled  using  the modules
-framework which gives an easy  and flexible mechanism to access to all
-of  the  CSCS  provided  compilers,  tools and  applications.   It  is
-particularly  useful for  testing code  portability  between different
-compilers  or for  changing  between different  versions  of the  same
-compiler.  INT2LM  and the COSMO-ART model are  implemented in Fortran
-90 for distributed memory parallel computers using the Message Passing
-Interface (MPI).  For  our initial benchmarking, we opted  for the GNU
-compiler (gcc/4.8.1) with the -O3  compiler flag as it generally gives
-a  good  level  of optimization  and  the  code  runs faster  in  this
-configuration  than  when  compiled  with  the  intel  compiler,  also
-available on  Monch.  Besides, we installed  the MPICH2 implementation
-of MPI (mvapich2/1.9) as well  as the commonly used HDF5 (hdf5/1.8.12)
-and NetCDF  (netcdf/4.3.1) libraries in favor of  the traditional GRIB
+INT2LM  and the  COSMO-ART model  are  implemented in  Fortran 90  for
+distributed  memory  parallel  computers  using  the  Message  Passing
+Interface (MPI).   The software stack on both  many-core platforms was
+controlled  using  the  modules  framework  which gives  an  easy  and
+flexible mechanism  to access to  all of the CSCS  provided compilers,
+tools and  applications.  For our  initial benchmarking, we  opted for
+the GNU compiler  (gcc/4.8.1 on Monch, gcc/4.8.2 on  Pilatus) with the
+-O3 compiler flag  as it generally gives a  good level of optimization
+and the code runs faster in this configuration than when compiled with
+the  intel  compiler  (14.0.1).   Besides,  we  installed  the  MPICH2
+implementation of MPI (mvapich2/1.9) as well as the commonly used HDF5
+(1.8.12) and NetCDF (4.3.1) libraries in favor of the traditional GRIB
 library  for  the  management  of  extremely large  and  complex  data
-collections.    Note   that   the  Linux   2.6.32-358.11.1.el6.x86\_64
-operating system was used for all the Monch compute nodes.
+collections.   The Linux  kernels  ``2.6.32-358.11.1.el6.x86\_64'' and
+``3.0.101-0.15-default'' were  used as compute nodes OS  for Monch and
+Pilatus respectively.
 
@@
 The COSMO-ART model uses NAMELIST-input to specify runtime parameters
-splitted into several groups:
+splitted into several groups (see Table~\ref{tab:1}).
 
-\begin{itemize}
-\item LMGRID: specifying the domain and the size of the grid,
-\item RUNCTL: parameters for the model run,
-\item TUNING: parameters for tuning physics and dynamics,
-\item DYNCTL: parameters for the adiabatic model,
-\item PHYCTL: parameters for the diabatic model,
-\item COSMO\_ART: parameters for gases and aerosols model,
-\item DIACTL: parameters for the diagnostic calculations,
-\item SATCTL: controlling computation of synthetic satellite images,
-\item IOCTL: controlling the I/O environment,
-\item GRIBIN: controlling the grib input,
-\item GRIBOUT: controlling the grib output,
-\end{itemize}
+\begin{table}[htbf]
+  \begin{center}
+    \caption{}
+    \label{tab:1}
+    \begin{tabular}{ll}
+      \hline\noalign{\smallskip} 
+      \textbf{Group} & \textbf{Description} \\
+      \noalign{\smallskip}\hline\noalign{\smallskip}
+      LMGRID & Grid domain and size parameters \\
+      RUNCTL & Model run parameters \\
+      TUNING & Physics and dynamics parameters \\
+      DYNCTL & Adiabatic model parameters \\
+      PHYCTL & Diabatic model parameters \\
+      COSMO\_ART & Gases and aerosols model parameters \\
+      DIACTL & Diagnostic calculations parameters \\
+      SATCTL & Synthetic satellite images parameters \\
+      IOCTL & I/O environment parameters \\
+      GRIBIN & GRIB input parameters \\
+      GRIBOUT & GRIB output parameters \\
+     \noalign{\smallskip}\hline
+    \end{tabular}
+  \end{center}
+\end{table}
 
 \noindent
-To run COSMO-ART the following input data are necessary:
+In  particular,  to  run   COSMO-ART  the  following  input  data  are
+necessary:
 \begin{itemize}
@@
 \label{subsec:3.3} 
-Measuring  the power  consumption on  one entire  Cray cabinet  is the
-easiest way to calculate energy-to-solution for given applications.
+The  computing performance  and  energy efficiency  evaluation of  the
+aforementioned   platforms  is   involving   two  important   metrics:
+\textit{time-to-solution}       and       \textit{energy-to-solution}.
+Time-to-Solution  refers to  the total  wall clock  time spent  by the
+application  to  obtain results,  which  in  our  case represents  the
+overall execution time of  COSMO-ART. Energy-to-solution is the amount
+of energy spent to achieve  results. We present here an approach which
+consists  in measuring the  power consumption  on one  entire cabinet.
+The  sampled  instantaneous power  during  execution  is averaged  and
+multiplied by the time-to-solution to determine energy-to-solution.
 
-\begin{figure*}
-  \includegraphics[width=0.4\textwidth]{Figs/NRJ_benchmark_Monch.eps}
-  \caption{Isola E1 Rack 2 Total Power and Isola E1 Total Power}
+\begin{figure}[htbf]
+  \begin{center}
+    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
+    \caption{Monch: Isola E1 Rack 2 Total Power and Isola E1 Total
+      Power}
     \label{fig:1}
-\end{figure*}
+  \end{center}
+\end{figure}
 
-The run  was issued successfully  twice, the start time  and execution
-time for both jobs are:
-\begin{itemize}
-\item start=12:51:03, end=13:20:46 $\Rightarrow$ T = 00:29:43 = 1783 s
-\item start=14:28:40, end=14:58:02 $\Rightarrow$ T = 00:29:22 = 1762 s
-\end{itemize}
+\begin{figure}[htbf]
+  \begin{center}
+    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
+    \caption{Pilatus: Isola HD Total Power}
+    \label{fig:2}
+  \end{center}
+\end{figure}
 
-The power  consumption for  both jobs is  calculated by  averaging the
-corresponding power  measurements from the Excel  file.  Power results
-account for the  Isola E1 Rack 2 Total Power. As  time resolution is 5
-minutes  for the  output  results, the  average  power consumption  is
-computed   by  considering  6   values  for   each  single   run  (see
-Table~\ref{tab:1}).
+Figures~\ref{fig:1} and  \ref{fig:2} account respectively  for Monch's
+Isola E1  Rack 2  and Pilatus' Isola  HD total power  measurements for
+1-day or 2-days simulations. On  the Intel Ivy Bridge EP based cluster
+(i.e. Monch), the 1-day simulation  was issued only twice due to usage
+restrictions. As time resolution was set to one update every 5 minutes
+for  power sampling,  the average  power consumption  was  computed by
+considering 6 values for each single  run.  On the Intel Xeon E5 based
+cluster (i.e.   Pilatus), the 1-day  simulation was issued  four times
+and a 2-days  run only once. Similarly, the  average power consumption
+was computed by  considering 4 values for each single  1-day run and 9
+values  for the  2-days  run. Corresponding  results  are gathered  in
+Table~\ref{tab:3}.
 
-\begin{table}
+\begin{table}[htbf]
   \begin{center}
-    \caption{}
-    \label{tab:1}
-    \begin{tabular}{lll}
+    \caption{Average power consumption (W) of the platforms}
+    \label{tab:3}
+    \begin{tabular}{ccc}
       \hline\noalign{\smallskip}
-      Time & Measured power (W) & Average power (W)  \\
+      \textbf{Simulation time} & \textbf{Xeon E5} & \textbf{Ivy Bridge EP} \\
       \noalign{\smallskip}\hline\noalign{\smallskip}
-      12:55:00 & 1.2355833333e+04 & 1.265852278e+04 \\ 
-      13:00:00 & 1.2820266667e+04 &  \\
-      13:05:00 & 1.2600743333e+04 &  \\ 
-      13:10:00 & 1.2811580000e+04 &  \\
-      13:15:00 & 1.2609680000e+04 &  \\
-      13:20:00 & 1.2753033333e+04 &  \\
+      \textbf{1 day} & 18122.01417 & 12658.52278 \\ 
+      & 17979.61083 & 12586.40833 \\
+      & 18065.45167 & - \\
+      & 17973.02833 & - \\
       \noalign{\smallskip}\hline\noalign{\smallskip}
-      14:30:00 & 1.2019266667e+04 & 1.258640833e+04 \\ 
-      14:35:00 & 1.2632253333e+04 &  \\
-      14:40:00 & 1.2779040000e+04 &  \\
-      14:45:00 & 1.2753180000e+04 &  \\
-      14:50:00 & 1.2610476667e+04 &  \\
-      14:55:00 & 1.2724233333e+04 &  \\
+      \textbf{2 days} & 17997.57815 & - \\
       \noalign{\smallskip}\hline
@@
 
-The  total  energy to  solution  is  the straightforward  calculation:\\
+In Figure~\ref{fig:3}, we compare both time-to-solution (right y-axis)
+and  energy-to-solution (left  y-axis) metrics  on both  platforms. As
+expected, the 1-day  simulation on Ivy Bridge EP  presents the highest
+execution time being roughly 1.3x  slower than Xeon E5. The reason for
+that is  twofold: (i) it has  lower clock frequency than  Xeon E5 (2.2
+GHz against 2.6 GHz); (ii)  Xeon E5 is a performance-centric processor
+that is  tuned far more speed  than for low power  consumption. In our
+experiments,  Ivy  Bridge   EP  showed  the  best  energy-to-solution,
+reducing the energy consumption of Xeon E5 by approximately $7\%$.
 
-Energy-to-solution  =  Integral  of  power consumption  over  the  job
-duration
-
-Often  the  average  power  consumption  times  the  job  duration  is
-sufficient.
-
-Thus, the total energy consumption corresponds to: 
-\begin{itemize}
-\item E  = 1783 x 12658.52278  = 22570146.11674 J $\sim$  22.57 MJ per
-  day of simulation
-\item E  = 1762 x 12586.40833  = 22177251.47746 J $\sim$  22.18 MJ per
-  day of simulation
-\end{itemize}
+\begin{figure}[htbf]
+  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART.eps}
+  \caption{Time and energy-to-solution comparison between Xeon E5 and
+    Ivy Bridge architectures}
+  \label{fig:3}
+\end{figure}

### 2014-04-09 trunk/paper_enahpc14/Sec2:Experimental-setup.tex
modifies "Evaluation" section; adds figure on T2S and E2S

--- 2014-04-07 paper_enahpc14/Sec2:Power-measurement.tex
@@
-\subsection{Hardware description}
+\subsection{Experimental Platforms}
 \label{subsec:2.1}
@@
 compute nodes.   Each standard compute node comprises  2 Intel 10-core
-Ivy Bridge EP E5-2660v2 processors operating at 2.2 GHz, offering 32GB
-of DDR3 1600MHz  RAM, themselves connected by a  high speed Infiniband
-network,  based on Mellanox  SX6036 managed  FDR (Fourteen  Data Rate)
-switches, with a 56 Gb/s speed.
+Ivy Bridge  EP E5-2660v2 processors  operating at 2.2  GHz, themselves
+connected by a high speed Infiniband network, based on Mellanox SX6036
+managed  FDR switches,  with a  56  Gb/s speed.  Each core  has 32  KB
+instruction and 32 KB  data L1 caches and 256 KB of  L2 cache. All the
+10 core  share a 25.6 MB  L3 cache and  the platform has 32GB  of DDR3
+1600MHz RAM.
 
@@
 Intel  8-core Intel  Xeon  E5-2670 processors  operating  at 2.6  GHz,
-offering  64GB of  DDR3 1600MHz  RAM, themselves  connected by  a high
-speed  Infiniband  network,  based  on  Mellanox  SX6036  managed  FDR
-(Fourteen Data Rate) switches, with a 56 Gb/s speed.
+themselves  connected by  a high  speed Infiniband  network,  based on
+Mellanox SX6036 managed FDR switches,  with a 56 Gb/s speed. Each core
+has  32 KB  instruction and  32 KB  data L1  caches and  256 KB  of L2
+cache. All the 8 core share a 20 MB L3 cache and the platform has 64GB
+of DDR3 1600MHz RAM.
 
-\subsection{Power-Performance measurement framework}
+\subsection{Power-Performance Measurement}
 \label{subsec:2.2}
@@
 \subsubsection{CSCS - ETH Zurich}
-E3METER  Intelligent Power Strips  (IPS) and  Monitors (IPM)  are high
-quality electricity meters which enable to monitor and optimize energy
-consumption of datacenters or large facilities.  A central data logger
-- the  E3METER Data Concentrator  - aggregates  all measured  data and
-makes them available in various forms (SNMP, Web, etc). Data transfers
-between E3METER  IPS and E3METER Concentrator  use reliable narrowband
-powerline  communication (PLC)  technology which  avoids the  need for
-extra cabling.  It has  one internal temperature sensor. Two dedicated
+Hardware platforms in our experiment include E3METER Intelligent Power
+Strips  (IPS) and Monitors  (IPM) which  are high  quality electricity
+meters  that enable  to  monitor and  optimize  energy consumption  of
+datacenters or large facilities.  A  central data logger - the E3METER
+Data  Concentrator  - aggregates  all  measured  data  and makes  them
+available in  various forms (SNMP,  Web, etc). Data  transfers between
+E3METER IPS and E3METER Concentrator use reliable narrowband powerline
+communication  (PLC)  technology  which  avoids  the  need  for  extra
+cabling.   It  has one  internal  temperature  sensor.  Two  dedicated
 extension  ports  can be  used  to  measure  external temperature  and

### 2014-04-09 trunk/paper_enahpc14/paper.tex
adjusts structure of the document
modifies "Evaluation" section; adds figure on T2S and E2S

@@
 \author{Joseph~Charles  \and William~Sawyer  \and  Manuel~F.~Dolz \and
-  Cristiano~Malossi}
+  Sandra~Catalan}
 
@@
 	\\ CH-6900 Lugano, Switzerland  
-	%\\ Tel: +41  (0) 91  610 8216
  	\\ E-mails:~{\{joseph.charles,william.sawyer\}@cscs.ch} 
         \and M.~F.~Dolz 
-        \at Dept. of Informatics, University of Hamburg
-	\\ D-22.527 Hamburg, Germany
-        %\\ Tel: +49 (0) 40 460094-404
+        \at Dept. of Informatics, University of Hamburg (UHAM)
+	\\ DE-22527 Hamburg, Germany
         \\ \email{manuel.dolz@informatik.uni-hamburg.de}  
-        \and C.~Malossi  
-        \at IBM Zurich Research Laboratory 
-        \\  CH-8000 Zurich, Switzerland
-        %\\  Tel: +41 (0) 44 724 8616
-        \\ \email{ACM@zurich.ibm.com}}
+        \and S.~Catalan
+        \at Jaume I University of Castell\'on (UJI)
+	\\ ES-12071 Castell\'on, Spain
+        \\ \email{catalans@icc.uji.es}}
 
@@
   systems is analysed  by a profiling study to  determine hotspots and
-  identify critical paths.   Moreover, we describe measurement devices
-  and  energy-aware   techniques  employed  to   evaluate  the  energy
+  identify  critical  CPU paths.   Moreover,  we describe  measurement
+  devices and energy-aware techniques  employed to evaluate the energy
   footprint of the considered application and to get detailed insights
@@
   chemistry  \and  Aerosols  modelling  \and  Profiling  methods  \and
-  Benchmark analysis \and COSMO-ART coupled model}
+  Benchmark    analysis   \and    COSMO-ART    coupled   model    \and
+  Time-to-solution \and Energy-to-solution \and Manycore}
 % \PACS{PACS code1 \and PACS code2 \and more}
@@
 
-\section{Power-performance measurement framework}
+\section{Experimental setup}
 \label{sec:2}
-\input{Sec2:Power-measurement.tex}
+\input{Sec2:Experimental-setup.tex}
 
@@
 Technologies (FET) Proactive Initiative: Minimising Energy Consumption
-of Computing to  the Limit (MINECC).  The authors  would like to thank
+of Computing  to the Limit  (MINECC).  We also  gratefully acknowledge
 the  High  Performance  and  High  Productivity  Computing  Initiative

### 2014-04-10 paper_enahpc14/paper.tex
adds latest modifications to the document
Manuel's changes

@@
 \title{Evaluating the Performance  and Energy Efficiency of COSMO-ART,
-  a Fully Online Coupled  Model System with Numerical Weather Forecast
-  and Chemical Transport Models }
+  a Coupled Numerical Weather Forecast and Chemical Transport Model}
 % \thanks{This work was supported by the EU Project FP7 318793 ``EXA2GREEN''}}
@@
 \author{Joseph~Charles  \and William~Sawyer  \and  Manuel~F.~Dolz \and
-  Cristiano~Malossi}
+  Sandra~Catal\'an}
 
@@
 	\\ CH-6900 Lugano, Switzerland  
-	%\\ Tel: +41  (0) 91  610 8216
  	\\ E-mails:~{\{joseph.charles,william.sawyer\}@cscs.ch} 
         \and M.~F.~Dolz 
-        \at Dept. of Informatics, University of Hamburg
-	\\ D-22.527 Hamburg, Germany
-        %\\ Tel: +49 (0) 40 460094-404
+        \at Dept. of Informatics, University of Hamburg (UHAM)
+	\\ DE-22527 Hamburg, Germany
         \\ \email{manuel.dolz@informatik.uni-hamburg.de}  
-        \and C.~Malossi  
-        \at IBM Zurich Research Laboratory 
-        \\  CH-8000 Zurich, Switzerland
-        %\\  Tel: +41 (0) 44 724 8616
-        \\ \email{ACM@zurich.ibm.com}}
+        \and S.~Catal\'an
+        \at Jaume I University of Castell\'on (UJI)
+	\\ ES-12071 Castell\'on, Spain
+        \\ \email{catalans@icc.uji.es}}
 
@@
 \section{Introduction}
-\label{intro}
-\input{Introduction.tex}
+\label{sec:1}
+\input{Sec1:Introduction.tex}
 
 \section{The COSMO-ART model system}
-\label{sec:1}
-\input{Sec1:COSMO-ART.tex}
+\label{sec:2}
+\input{Sec2:COSMO-ART.tex}
 
-\section{Power-performance measurement framework}
-\label{sec:2}
-\input{Sec2:Power-measurement.tex}
+\section{Experimental setup}
+\label{sec:3}
+\input{Sec3:Power-measurement.tex}
 
 \section{Evaluation}
-\label{sec:3}
-\input{Sec3:Evaluation.tex}
+\label{sec:4}
+\input{Sec4:Evaluation.tex}
 
 \section{Related work}
-\label{sec:4}
-\input{Sec4:Related.tex}
+\label{sec:5}
+\input{Sec5:Related.tex}
 
 \section{Conclusion}
-\label{concl}
+\label{sec:6}
+\label{Sec6:Conclusions.tex}
 \input{Conclusion.tex}
@@
 %\paragraph{Paragraph headings} Use paragraph headings as needed.
-
-%\begin{figure*}
-%  \includegraphics[width=0.4\textwidth]{Figs/earth.eps}
-%  \caption{Please write your figure caption here}
-%  \label{fig:1}
-%\end{figure*}
-
-%\begin{table}
-%  \caption{Please write your table caption here}
-%  \label{tab:1}
-%  \begin{tabular}{lll}
-%    \hline\noalign{\smallskip}
-%    first & second & third  \\
-%    \noalign{\smallskip}\hline\noalign{\smallskip}
-%    number & number & number \\
-%    number & number & number \\
-%    \noalign{\smallskip}\hline
-%  \end{tabular}
-%\end{table}
 

### 2014-04-11 trunk/paper_enahpc14/paper.tex
restructures paper; improves contents

@@
   Benchmark    analysis   \and    COSMO-ART    coupled   model    \and
-  Time-to-solution \and Energy-to-solution \and Manycore}
+  Time-to-solution \and Energy-to-solution \and Xeon processor}
 % \PACS{PACS code1 \and PACS code2 \and more}
@@
 
+\section{Related work}
+\label{sec:1}
+\input{Sec1:Related.tex}
+
 \section{The COSMO-ART model system}
-\label{sec:1}
-\input{Sec1:COSMO-ART.tex}
+\label{sec:2}
+\input{Sec2:COSMO-ART.tex}
 
-\section{Experimental setup}
-\label{sec:2}
-\input{Sec2:Experimental-setup.tex}
+\section{Power-performance measurement framework}
+\label{sec:3}
+\input{Sec3:Measurement-framework.tex}
 
 \section{Evaluation}
-\label{sec:3}
-\input{Sec3:Evaluation.tex}
-
-\section{Related work}
 \label{sec:4}
-\input{Sec4:Related.tex}
+\input{Sec4:Evaluation.tex}
 
@@
 \begin{acknowledgements}
-This  research  was  supported   by  the  Exa2Green  research  project
-co-funded  under the  EU  7th Framework  Program  Future and  Emerging
-Technologies (FET) Proactive Initiative: Minimising Energy Consumption
-of Computing  to the Limit  (MINECC).  We also  gratefully acknowledge
-the  High  Performance  and  High  Productivity  Computing  Initiative
-(\url{www.hp2c.ch}) for  results that will be  leveraged in subsequent
-code refactoring.
+The research  leading to these  results is supported by  the Exa2Green
+project  co-financed by  the European  Commission under  7th Framework
+Programme Future and Emerging Technologies (FET) Proactive Initiative:
+Minimising Energy  Consumption of Computing to the  Limit (MINECC). We
+also gratefully acknowledge the High Performance and High Productivity
+Computing  Initiative  (\url{www.hp2c.ch}) for  results  that will  be
+leveraged in subsequent code refactoring.
 \end{acknowledgements}

### 2014-04-11 paper_enahpc14/Sec2:COSMO-ART.tex
restructures paper; improves contents

--- 2014-04-08 paper_enahpc14/Sec1:COSMO-ART.tex
@@
 $0.22\,^{\circ}$,  i.e.  50  km in  both directions,  and  40 vertical
-layers.   Boundary  and initial  conditions,  emmissions and  external
+layers.   Boundary  and  initial  conditions, emissions  and  external
 parameters  were obtained by  the INT2LM  pre-processing interpolation
@@
 initial and boundary conditions  and consists mostly of the definition
-of a large number of  tracers.  The meteorological initial and bounday
-conditions are  obtained by  the the ECMWF  global spectral  model IFS
-with an update  frequency of 3h.  Boundary data  for gas-phase species
-are taken from IFS-MOZART output at 6h temporal resolution.  The model
-setup incorporates  34 2-d and 45  3-d fields to be  written out every
-hour and doesn't include data assimilation methods.
+of a large number of  tracers. In particular when using COSMO-ART, the
+following input data are necessary:
+
+\begin{itemize}
+\item  Gas phase:  Anthropogenic emissions  for different  species and
+  land use data for biogenic emissions and deposition,
+\item Aerosol particles: Anthropogenic emissions,
+\item Mineral dust: Soil specific land use data.
+\end{itemize}
+
+The meteorological initial and  bounday conditions are obtained by the
+the ECMWF  global spectral model IFS  with an update  frequency of 3h.
+Boundary data  for gas-phase species are taken  from IFS-MOZART output
+at   6h  temporal   resolution.   NAMELIST-input   specifying  runtime
+parameters    splitted   into    several   groups    are    given   in
+Table~\ref{tab:1}.  The model  setup incorporates  34 2-d  and  45 3-d
+fields  to  be  written  out  every  hour  and  doesn't  include  data
+assimilation methods.
+
+\begin{table}[htbf]
+  \begin{center}
+    \caption{}
+    \label{tab:1}
+    \begin{tabular}{ll}
+      \hline\noalign{\smallskip} 
+      \textbf{Group} & \textbf{Description} \\
+      \noalign{\smallskip}\hline\noalign{\smallskip}
+      LMGRID & Grid domain and size parameters \\
+      RUNCTL & Model run parameters \\
+      TUNING & Physics and dynamics parameters \\
+      DYNCTL & Adiabatic model parameters \\
+      PHYCTL & Diabatic model parameters \\
+      COSMO\_ART & Gases and aerosols model parameters \\
+      DIACTL & Diagnostic calculations parameters \\
+      SATCTL & Synthetic satellite images parameters \\
+      IOCTL & I/O environment parameters \\
+      GRIBIN & GRIB input parameters \\
+      GRIBOUT & GRIB output parameters \\
+     \noalign{\smallskip}\hline
+    \end{tabular}
+  \end{center}
+\end{table}
 

### 2014-04-11 paper_enahpc14/Sec3:Measurement-framework.tex
restructures paper; improves contents

--- 2014-04-04 paper_enahpc14/Sec2:Power-measurement.tex
@@
-\subsection{Hardware description}
-\label{subsec:2.1}
+In this section,  we present two different frameworks  deployed on HPC
+systems  to  measure the  power  consumption  and  performance of  the
+baseline execution.
 
-\subsubsection{Monch (CSCS - ETH Zurich)}
-While hardware  platforms mature  and are replaced,  we have  chosen a
-state-of-the-art  dual-socket Intel  Ivybridge-based  platform (called
-``Monch'')  for our power  measurements, as  it is  slated to  stay in
-service without upgrade for the duration of the Exa2Green project.  In
-principle at least,  this architecture could be recreated  or found in
-an identical configuration beyond  the lifetime of the project.  Given
-that  the baseline  benchmark  can be  reproduced  within an  expected
-variance (see subsection~\ref{subsec:1.2}),  and that the baseline run
-configuration can be  used in all future versions of  the code, a fair
-comparison  will  be  made  between  the baseline  and  the  milestone
-versions of COSMO-ART.
+\subsection{Framework description}
+\label{subsec:3.1}
 
-Monch, which is installed  at the Swiss National Supercomputing Center
-(CSCS)  of  the  ETH  Zurich,  is a  10  rack  NEC-provided  (LX-2400)
-Intel-based cluster, utilised  from people that are part  of the Swiss
-Platform  for  Advanced Scientific  Computing  (PASC),  which aims  at
-positioning  Swiss  computational sciences  in  the emerging  exascale
-era. It is  composed of 16 administration nodes,  312 standard compute
-nodes, 24 large-memory compute nodes, 24 huge-memory compute nodes and
-8  servers to  provide  I/O  to Lustre.   Each  standard compute  node
-comprises 2 Intel 10-core Ivy Bridge EP E5-2660v2 processors operating
-at  2.2 GHz,  offering 32GB  of DDR3  1600MHz RAM,  with a  base clock
-frequency of 2.2 GHz, themselves  connected by a high speed Infiniband
-network,  based on Mellanox  SX6036 managed  FDR (Fourteen  Data Rate)
-switches, with a 56 Gb/s speed.
-
-\subsection{Power-Performance measurement framework}
-\label{subsec:2.2}
-
-\subsubsection{CSCS - ETH Zurich}
-E3METER  Intelligent Power Strips  (IPS) and  Monitors (IPM)  are high
-quality electricity meters which enable to monitor and optimize energy
-consumption of datacenters or large facilities.  A central data logger
-- the  E3METER Data Concentrator  - aggregates  all measured  data and
-makes them available in various forms (SNMP, Web, etc). Data transfers
-between E3METER  IPS and E3METER Concentrator  use reliable narrowband
-powerline  communication (PLC)  technology which  avoids the  need for
-extra cabling.  It has  one internal temperature sensor. Two dedicated
-extension  ports  can be  used  to  measure  external temperature  and
-humidity  through E3METER  remote sensors.  It  features non-intrusive
-voltage, current and frequency measurement through current transformer
-with an accuracy of 1\%.
+\subsubsection{E3METER metering products}
+Supercomputer  clusters considered  for our  experiments at  the Swiss
+National Supercomputing Center of  ETH Zurich (CSCS) are equipped with
+E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
+high  quality  electricity meters  released  by  Riedo Networks,  that
+enable to monitor  and log power consumption of  the IT infrastructure
+as  well as  constantly analyze  line voltage,  current, power-factor,
+frequency  with  1\% accuracy.   Using  reliable narrowband  powerline
+communication (PLC)  technology, all  metering and power  quality data
+from   each  IPS  are   centrally  collected   by  the   E3METER  Data
+Concentrator, via the existing power cables thus avoiding the need for
+extra cabling.   This data  is made available  via SNMP,  HTTP, TELNET
+through  the built-in  Fast  Ethernet port.   Time synchronization  is
+guaranteed by  using NTP servers.   Measured data is  accessed through
+the open source  Cacti software including the E3METER  Cacti Plugin to
+scan the entire PLC network  and monitor the power usage of individual
+rack in real-time.
 

### 2014-04-11 paper_enahpc14/Sec4:Evaluation.tex
restructures paper; improves contents

In  this  section, we  present  a  performance  and energy  efficiency
evaluation  of  different  achitectures  when  running  the  COSMO-ART
baseline.  We  start by  specifying our measurement  methodology along
with the metrics used to  analyse the results on all platforms.  Then,
we detail the environment  setup gathering the considered HPC systems,
the software  environment and the runtime  configuration.  Finally, we
discuss  benchmark   results  and  power-performance   traces  of  the
application.

\subsection{Measurement methodology}
\label{subsec:4.1}
To analyse the energy  footprint and overall performance of COSMO-ART,
two  important metrics  were  examined: \textit{time-to-solution}  and
\textit{energy-to-solution}.   Time-to-Solution  refers  to the  total
wall clock time of  the application execution time. Energy-to-solution
is the amount  of energy spent to achieve  results. Energy consumption
of CSCS clusters is assessed  by sampling the instantaneous peak power
during  execution  which  is  then  averaged  and  multiplied  by  the
time-to-solution  to determine energy-to-solution.  Whenever possible,
multiple production runs of COSMO-ART were performed to illustrate the
reproducibility  of   the  baseline,  and   quantify  the  significant
uncertainties in  the power measurement, as dictated  by the available
technology.

\subsection{Environment setup}
\label{subsec:4.2}
While hardware  platforms mature  and are replaced,  we have  chosen a
state-of-the-art  Intel's  third-generation   Core  (aka  Ivy  Bridge)
processing platform (called ``Monch'')  for our power measurements, as
it  is slated  to stay  in service  without hardware  upgrade  for the
duration  of  the Exa2Green  project.   In  principle  at least,  this
architecture could be recreated or found in an identical configuration
beyond the lifetime of the project.  Given that the baseline benchmark
can be reproduced  within an expected variance, and  that the baseline
run configuration  can be used in  all future versions of  the code, a
fair comparison  will be made  between the baseline and  the milestone
versions   of    COSMO-ART.    In   this    study,   a   complementary
energy-to-solution  benchmarking  comparison  is  carried out  on  the
``Pilatus'' cluster based  on Intel's previous-generation Sandy Bridge
processors, conventional  in HPC  systems and known  to be  more power
consuming.

\subsubsection{Monch (CSCS - ETH Zurich)}
Monch, which is installed  at the Swiss National Supercomputing Center
(CSCS) of ETH Zurich, is  a 10 rack NEC-provided and dual-socket Intel
Ivy Bridge-EP based cluster, utilised from people that are part of the
Swiss    Platform   for    Advanced   Scientific    Computing   (PASC,
\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
Each standard  compute node comprises two Intel  Ivy Bridge-EP E5-2660
v2 ten-core processors operating at 2.2 GHz, themselves connected by a
high speed  InfiniBand network, based  on Mellanox SX6036  managed FDR
switches, with a  56 Gb/s speed.  Each core has  32 KB instruction and
32 KB data L1  caches and 256 KB of L2 cache. All  the 10 core share a
25 MB L3 cache and the platform has 32GB of DDR3 1600 MHz RAM. For our
energy-to-solution benchmark,  a full rack of Monch  constituted of 52
standard  compute   nodes  (monchc[029-080])  was   considered.

\subsubsection{Pilatus (CSCS - ETH Zurich)}
Pilatus,  which  is installed  at  the  Swiss National  Supercomputing
Center (CSCS)  of ETH Zurich,  is a dual-socket Intel  Sandy Bridge-EP
based cluster  used as Piz  Daint pre-post processing cluster.   It is
composed of 42 compute nodes  and has 2 high-speed interconnects based
on FDR:  the first is dedicated to  the MPI traffic and  the second to
the storage high speed traffic.  The 2 login nodes and the 42 computes
nodes  consists in  11  twin-pair Intel  E5-Series  DALCO r2264i4t  2U
scalable compute modules.  Each  module contains 4 compute nodes based
on two Intel Xeon E5-2670  eight-core processors operating at 2.6 GHz,
themselves  connected by  a high  speed InfiniBand  network,  based on
Mellanox SX6036 managed FDR switches,  with a 56 Gb/s speed. Each core
has 32 KB instruction and 32 KB data L1 caches and 256 KB of L2 cache.
All the  8 core share a  20 MB L3 cache  and the platform  has 64GB of
DDR3 1600 MHz  RAM. For our energy-to-solution benchmark,  a full rack
of Pilatus  constituted of 42 standard  compute nodes (pilatus[03-44])
was considered.

\subsubsection{Sofware environment}
INT2LM  and the  COSMO-ART model  are  implemented in  Fortran 90  for
distributed  memory  parallel  computers  using  the  Message  Passing
Interface  (MPI) and is  purely MPI-parallel.   The software  stack on
both many-core  platforms was  controlled using the  modules framework
which gives  an easy and  flexible mechanism to  access to all  of the
CSCS  provided compilers,  tools  and applications.   For our  initial
benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,
gcc/4.8.2 on Pilatus) with the -O3 compiler flag as it generally gives
a  good  level  of optimization  and  the  code  runs faster  in  this
configuration  than when  compiled with  the intel  compiler (14.0.1).
Besides, we installed the  MPICH2 implementation of MPI (mvapich2/1.9)
as  well  as  the  commonly  used HDF5  (1.8.12)  and  NetCDF  (4.3.1)
libraries in favor of the  traditional GRIB library for the management
of extremely  large and complex data collections.   All computes nodes
have    an   operating   system    based   on    GNU/Linux   featuring
``2.6.32-358.11.1.el6.x86\_64''  and  ``3.0.101-0.15-default'' kernels
for Monch and Pilatus respectively.

\subsubsection{Run configuration}
A snapshot of the code, which includes, at least conceptually, all the
information needed  to reproduce the  energy-to-solution benchmarks of
COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
Pilatus.   The  calculated  region  was mapped  to  the  participating
processors using  a 2D-partitioning strategy.   The distribution along
the  x and  y  coordinates  was defined  by  setting: $nprocx=40$  and
$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
usually  kept  bigger  than  $nprocy$.   Besides as  this  version  of
COSMO-ART  doesn't  make  use   of  the  GRIB  library,  we  specified
$nprocio=0$ for  GRIB I/O.  Hyperthreading  is not considered  in this
study as previous  attempts of its use revealed that  it always led to
higher energy-to-solution.\\

\subsection{Experimental results}
\label{subsec:4.3}

\begin{figure}[htbf]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
    \caption{Monch: Isola E1 Rack 2 Total Power}
    \label{fig:1}
  \end{center}
\end{figure}

\begin{figure}[htbf]
  \begin{center}
    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
    \caption{Pilatus: Isola HD Total Power}
    \label{fig:2}
  \end{center}
\end{figure}

Figures~\ref{fig:1} and  \ref{fig:2} account respectively  for Monch's
Isola E1  Rack 2  and Pilatus' Isola  HD total power  measurements for
1-day or 2-days simulations. On  the Intel Ivy Bridge EP based cluster
(i.e. Monch), the 1-day simulation  was issued only twice due to usage
restrictions. As time resolution was set to one update every 5 minutes
for  power sampling,  the average  power consumption  was  computed by
considering 6 values for each single  run.  On the Intel Xeon E5 based
cluster (i.e.   Pilatus), the 1-day  simulation was issued  four times
and a 2-days  run only once. Similarly, the  average power consumption
was computed by  considering 4 values for each single  1-day run and 9
values  for the  2-days  run. Corresponding  results  are gathered  in
Table~\ref{tab:3}.

\begin{table}[htbf]
  \begin{center}
    \caption{Average power consumption (W) of the platforms}
    \label{tab:3}
    \begin{tabular}{ccc}
      \hline\noalign{\smallskip}
      \textbf{Simulation time} & \textbf{Xeon E5} & \textbf{Ivy Bridge EP} \\
      \noalign{\smallskip}\hline\noalign{\smallskip}
      \textbf{1 day} & 18122.01417 & 12658.52278 \\ 
      & 17979.61083 & 12586.40833 \\
      & 18065.45167 & - \\
      & 17973.02833 & - \\
      \noalign{\smallskip}\hline\noalign{\smallskip}
      \textbf{2 days} & 17997.57815 & - \\
      \noalign{\smallskip}\hline
    \end{tabular}
  \end{center}
\end{table}

In Figure~\ref{fig:3}, we compare both time-to-solution (right y-axis)
and  energy-to-solution (left  y-axis) metrics  on both  platforms. As
expected,  Xeon  E5 outperforms  Ivy  Bridge  EP,  being roughly  1.3x
faster.  The reason for that is twofold: it has higher clock frequency
than Xeon E5 (2.6 GHz against  2.2 GHz) and it aims at computing speed
regardless to  energy consumption. In  our experiments, Ivy  Bridge EP
showed the best energy-to-solution, reducing the energy consumption of
Xeon E5 by approximately $7\%$.

\begin{figure}[htbf]
  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART.eps}
  \caption{Time-to-solution and  energy-to-solution comparison between
    Xeon E5 and Ivy Bridge-EP architectures}
  \label{fig:3}
\end{figure}

### 2014-04-11 paper_enahpc14/Introduction.tex
WS: significant pruning to make way for results from Manuel and Sandra

@@
 temperatures,  widespread melting  of  snow and  ice,  and rising  sea
-levels.   The last report  of the  Intergovernmental Panel  on Climate
-Change, released  in September 2013  \citep{IPCC-2013}, concluded that
-most of the observed temperature  increases since the mid 20th Century
-have  been caused  by  increasing concentrations  of greenhouse  gases
-resulting from  anthropogenic activities  such as fossil  fuel burning
-and  deforestation.
-
-The Earth's  surface temperature is determined by  the balance between
-incoming  solar radiation and  outgoing infrared  radiation. Radiative
-forcing describes any imbalance  in Earth's radiation budget caused by
-human  or natural interventions  in the  climate system.   It directly
-measures the capacity of a gas  or other forcing agents to affect that
-energy balance, thereby contributing to climate change.  The intensity
-of  solar energy,  reflectivity  or absorption  of  gases or  surfaces
-(forests, grasslands, oceans, deserts, cloud cover, ice caps, cities),
-emission of heat by various  materials, all differ in their warming or
-cooling contributions to climate forcing.
-
-The radiative forcing  potential of a greenhouse gas  is determined by
-its  atmospheric concentration, warming  capacity, residence  time and
-spatial  distribution.   Long-lived greenhouse  gases  such as  carbon
-dioxide, methane and nitrous oxide, which linger chemically stable and
-homogeneously well-mixed over time scales  of a decade to thousands of
-years  in the  atmosphere,  are major  human-induced climate  drivers;
-whereas  short-lived gases such  as water,  suphur dioxide  and carbon
-monoxide,  which are  abundant heat-trapping  gases but  physically or
-chemically  reactive  to changes  in  temperature,  cause positive  or
-negative  feedbacks effects which  may vary  with forcing  history and
-climate  state. On a  per molecule  basis, other  gases can  have more
-potent  heat-trapping ability  than carbon  dioxide  (e.g. fluorinated
-gases  are 23000  times more  effective \citep{Zehner-2012}),  but are
-fortunately far less abundant in the atmosphere.
+levels.  The suspected culprit for this change are human-induced or anthropogenic
+factors, such as the greenhouse gas emissions from the burning of 
+fossil fuels, which may be trapping more incoming solar radiation
+in the earth system.
 
@@
 forcing     \citep{Charlson-1991,    Haywood-2000,    Ramanathan-2001,
-Liao-2005,  Bangert-2012,  Lundgren-2013}.   Depending on  their  size
-distribution  and chemical  composition, they  can also  act  as cloud
-condensation  and   ice  nuclei,  and  thereby   influence  the  cloud
-microphysical   processes  and   optical   properties  (cloud   albedo
-effect,  \citealp{Twomey-1977})  through  the  contribution  to  cloud
-formation.      This     results     in     changes     in     droplet
-concentrations                \citep{Albrecht-1989}                and
-precipitation     \citep{Rosenfeld-2000,    Khain-2008,    Khain-2009,
-Pruppacher-2010,   Seifert-2012,  Tao-2012,  Lee-2012},   producing  a
-so-called  negative  indirect  radiative forcing  \citep{Haywood-2000,
-Lohmann-2005,  VandenHeever-2011, Rosenfeld-2013}.  Wet  scavenging of
-aerosol  particles  represents  a  major  removal  mechanism  for  air
-pollutants from  the atmosphere, however the  effects of anthropogenic
-aerosols on  clouds and  the hydrological cycle  as well as  the cloud
-lifetime    effect    are    especially    hard    to    assess    and
-quantify   \citep{IPCC-2013},   and   remain   one  of   the   largest
-uncertainties in climate modeling and in climate change prediction due
-to  the lack  of  knowledge in  cloud feedbacks  \citep{Sherwood-2013,
-Rosenfeld-2013, Lee-2013}.
+Liao-2005,  Bangert-2012,  Lundgren-2013}.
 
@@
 underlying chemical  processes through models.  These  models, such as
-ART (Aerosols  and Reactive Trace  gases) extension, developed  at the
+ART (Aerosol Reactive Transport), developed  at the
 Karlsruhe Institute  of Technology (KIT),  offer a key  opportunity to

### 2014-04-11 paper_enahpc14/Sec2:COSMO-ART.tex
WS: significant pruning to make way for results from Manuel and Sandra

@@
 photochemistry and aerosol dynamics using the modal approach.
-
-Aerosols dynamics  are simulated  with the modal  aerosol microphysics
-module  MADE \citep{Ackermann-1998}, expanded  in MADEsoot  to feature
-explicit  treatment of  soot aging  through condensation  of inorganic
-\citep{Riemer-2003} and organic substances. MADE describes the aerosol
-population  through  five   modes  representing  sub-micron  particles
-consisting of sulphate, ammonium, nitrate, particulate organic matter,
-water and soot  \citep{Riemer-2004} in a range of  mixing state. These
-modes are coupled  with the gas phase by  condensation and nucleation,
-and are  strongly influenced by  anthropogenic emissions of  gases and
-particles. MADEsoot  describes the sub-micrometer  aerosol population,
-composed  of sea  salt \citep{Lundgren-2013},  mineral  dust particles
-\citep{Vogel-2006,  Stanelle-2010},   by  means  of   six  interacting
-log-normal modes.  For each  mode, mass contributions and total number
-concentration are prognostic  quantities, while the standard deviation
-is fixed.
-
-Specific  modules are included  to simulate  the dispersion  of pollen
-grains    \citep{Vogel-2008}   and    other    biological   particles.
-Meteorologically-influenced emissions  are also online  coupled within
-the  model  system.  The  biogenic  VOC  (volatile organic  compounds)
-emissions are  calculated as functions of  the land use  type based on
-the Global  Land Cover 2000  dataset and the modeled  temperatures and
-radiative fluxes \citep{Vogel-1995}.
-
-The gaseous chemistry in COSMO-ART  is solved by a modified version of
-the  Regional  Acid  Deposition  Model, Version  2  (RADM2)  mechanism
-\citep{Stockwell-1990},  which  is   extended  to  describe  secondary
-organic aerosol formation based on a volatile basis set (VBS) approach
-\citep{Athanasopoulou-2013}  and  hydroxyl  radical recycling  due  to
-isoprene chemistry \citep{Geiger-2003}.  The thermodynamic equilibrium
-between the  gas and particulate  phases of the inorganic  material is
-achieved through the  ISORROPIA II module \citep{Fountoukis-2007}.
-
-COSMO-ART  is  fully  online-coupled,  and  allows  for  feedbacks  of
-aerosols  on  temperature,  radiation  and cloud  condensation  nuclei
-(CCN).   Analytical  description  of  modules  incorporated  for  this
-purpose  exists in  \citep{Vogel-2009,  Bangert-2011}.  The  radiation
-scheme used  within the  model to calculate  the vertical  profiles of
-shortwave and longwave radiative fluxes is GRAALS \citep{Ritter-1992}.
-In order to account for  the interaction of aerosol particles with the
-cloud microphysics and radiation,  COSMO-ART uses the two-moment cloud
-microphysics  scheme of  Seifert and  Beheng  \citep{Seifert-2006} and
-comprehensive parameterizations for cloud condensation and ice crystal
-nucleation  \citep{Bangert-2011, Bangert-2012}.   The system  of stiff
-ordinary differential equations described by chemical reactions in the
-aqueous-phase together with the transfer reactions is solved using the
-Kinetic PreProcessor (KPP, \citealp{Damian-2002}).
 
@@
 $0.22\,^{\circ}$,  i.e.  50  km in  both directions,  and  40 vertical
-layers.   Boundary  and  initial  conditions, emissions  and  external
-parameters  were obtained by  the INT2LM  pre-processing interpolation
-tool, which interpolate  data from a coarse-grid driving  model to the
-rotated    latitude-longitude   grid    of   the    COSMO-Model,   and
-INT2LM\_trcr\_ART,  which  interpolates  aerosol  or  gaseous  species
-initial and boundary conditions  and consists mostly of the definition
-of a large number of  tracers. In particular when using COSMO-ART, the
-following input data are necessary:
+layers.   COSMO-ART requires the following input data:
 

### 2014-04-11 trunk/paper_enahpc14/Sec4:Evaluation.tex
adds minor changes

--- 2014-04-11 paper_enahpc14/Sec4:Evaluation.tex
@@
 distributed  memory  parallel  computers  using  the  Message  Passing
-Interface  (MPI) and is  purely MPI-parallel.   The software  stack on
+Interface (MPI)  and are purely  MPI-parallel.  The software  stack on
 both many-core  platforms was  controlled using the  modules framework

### 2014-04-11 trunk/paper_enahpc14/paper.tex
corrects Sandra's family name
adds minor changes

@@
 \author{Joseph~Charles  \and William~Sawyer  \and  Manuel~F.~Dolz \and
-  Sandra~Catalan}
+  Sandra~Catal\'an}
 
@@
         \\ \email{manuel.dolz@informatik.uni-hamburg.de}  
-        \and S.~Catalan
+        \and S.~Catal\'an
         \at Jaume I University of Castell\'on (UJI)
@@
 \begin{abstract}
-  In this paper we present  COSMO-ART, an extension of the operational
+  In  this paper  we present  the energy-to-solution  of  COSMO-ART on
+  various platforms.   This model is  an extension of  the operational
   weather  forecast  model  of   the  German  weather  service  (DWD),
@@
   about power bottlenecks.  Our motivation is to improve corresponding
-  code   sections  to  sustain   high  performance   while  minimizing
+  code    sections   to    sustain   performance    while   minimizing
   energy-to-solution.   This  preliminary  work  sets  the  basis  for

### 2014-04-11 paper_enahpc14/Introduction.tex
adds minor changes

@@
 temperatures,  widespread melting  of  snow and  ice,  and rising  sea
-levels.  The suspected culprit for this change are human-induced or anthropogenic
-factors, such as the greenhouse gas emissions from the burning of 
-fossil fuels, which may be trapping more incoming solar radiation
-in the earth system.
+levels.  The  suspected culprit for  this change are  human-induced or
+anthropogenic factors,  such as the greenhouse gas  emissions from the
+burning of  fossil fuels,  which may be  trapping more  incoming solar
+radiation in the earth system.
 
@@
 underlying chemical  processes through models.  These  models, such as
-ART (Aerosol Reactive Transport), developed  at the
-Karlsruhe Institute  of Technology (KIT),  offer a key  opportunity to
-reduce   the  climate  uncertainty,   particularly  on   the  regional
-scale \citep{Knote-2011, Bangert-2011,  Knote-2013}.  These models can
-be coupled with climate models,  such as the regional weather forecast
-model COSMO (Consortium  for Small-scale Modelling), jointly developed
-by  a consortium  of  European weather  centers  including the  German
-weather service  DWD and MeteoSwiss,  and used in the  climate version
-(COSMO-CLM)  by a  wide  research community.   The extended  COSMO-ART
-model provides  a detailed description of air  pollution chemistry and
-aerosol processes,  and is  mainly designed to  study air  quality and
-aerosol  meteorology  feedbacks  on  short, episodic  to  annual  time
-scales.  It is capable of  simulating aerosol distributions as well as
-their interactions over Europe as well as other regional domains.
+ART (Aerosol Reactive Transport), developed at the Karlsruhe Institute
+of Technology  (KIT), offer  a key opportunity  to reduce  the climate
+uncertainty,  particularly on  the  regional scale  \citep{Knote-2011,
+Bangert-2011, Knote-2013}.   These models can be  coupled with climate
+models, such as the  regional weather forecast model COSMO (Consortium
+for  Small-scale  Modelling), jointly  developed  by  a consortium  of
+European weather centers including  the German weather service DWD and
+MeteoSwiss,  and used  in the  climate version  (COSMO-CLM) by  a wide
+research community.  The extended  COSMO-ART model provides a detailed
+description of  air pollution chemistry and aerosol  processes, and is
+mainly designed to study air quality and aerosol meteorology feedbacks
+on short, episodic to annual time scales.  It is capable of simulating
+aerosol  distributions as well  as their  interactions over  Europe as
+well as other regional domains.
 
@@
 Performance  and  High  Productivity  Computing (HP2C)  Initiative  to
-optimize it  for computational and energy  efficiency.  Although these
-developments will  facilitate the  application of COSMO  for numerical
-weather prediction and climate  simulations, they do little to address
-the  coupling with  the  ART model  extension,  for which  significant
-investments  are still required  to take  it to  a similar  level. The
-efficiency of ART is being  addressed in the EU Exa2Green project. The
-ultimate goal  of the  project is to  deliver a prototype  code, which
-provides an energy  efficiency of at least five  times of the baseline
-value.   Such   an  implementation   would  allow  the   community  to
-investigate critical  questions at  higher resolution and  over longer
-periods, at reduced cost  to the environment.
+optimize    it    for     computational    and    energy    efficiency
+(\url{http://www.hp2c.ch/}).    Although   these   developments   will
+facilitate the  application of COSMO for  numerical weather prediction
+and climate simulations,  they do little to address  the coupling with
+the ART  model extension, for which significant  investments are still
+required to take it to a similar level. The efficiency of ART is being
+addressed        in       the        EU        Exa2Green       project
+(\url{http://exa2green.eu/}). The  ultimate goal of the  project is to
+deliver a  prototype code, which  provides an energy efficiency  of at
+least five times of the  baseline value.  Such an implementation would
+allow  the  community  to  investigate critical  questions  at  higher
+resolution  and   over  longer  periods,   at  reduced  cost   to  the
+environment.
 
-In  Sec.~\ref{sec:1},  we  introduce  the regional  atmospheric  model
-COSMO-ART  \citep{Vogel-2009}  that  accounts  for  feedbacks  between
+In  Sec.~\ref{sec:1}, we  briefly introduce  the  regional atmospheric
+model COSMO-ART \citep{Vogel-2009} that accounts for feedbacks between
 chemistry,  aerosols, radiation, and  clouds and  additionally specify

### 2014-04-11 paper_enahpc14/Sec2:COSMO-ART.tex
adds minor changes

@@
 diffusion,  and  dry and  wet  deposition  are  treated together  with
-photochemistry and aerosol dynamics using the modal approach.
-
-Detailed  model  description  can   be  found  in  the  aforementioned
-publications  as   well  as  in   \citep{Stanelle-2010,  Bangert-2012,
-  Knote-2011, Knote-2013}.
+photochemistry and aerosol dynamics using the modal approach. Detailed
+model description  can be found in the  aforementioned publications as
+well    as   in   \citep{Stanelle-2010,    Bangert-2012,   Knote-2011,
+  Knote-2013}.
 
@@
 
-This COSMO-ART  version is configured  to deal with a  semi Lagrangian
-horizontal advection scheme  with tricubic interpolation and selective
-filling diffusion option in combination with the Runge-Kutta dynamical
-core. It  also makes  use of the  Kinetic PreProcessor solver  for the
-resolution of  atmospheric chemistry ordinary  differential equations.
-Concerning the  modelling of wet deposition in  aerosols, the baseline
-has  only  indirect  cloud  feedbacks  but  doesn't  include  in-cloud
-scavenging  (rainout)   and  below-cloud  scavenging   (washout)  yet.
-Amongst   physical  parameterizations,   precipitation   formation  is
-performed  by a  two-moment  cloud microphysics  scheme  instead of  a
-classical  bulk microphysics  scheme.
+This COSMO-ART version is configured with a semi Lagrangian horizontal
+advection  scheme with  tricubic interpolation  and  selective filling
+diffusion  option  in  combination   with  the  dynamical  core  using
+Runge-Kutta time stepping \citep{}.  It  also makes use of the Kinetic
+PreProcessor  solver  \citep{}   for  the  resolution  of  atmospheric
+chemistry ordinary  differential equations.  Concerning  the modelling
+of wet  deposition in aerosols,  the baseline has only  indirect cloud
+feedbacks  but  doesn't  include  in-cloud  scavenging  (rainout)  and
+below-cloud    scavenging    (washout)    yet.     Amongst    physical
+parameterisations,   precipitation  formation   is   performed  by   a
+two-moment  cloud  microphysics scheme  instead  of  a classical  bulk
+microphysics scheme.

### 2014-04-12 paper_enahpc14/Introduction.tex
updates Introduction; refreshes bibliography

--- 2014-04-11 paper_enahpc14/Introduction.tex
@@
-Comprehensive  scientific assessments  of climate-warming  trends over
-the  past 50  years have  shown  that the  state and  dynamics of  the
-Earth's climate system have  undergone unprecedented major changes, as
-evidenced  by  increases  in   global  average  atmosphere  and  ocean
-temperatures,  widespread melting  of  snow and  ice,  and rising  sea
-levels.  The suspected culprit for this change are human-induced or anthropogenic
-factors, such as the greenhouse gas emissions from the burning of 
-fossil fuels, which may be trapping more incoming solar radiation
-in the earth system.
-
-One  factor   overwhelmingly  affects  the   uncertainty  in  defining
-anthropogenic  radiative  forcing:  the  effects  of  aerosols  (small
-airborne liquid,  solid or mixed  particles, \citealp{Putaud-2010}) in
-the  atmosphere.  Although  they  are not  considered a  heat-trapping
-greenhouse   gas  and   have  shorter   atmospheric   lifetimes,  they
-significantly affected the atmospheric  radiative fluxes over the last
-150    years   \citep{Koch-2009,   IPCC-2013}.     Enhanced   aerosols
-concentrations can impact the climate system by reflecting (e.g.  pure
-sulfates  and  nitrates)  or  absorbing  (e.g.   black  carbon)  solar
-radiation  and  thereby exert  a  cooling  or  warming effect  on  the
-Earth-atmosphere   system,  causing   a  so-called   direct  radiative
-forcing     \citep{Charlson-1991,    Haywood-2000,    Ramanathan-2001,
-Liao-2005,  Bangert-2012,  Lundgren-2013}.
+While anthropogenic greenhouse gas emissions are driving unprecedented
+major  climate  changes  since   the  mid  20th  Century,  one  factor
+overwhelmingly  affects  the  uncertainty  in  defining  human-induced
+radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
+Although they  are not considered  a heat-trapping greenhouse  gas and
+have  shorter  atmospheric lifetimes,  they  significantly modify  the
+global   radiation   budget   \citep{IPCC-2013}.   Enhanced   aerosols
+concentrations impact the climate  system by scattering and absorption
+of   solar   radiation,    thereby   exerting   a   direct   radiative
+forcing  \citep{Liao-2005,  Bangert-2012,  Lundgren-2013}, or  by  the
+modification of  cloud properties, cloud fraction  and surface albedo,
+causing  a negative  indirect  radiative forcing  \citep{Haywood-2000,
+Lohmann-2005,     VandenHeever-2011,     Rosenfeld-2013}.      Despite
+considerable  progress in  global aerosol  modelling \citep{Mann-2013}
+and    measurement-based    assessments   \citep{Myhre-2009},    large
+uncertainties    remain    in    current    estimates    of    aerosol
+forcing   \citep{Myhre-2013,    IPCC-2013,   Lee-2013,   Randles-2013,
+Rosenfeld-2013, Sherwood-2013, Stier-2013}.
 
 Hence to  improve our understanding of  aerosol-cloud interactions and
-reduce  uncertainties  of aerosol  effects  in  climate, the  research
-community is making a  concerted international effort to represent the
-underlying chemical  processes through models.  These  models, such as
-ART (Aerosol Reactive Transport), developed  at the
-Karlsruhe Institute  of Technology (KIT),  offer a key  opportunity to
-reduce   the  climate  uncertainty,   particularly  on   the  regional
-scale \citep{Knote-2011, Bangert-2011,  Knote-2013}.  These models can
-be coupled with climate models,  such as the regional weather forecast
-model COSMO (Consortium  for Small-scale Modelling), jointly developed
-by  a consortium  of  European weather  centers  including the  German
-weather service  DWD and MeteoSwiss,  and used in the  climate version
-(COSMO-CLM)  by a  wide  research community.   The extended  COSMO-ART
-model provides  a detailed description of air  pollution chemistry and
-aerosol processes,  and is  mainly designed to  study air  quality and
-aerosol  meteorology  feedbacks  on  short, episodic  to  annual  time
-scales.  It is capable of  simulating aerosol distributions as well as
-their interactions over Europe as well as other regional domains.
+reduce  uncertainties  in  aerosol  radiative forcing  estimates,  the
+research  community  is making  a  concerted  international effort  to
+represent  the  underlying physical,  chemical  and aerosol  dynamical
+processes through numerical chemical  transport models (CTMs), such as
+ART (Aerosols  and Reactive Trace  gases), developed at  the Karlsruhe
+Institute   of  Technology   (KIT)   \citep{Vogel-2009,  Bangert-2011,
+Knote-2013}. This  regional scale model system is  online coupled with
+the  operational weather  forecast  model COSMO  \citep{Baldauf-2011},
+jointly developed by a consortium of European weather centers and used
+in the climate version (COSMO-CLM)  by a wide research community.
 
-COSMO-ART is  computationally much more demanding than  the COSMO core
-since a  large number of additional  tracers and processes  have to be
-considered.  Thus this model is currently severely-limited in terms of
-applicability and expensive in terms of energy consumption.  COSMO has
-recently  been  ported  to  GPUs  within the  framework  of  the  High
-Performance  and  High  Productivity  Computing (HP2C)  Initiative  to
-optimize it  for computational and energy  efficiency.  Although these
-developments will  facilitate the  application of COSMO  for numerical
-weather prediction and climate  simulations, they do little to address
-the  coupling with  the  ART model  extension,  for which  significant
-investments  are still required  to take  it to  a similar  level. The
-efficiency of ART is being  addressed in the EU Exa2Green project. The
-ultimate goal  of the  project is to  deliver a prototype  code, which
-provides an energy  efficiency of at least five  times of the baseline
-value.   Such   an  implementation   would  allow  the   community  to
-investigate critical  questions at  higher resolution and  over longer
-periods, at reduced cost  to the environment.
+The extended  COSMO-ART model  is computationally much  more demanding
+than the  COSMO core  since a large  number of additional  tracers and
+processes  have  to  be  considered.   Thus this  model  is  currently
+severely-limited in  terms of applicability and expensive  in terms of
+energy consumption.   Although COSMO has recently been  ported to GPUs
+within  the framework of  the High  Performance and  High Productivity
+Computing  (HP2C)  Initiative to  optimize  it  for computational  and
+energy efficiency (\url{http://www.hp2c.ch/}), significant investments
+in  ART  are still  required  to  take it  to  a  similar level.   The
+efficiency of COSMO-ART is being addressed in the EU Exa2Green project
+(\url{http://exa2green.eu/}). The  ultimate goal of the  project is to
+deliver a  prototype code, which  provides an energy efficiency  of at
+least five times of the  baseline value.  Such an implementation would
+allow  the  community  to  investigate critical  questions  at  higher
+resolution  and   over  longer  periods,   at  reduced  cost   to  the
+environment.
 
-In  Sec.~\ref{sec:1},  we  introduce  the regional  atmospheric  model
-COSMO-ART  \citep{Vogel-2009}  that  accounts  for  feedbacks  between
-chemistry,  aerosols, radiation, and  clouds and  additionally specify
-its technical setup related to the investigated performance and energy
-evaluation  methods.   Sec.~\ref{sec:2}  describes the  different  HPC
-systems   and  associated  power   measurement  equipments   used  for
-conducting the benchmarking study and presents energy-aware techniques
-employed  to   determine  the  energy  footprint   of  the  considered
-application.   Sec.~\ref{sec:3}  gathers  benchmark results  from  the
-Exa2Green  project and  highlights  areas where  improvements will  be
-necessary for the subsequent  baseline refactoring.  Some related work
-will  be  discussed  in  Sec.~\ref{sec:4} Finally,  we  conclude  with
-practical  guidelines to identifying  energy saving  opportunities and
-give  directions for further  developments of  the modeling  system in
-Sec.~\ref{concl}.
+Some  related work  will  be discussed  in  Sec.~\ref{sec:1}. Then  in
+Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
+COSMO-ART \citep{Vogel-2009}  and specify its  technical setup related
+to  the  investigated   performance  and  energy  evaluation  methods.
+Sec.~\ref{sec:3} describes different power measurement equipments used
+for  conducting  this  benchmarking  study and  presents  energy-aware
+techniques  employed   to  determine  the  energy   footprint  of  the
+considered  application.  Sec.~\ref{sec:4}  gathers  benchmark results
+from  the Exa2Green  project and  highlights areas  where improvements
+will be  necessary for the subsequent  baseline refactoring.  Finally,
+we  conclude with  practical guidelines  to identifying  energy saving
+opportunities  and give  directions  for further  developments of  the
+modeling system in Sec.~\ref{concl}.

### 2014-04-12 paper_enahpc14/Sec2:COSMO-ART.tex
updates Introduction; refreshes bibliography

@@
 \label{subsec:1.1}
-COSMO-ART  (ART  stands  for   Aerosols  and  Reactive  Trace  gases),
-developed  at  KIT  Karlsruhe  \citep{Vogel-2009}, is  a  regional  to
-continental scale model coupled  online to the COSMO numerical weather
-prediction (NWP) and climate model \citep{Baldauf-2011}. It allows the
-online calculation of reactive  trace substances and their interaction
-with  the atmosphere.   Physical processes  like  transport, turbulent
-diffusion,  and  dry and  wet  deposition  are  treated together  with
-photochemistry and aerosol dynamics using the modal approach. Detailed
-model description  can be found in the  aforementioned publications as
-well    as   in   \citep{Stanelle-2010,    Bangert-2012,   Knote-2011,
-  Knote-2013}.
+The  model  system COSMO-ART  described  in  \citep{Vogel-2009}, is  a
+regional  to  continental scale  model  coupled  online  to the  COSMO
+numerical weather  prediction and climate  model \citep{Baldauf-2011}.
+It  incorporates  sophisticated  modules  for  gaseous  chemistry  and
+aerosol dynamics  and allows the online calculation  of reactive trace
+substances    and    their    interaction    with    the    atmosphere
+\citep{Knote-2011}.  Detailed  model   description  can  be  found  in
+\citep{Stanelle-2010, Bangert-2012, Knote-2011, Knote-2013}.
 
@@
 Runge-Kutta time stepping \citep{}.  It  also makes use of the Kinetic
-PreProcessor  solver  \citep{}   for  the  resolution  of  atmospheric
-chemistry ordinary  differential equations.  Concerning  the modelling
-of wet  deposition in aerosols,  the baseline has only  indirect cloud
-feedbacks  but  doesn't  include  in-cloud  scavenging  (rainout)  and
-below-cloud    scavenging    (washout)    yet.     Amongst    physical
+PreProcessor solver (KPP) for  the resolution of atmospheric chemistry
+ordinary  differential equations \citep{Damian-2002}.   Concerning the
+modelling  of  wet  deposition  in  aerosols, the  baseline  has  only
+indirect  cloud  feedbacks  but  doesn't include  in-cloud  scavenging
+(rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
 parameterisations,   precipitation  formation   is   performed  by   a

### 2014-04-12 paper_enahpc14/Sec3:Measurement-framework.tex
updates Introduction; refreshes bibliography

@@
 E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
-high  quality  electricity meters  released  by  Riedo Networks,  that
-enable to monitor  and log power consumption of  the IT infrastructure
-as  well as  constantly analyze  line voltage,  current, power-factor,
-frequency  with  1\% accuracy.   Using  reliable narrowband  powerline
-communication (PLC)  technology, all  metering and power  quality data
-from   each  IPS  are   centrally  collected   by  the   E3METER  Data
-Concentrator, via the existing power cables thus avoiding the need for
-extra cabling.   This data  is made available  via SNMP,  HTTP, TELNET
-through  the built-in  Fast  Ethernet port.   Time synchronization  is
-guaranteed by  using NTP servers.   Measured data is  accessed through
-the open source  Cacti software including the E3METER  Cacti Plugin to
-scan the entire PLC network  and monitor the power usage of individual
-rack in real-time.
+high   quality   electricity  meters   released   by  Riedo   Networks
+(\url{http://riedonetworks.com/}),  that  enable  to monitor  and  log
+power  consumption of  the  IT infrastructure  as  well as  constantly
+analyze  line  voltage,  current,  power-factor,  frequency  with  1\%
+accuracy.   Using reliable  narrowband  powerline communication  (PLC)
+technology,  all metering  and power  quality data  from each  IPS are
+centrally collected by the E3METER Data Concentrator, via the existing
+power cables thus  avoiding the need for extra  cabling.  This data is
+made  available  via SNMP,  HTTP,  TELNET  through  the built-in  Fast
+Ethernet  port.   Time  synchronization  is guaranteed  by  using  NTP
+servers.   Measured data  is accessed  through the  open  source Cacti
+software including  the E3METER  Cacti Plugin to  scan the  entire PLC
+network and monitor the power usage of individual rack in real-time.
 

### 2014-04-12 trunk/paper_enahpc14/Sec4:Evaluation.tex
updates Introduction; refreshes bibliography

@@
 \label{subsec:4.1}
-To analyse the energy  footprint and overall performance of COSMO-ART,
-two  important metrics  were  examined: \textit{time-to-solution}  and
-\textit{energy-to-solution}.   Time-to-Solution  refers  to the  total
-wall clock time of  the application execution time. Energy-to-solution
-is the amount  of energy spent to achieve  results. Energy consumption
-of CSCS clusters is assessed  by sampling the instantaneous peak power
-during  execution  which  is  then  averaged  and  multiplied  by  the
-time-to-solution  to determine energy-to-solution.  Whenever possible,
-multiple production runs of COSMO-ART were performed to illustrate the
-reproducibility  of   the  baseline,  and   quantify  the  significant
-uncertainties in  the power measurement, as dictated  by the available
-technology.
+We  approach  the  assessment  of  the energy  footprint  and  overall
+performance    of    COSMO-ART    with    two    important    metrics:
+\textit{time-to-solution}       and       \textit{energy-to-solution}.
+Time-to-Solution  refers   to  the  total  wall  clock   time  of  the
+application execution time. Energy-to-solution is the amount of energy
+spent  to achieve  results.  Energy  consumption of  CSCS  clusters is
+assessed  by sampling  the instantaneous  peak power  during execution
+which  is then  averaged  and multiplied  by  the time-to-solution  to
+determine energy-to-solution.   Whenever possible, multiple production
+runs of COSMO-ART were  performed to illustrate the reproducibility of
+the baseline, and quantify  the significant uncertainties in the power
+measurement, as dictated by the available technology.
 

### 2014-04-12 trunk/paper_enahpc14/Introduction.tex
adds minor change

@@
-Comprehensive  scientific assessments  of climate-warming  trends over
-the  past 50  years have  shown  that the  state and  dynamics of  the
-Earth's climate system have  undergone unprecedented major changes, as
-evidenced  by  increases  in   global  average  atmosphere  and  ocean
-temperatures,  widespread melting  of  snow and  ice,  and rising  sea
-levels.   The last report  of the  Intergovernmental Panel  on Climate
-Change, released  in September 2013  \citep{IPCC-2013}, concluded that
-most of the observed temperature  increases since the mid 20th Century
-have  been caused  by  increasing concentrations  of greenhouse  gases
-resulting from  anthropogenic activities  such as fossil  fuel burning
-and  deforestation.
-
-The Earth's  surface temperature is determined by  the balance between
-incoming  solar radiation and  outgoing infrared  radiation. Radiative
-forcing describes any imbalance  in Earth’s radiation budget caused by
-human  or natural interventions  in the  climate system.   It directly
-measures the capacity of a gas  or other forcing agents to affect that
-energy balance, thereby contributing to climate change.  The intensity
-of  solar energy,  reflectivity  or absorption  of  gases or  surfaces
-(forests, grasslands, oceans, deserts, cloud cover, ice caps, cities),
-emission of heat by various  materials, all differ in their warming or
-cooling contributions to climate forcing.
-
-The radiative forcing  potential of a greenhouse gas  is determined by
-its  atmospheric concentration, warming  capacity, residence  time and
-spatial  distribution.   Long-lived greenhouse  gases  such as  carbon
-dioxide, methane and nitrous oxide, which linger chemically stable and
-homogeneously well-mixed over time scales  of a decade to thousands of
-years  in the  atmosphere,  are major  human-induced climate  drivers;
-whereas  short-lived gases such  as water,  suphur dioxide  and carbon
-monoxide,  which are  abundant heat-trapping  gases but  physically or
-chemically  reactive  to changes  in  temperature,  cause positive  or
-negative  feedbacks effects which  may vary  with forcing  history and
-climate  state. On a  per molecule  basis, other  gases can  have more
-potent  heat-trapping ability  than carbon  dioxide  (e.g. fluorinated
-gases  are 23000  times more  effective \citep{Zehner-2012}),  but are
-fortunately far less abundant in the atmosphere.
-
-One  factor   overwhelmingly  affects  the   uncertainty  in  defining
-anthropogenic  radiative  forcing:  the  effects  of  aerosols  (small
-airborne liquid,  solid or mixed  particles, \citealp{Putaud-2010}) in
-the  atmosphere.  Although  they  are not  considered a  heat-trapping
-greenhouse   gas  and   have  shorter   atmospheric   lifetimes,  they
-significantly affected the atmospheric  radiative fluxes over the last
-150    years   \citep{Koch-2009,   IPCC-2013}.     Enhanced   aerosols
-concentrations can impact the climate system by reflecting (e.g.  pure
-sulfates  and  nitrates)  or  absorbing  (e.g.   black  carbon)  solar
-radiation  and  thereby exert  a  cooling  or  warming effect  on  the
-Earth-atmosphere   system,  causing   a  so-called   direct  radiative
-forcing     \citep{Charlson-1991,    Haywood-2000,    Ramanathan-2001,
-Liao-2005,  Bangert-2012,  Lundgren-2013}.   Depending on  their  size
-distribution  and chemical  composition, they  can also  act  as cloud
-condensation  and   ice  nuclei,  and  thereby   influence  the  cloud
-microphysical   processes  and   optical   properties  (cloud   albedo
-effect,  \citealp{Twomey-1977})  through  the  contribution  to  cloud
-formation.      This     results     in     changes     in     droplet
-concentrations                \citep{Albrecht-1989}                and
-precipitation     \citep{Rosenfeld-2000,    Khain-2008,    Khain-2009,
-Pruppacher-2010,   Seifert-2012,  Tao-2012,  Lee-2012},   producing  a
-so-called  negative  indirect  radiative forcing  \citep{Haywood-2000,
-Lohmann-2005,  VandenHeever-2011, Rosenfeld-2013}.  Wet  scavenging of
-aerosol  particles  represents  a  major  removal  mechanism  for  air
-pollutants from  the atmosphere, however the  effects of anthropogenic
-aerosols on  clouds and  the hydrological cycle  as well as  the cloud
-lifetime    effect    are    especially    hard    to    assess    and
-quantify   \citep{IPCC-2013},   and   remain   one  of   the   largest
-uncertainties in climate modeling and in climate change prediction due
-to  the lack  of  knowledge in  cloud feedbacks  \citep{Sherwood-2013,
-Rosenfeld-2013, Lee-2013}.
+While anthropogenic greenhouse gas emissions are driving unprecedented
+major  climate  changes  since   the  mid  20th  Century,  one  factor
+overwhelmingly  affects  the  uncertainty  in  defining  human-induced
+radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
+Although they  are not considered  a heat-trapping greenhouse  gas and
+have  shorter  atmospheric lifetimes,  they  significantly modify  the
+global   radiation   budget   \citep{IPCC-2013}.   Enhanced   aerosols
+concentrations impact the climate  system by scattering and absorption
+of   solar   radiation,    thereby   exerting   a   direct   radiative
+forcing  \citep{Liao-2005,  Bangert-2012,  Lundgren-2013}; or  by  the
+modification of  cloud properties, cloud fraction  and surface albedo,
+causing  a negative  indirect  radiative forcing  \citep{Haywood-2000,
+Lohmann-2005,     VandenHeever-2011,     Rosenfeld-2013}.      Despite
+considerable  progress in  global aerosol  modelling \citep{Mann-2013}
+and    measurement-based    assessments   \citep{Myhre-2009},    large
+uncertainties  remain  in   current  estimates  of  aerosol  radiative
+forcing   \citep{Myhre-2013,    IPCC-2013,   Lee-2013,   Randles-2013,
+Rosenfeld-2013, Sherwood-2013, Stier-2013}.
 
 Hence to  improve our understanding of  aerosol-cloud interactions and
-reduce  uncertainties  of aerosol  effects  in  climate, the  research
-community is making a  concerted international effort to represent the
-underlying chemical  processes through models.  These  models, such as
-ART (Aerosols  and Reactive Trace  gases) extension, developed  at the
-Karlsruhe Institute  of Technology (KIT),  offer a key  opportunity to
-reduce   the  climate  uncertainty,   particularly  on   the  regional
-scale \citep{Knote-2011, Bangert-2011,  Knote-2013}.  These models can
-be coupled with climate models,  such as the regional weather forecast
-model COSMO (Consortium  for Small-scale Modelling), jointly developed
-by  a consortium  of  European weather  centers  including the  German
-weather service  DWD and MeteoSwiss,  and used in the  climate version
-(COSMO-CLM)  by a  wide  research community.   The extended  COSMO-ART
-model provides  a detailed description of air  pollution chemistry and
-aerosol processes,  and is  mainly designed to  study air  quality and
-aerosol  meteorology  feedbacks  on  short, episodic  to  annual  time
-scales.  It is capable of  simulating aerosol distributions as well as
-their interactions over Europe as well as other regional domains.
+reduce  these  uncertainties,  the  research  community  is  making  a
+concerted international  effort to represent  the underlying physical,
+chemical  and aerosol dynamical  processes through  numerical chemical
+transport  models (CTMs)  such  as ART  (Aerosols  and Reactive  Trace
+gases),   developed   at  the   Karlsruhe   Institute  of   Technology
+(KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
+scale  model system  is online  coupled with  the  operational weather
+forecast  model  COSMO \citep{Baldauf-2011},  jointly  developed by  a
+consortium of European weather centers and used in the climate version
+(COSMO-CLM) by a wide research community.
 
-COSMO-ART is  computationally much more demanding than  the COSMO core
-since a  large number of additional  tracers and processes  have to be
-considered.  Thus this model is currently severely-limited in terms of
-applicability and expensive in terms of energy consumption.  COSMO has
-recently  been  ported  to  GPUs  within the  framework  of  the  High
-Performance  and  High  Productivity  Computing (HP2C)  Initiative  to
-optimize it  for computational and energy  efficiency.  Although these
-developments will  facilitate the  application of COSMO  for numerical
-weather prediction and climate  simulations, they do little to address
-the  coupling with  the  ART model  extension,  for which  significant
-investments  are still required  to take  it to  a similar  level. The
-efficiency of ART is being  addressed in the EU Exa2Green project. The
-ultimate goal  of the  project is to  deliver a prototype  code, which
-provides an energy  efficiency of at least five  times of the baseline
-value.   Such   an  implementation   would  allow  the   community  to
-investigate critical  questions at  higher resolution and  over longer
-periods, at reduced cost  to the environment.
+The extended  COSMO-ART model  is computationally much  more demanding
+than the  COSMO core  since a large  number of additional  tracers and
+processes  have  to  be  considered.   Thus this  model  is  currently
+severely-limited in  terms of applicability and expensive  in terms of
+energy consumption.   Although COSMO has recently been  ported to GPUs
+within  the framework of  the High  Performance and  High Productivity
+Computing  (HP2C)  Initiative to  optimize  it  for computational  and
+energy efficiency (\url{http://www.hp2c.ch/}), significant investments
+in  ART  are still  required  to  take it  to  a  similar level.   The
+efficiency of COSMO-ART is being addressed in the EU Exa2Green project
+(\url{http://exa2green.eu/}). The  ultimate goal of the  project is to
+deliver a  prototype code, which  provides an energy efficiency  of at
+least five times of the  baseline value.  Such an implementation would
+allow  the  community  to  investigate critical  questions  at  higher
+resolution  and   over  longer  periods,   at  reduced  cost   to  the
+environment.
 
-In  Sec.~\ref{sec:1},  we  introduce  the regional  atmospheric  model
-COSMO-ART  \citep{Vogel-2009}  that  accounts  for  feedbacks  between
-chemistry,  aerosols, radiation, and  clouds and  additionally specify
-its technical setup related to the investigated performance and energy
-evaluation  methods.   Sec.~\ref{sec:2}  describes the  different  HPC
-systems   and  associated  power   measurement  equipments   used  for
-conducting the benchmarking study and presents energy-aware techniques
-employed  to   determine  the  energy  footprint   of  the  considered
-application.   Sec.~\ref{sec:3}  gathers  benchmark results  from  the
-Exa2Green  project and  highlights  areas where  improvements will  be
-necessary for the subsequent  baseline refactoring.  Some related work
-will  be  discussed  in  Sec.~\ref{sec:4} Finally,  we  conclude  with
-practical  guidelines to identifying  energy saving  opportunities and
-give  directions for further  developments of  the modeling  system in
-Sec.~\ref{concl}.
+Some  related work  will  be discussed  in  Sec.~\ref{sec:1}. Then  in
+Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
+COSMO-ART \citep{Vogel-2009}  and specify its  technical setup related
+to  the  investigated   performance  and  energy  evaluation  methods.
+Sec.~\ref{sec:3} describes different power measurement equipments used
+for  conducting  this  benchmarking  study and  presents  energy-aware
+techniques  employed   to  determine  the  energy   footprint  of  the
+considered  application.  Sec.~\ref{sec:4}  gathers  benchmark results
+from  the Exa2Green  project and  highlights areas  where improvements
+will be  necessary for the subsequent  baseline refactoring.  Finally,
+we  conclude with  practical guidelines  to identifying  energy saving
+opportunities  and give  directions  for further  developments of  the
+modeling system in Sec.~\ref{concl}.

### 2014-04-12 trunk/paper_enahpc14/Sec3:Measurement-framework.tex
adds minor changes

--- 2014-04-12 paper_enahpc14/Sec3:Measurement-framework.tex
@@
 software including  the E3METER  Cacti Plugin to  scan the  entire PLC
-network and monitor the power usage of individual rack in real-time.
+network and monitor  in real-time the power usage  of individual rack,
+recorded in 5 minutes interval periods.
 

### 2014-04-12 trunk/paper_enahpc14/Sec4:Evaluation.tex
adds minor changes

@@
 have    an   operating   system    based   on    GNU/Linux   featuring
-``2.6.32-358.11.1.el6.x86\_64''  and  ``3.0.101-0.15-default'' kernels
-for Monch and Pilatus respectively.
+``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
+``3.0.101-0.15-default'' kernel in Pilatus.
 

### 2014-04-15 paper_enahpc14/Introduction.tex
updates Section 2: "The COSMO-ART model system"

@@
 of   solar   radiation,    thereby   exerting   a   direct   radiative
-forcing  \citep{Liao-2005,  Bangert-2012,  Lundgren-2013}, or  by  the
+forcing  \citep{Liao-2005,  Bangert-2012,  Lundgren-2013}; or  by  the
 modification of  cloud properties, cloud fraction  and surface albedo,
@@
 and    measurement-based    assessments   \citep{Myhre-2009},    large
-uncertainties    remain    in    current    estimates    of    aerosol
+uncertainties  remain  in   current  estimates  of  aerosol  radiative
 forcing   \citep{Myhre-2013,    IPCC-2013,   Lee-2013,   Randles-2013,
@@
 Hence to  improve our understanding of  aerosol-cloud interactions and
-reduce  uncertainties  in  aerosol  radiative forcing  estimates,  the
-research  community  is making  a  concerted  international effort  to
-represent  the  underlying physical,  chemical  and aerosol  dynamical
-processes through numerical chemical  transport models (CTMs), such as
-ART (Aerosols  and Reactive Trace  gases), developed at  the Karlsruhe
-Institute   of  Technology   (KIT)   \citep{Vogel-2009,  Bangert-2011,
-Knote-2013}. This  regional scale model system is  online coupled with
-the  operational weather  forecast  model COSMO  \citep{Baldauf-2011},
-jointly developed by a consortium of European weather centers and used
-in the climate version (COSMO-CLM)  by a wide research community.
+reduce  these  uncertainties,  the  research  community  is  making  a
+concerted international  effort to represent  the underlying physical,
+chemical  and aerosol dynamical  processes through  numerical chemical
+transport  models (CTMs)  such  as ART  (Aerosols  and Reactive  Trace
+gases),   developed   at  the   Karlsruhe   Institute  of   Technology
+(KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
+scale  model system  is online  coupled with  the  operational weather
+forecast  model  COSMO \citep{Baldauf-2011},  jointly  developed by  a
+consortium of European weather centers and used in the climate version
+(COSMO-CLM) by a wide research community.
 
@@
 Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
-COSMO-ART \citep{Vogel-2009}  and specify its  technical setup related
-to  the  investigated   performance  and  energy  evaluation  methods.
-Sec.~\ref{sec:3} describes different power measurement equipments used
-for  conducting  this  benchmarking  study and  presents  energy-aware
-techniques  employed   to  determine  the  energy   footprint  of  the
-considered  application.  Sec.~\ref{sec:4}  gathers  benchmark results
-from  the Exa2Green  project and  highlights areas  where improvements
-will be  necessary for the subsequent  baseline refactoring.  Finally,
-we  conclude with  practical guidelines  to identifying  energy saving
-opportunities  and give  directions  for further  developments of  the
-modeling system in Sec.~\ref{concl}.
+COSMO-ART and specify its  technical setup related to the investigated
+performance and energy evaluation methods.  Sec.~\ref{sec:3} describes
+different  power  measurement  equipments  used  for  conducting  this
+benchmarking  study and presents  energy-aware techniques  employed to
+determine  the   energy  footprint  of   the  considered  application.
+Sec.~\ref{sec:4} gathers benchmark  results from the Exa2Green project
+and  highlights areas  where improvements  will be  necessary  for the
+subsequent baseline refactoring.   Finally, we conclude with practical
+guidelines  to  identifying   energy  saving  opportunities  and  give
+directions  for  further  developments   of  the  modeling  system  in
+Sec.~\ref{concl}.

### 2014-04-15 paper_enahpc14/Sec2:COSMO-ART.tex
updates Section 2: "The COSMO-ART model system"

@@
 aerosol dynamics  and allows the online calculation  of reactive trace
-substances    and    their    interaction    with    the    atmosphere
-\citep{Knote-2011}.  Detailed  model   description  can  be  found  in
-\citep{Stanelle-2010, Bangert-2012, Knote-2011, Knote-2013}.
+substances and their interaction  with the atmosphere.  Detailed model
+description   can  be   found   in  \citep{Bangert-2012,   Knote-2011,
+  Knote-2013}.
 
@@
 the  absolute necessity that  results must  be reproducible  within an
-expected  variance for the  duration of  the Exa2Green  project. After
-consultation  with  the  climate  community  to  properly  define  the
-baseline,  it was  necessary to  find a  run-configuration  capable of
-being recreated in all subsequent versions of the code.
+expected variance for the duration of the Exa2Green project. To define
+a baseline,  it was necessary  to find a run-configuration  capable of
+being  recreated in  lal subsequent  versions.  Here  we give  a brief
+overview of the model setup  for our performance and energy efficiency
+evaluation.
 
-Thus, we give here a brief  overview of the main features of the model
-setup for  our benchmark's daily  runs.  Three-dimensional simulations
-are performed over  Europe for April 13th 2010, which  is close to the
-equinox and thus  brings benefits of having approximately  half day of
-night  and half  day  of  sun exposure,  therefore  ensuring a  proper
-activation of the chemistry cycle. We consider daily 24-hour forecasts
-without  any  previous spin-up  simulations.   The calculation  domain
-corresponds to  the CORDEX-EU-44  domain and is  covered by a  grid of
-$222\times   216$    points   with   a    horizontal   resolution   of
-$0.22\,^{\circ}$,  i.e.  50  km in  both directions,  and  40 vertical
-layers.  COSMO-ART requires the following input data:
-
-\begin{itemize}
-\item  Gas phase:  Anthropogenic emissions  for different  species and
-  land use data for biogenic emissions and deposition,
-\item Aerosol particles: Anthropogenic emissions,
-\item Mineral dust: Soil specific land use data.
-\end{itemize}
+Three-dimensional simulations are performed over large parts of Europe
+and the Mediterranean Sea for April 13th 2010 in order to benefit from
+around  half  day  of  sun  exposure and  therefore  ensure  a  proper
+activation of the chemistry cycle.  The calculation domain corresponds
+to the CORDEX-EU-44 domain and is covered by a grid of $222\times 216$
+points with  a horizontal resolution of $0.22\,^{\circ}$,  i.e.  50 km
+in  both directions and  40 vertical  layers.  These  24-hour forecast
+simulations  are not  preceded  by  a spin-up  phase  to build-up  the
+simulated gaseous and  aerosol concentrations.  This condition entails
+that at the  model initialization there is a  period of adjustment for
+the model  to reach its equilibrium  state and minimize  the effect of
+initial conditions for gases and aerosols on the model predictions.
 
@@
 Boundary data  for gas-phase species are taken  from IFS-MOZART output
-at   6h  temporal   resolution.   NAMELIST-input   specifying  runtime
-parameters    splitted   into    several   groups    are    given   in
-Table~\ref{tab:1}.   The model setup  incorporates 34  2-d and  45 3-d
-fields  to  be  written  out  every  hour  and  doesn't  include  data
+at 6h temporal resolution. The  model setup incorporates 34 2-d and 45
+3-d  fields  to be  written  out  every hour  and  is  devoid of  data
 assimilation methods.
 
-\begin{table}[htbf]
-  \begin{center}
-    \caption{}
-    \label{tab:1}
-    \begin{tabular}{ll}
-      \hline\noalign{\smallskip} 
-      \textbf{Group} & \textbf{Description} \\
-      \noalign{\smallskip}\hline\noalign{\smallskip}
-      LMGRID & Grid domain and size parameters \\
-      RUNCTL & Model run parameters \\
-      TUNING & Physics and dynamics parameters \\
-      DYNCTL & Adiabatic model parameters \\
-      PHYCTL & Diabatic model parameters \\
-      COSMO\_ART & Gases and aerosols model parameters \\
-      DIACTL & Diagnostic calculations parameters \\
-      SATCTL & Synthetic satellite images parameters \\
-      IOCTL & I/O environment parameters \\
-      GRIBIN & GRIB input parameters \\
-      GRIBOUT & GRIB output parameters \\
-     \noalign{\smallskip}\hline
-    \end{tabular}
-  \end{center}
-\end{table}
-
-This COSMO-ART version is configured with a semi Lagrangian horizontal
-advection  scheme with  tricubic interpolation  and  selective filling
-diffusion  option  in  combination   with  the  dynamical  core  using
-Runge-Kutta time stepping \citep{}.  It  also makes use of the Kinetic
-PreProcessor solver (KPP) for  the resolution of atmospheric chemistry
-ordinary  differential equations \citep{Damian-2002}.   Concerning the
-modelling  of  wet  deposition  in  aerosols, the  baseline  has  only
-indirect  cloud  feedbacks  but  doesn't include  in-cloud  scavenging
-(rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
-parameterisations,   precipitation  formation   is   performed  by   a
-two-moment  cloud  microphysics scheme  instead  of  a classical  bulk
-microphysics scheme.
+The  considered  COSMO-ART model  system  is  configured  with a  semi
+Lagrangian horizontal advection scheme with tricubic interpolation and
+selective filling  diffusion option in combination  with the dynamical
+core  using Runge-Kutta  time  stepping \citep{COSMO-PartI-2011}.   It
+also  makes use  of  the  Kinetic PreProcessor  solver  (KPP) for  the
+resolution  of atmospheric  chemistry ordinary  differential equations
+\citep{Damian-2002}.   Concerning the modelling  of wet  deposition in
+aerosols, the  baseline has only indirect cloud  feedbacks but doesn't
+include  in-cloud  scavenging  (rainout)  and  below-cloud  scavenging
+(washout)  yet.   Amongst  physical  parameterisations,  precipitation
+formation is performed by a two-moment cloud microphysics.

### 2014-04-16 paper_enahpc14/Sec3:Measurement-framework.tex
updates "Power-performance measurement framework" & "Evaluation" sections; update bibliography

@@
 employ   a    version   of   the    integrated   framework   presented
-in~\cite{energy13}  that  works in  combination  with VampirTrace  and
-Vampir,   which  are   profiling/tracing   and  visualization   tools,
+in~\cite{energy13}  that  works in  combination  with Extrae  and
+Paraver,   which  are   profiling/tracing   and  visualization   tools,
 respectively.
@@
 %the Lustre architecture; the right depicts the tracing and profiling framework.
-To  use our  approach,  COSMO-ART is  compiled  using the  VampirTrace
+To  use our  approach,  COSMO-ART is  compiled  using the  Extrae
 compiler wrappers, which automatically  instrument the Fortran code of
 the  model. Next,  COSMO-ART is  run  on the  nodes, thus  dissipating
-certain  amount of  power.  The  server nodes  are connected  to power
+certain  amount of  power.  These  nodes are connected  to power 
 measurement  devices that  account for  the dissipated power/consumed 
-energy and  send the power data  to the tracing  server.  The attached
-VampirTrace \pmlib plugin employs the client API that sends start/stop
-primitives in order to gather captured data by the wattmeters onto the
-tracing server,  where an  instance of the  \pmlib server  is running.
-Once COSMO-ART run is finished, the VampirTrace \pmlib plugin receives
-the  power   data  from  the  tracing   server.   The  instrumentation
-post-processing generates  the performance trace files  and the \pmlib
-plugin inserts the power data into them.
+energy and  send the power data  to the tracing  server. The client, 
+meanwhile, sends start/stop primitives in order to gather captured data
+by the wattmeters onto the tracing server, where an  instance of the 
+\pmlib server  is running.
 
-In  addition  to the  power  measurements,  we  also account  for  the
-resource utilization values  of the nodes: CPU load,  memory usage and
-storage device utilization. % and network utilization.
+Once COSMO-ART run is finished, a file containing the power profile is 
+created using the  power   data  received from  the  tracing   server and
+the  instrumentation post-processing generates  the performance trace files.
+All these files are combined then in Paraver which allows us to visualize 
+the performance trace and the power profile of COSMO-ART all together.
 
-We  run special  \pmlib  server  instances on  the  server nodes  that
-retrieve these  values from the \texttt{proc}  file system (leveraging
-the  \texttt{psutil} Python library).   Thus, \pmlib  plugin instances
-running  with the  instrumented  application connect  with the  \pmlib
-servers.   Finally,   using  the   Vampir   visualization  tool,   the
-power-performance traces  can be easily  analyzed through a  series of
-plots and statistics.
+%Once COSMO-ART run is finished, the VampirTrace \pmlib plugin receives
+%the  power   data  from  the  tracing   server.   The  instrumentation
+%post-processing generates  the performance trace files  and the \pmlib
+%plugin inserts the power data into them.
+
+%In  addition  to the  power  measurements,  we  also account  for  the
+%resource utilization values  of the nodes: CPU load,  memory usage and
+%storage device utilization. % and network utilization.
+
+%We  run special  \pmlib  server  instances on  the  server nodes  that
+%retrieve these  values from the \texttt{proc}  file system (leveraging
+%the  \texttt{psutil} Python library).   Thus, \pmlib  plugin instances
+%running  with the  instrumented  application connect  with the  \pmlib
+%servers.   Finally,   using  the   Vampir   visualization  tool,   the
+%power-performance traces  can be easily  analyzed through a  series of
+%plots and statistics.

### 2014-04-16 paper_enahpc14/Sec4:Evaluation.tex
updates "Power-performance measurement framework" & "Evaluation" sections; update bibliography

@@
 
+\subsubsection{Tintorrum (UJI)}
+Tintorrum, which is installed at the Jaume I University (UJI),  is an 
+heterogeneous cluster composed of 28 compute nodes. For our experiments
+only a subset of these nodes were considered. This set is composed of 
+16 nodes, each of which includes two Intel Xeon E5645 hexa-core processors at 
+2.40 GHz connected via Infiniband QDR (Mellanox MTS3600 switch). 
+Each core has 32 KB instruction and 32 KB data L1 caches and 
+256 KB L2 cache. The 6 cores share a 12 MB L3 cache and the platform has 
+288 GB of DDR3 1333 MHz.
+
 \subsubsection{Sofware environment}

### 2014-04-16 trunk/paper_enahpc14/Sec3:Measurement-framework.tex
add minor changes to the bibliography, section 3 & section 4

@@
 employ   a    version   of   the    integrated   framework   presented
-in~\cite{energy13}  that  works in  combination  with VampirTrace  and
-Vampir,   which  are   profiling/tracing   and  visualization   tools,
-respectively.
+in~\cite{energy13} that works in  combination with Extrae and Paraver,
+which are profiling/tracing and visualization tools, respectively.
+
 %The left part of the \vref{fig:Lustre} offers a graphical representation of
 %the Lustre architecture; the right depicts the tracing and profiling framework.
-To  use our  approach,  COSMO-ART is  compiled  using the  VampirTrace
-compiler wrappers, which automatically  instrument the Fortran code of
-the  model. Next,  COSMO-ART is  run  on the  nodes, thus  dissipating
-certain  amount of  power.  The  server nodes  are connected  to power
-measurement  devices that  account for  the  dissipated power/consumed
-energy and  send the power data  to the tracing  server.  The attached
-VampirTrace \pmlib plugin employs the client API that sends start/stop
-primitives in order to gather captured data by the wattmeters onto the
-tracing server,  where an  instance of the  \pmlib server  is running.
-Once COSMO-ART run is finished, the VampirTrace \pmlib plugin receives
-the  power   data  from  the  tracing   server.   The  instrumentation
-post-processing generates  the performance trace files  and the \pmlib
-plugin inserts the power data into them.
+To use our  approach, COSMO-ART is compiled using  the Extrae compiler
+wrappers,  which  automatically instrument  the  Fortran  code of  the
+model. Next, COSMO-ART  is run on the nodes,  thus dissipating certain
+amount  of power.   These  nodes are  connected  to power  measurement
+devices that account for the dissipated power/consumed energy and send
+the power  data to  the tracing server.  The client,  meanwhile, sends
+start/stop  primitives  in  order  to  gather  captured  data  by  the
+wattmeters onto  the tracing server,  where an instance of  the \pmlib
+server is running.
 
-In  addition  to the  power  measurements,  we  also account  for  the
-resource utilization values  of the nodes: CPU load,  memory usage and
-storage device utilization. % and network utilization.
+Once COSMO-ART run is finished, a file containing the power profile is
+created using the power data  received from the tracing server and the
+instrumentation post-processing generates the performance trace files.
+All  these files  are  combined then  in  Paraver which  allows us  to
+visualize the performance trace and the power profile of COSMO-ART all
+together.
 
-We  run special  \pmlib  server  instances on  the  server nodes  that
-retrieve these  values from the \texttt{proc}  file system (leveraging
-the  \texttt{psutil} Python library).   Thus, \pmlib  plugin instances
-running  with the  instrumented  application connect  with the  \pmlib
-servers.   Finally,   using  the   Vampir   visualization  tool,   the
-power-performance traces  can be easily  analyzed through a  series of
-plots and statistics.
+%Once COSMO-ART run is finished, the VampirTrace \pmlib plugin receives
+%the  power   data  from  the  tracing   server.   The  instrumentation
+%post-processing generates  the performance trace files  and the \pmlib
+%plugin inserts the power data into them.
+
+%In  addition  to the  power  measurements,  we  also account  for  the
+%resource utilization values  of the nodes: CPU load,  memory usage and
+%storage device utilization. % and network utilization.
+
+%We  run special  \pmlib  server  instances on  the  server nodes  that
+%retrieve these  values from the \texttt{proc}  file system (leveraging
+%the  \texttt{psutil} Python library).   Thus, \pmlib  plugin instances
+%running  with the  instrumented  application connect  with the  \pmlib
+%servers.   Finally,   using  the   Vampir   visualization  tool,   the
+%power-performance traces  can be easily  analyzed through a  series of
+%plots and statistics.

### 2014-04-16 paper_enahpc14/Sec4:Evaluation.tex
add minor changes to the bibliography, section 3 & section 4

@@
-In  this  section, we  present  a  performance  and energy  efficiency
+In  this section,  we  describe a  performance  and energy  efficiency
 evaluation  of  different  achitectures  when  running  the  COSMO-ART
@@
 we detail the environment  setup gathering the considered HPC systems,
-the software  environment and the runtime  configuration.  Finally, we
-discuss  benchmark   results  and  power-performance   traces  of  the
-application.
+the  software  environment and  the  run  configuration.  Finally,  we
+discuss benchmark  results and  power-performance traces of  the model
+system.
 
@@
 \textit{time-to-solution}       and       \textit{energy-to-solution}.
-Time-to-Solution  refers   to  the  total  wall  clock   time  of  the
+Time-to-solution  refers   to  the  total  wall  clock   time  of  the
 application execution time. Energy-to-solution is the amount of energy
@@
 \subsubsection{Tintorrum (UJI)}
-Tintorrum, which is installed at the Jaume I University (UJI),  is an 
-heterogeneous cluster composed of 28 compute nodes. For our experiments
-only a subset of these nodes were considered. This set is composed of 
-16 nodes, each of which includes two Intel Xeon E5645 hexa-core processors at 
-2.40 GHz connected via Infiniband QDR (Mellanox MTS3600 switch). 
-Each core has 32 KB instruction and 32 KB data L1 caches and 
-256 KB L2 cache. The 6 cores share a 12 MB L3 cache and the platform has 
-288 GB of DDR3 1333 MHz.
+Tintorrum, which  is installed at the  Jaume I University  (UJI), is a
+heterogeneous  cluster   composed  of  28  compute   nodes.   For  our
+experiments only a subset of  these nodes were considered. This set is
+composed  of 16 nodes,  each of  which includes  two Intel  Xeon E5645
+hexa-core  processors  at  2.40   GHz  connected  via  Infiniband  QDR
+(Mellanox MTS3600 switch).  Each core  has 32 KB instruction and 32 KB
+data L1 caches and 256 KB L2 cache. The 6 cores share a 12 MB L3 cache
+and the platform has 288 GB of DDR3 1333 MHz.
 

### 2014-04-18 paper_enahpc14/Introduction.tex
adds very few changes; adds Cristiano Malossi as a co-author

@@
 subsequent baseline refactoring.   Finally, we conclude with practical
-guidelines  to  identifying   energy  saving  opportunities  and  give
-directions  for  further  developments   of  the  modeling  system  in
-Sec.~\ref{concl}.
+guidelines  to identifying  energy  saving opportunities  and give  an
+outlook to future research in Sec.~\ref{concl}.

### 2014-04-18 paper_enahpc14/Sec4:Evaluation.tex
adds very few changes; adds Cristiano Malossi as a co-author

@@
 processors, conventional  in HPC  systems and known  to be  more power
-consuming.
+consuming.   Finally a  full tracing  experiment is  conducted  on the
+``Tintorrum''  cluster using  an integrated  framework, to  capture an
+overall  power profile  at a  much finer  resolution and  to  get more
+insights  into  the  MPI  blocking  and polling  influences  on  power
+savings.
 
@@
 nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
-Each standard  compute node comprises two Intel  Ivy Bridge-EP E5-2660
-v2 ten-core processors operating at 2.2 GHz, themselves connected by a
-high speed  InfiniBand network, based  on Mellanox SX6036  managed FDR
-switches, with a  56 Gb/s speed.  Each core has  32 KB instruction and
-32 KB data L1  caches and 256 KB of L2 cache. All  the 10 core share a
-25 MB L3 cache and the platform has 32GB of DDR3 1600 MHz RAM. For our
-energy-to-solution benchmark,  a full rack of Monch  constituted of 52
-standard  compute   nodes  (monchc[029-080])  was   considered.
+Each standard  compute node comprises  two Intel Ivy  Bridge Efficient
+Performance (EP)  E5-2660 v2 ten-core processors operating  at 2.2 GHz
+base  clock speed,  themselves connected  by a  high  speed InfiniBand
+network based on Mellanox SX6036  managed FDR switches, with a 56 Gb/s
+speed.  Each core  has 32 KB instruction and 32 KB  data L1 caches and
+256 KB  of L2 cache. All  the 10 core share  a 25 MB L3  cache and the
+platform has  32GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
+benchmark, a  full rack  of Monch constituted  of 52  standard compute
+nodes (monchc[029-080]) was considered.
 
@@
 Pilatus,  which  is installed  at  the  Swiss National  Supercomputing
-Center (CSCS)  of ETH Zurich,  is a dual-socket Intel  Sandy Bridge-EP
-based cluster  used as Piz  Daint pre-post processing cluster.   It is
-composed of 42 compute nodes  and has 2 high-speed interconnects based
-on FDR:  the first is dedicated to  the MPI traffic and  the second to
-the storage high speed traffic.  The 2 login nodes and the 42 computes
-nodes  consists in  11  twin-pair Intel  E5-Series  DALCO r2264i4t  2U
-scalable compute modules.  Each  module contains 4 compute nodes based
-on two Intel Xeon E5-2670  eight-core processors operating at 2.6 GHz,
-themselves  connected by  a high  speed InfiniBand  network,  based on
-Mellanox SX6036 managed FDR switches,  with a 56 Gb/s speed. Each core
-has 32 KB instruction and 32 KB data L1 caches and 256 KB of L2 cache.
-All the  8 core share a  20 MB L3 cache  and the platform  has 64GB of
-DDR3 1600 MHz  RAM. For our energy-to-solution benchmark,  a full rack
-of Pilatus  constituted of 42 standard  compute nodes (pilatus[03-44])
-was considered.
+Center (CSCS) of  ETH Zurich, is a dual-socket  eight-core Intel Sandy
+Bridge EP based cluster used as Piz Daint pre-post processing cluster.
+It is composed of 42  compute nodes and has 2 high-speed interconnects
+based on FDR: the first is dedicated to the MPI traffic and the second
+to  the storage  high speed  traffic.  The  2 login  nodes and  the 42
+computes nodes consists in 11 twin-pair Intel E5-Series DALCO r2264i4t
+2U  scalable compute modules.   Each module  contains 4  compute nodes
+based on two  Intel Xeon E5-2670 processors operating  at 2.6 GHz base
+clock speed,  themselves connected by a high  speed InfiniBand network
+based  on  Mellanox  SX6036  managed  FDR switches,  with  a  56  Gb/s
+speed. Each  core has 32 KB instruction  and 32 KB data  L1 caches and
+256 KB  of L2 cache.  All  the 8 core share  a 20 MB L3  cache and the
+platform has  64GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
+benchmark, a full  rack of Pilatus constituted of  42 standard compute
+nodes (pilatus[03-44]) was considered.
 
@@
 Interface (MPI)  and are purely  MPI-parallel.  The software  stack on
-both many-core  platforms was  controlled using the  modules framework
-which gives  an easy and  flexible mechanism to  access to all  of the
-CSCS  provided compilers,  tools  and applications.   For our  initial
+both  many-core  CSCS  platforms  was  controlled  using  the  modules
+framework which gives an easy  and flexible mechanism to access to all
+of the  provided compilers, tools  and applications.  For  our initial
 benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,

### 2014-04-18 paper_enahpc14/paper.tex
adds very few changes; adds Cristiano Malossi as a co-author

@@
 \author{Joseph~Charles \and William~Sawyer \and Manuel~F.~Dolz \and
-  Sandra~Catal\'an}
+  Sandra~Catal\'an \and Cristiano~Malossi }
 
@@
 	\\ ES-12071 Castell\'on, Spain
-        \\ \email{catalans@icc.uji.es}}
+        \\ \email{catalans@icc.uji.es}
+        \and C.~Malossi  
+        \at IBM Zurich Research Laboratory 
+        \\ CH-8000 Zurich, Switzerland
+        \\ \email{ACM@zurich.ibm.com}
+}
 
@@
 \section{Introduction}
+\label{intro}
+\input{Introduction.tex}
+
+\section{Related work}
 \label{sec:1}
-\input{Sec1:Introduction.tex}
+\input{Sec1:Related.tex}
 
@@
 
-\section{Experimental setup}
+\section{Power-performance measurement framework}
 \label{sec:3}
-\input{Sec3:Power-measurement.tex}
+\input{Sec3:Measurement-framework.tex}
 
@@
 
-\section{Related work}
-\label{sec:5}
-\input{Sec5:Related.tex}
-
 \section{Conclusion}
-\label{sec:6}
-\label{Sec6:Conclusions.tex}
+\label{concl}
 \input{Conclusion.tex}

### 2014-04-19 trunk/paper_enahpc14/Introduction.tex
New version

@@
 
-Some  related work  will  be discussed  in  Sec.~\ref{sec:1}. Then  in
+This work is organized as follows: in Sec.~\ref{sec:1} we describe related work, then in
 Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
-COSMO-ART \citep{Vogel-2009}  and specify its  technical setup related
-to  the  investigated   performance  and  energy  evaluation  methods.
-Sec.~\ref{sec:3} describes different power measurement equipments used
-for  conducting  this  benchmarking  study and  presents  energy-aware
-techniques  employed   to  determine  the  energy   footprint  of  the
-considered  application.  Sec.~\ref{sec:4}  gathers  benchmark results
-from  the Exa2Green  project and  highlights areas  where improvements
-will be  necessary for the subsequent  baseline refactoring.  Finally,
-we  conclude with  practical guidelines  to identifying  energy saving
-opportunities  and give  directions  for further  developments of  the
-modeling system in Sec.~\ref{concl}.
+COSMO-ART and specify its  technical setup related to the investigated
+performance and energy evaluation methods.  Sec.~\ref{sec:5} describes
+the HPC supercomputers used to compute the baseline. Sec.~\ref{sec:3} describes
+different  power  measurement  equipment  used  for  conducting  this
+benchmarking  study and presents  energy-aware techniques  employed to
+determine  the   energy  footprint  of   the  considered  application.
+Sec.~\ref{sec:4} gathers benchmark  results from the Exa2Green project
+and  highlights areas  where improvements  will be  necessary  for the
+subsequent baseline refactoring.  Finally, we conclude with practical
+guidelines  to identifying  energy  saving opportunities  and give  an
+outlook to future research in Sec.~\ref{concl}.

### 2014-04-19 trunk/paper_enahpc14/Sec3:Measurement-framework.tex
New version

@@
-In this section,  we present two different frameworks  deployed on HPC
+In this section,  we present three different frameworks  deployed on HPC
 systems  to  measure the  power  consumption  and  performance of  the
@@
 
-\subsection{Framework description}
-\label{subsec:3.1}
+%\subsection{Framework description}
+%\label{subsec:3.1}
 
-\subsubsection{E3METER metering products}
+\subsection{E3METER metering products}
 Supercomputer  clusters considered  for our  experiments at  the Swiss
@@
 
-\subsubsection{University of Hamburg}
+\textbf{MALOSSI: I MOVED THIS PART HERE, SINCE IT IS RELATED TO CSCS MACHINES: IT NEEDS TO BE INTEGRATED IN THE TEXT ABOVE}
+Energy  consumption of  CSCS  clusters is
+assessed  by sampling  the instantaneous  peak power  during execution
+which  is then  averaged  and multiplied  by  the time-to-solution  to
+determine energy-to-solution.
+
+\subsection{University of Hamburg}
 To assess the  performance and the energy efficiency  of COSMO-ART, we
@@
 %plots and statistics.
+
+\subsection{IBM BG/Q integrated power measurement system}
+The IBM Blue~Gene/Q~(BG/Q) supercomputer is equipped with internal hardware sensors
+that provide 8~power measures for each node board (32~node cards, 512~cores), including total power, chip power, DDR~DIMM power, and network power.
+Therefore, no additional hardware tools are required to obtain power measurements.
+In order to instrument the code and collect power measurements in a file we use the MonEQ library~\cite{SeanWallace2013},
+which provides a simple interface to start and stop power measurement in specific regions of the code.
+Energy is computed a posteriori as the product of average power and time-to-solution.

### 2014-04-19 trunk/paper_enahpc14/Sec4:Evaluation.tex
New version

@@
-In  this  section, we  present  a  performance  and energy  efficiency
+In  this section,  we  describe a  performance  and energy  efficiency
 evaluation  of  different  achitectures  when  running  the  COSMO-ART
-baseline.  We  start by  specifying our measurement  methodology along
-with the metrics used to  analyse the results on all platforms.  Then,
-we detail the environment  setup gathering the considered HPC systems,
-the software  environment and the runtime  configuration.  Finally, we
-discuss  benchmark   results  and  power-performance   traces  of  the
-application.
+baseline.  We  start by  specifying the metrics used to analyse the results on all platforms.
+Then, we detail the environment  setup gathering the considered HPC systems,
+the  software  environment and  the  run  configuration.  Finally,  we
+discuss benchmark  results and  power-performance traces of  the model
+system.
 
@@
 \textit{time-to-solution}       and       \textit{energy-to-solution}.
-Time-to-Solution  refers   to  the  total  wall  clock   time  of  the
+Time-to-solution  refers   to  the  total  wall  clock   time  of  the
 application execution time. Energy-to-solution is the amount of energy
-spent  to achieve  results.  Energy  consumption of  CSCS  clusters is
-assessed  by sampling  the instantaneous  peak power  during execution
-which  is then  averaged  and multiplied  by  the time-to-solution  to
-determine energy-to-solution.   Whenever possible, multiple production
+spent  to achieve  results.  Whenever possible, multiple production
 runs of COSMO-ART were  performed to illustrate the reproducibility of
@@
 
-\subsection{Environment setup}
+\subsection{Sofware environment}
 \label{subsec:4.2}
-While hardware  platforms mature  and are replaced,  we have  chosen a
-state-of-the-art  Intel's  third-generation   Core  (aka  Ivy  Bridge)
-processing platform (called ``Monch'')  for our power measurements, as
-it  is slated  to stay  in service  without hardware  upgrade  for the
-duration  of  the Exa2Green  project.   In  principle  at least,  this
-architecture could be recreated or found in an identical configuration
-beyond the lifetime of the project.  Given that the baseline benchmark
-can be reproduced  within an expected variance, and  that the baseline
-run configuration  can be used in  all future versions of  the code, a
-fair comparison  will be made  between the baseline and  the milestone
-versions   of    COSMO-ART.    In   this    study,   a   complementary
-energy-to-solution  benchmarking  comparison  is  carried out  on  the
-``Pilatus'' cluster based  on Intel's previous-generation Sandy Bridge
-processors, conventional  in HPC  systems and known  to be  more power
-consuming.
-
-\subsubsection{Monch (CSCS - ETH Zurich)}
-Monch, which is installed  at the Swiss National Supercomputing Center
-(CSCS) of ETH Zurich, is  a 10 rack NEC-provided and dual-socket Intel
-Ivy Bridge-EP based cluster, utilised from people that are part of the
-Swiss    Platform   for    Advanced   Scientific    Computing   (PASC,
-\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
-nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
-Each standard  compute node comprises two Intel  Ivy Bridge-EP E5-2660
-v2 ten-core processors operating at 2.2 GHz, themselves connected by a
-high speed  InfiniBand network, based  on Mellanox SX6036  managed FDR
-switches, with a  56 Gb/s speed.  Each core has  32 KB instruction and
-32 KB data L1  caches and 256 KB of L2 cache. All  the 10 core share a
-25 MB L3 cache and the platform has 32GB of DDR3 1600 MHz RAM. For our
-energy-to-solution benchmark,  a full rack of Monch  constituted of 52
-standard  compute   nodes  (monchc[029-080])  was   considered.
-
-\subsubsection{Pilatus (CSCS - ETH Zurich)}
-Pilatus,  which  is installed  at  the  Swiss National  Supercomputing
-Center (CSCS)  of ETH Zurich,  is a dual-socket Intel  Sandy Bridge-EP
-based cluster  used as Piz  Daint pre-post processing cluster.   It is
-composed of 42 compute nodes  and has 2 high-speed interconnects based
-on FDR:  the first is dedicated to  the MPI traffic and  the second to
-the storage high speed traffic.  The 2 login nodes and the 42 computes
-nodes  consists in  11  twin-pair Intel  E5-Series  DALCO r2264i4t  2U
-scalable compute modules.  Each  module contains 4 compute nodes based
-on two Intel Xeon E5-2670  eight-core processors operating at 2.6 GHz,
-themselves  connected by  a high  speed InfiniBand  network,  based on
-Mellanox SX6036 managed FDR switches,  with a 56 Gb/s speed. Each core
-has 32 KB instruction and 32 KB data L1 caches and 256 KB of L2 cache.
-All the  8 core share a  20 MB L3 cache  and the platform  has 64GB of
-DDR3 1600 MHz  RAM. For our energy-to-solution benchmark,  a full rack
-of Pilatus  constituted of 42 standard  compute nodes (pilatus[03-44])
-was considered.
-
-\subsubsection{Sofware environment}
 INT2LM  and the  COSMO-ART model  are  implemented in  Fortran 90  for
 distributed  memory  parallel  computers  using  the  Message  Passing
-Interface (MPI)  and are purely  MPI-parallel.  The software  stack on
-both many-core  platforms was  controlled using the  modules framework
-which gives  an easy and  flexible mechanism to  access to all  of the
-CSCS  provided compilers,  tools  and applications.   For our  initial
+Interface (MPI)  and are purely  MPI-parallel.
+
+\textbf{MALOSSI: Here I suggest to create small subsections for each machine to avoid confusion especially because we need to put details of 4 different machines\ldots}
+
+The software  stack on
+both  many-core  CSCS  platforms  was  controlled  using  the  modules
+framework which gives an easy  and flexible mechanism to access to all
+of the  provided compilers, tools  and applications.  For  our initial
 benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,
@@
 
-\subsubsection{Run configuration}
+\subsubsection{IBM BG/Q}
+\textbf{THIS PART NEEDS TO BE COMPLETED. IT IS JUST A PLACEHOLDER}
+All tests are compiled with the IBM~XL~C/\cpp compiler, with the options: {\ttfamily\scriptsize -qarch=qp -qtune=qp -O3}, and involve double-precision arithmetic.
+
+\subsection{Run configuration}
+\label{subsec:4.3}
 A snapshot of the code, which includes, at least conceptually, all the
@@
 COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
-node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
-Pilatus.   The  calculated  region  was mapped  to  the  participating
-processors using  a 2D-partitioning strategy.   The distribution along
-the  x and  y  coordinates  was defined  by  setting: $nprocx=40$  and
-$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
+node on  Monch, on a  1344 cores using  16 MPI  tasks per  node on
+Pilatus, and on 512 cores using 32 MPI tasks per node on the IBM BG/Q.
+The  calculated  region  was mapped  to  the  participating
+processors using  a 2-D partitioning strategy.   \textbf{MALOSSI: I suggest to remove the following part. It adds low level details that probably are not useful for the reader at this level.}
+The distribution along
+the~$x$ and~$y$~coordinates  was defined  by  setting: $nprocx=40$  and
+$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$ \textbf{(for Pilatus?)} as $nprocx$ is
 usually  kept  bigger  than  $nprocy$.   Besides as  this  version  of
@@
 study as previous  attempts of its use revealed that  it always led to
-higher energy-to-solution.\\
+higher energy-to-solution \textbf{Also this last sentence is quite strong: I would remove it because we might reconsider this opportunity in a future stage of optimization}.\\
 
 \subsection{Experimental results}
-\label{subsec:4.3}
+\label{subsec:4.4}
 

### 2014-04-19 trunk/paper_enahpc14/Sec5:Supercomputers.tex
New version

In this section we provide a brief description of the HPC supercomputers
used to obtain the COSMO-ART baseline. For our power measurement we have chosen state-of-the-art Intel and IBM platforms;
both architectures could be recreated or found in an identical configuration
beyond the lifetime of the project.  Given that the baseline benchmark
can be reproduced  within an expected variance, and  that the baseline
run configuration  can be used in  all future versions of  the code, a
fair comparison  will be made  between the baseline and  the future milestone
versions   of    COSMO-ART. In addition, a   complementary energy-to-solution  benchmarking  comparison  is  carried out  on  the
``Pilatus'' cluster based  on Intel's previous-generation Sandy Bridge
processors, conventional  in HPC  systems and known  to be  more power
consuming.   Finally a  full tracing  experiment is  conducted  on the
``Tintorrum''  cluster using  an integrated  framework, to  capture an
overall  power profile  at a  much finer  resolution and  to  get more
insights  into  the  MPI  blocking  and polling  influences  on  power
savings.

% In  this section,  we  describe a  performance  and energy  efficiency
% evaluation  of  different  achitectures  when  running  the  COSMO-ART
% baseline.  We  start by  specifying our measurement  methodology along
% with the metrics used to  analyse the results on all platforms.  Then,
% we detail the environment  setup gathering the considered HPC systems,
% the  software  environment and  the  run  configuration.  Finally,  we
% discuss benchmark  results and  power-performance traces of  the model
% system.

% \subsection{Environment setup}
% \label{subsec:4.2}
% While hardware  platforms mature  and are replaced,  we have  chosen a
% state-of-the-art  Intel's  third-generation   Core  (aka  Ivy  Bridge)
% processing platform (called ``Monch'')  for our power measurements, as
% it  is slated  to stay  in service  without hardware  upgrade  for the
% duration  of  the Exa2Green  project.   In  principle  at least,  this
% architecture could be recreated or found in an identical configuration
% beyond the lifetime of the project.  Given that the baseline benchmark
% can be reproduced  within an expected variance, and  that the baseline
% run configuration  can be used in  all future versions of  the code, a
% fair comparison  will be made  between the baseline and  the milestone
% versions   of    COSMO-ART.    In   this    study,   a   complementary
% energy-to-solution  benchmarking  comparison  is  carried out  on  the
% ``Pilatus'' cluster based  on Intel's previous-generation Sandy Bridge
% processors, conventional  in HPC  systems and known  to be  more power
% consuming.   Finally a  full tracing  experiment is  conducted  on the
% ``Tintorrum''  cluster using  an integrated  framework, to  capture an
% overall  power profile  at a  much finer  resolution and  to  get more
% insights  into  the  MPI  blocking  and polling  influences  on  power
% savings.

\subsection{Monch (CSCS - ETH Zurich)}
Monch, which is installed  at the Swiss National Supercomputing Center
(CSCS) of ETH Zurich, is  a 10 rack NEC-provided and dual-socket Intel
Ivy Bridge-EP based cluster, utilised from people that are part of the
Swiss    Platform   for    Advanced   Scientific    Computing   (PASC,
\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
Each standard  compute node comprises  two Intel Ivy  Bridge Efficient
Performance (EP)  E5-2660 v2 ten-core processors operating  at 2.2 GHz
base  clock speed,  themselves connected  by a  high  speed InfiniBand
network based on Mellanox SX6036  managed FDR switches, with a 56 Gb/s
speed.  Each core  has 32 KB instruction and 32 KB  data L1 caches and
256 KB  of L2 cache. All  the 10 core share  a 25 MB L3  cache and the
platform has  32GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
benchmark, a  full rack  of Monch constituted  of 52  standard compute
nodes (monchc[029-080]) was considered.

\subsection{Pilatus (CSCS - ETH Zurich)}
Pilatus,  which  is installed  at  the  Swiss National  Supercomputing
Center (CSCS) of  ETH Zurich, is a dual-socket  eight-core Intel Sandy
Bridge EP based cluster used as Piz Daint pre-post processing cluster.
It is composed of 42  compute nodes and has 2 high-speed interconnects
based on FDR: the first is dedicated to the MPI traffic and the second
to  the storage  high speed  traffic.  The  2 login  nodes and  the 42
computes nodes consists in 11 twin-pair Intel E5-Series DALCO r2264i4t
2U  scalable compute modules.   Each module  contains 4  compute nodes
based on two  Intel Xeon E5-2670 processors operating  at 2.6 GHz base
clock speed,  themselves connected by a high  speed InfiniBand network
based  on  Mellanox  SX6036  managed  FDR switches,  with  a  56  Gb/s
speed. Each  core has 32 KB instruction  and 32 KB data  L1 caches and
256 KB  of L2 cache.  All  the 8 core share  a 20 MB L3  cache and the
platform has  64GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
benchmark, a full  rack of Pilatus constituted of  42 standard compute
nodes (pilatus[03-44]) was considered.

\subsection{Tintorrum (UJI)}
Tintorrum, which  is installed at the  Jaume I University  (UJI), is a
heterogeneous  cluster   composed  of  28  compute   nodes.   For  our
experiments only a subset of  these nodes were considered. This set is
composed  of 16 nodes,  each of  which includes  two Intel  Xeon E5645
hexa-core  processors  at  2.40   GHz  connected  via  Infiniband  QDR
(Mellanox MTS3600 switch).  Each core  has 32 KB instruction and 32 KB
data L1 caches and 256 KB L2 cache. The 6 cores share a 12 MB L3 cache
and the platform has 288 GB of DDR3 1333 MHz.

\subsection{Blue Gene/Q (IBM Research -- Zurich)}
The IBM Blue~Gene/Q~(BG/Q) supercomputer~\cite{IBMRedBookBGQHardware} is the
third-generation computer architecture in the Blue~Gene family. It is composed
by one or more compute racks, where each rack contains 1024 16-core compute cards,
with a peak performance of 209~TFLOPS per rack. Each compute card contains 16~IBM~Blue~Gene/Q
PowerPC~A2 core processors and 16 GB of memory. The chip has 16~cores (with
up to 4~threads per core) at~1.60~GHz, with 16~kB~L1 cache each. In addition,
it has 32~MB~L2 shared cache, and 2~controllers which provide access to 2~RAM banks
of 8~GB each, both at~1333~MHz. Moreover, each core has an L1~prefetcher that is shared by
the threads on the core. The network topology for Blue Gene/Q is a five-dimensional torus,
with direct links between the nearest neighbors in the $\pm$A, $\pm$B, $\pm$C, $\pm$D, and $\pm$E
directions and 40~GBps bandwidth.

### 2014-04-19 paper_enahpc14/Sec2:COSMO-ART.tex
New version

@@
 Three-dimensional simulations are performed over large parts of Europe
-and the Mediterranean Sea for April 13th 2010 in order to benefit from
+and the Mediterranean~Sea for April 13th 2010 in order to benefit from
 around  half  day  of  sun  exposure and  therefore  ensure  a  proper
@@
 to the CORDEX-EU-44 domain and is covered by a grid of $222\times 216$
-points with  a horizontal resolution of $0.22\,^{\circ}$,  i.e.  50 km
-in  both directions and  40 vertical  layers.  These  24-hour forecast
+points with  a horizontal resolution of $0.22\,^{\circ}$,  i.e.,  50~km
+in  both directions and  40~vertical  layers.  These  24-hour forecast
 simulations  are not  preceded  by  a spin-up  phase  to build-up  the
@@
 
-The meteorological initial and  bounday conditions are obtained by the
+The meteorological initial and  boundary conditions are obtained by the
 the ECMWF  global spectral model IFS  with an update  frequency of 3h.
 Boundary data  for gas-phase species are taken  from IFS-MOZART output
-at 6h temporal resolution. The  model setup incorporates 34 2-d and 45
-3-d  fields  to be  written  out  every hour  and  is  devoid of  data
+at 6h temporal resolution. The  model setup incorporates 34~2-D and
+45~3-D  fields  to be  written  out  every hour  and  is  devoid of  data
 assimilation methods.
@@
 \citep{Damian-2002}.   Concerning the modelling  of wet  deposition in
-aerosols, the  baseline has only indirect cloud  feedbacks but doesn't
+aerosols, the  baseline has only indirect cloud  feedbacks but does not
 include  in-cloud  scavenging  (rainout)  and  below-cloud  scavenging

### 2014-04-19 paper_enahpc14/paper.tex
New version

@@
 \usepackage{graphicx}
+\usepackage[USenglish]{babel}
+\usepackage[T1]{fontenc}
 \usepackage{hyperref}
@@
 \usepackage{xspace}
+%\usepackage{cite}
 %
@@
 %
+
+%
+\def\plusNoSpace{\texttt{+}}
+\def\plus{\texttt{+}\xspace}
+\def\minusNoSpace{\texttt{-}}
+\def\minus{\texttt{-}\xspace}
+\def\cpp{C{}\plusNoSpace\plusNoSpace\xspace}
+
 \begin{document}
@@
 \title{Evaluating the Performance  and Energy Efficiency of COSMO-ART,
-  a Coupled Numerical Weather Forecast and Chemical Transport Model}
+  a Coupled Numerical Weather Forecast and Chemical Transport Model\thanks{IBM, the IBM~logo, ibm.com, and Blue~Gene/Q are trademarks or registered trademarks of International Business Machines Corporation in the United~States, other countries, or both.}}
 % \thanks{This work was supported by the EU Project FP7 318793 ``EXA2GREEN''}}
@@
 \author{Joseph~Charles \and William~Sawyer \and Manuel~F.~Dolz \and
-  Sandra~Catal\'an \and Cristiano~Malossi }
+  Sandra~Catal\'an \and A.~Cristiano~I.~Malossi \and Costas~Bekas}
 
@@
         \\ \email{catalans@icc.uji.es}
-        \and C.~Malossi  
-        \at IBM Zurich Research Laboratory 
-        \\ CH-8000 Zurich, Switzerland
-        \\ \email{ACM@zurich.ibm.com}
+        \and A.~C.~I.~Malossi, C.~Bekas
+        \at IBM Research -- Zurich
+        \\ CH-8803 R\"uschlikon, Switzerland
+        \\ E-mails:~{\{acm,bek\}@zurich.ibm.com}
 }
@@
 
+\section{HPC Supercomputers}
+\label{sec:5}
+\input{Sec5:Supercomputers.tex}
+
 \section{Power-performance measurement framework}

### 2014-04-22 paper_enahpc14/Sec3:Measurement-framework.tex
cancels previous commits of Cristiano as IBM Zurich is actually not going to contribute to this work

@@
 employ   a    version   of   the    integrated   framework   presented
-in~\cite{energy13}  that  works in  combination  with Extrae  and
-Paraver,   which  are   profiling/tracing   and  visualization   tools,
-respectively.
+in~\cite{energy13} that works in  combination with Extrae and Paraver,
+which are profiling/tracing and visualization tools, respectively.
+
 %The left part of the \vref{fig:Lustre} offers a graphical representation of
 %the Lustre architecture; the right depicts the tracing and profiling framework.
-To  use our  approach,  COSMO-ART is  compiled  using the  Extrae
-compiler wrappers, which automatically  instrument the Fortran code of
-the  model. Next,  COSMO-ART is  run  on the  nodes, thus  dissipating
-certain  amount of  power.  These  nodes are connected  to power 
-measurement  devices that  account for  the dissipated power/consumed 
-energy and  send the power data  to the tracing  server. The client, 
-meanwhile, sends start/stop primitives in order to gather captured data
-by the wattmeters onto the tracing server, where an  instance of the 
-\pmlib server  is running.
+To use our  approach, COSMO-ART is compiled using  the Extrae compiler
+wrappers,  which  automatically instrument  the  Fortran  code of  the
+model. Next, COSMO-ART  is run on the nodes,  thus dissipating certain
+amount  of power.   These  nodes are  connected  to power  measurement
+devices that account for the dissipated power/consumed energy and send
+the power  data to  the tracing server.  The client,  meanwhile, sends
+start/stop  primitives  in  order  to  gather  captured  data  by  the
+wattmeters onto  the tracing server,  where an instance of  the \pmlib
+server is running.
 
 Once COSMO-ART run is finished, a file containing the power profile is
-created using the  power   data  received from  the  tracing   server and
-the  instrumentation post-processing generates  the performance trace files.
-All these files are combined then in Paraver which allows us to visualize 
-the performance trace and the power profile of COSMO-ART all together.
+created using the power data  received from the tracing server and the
+instrumentation post-processing generates the performance trace files.
+All  these files  are  combined then  in  Paraver which  allows us  to
+visualize the performance trace and the power profile of COSMO-ART all
+together.
 

### 2014-04-22 paper_enahpc14/paper.tex
cancels previous commits of Cristiano as IBM Zurich is actually not going to contribute to this work

--- 2014-04-18 paper_enahpc14/paper.tex
@@
 \usepackage{graphicx}
+\usepackage[T1]{fontenc}
 \usepackage{hyperref}
@@
 \title{Evaluating the Performance and Energy Efficiency of COSMO-ART,
-  a Coupled Numerical Weather Forecast and Chemical Transport Model}
-% \thanks{This work was supported by the EU Project FP7 318793 ``EXA2GREEN''}}
-%\subtitle{Write here subtitle}
+  a Coupled Numerical Weather Forecast and Chemical Transport Model.}
+%\thanks{}
+%\subtitle{}
 
@@
 \author{Joseph~Charles \and William~Sawyer \and Manuel~F.~Dolz \and
-  Sandra~Catal\'an \and Cristiano~Malossi }
+  Sandra~Catal\'an}
 
@@
         \\ \email{catalans@icc.uji.es}
-        \and C.~Malossi  
-        \at IBM Zurich Research Laboratory 
-        \\ CH-8000 Zurich, Switzerland
-        \\ \email{ACM@zurich.ibm.com}
 }

### 2014-04-22 trunk/paper_enahpc14/Sec3:Measurement-framework.tex
commits Will's slight corrections

--- 2014-04-16 trunk/paper_enahpc14/Sec3:Measurement-framework.tex
@@
 
-\subsubsection{E3METER metering products}
+\subsubsection{CSCS - E3METER metering products}
 Supercomputer  clusters considered  for our  experiments at  the Swiss
@@
 made  available  via SNMP,  HTTP,  TELNET  through  the built-in  Fast
-Ethernet  port.   Time  synchronization  is guaranteed  by  using  NTP
+Ethernet  port.   Time  synchronisation  is guaranteed  by  using  NTP
 servers.   Measured data  is accessed  through the  open  source Cacti
@@
 start/stop  primitives  in  order  to  gather  captured  data  by  the
-wattmeters onto  the tracing server,  where an instance of  the \pmlib
+watt-meters onto the  tracing server, where an instance  of the \pmlib
 server is running.
@@
 All  these files  are  combined then  in  Paraver which  allows us  to
-visualize the performance trace and the power profile of COSMO-ART all
+visualise the performance trace and the power profile of COSMO-ART all
 together.

### 2014-04-22 trunk/paper_enahpc14/paper.tex
commits Will's slight corrections

@@
 	\\ ES-12071 Castell\'on, Spain
-        \\ \email{catalans@icc.uji.es}}
+        \\ \email{catalans@icc.uji.es}
+}
 
@@
 \begin{abstract}
-  In  this paper  we present  the energy-to-solution  of  COSMO-ART on
-  various platforms.   This model is  an extension of  the operational
-  weather  forecast  model  of   the  German  weather  service  (DWD),
-  developed for  the evaluation of the interactions  of reactive gases
-  and aerosol particles  with the state of atmosphere  at the regional
-  scale. It  includes secondary aerosols,  directly emitted components
-  like  soot,  mineral  dust,  sea  salt and  biological  material  as
-  pollen. Processes such  as emissions, coagulation, condensation, dry
-  deposition,  wet removal,  and sedimentation  of aerosols  are taken
-  into account.   The overall performance  of this application  on HPC
+  In  this paper we  present the  energy-to-solution and  profiling of
+  COSMO-ART on various  platforms.  This model is an  extension of the
+  operational  weather forecast  model of  the German  weather service
+  (DWD), developed for the  evaluation of the interactions of reactive
+  gases  and aerosol  particles with  the state  of atmosphere  at the
+  regional scale.  The overall  performance of this application on HPC
   systems is analysed  by a profiling study to  determine hotspots and
@@
   about power bottlenecks.  Our motivation is to improve corresponding
-  code    sections   to    sustain   performance    while   minimizing
+  code    sections   to    sustain   performance    while   minimising
   energy-to-solution.   This  preliminary  work  sets  the  basis  for

### 2014-04-22 paper_enahpc14/Introduction.tex
commits Will's slight corrections

@@
 (KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
-scale  model system  is online  coupled with  the  operational weather
-forecast  model  COSMO \citep{Baldauf-2011},  jointly  developed by  a
-consortium of European weather centers and used in the climate version
-(COSMO-CLM) by a wide research community.
+scale model  system is coupled  with the operational  weather forecast
+model COSMO \citep{Baldauf-2011}, jointly developed by a consortium of
+European weather  centers and used in the  climate version (COSMO-CLM)
+by a wide research community.
 
 The extended  COSMO-ART model  is computationally much  more demanding
-than the  COSMO core  since a large  number of additional  tracers and
-processes  have  to  be  considered.   Thus this  model  is  currently
-severely-limited in  terms of applicability and expensive  in terms of
-energy consumption.   Although COSMO has recently been  ported to GPUs
-within  the framework of  the High  Performance and  High Productivity
-Computing  (HP2C)  Initiative to  optimize  it  for computational  and
-energy efficiency (\url{http://www.hp2c.ch/}), significant investments
-in  ART  are still  required  to  take it  to  a  similar level.   The
-efficiency of COSMO-ART is being addressed in the EU Exa2Green project
+than COSMO  since a large  number of additional tracers  and processes
+have to  be considered.  Thus the model  is currently severely-limited
+in  terms   of  applicability  and   expensive  in  terms   of  energy
+consumption.    Although   COSMO   has   recently   been   ported   to
+GPUs  \citep{Gysi-2014, Lapillonne-2014} within  the framework  of the
+High Performance and High  Productivity Computing (HP2C) Initiative to
+optimise    it    for     computational    and    energy    efficiency
+(\url{http://www.hp2c.ch/}), significant investments  in ART are still
+required to take  it to a similar level.   The efficiency of COSMO-ART
+is     being    addressed    in     the    EU     Exa2Green    project
 (\url{http://exa2green.eu/}). The  ultimate goal of the  project is to
@@
 
-Some  related work  will  be discussed  in  Sec.~\ref{sec:1}. Then  in
-Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
-COSMO-ART and specify its  technical setup related to the investigated
-performance and energy evaluation methods.  Sec.~\ref{sec:3} describes
-different  power  measurement  equipments  used  for  conducting  this
-benchmarking  study and presents  energy-aware techniques  employed to
-determine  the   energy  footprint  of   the  considered  application.
-Sec.~\ref{sec:4} gathers benchmark  results from the Exa2Green project
-and  highlights areas  where improvements  will be  necessary  for the
-subsequent baseline refactoring.   Finally, we conclude with practical
-guidelines  to identifying  energy  saving opportunities  and give  an
-outlook to future research in Sec.~\ref{concl}.
+This  work is organized  as follows:  in Sec.~\ref{sec:1}  we describe
+related  work,  then in  Sec.~\ref{sec:2},  we  briefly introduce  the
+regional atmospheric  model COSMO-ART and specify  its technical setup
+related to the investigated performance and energy evaluation methods.
+Sec.~\ref{sec:5} describes the HPC  supercomputers used to compute the
+baseline.  Sec.~\ref{sec:3}   describes  different  power  measurement
+equipment  used for  conducting this  benchmarking study  and presents
+energy-aware techniques employed to  determine the energy footprint of
+the   considered  application.   Sec.~\ref{sec:4}   gathers  benchmark
+results  from  the  Exa2Green   project  and  highlights  areas  where
+improvements   will   be  necessary   for   the  subsequent   baseline
+refactoring.   Finally,  we  conclude  with  practical  guidelines  to
+identifying energy saving opportunities  and give an outlook to future
+research in Sec.~\ref{concl}.

### 2014-04-22 paper_enahpc14/Sec2:COSMO-ART.tex
commits Will's slight corrections

@@
 a baseline,  it was necessary  to find a run-configuration  capable of
-being  recreated in  lal subsequent  versions.  Here  we give  a brief
+being  recreated in  all subsequent  versions.  Here  we give  a brief
 overview of the model setup  for our performance and energy efficiency
@@
 
-The  considered  COSMO-ART model  system  is  configured  with a  semi
-Lagrangian horizontal advection scheme with tricubic interpolation and
-selective filling  diffusion option in combination  with the dynamical
-core  using Runge-Kutta  time  stepping \citep{COSMO-PartI-2011}.   It
-also  makes use  of  the  Kinetic PreProcessor  solver  (KPP) for  the
-resolution  of atmospheric  chemistry ordinary  differential equations
-\citep{Damian-2002}.   Concerning the modelling  of wet  deposition in
-aerosols, the  baseline has only indirect cloud  feedbacks but does not
-include  in-cloud  scavenging  (rainout)  and  below-cloud  scavenging
-(washout)  yet.   Amongst  physical  parameterisations,  precipitation
-formation is performed by a two-moment cloud microphysics.
+The   considered  COSMO-ART   model  system   is  configured   with  a
+semi-Lagrangian    horizontal   advection    scheme    with   tricubic
+interpolation  and selective filling  diffusion option  in combination
+with   the   dynamical    core   using   Runge-Kutta   time   stepping
+\citep{COSMO-PartI-2011}.    It  also   makes  use   of   the  Kinetic
+PreProcessor solver (KPP) for  the resolution of atmospheric chemistry
+ordinary  differential equations \citep{Damian-2002}.   Concerning the
+modelling  of  wet  deposition  in  aerosols, the  baseline  has  only
+indirect  cloud feedbacks  but  does not  include in-cloud  scavenging
+(rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
+parameterisations,   precipitation  formation   is   performed  by   a
+two-moment cloud microphysics.

### 2014-04-22 paper_enahpc14/Sec4:Evaluation.tex
cancels previous commits of Cristiano as IBM Zurich is actually not going to contribute to this work
commits Will's slight corrections

@@
 \label{subsec:4.2}
-While hardware  platforms mature  and are replaced,  we have  chosen a
-state-of-the-art  Intel's  third-generation   Core  (aka  Ivy  Bridge)
-processing platform (called ``Monch'')  for our power measurements, as
-it  is slated  to stay  in service  without hardware  upgrade  for the
-duration  of  the Exa2Green  project.   In  principle  at least,  this
-architecture could be recreated or found in an identical configuration
-beyond the lifetime of the project.  Given that the baseline benchmark
-can be reproduced  within an expected variance, and  that the baseline
-run configuration  can be used in  all future versions of  the code, a
-fair comparison  will be made  between the baseline and  the milestone
-versions   of    COSMO-ART.    In   this    study,   a   complementary
+We have  chosen a state-of-the-art Intel's  third-generation Core (aka
+Ivy  Bridge)  processing platform  (called  ``Monch'')  for our  power
+measurements,  which is  slated to  stay in  service  without hardware
+upgrade for  the duration of  the Exa2Green project.  In  principle at
+least, this architecture  could be recreated or found  in an identical
+configuration beyond the lifetime  of the project.  Since the baseline
+benchmark can be reproduced within  an expected variance, and that the
+baseline run configuration  can be used in all  future versions of the
+code,  a fair comparison  will be  made between  the baseline  and the
+milestone  versions  of COSMO-ART.   In  this  study, a  complementary
 energy-to-solution  benchmarking  comparison  is  carried out  on  the
@@
 (CSCS) of ETH Zurich, is  a 10 rack NEC-provided and dual-socket Intel
-Ivy Bridge-EP based cluster, utilised from people that are part of the
-Swiss    Platform   for    Advanced   Scientific    Computing   (PASC,
+Ivy Bridge-EP  based cluster, utilised  by scientists who are  part of
+the   Swiss  Platform   for  Advanced   Scientific   Computing  (PASC,
 \url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
@@
 speed.  Each core  has 32 KB instruction and 32 KB  data L1 caches and
-256 KB  of L2 cache. All  the 10 core share  a 25 MB L3  cache and the
+256  KB of L2  cache. All  10 cores  share a  25 MB  L3 cache  and the
 platform has  32GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
@@
 gcc/4.8.2 on Pilatus) with the -O3 compiler flag as it generally gives
-a  good  level  of optimization  and  the  code  runs faster  in  this
+a  good  level  of optimisation  and  the  code  runs faster  in  this
 configuration  than when  compiled with  the intel  compiler (14.0.1).
@@
 processors using  a 2D-partitioning strategy.   The distribution along
-the  x and  y  coordinates  was defined  by  setting: $nprocx=40$  and
+the $x$  and $y$ coordinates  was defined by setting:  $nprocx=40$ and
 $nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
-usually  kept  bigger  than  $nprocy$.   Besides as  this  version  of
+usually  kept  bigger than  $nprocy$.   Besides,  as  this version  of
 COSMO-ART  doesn't  make  use   of  the  GRIB  library,  we  specified
@@
 expected,  Xeon  E5 outperforms  Ivy  Bridge  EP,  being roughly  1.3x
-faster.  The reason for that is twofold: it has higher clock frequency
-than Xeon E5 (2.6 GHz against  2.2 GHz) and it aims at computing speed
-regardless to  energy consumption. In  our experiments, Ivy  Bridge EP
-showed the best energy-to-solution, reducing the energy consumption of
-Xeon E5 by approximately $7\%$.
+faster.   The  reason  for  that  is two-fold:  it  has  higher  clock
+frequency than  Ivy Bridge (2.6  GHz against 2.2  GHz) and it  aims at
+computing speed regardless to  energy consumption. In our experiments,
+Ivy Bridge EP showed  the best energy-to-solution, reducing the energy
+consumption of Xeon E5 by approximately $7\%$.
 

### 2014-04-23 paper_enahpc14/Sec1:Related.tex
adds minor changes; updates paper.bib; includes references for related work

\begin{itemize}
\item Application centric energy-efficiency study of distributed
  multi-core and hybrid {CPU-GPU} systems \citep{Cumming-2014}
\item Stella: A domain-specific language and tool for structured grid
  methods \citep{Gysi-2014}
\item Using compiler directives to port large scientific applications
  to GPUs: An example from atmospherci science \citep{Lapillonne-2014}
\item An analysis of energy-optimized lattice-Boltzmann CFD
  simulations from the chip to the highly parallel level
  \citep{Wittmann-2013}
\item Energy efficiency vs. performance of the numerical solution of
  PDEs: An application study on a low-power ARM-based cluster
  \citep{Goddeke-2013}
\item Analysis of Computing and Energy Performance of Multicore, NUMA,
  and Manycore Platforms for an Irregular Application
  \citep{Castro-2013}
\item Exploring performance and power properties of modern multi-core
  chips via simple machine models \citep{Hager-2013}
\end{itemize}

### 2014-04-23 paper_enahpc14/Sec3:Measurement-framework.tex
adds minor changes; updates paper.bib; includes references for related work

--- 2014-04-16 paper_enahpc14/Sec3:Measurement-framework.tex
@@
 
-\subsubsection{E3METER metering products}
+\subsubsection{CSCS - E3METER metering products}
 Supercomputer  clusters considered  for our  experiments at  the Swiss
-National Supercomputing Center of  ETH Zurich (CSCS) are equipped with
+National Supercomputing Center (CSCS)  of ETH Zurich are equipped with
 E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
@@
 
-\subsubsection{University of Hamburg}
+\subsubsection{UHAM - UJI}
 To assess the  performance and the energy efficiency  of COSMO-ART, we
 employ   a    version   of   the    integrated   framework   presented
-in~\cite{energy13}  that  works in  combination  with Extrae  and
-Paraver,   which  are   profiling/tracing   and  visualization   tools,
-respectively.
+in~\cite{energy13} that works in  combination with Extrae and Paraver,
+which are profiling/tracing and visualization tools, respectively.
+
 %The left part of the \vref{fig:Lustre} offers a graphical representation of
 %the Lustre architecture; the right depicts the tracing and profiling framework.
-To  use our  approach,  COSMO-ART is  compiled  using the  Extrae
-compiler wrappers, which automatically  instrument the Fortran code of
-the  model. Next,  COSMO-ART is  run  on the  nodes, thus  dissipating
-certain  amount of  power.  These  nodes are connected  to power 
-measurement  devices that  account for  the dissipated power/consumed 
-energy and  send the power data  to the tracing  server. The client, 
-meanwhile, sends start/stop primitives in order to gather captured data
-by the wattmeters onto the tracing server, where an  instance of the 
-\pmlib server  is running.
+To use our  approach, COSMO-ART is compiled using  the Extrae compiler
+wrappers,  which  automatically instrument  the  Fortran  code of  the
+model. Next, COSMO-ART  is run on the nodes,  thus dissipating certain
+amount  of power.   These  nodes are  connected  to power  measurement
+devices that account for the dissipated power/consumed energy and send
+the power  data to the  tracing server.  The client,  meanwhile, sends
+start/stop  primitives  in  order  to  gather  captured  data  by  the
+watt-meters onto the  tracing server, where an instance  of the \pmlib
+server is running.
 
 Once COSMO-ART run is finished, a file containing the power profile is
-created using the  power   data  received from  the  tracing   server and
-the  instrumentation post-processing generates  the performance trace files.
-All these files are combined then in Paraver which allows us to visualize 
-the performance trace and the power profile of COSMO-ART all together.
+created using the power data  received from the tracing server and the
+instrumentation post-processing generates the performance trace files.
+All  these files  are  combined then  in  Paraver which  allows us  to
+visualise the performance trace and the power profile of COSMO-ART all
+together.
 

### 2014-04-23 trunk/paper_enahpc14/Introduction.tex
adds minor changes; updates paper.bib; includes references for related work

@@
 (KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
-scale  model system  is online  coupled with  the  operational weather
-forecast  model  COSMO \citep{Baldauf-2011},  jointly  developed by  a
-consortium of European weather centers and used in the climate version
-(COSMO-CLM) by a wide research community.
+scale model  system is coupled  with the operational  weather forecast
+model COSMO \citep{Baldauf-2011}, jointly developed by a consortium of
+European weather  centers and used in the  climate version (COSMO-CLM)
+by a wide research community.
 
-The extended  COSMO-ART model  is computationally much  more demanding
-than the  COSMO core  since a large  number of additional  tracers and
-processes  have  to  be  considered.   Thus this  model  is  currently
+The extended atmospheric model  COSMO-ART is computationally much more
+demanding than  COSMO since a  large number of additional  tracers and
+processes  have  to  be  considered.   Thus  the  model  is  currently
 severely-limited in  terms of applicability and expensive  in terms of
-energy consumption.   Although COSMO has recently been  ported to GPUs
-within  the framework of  the High  Performance and  High Productivity
-Computing  (HP2C)  Initiative to  optimize  it  for computational  and
-energy efficiency (\url{http://www.hp2c.ch/}), significant investments
-in  ART  are still  required  to  take it  to  a  similar level.   The
-efficiency of COSMO-ART is being addressed in the EU Exa2Green project
+energy  consumption.   Although  COSMO  has recently  been  ported  to
+GPUs  \citep{Gysi-2014, Lapillonne-2014} within  the framework  of the
+High Performance and High  Productivity Computing (HP2C) Initiative to
+optimise    it    for     computational    and    energy    efficiency
+(\url{http://www.hp2c.ch/}), significant investments  in ART are still
+required to take  it to a similar level.   The efficiency of COSMO-ART
+is     being    addressed    in     the    EU     Exa2Green    project
 (\url{http://exa2green.eu/}). The  ultimate goal of the  project is to
@@
 
-This work is organized as follows: in Sec.~\ref{sec:1} we describe related work, then in
-Sec.~\ref{sec:2}, we briefly  introduce the regional atmospheric model
-COSMO-ART and specify its  technical setup related to the investigated
-performance and energy evaluation methods.  Sec.~\ref{sec:5} describes
-the HPC supercomputers used to compute the baseline. Sec.~\ref{sec:3} describes
-different  power  measurement  equipment  used  for  conducting  this
-benchmarking  study and presents  energy-aware techniques  employed to
-determine  the   energy  footprint  of   the  considered  application.
-Sec.~\ref{sec:4} gathers benchmark  results from the Exa2Green project
-and  highlights areas  where improvements  will be  necessary  for the
+This  work is  organized as  follows: in  Sec.~\ref{sec:1} we  give an
+overview  of  related  work,   then  in  Sec.~\ref{sec:2}  we  briefly
+introduce  COSMO-ART and specify  its technical  setup related  to the
+investigated  performance  and  energy  evaluation  methods.   In  the
+subsequent Sec.~\ref{sec:3},  we describe different  power measurement
+equipments as well as energy-aware techniques employed to conduct this
+benchmarking  study.  Sec.~\ref{sec:4}  presents  the performance  and
+energy  requirements  of the  baseline  on  various architectures  and
+highlights  areas  where  improvements   will  be  necessary  for  the
 subsequent baseline refactoring.   Finally, we conclude with practical

### 2014-04-23 trunk/paper_enahpc14/Sec4:Evaluation.tex
adds minor changes; updates paper.bib; includes references for related work

--- 2014-04-12 trunk/paper_enahpc14/Sec4:Evaluation.tex
@@
 performance    of    COSMO-ART    with    two    important    metrics:
-\textit{time-to-solution}       and       \textit{energy-to-solution}.
-Time-to-Solution  refers   to  the  total  wall  clock   time  of  the
-application execution time. Energy-to-solution is the amount of energy
-spent  to achieve  results.  Energy  consumption of  CSCS  clusters is
-assessed  by sampling  the instantaneous  peak power  during execution
-which  is then  averaged  and multiplied  by  the time-to-solution  to
-determine energy-to-solution.   Whenever possible, multiple production
-runs of COSMO-ART were  performed to illustrate the reproducibility of
-the baseline, and quantify  the significant uncertainties in the power
+\textit{time-to-solution} (TTS) and \textit{energy-to-solution} (ETS).
+TTS refers to  the total wall clock time  of the application execution
+time. ETS  is the amount of  energy spent to  achieve results.  Energy
+consumption  is  assessed by  sampling  the  instantaneous peak  power
+during execution which  is then averaged and multiplied  by the TTS to
+determine  ETS.   Whenever   possible,  multiple  production  runs  of
+COSMO-ART  were performed  to  illustrate the  reproducibility of  the
+baseline,  and quantify  the  significant uncertainties  in the  power
 measurement, as dictated by the available technology.
@@
 \label{subsec:4.2}
-While hardware  platforms mature  and are replaced,  we have  chosen a
-state-of-the-art  Intel's  third-generation   Core  (aka  Ivy  Bridge)
-processing platform (called ``Monch'')  for our power measurements, as
-it  is slated  to stay  in service  without hardware  upgrade  for the
-duration  of  the Exa2Green  project.   In  principle  at least,  this
-architecture could be recreated or found in an identical configuration
-beyond the lifetime of the project.  Given that the baseline benchmark
-can be reproduced  within an expected variance, and  that the baseline
-run configuration  can be used in  all future versions of  the code, a
-fair comparison  will be made  between the baseline and  the milestone
-versions   of    COSMO-ART.    In   this    study,   a   complementary
-energy-to-solution  benchmarking  comparison  is  carried out  on  the
-``Pilatus'' cluster based  on Intel's previous-generation Sandy Bridge
-processors, conventional  in HPC  systems and known  to be  more power
-consuming.
+We have  chosen a state-of-the-art Intel's  third-generation Core (aka
+Ivy  Bridge) processing platform  at CSCS  (called ``Monch'')  for our
+power  measurements,  which  is  slated  to stay  in  service  without
+hardware  upgrade  for the  duration  of  the  Exa2Green project.   In
+principle at least,  this architecture could be recreated  or found in
+an identical configuration beyond  the lifetime of the project.  Since
+the baseline benchmark can  be reproduced within an expected variance,
+and  that the baseline  run configuration  can be  used in  all future
+versions  of the  code, a  fair comparison  will be  made  between the
+baseline and  the milestone versions  of COSMO-ART.  In this  study, a
+complementary  energy-to-solution benchmarking  comparison  is carried
+out   on  the   ``Pilatus''  cluster   at  CSCS,   based   on  Intel's
+previous-generation  Sandy  Bridge  processors,  conventional  in  HPC
+systems and known to be  more power consuming.  Finally a full tracing
+experiment is  conducted on the  ``Tintorrum'' cluster at the  Jaume I
+University (UJI) using an  integrated framework, to capture an overall
+power profile at a much finer resolution and to get more insights into
+the MPI blocking and polling influences on power savings.
 
-\subsubsection{Monch (CSCS - ETH Zurich)}
-Monch, which is installed  at the Swiss National Supercomputing Center
-(CSCS) of ETH Zurich, is  a 10 rack NEC-provided and dual-socket Intel
-Ivy Bridge-EP based cluster, utilised from people that are part of the
-Swiss    Platform   for    Advanced   Scientific    Computing   (PASC,
+\subsubsection{CSCS - Monch}
+Monch is  a 10 rack  NEC-provided and dual-socket Intel  Ivy Bridge-EP
+based  cluster, utilised  by  scientists  who are  part  of the  Swiss
+Platform     for      Advanced     Scientific     Computing     (PASC,
 \url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
 nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
-Each standard  compute node comprises two Intel  Ivy Bridge-EP E5-2660
-v2 ten-core processors operating at 2.2 GHz, themselves connected by a
-high speed  InfiniBand network, based  on Mellanox SX6036  managed FDR
-switches, with a  56 Gb/s speed.  Each core has  32 KB instruction and
-32 KB data L1  caches and 256 KB of L2 cache. All  the 10 core share a
-25 MB L3 cache and the platform has 32GB of DDR3 1600 MHz RAM. For our
-energy-to-solution benchmark,  a full rack of Monch  constituted of 52
-standard  compute   nodes  (monchc[029-080])  was   considered.
+Each standard  compute node comprises  two Intel Ivy  Bridge Efficient
+Performance (EP)  E5-2660 v2 ten-core processors operating  at 2.2 GHz
+base  clock speed,  themselves connected  by a  high  speed InfiniBand
+network based on Mellanox SX6036  managed FDR switches, with a 56 Gb/s
+speed.  Each core  has 32 KB instruction and 32 KB  data L1 caches and
+256  KB of L2  cache. All  10 cores  share a  25 MB  L3 cache  and the
+platform has  32GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
+benchmark, a  full rack  of Monch constituted  of 52  standard compute
+nodes (monchc[029-080]) was considered.
 
-\subsubsection{Pilatus (CSCS - ETH Zurich)}
-Pilatus,  which  is installed  at  the  Swiss National  Supercomputing
-Center (CSCS)  of ETH Zurich,  is a dual-socket Intel  Sandy Bridge-EP
-based cluster  used as Piz  Daint pre-post processing cluster.   It is
-composed of 42 compute nodes  and has 2 high-speed interconnects based
-on FDR:  the first is dedicated to  the MPI traffic and  the second to
-the storage high speed traffic.  The 2 login nodes and the 42 computes
+\subsubsection{CSCS - Pilatus}
+Pilatus  is  a dual-socket  eight-core  Intel  Sandy  Bridge EP  based
+cluster used as Piz Daint pre-post processing cluster.  It is composed
+of 42 compute  nodes and has 2 high-speed  interconnects based on FDR:
+the  first is  dedicated to  the  MPI traffic  and the  second to  the
+storage high  speed traffic.   The 2 login  nodes and the  42 computes
 nodes  consists in  11  twin-pair Intel  E5-Series  DALCO r2264i4t  2U
 scalable compute modules.  Each  module contains 4 compute nodes based
-on two Intel Xeon E5-2670  eight-core processors operating at 2.6 GHz,
-themselves  connected by  a high  speed InfiniBand  network,  based on
-Mellanox SX6036 managed FDR switches,  with a 56 Gb/s speed. Each core
-has 32 KB instruction and 32 KB data L1 caches and 256 KB of L2 cache.
-All the  8 core share a  20 MB L3 cache  and the platform  has 64GB of
-DDR3 1600 MHz  RAM. For our energy-to-solution benchmark,  a full rack
-of Pilatus  constituted of 42 standard  compute nodes (pilatus[03-44])
-was considered.
+on two Intel  Xeon E5-2670 processors operating at  2.6 GHz base clock
+speed, themselves  connected by a high speed  InfiniBand network based
+on Mellanox  SX6036 managed FDR switches,  with a 56  Gb/s speed. Each
+core has 32 KB  instruction and 32 KB data L1 caches  and 256 KB of L2
+cache.  All  the 8 core share  a 20 MB  L3 cache and the  platform has
+64GB of  DDR3 1600 MHz  RAM.  For our energy-to-solution  benchmark, a
+full  rack  of  Pilatus  constituted  of  42  standard  compute  nodes
+(pilatus[03-44]) was considered.
+
+\subsubsection{UJI - Tintorrum}
+Tintorrum  is a heterogeneous  cluster composed  of 28  compute nodes.
+For our experiments only a subset of these nodes were considered. This
+set is  composed of 16  nodes, each of  which includes two  Intel Xeon
+E5645 hexa-core  processors at 2.40  GHz connected via  Infiniband QDR
+(Mellanox MTS3600 switch).  Each core  has 32 KB instruction and 32 KB
+data L1 caches and 256 KB L2 cache. The 6 cores share a 12 MB L3 cache
+and the platform has 288 GB of DDR3 1333 MHz.
 
@@
 Interface (MPI)  and are purely  MPI-parallel.  The software  stack on
-both many-core  platforms was  controlled using the  modules framework
-which gives  an easy and  flexible mechanism to  access to all  of the
-CSCS  provided compilers,  tools  and applications.   For our  initial
+both  many-core  CSCS  platforms  was  controlled  using  the  modules
+framework which gives an easy  and flexible mechanism to access to all
+of the  provided compilers, tools  and applications.  For  our initial
 benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,
 gcc/4.8.2 on Pilatus) with the -O3 compiler flag as it generally gives
-a  good  level  of optimization  and  the  code  runs faster  in  this
+a  good  level  of optimisation  and  the  code  runs faster  in  this
 configuration  than when  compiled with  the intel  compiler (14.0.1).

### 2014-04-23 trunk/paper_enahpc14/Introduction.tex
shortens paper length with the deletion of some references

@@
 concentrations impact the climate  system by scattering and absorption
-of   solar   radiation,    thereby   exerting   a   direct   radiative
-forcing  \citep{Liao-2005,  Bangert-2012,  Lundgren-2013}; or  by  the
-modification of  cloud properties, cloud fraction  and surface albedo,
-causing  a negative  indirect  radiative forcing  \citep{Haywood-2000,
-Lohmann-2005,     VandenHeever-2011,     Rosenfeld-2013}.      Despite
+of solar radiation, thereby exerting a direct radiative forcing; or by
+the  modification  of cloud  properties,  cloud  fraction and  surface
+albedo,  causing  a  negative  indirect  radiative  forcing.   Despite
 considerable  progress in  global aerosol  modelling \citep{Mann-2013}

### 2014-04-23 trunk/paper_enahpc14/Sec4:Evaluation.tex
shortens paper length with the deletion of some references

@@
 \subsubsection{Sofware environment}
-INT2LM  and the  COSMO-ART model  are  implemented in  Fortran 90  for
-distributed  memory  parallel  computers  using  the  Message  Passing
-Interface (MPI)  and are purely  MPI-parallel.  The software  stack on
-both  many-core  CSCS  platforms  was  controlled  using  the  modules
-framework which gives an easy  and flexible mechanism to access to all
-of the  provided compilers, tools  and applications.  For  our initial
-benchmarking,  we opted  for  the GNU  compiler  (gcc/4.8.1 on  Monch,
-gcc/4.8.2 on Pilatus) with the -O3 compiler flag as it generally gives
-a  good  level  of optimisation  and  the  code  runs faster  in  this
-configuration  than when  compiled with  the intel  compiler (14.0.1).
-Besides, we installed the  MPICH2 implementation of MPI (mvapich2/1.9)
-as  well  as  the  commonly  used HDF5  (1.8.12)  and  NetCDF  (4.3.1)
-libraries in favor of the  traditional GRIB library for the management
-of extremely  large and complex data collections.   All computes nodes
-have    an   operating   system    based   on    GNU/Linux   featuring
+The COSMO-ART baseline  is a pure MPI based  Fortran~90 code currently
+running on distributed multi-core systems only.  The software stack on
+both CSCS  platforms was controlled using the  modules framework which
+gives an easy and flexible mechanism  to access to all of the provided
+compilers, tools  and applications.  For our  initial benchmarking, we
+opted for the GNU compiler  (gcc/4.8.1 on Monch, gcc/4.8.2 on Pilatus)
+using the -O3  compiler flag in favor of  the intel compiler (14.0.1),
+delivering inferior performance.  In addition, we installed the MPICH2
+implementation of MPI (mvapich2/1.9) as well as the commonly used HDF5
+(1.8.12) and NetCDF (4.3.1)  libraries for the management of extremely
+large  and  complex data  collections.   All  computes  nodes have  an
+operating      system      based      on      GNU/Linux      featuring
 ``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
@@
 usually  kept  bigger than  $nprocy$.   Besides,  as  this version  of
-COSMO-ART  doesn't  make  use   of  the  GRIB  library,  we  specified
-$nprocio=0$ for  GRIB I/O.  Hyperthreading  is not considered  in this
-study as previous  attempts of its use revealed that  it always led to
-higher energy-to-solution.\\
+COSMO-ART  doesn't  make  use  of  the traditional  GRIB  library,  we
+specified $nprocio=0$ for GRIB  I/O.  Hyperthreading is not considered
+in this study as previous attempts  of its use revealed that it always
+led to higher energy-to-solution.\\
 
@@
 expected,  Xeon  E5 outperforms  Ivy  Bridge  EP,  being roughly  1.3x
-faster.   The  reason  for  that  is two-fold:  it  has  higher  clock
-frequency than  Ivy Bridge (2.6  GHz against 2.2  GHz) and it  aims at
-computing speed regardless to  energy consumption. In our experiments,
-Ivy Bridge EP showed  the best energy-to-solution, reducing the energy
-consumption of Xeon E5 by approximately $7\%$.
+faster.   The reason for  that is  two-fold: (1)  it has  higher clock
+frequency than Ivy  Bridge (2.6 GHz against 2.2 GHz),  and (2) it aims
+at  computing   speed  regardless   to  energy  consumption.   In  our
+experiments,  Ivy  Bridge   EP  showed  the  best  energy-to-solution,
+reducing the energy consumption of Xeon E5 by approximately $7\%$.
 

### 2014-04-23 trunk/paper_enahpc14/paper.tex
shortens the paper's title
adds minor changes; updates paper.bib; includes references for related work
updates abstract

@@
 
-\title{Evaluating the Performance and Energy Efficiency of COSMO-ART,
-  a Coupled Numerical Weather Forecast and Chemical Transport Model.}
+\title{Evaluating the Performance and Energy Efficiency of the
+  COSMO-ART model system}
 %\thanks{}
@@
 \begin{abstract}
-  In  this paper we  present the  energy-to-solution and  profiling of
-  COSMO-ART on various  platforms.  This model is an  extension of the
-  operational  weather forecast  model of  the German  weather service
-  (DWD), developed for the  evaluation of the interactions of reactive
-  gases  and aerosol  particles with  the state  of atmosphere  at the
-  regional scale.  The overall  performance of this application on HPC
-  systems is analysed  by a profiling study to  determine hotspots and
-  identify  critical  CPU paths.   Moreover,  we describe  measurement
-  devices and energy-aware techniques  employed to evaluate the energy
-  footprint of the considered application and to get detailed insights
-  about power bottlenecks.  Our motivation is to improve corresponding
-  code    sections   to    sustain   performance    while   minimising
-  energy-to-solution.   This  preliminary  work  sets  the  basis  for
-  subsequent studies to tackle  challenges related to energy efficient
-  high performance computing in the framework of the Exa2Green project
-  (\url{http://exa2green.eu/}).
+  In this  paper we investigate  the energy footprint  and performance
+  profiling of COSMO-ART  on various HPC platforms.  This  model is an
+  extension of  the operational weather  forecast model of  the German
+  weather  service   (DWD),  developed  for  the   evaluation  of  the
+  interactions of reactive gases  and aerosol particles with the state
+  of atmosphere at the  regional scale.  Different measurement devices
+  and energy-aware techniques are  described to evaluate both time and
+  energy  to  solution  of  the  considered application  and  to  gain
+  detailed  insights  into power  and  performance requirements.   Our
+  motivation  is to  improve  corresponding code  sections to  sustain
+  performance  while minimising energy-to-solution.   This preliminary
+  work  sets the  basis for  subsequent studies  to  tackle challenges
+  related  to  energy  efficient  high performance  computing  in  the
+  framework of the Exa2Green project (\url{http://exa2green.eu/}).
 
@@
 
-\section{Related work}
+\section{Related work \emph{(to be completed)}}
 \label{sec:1}
@@
 
-\section{Conclusion}
+\section{Conclusion \emph{(to be completed)}}
 \label{concl}

### 2014-04-23 paper_enahpc14/Sec4:Evaluation.tex
creates subsubsubsection; adds new figures

@@
 performance    of    COSMO-ART    with    two    important    metrics:
-\textit{time-to-solution}       and       \textit{energy-to-solution}.
-Time-to-solution  refers   to  the  total  wall  clock   time  of  the
-application execution time. Energy-to-solution is the amount of energy
-spent  to achieve  results.  Energy  consumption of  CSCS  clusters is
-assessed  by sampling  the instantaneous  peak power  during execution
-which  is then  averaged  and multiplied  by  the time-to-solution  to
-determine energy-to-solution.   Whenever possible, multiple production
-runs of COSMO-ART were  performed to illustrate the reproducibility of
-the baseline, and quantify  the significant uncertainties in the power
+\textit{time-to-solution} (TTS) and \textit{energy-to-solution} (ETS).
+TTS refers to  the total wall clock time  of the application execution
+time. ETS  is the amount of  energy spent to  achieve results.  Energy
+consumption  is  assessed by  sampling  the  instantaneous peak  power
+during execution which  is then averaged and multiplied  by the TTS to
+determine  ETS.   Whenever   possible,  multiple  production  runs  of
+COSMO-ART  were performed  to  illustrate the  reproducibility of  the
+baseline,  and quantify  the  significant uncertainties  in the  power
 measurement, as dictated by the available technology.
@@
 \label{subsec:4.2}
+
+\subsubsection{HPC platforms}
 We have  chosen a state-of-the-art Intel's  third-generation Core (aka
-Ivy  Bridge)  processing platform  (called  ``Monch'')  for our  power
-measurements,  which is  slated to  stay in  service  without hardware
-upgrade for  the duration of  the Exa2Green project.  In  principle at
-least, this architecture  could be recreated or found  in an identical
-configuration beyond the lifetime  of the project.  Since the baseline
-benchmark can be reproduced within  an expected variance, and that the
-baseline run configuration  can be used in all  future versions of the
-code,  a fair comparison  will be  made between  the baseline  and the
-milestone  versions  of COSMO-ART.   In  this  study, a  complementary
-energy-to-solution  benchmarking  comparison  is  carried out  on  the
-``Pilatus'' cluster based  on Intel's previous-generation Sandy Bridge
-processors, conventional  in HPC  systems and known  to be  more power
-consuming.   Finally a  full tracing  experiment is  conducted  on the
-``Tintorrum''  cluster using  an integrated  framework, to  capture an
-overall  power profile  at a  much finer  resolution and  to  get more
-insights  into  the  MPI  blocking  and polling  influences  on  power
-savings.
+Ivy  Bridge) processing platform  at CSCS  (called ``Monch'')  for our
+power  measurements,  which  is  slated  to stay  in  service  without
+hardware  upgrade  for the  duration  of  the  Exa2Green project.   In
+principle at least,  this architecture could be recreated  or found in
+an identical configuration beyond  the lifetime of the project.  Since
+the baseline benchmark can  be reproduced within an expected variance,
+and  that the baseline  run configuration  can be  used in  all future
+versions  of the  code, a  fair comparison  will be  made  between the
+baseline and  the milestone versions  of COSMO-ART.  In this  study, a
+complementary  energy-to-solution benchmarking  comparison  is carried
+out   on  the   ``Pilatus''  cluster   at  CSCS,   based   on  Intel's
+previous-generation  Sandy  Bridge  processors,  conventional  in  HPC
+systems and known to be  more power consuming.  Finally a full tracing
+experiment is  conducted on the  ``Tintorrum'' cluster at the  Jaume I
+University (UJI) using an  integrated framework, to capture an overall
+power profile at a much finer resolution and to get more insights into
+the MPI blocking and polling influences on power savings.
 
-\subsubsection{Monch (CSCS - ETH Zurich)}
-Monch, which is installed  at the Swiss National Supercomputing Center
-(CSCS) of ETH Zurich, is  a 10 rack NEC-provided and dual-socket Intel
-Ivy Bridge-EP  based cluster, utilised  by scientists who are  part of
-the   Swiss  Platform   for  Advanced   Scientific   Computing  (PASC,
+\paragraph{CSCS - Monch} 
+~\\
+\vskip -1em
+Monch is  a 10 rack  NEC-provided and dual-socket Intel  Ivy Bridge-EP
+based  cluster, utilised  by  scientists  who are  part  of the  Swiss
+Platform     for      Advanced     Scientific     Computing     (PASC,
 \url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
@@
 
-\subsubsection{Pilatus (CSCS - ETH Zurich)}
-Pilatus,  which  is installed  at  the  Swiss National  Supercomputing
-Center (CSCS) of  ETH Zurich, is a dual-socket  eight-core Intel Sandy
-Bridge EP based cluster used as Piz Daint pre-post processing cluster.
-It is composed of 42  compute nodes and has 2 high-speed interconnects
-based on FDR: the first is dedicated to the MPI traffic and the second
-to  the storage  high speed  traffic.  The  2 login  nodes and  the 42
-computes nodes consists in 11 twin-pair Intel E5-Series DALCO r2264i4t
-2U  scalable compute modules.   Each module  contains 4  compute nodes
-based on two  Intel Xeon E5-2670 processors operating  at 2.6 GHz base
-clock speed,  themselves connected by a high  speed InfiniBand network
-based  on  Mellanox  SX6036  managed  FDR switches,  with  a  56  Gb/s
-speed. Each  core has 32 KB instruction  and 32 KB data  L1 caches and
-256 KB  of L2 cache.  All  the 8 core share  a 20 MB L3  cache and the
-platform has  64GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
-benchmark, a full  rack of Pilatus constituted of  42 standard compute
-nodes (pilatus[03-44]) was considered.
+\paragraph{CSCS - Pilatus} 
+~\\
+\vskip -1em
+Pilatus  is  a dual-socket  eight-core  Intel  Sandy  Bridge EP  based
+cluster used as Piz Daint pre-post processing cluster.  It is composed
+of 42 compute  nodes and has 2 high-speed  interconnects based on FDR:
+the  first is  dedicated to  the  MPI traffic  and the  second to  the
+storage high  speed traffic.   The 2 login  nodes and the  42 computes
+nodes  consists in  11  twin-pair Intel  E5-Series  DALCO r2264i4t  2U
+scalable compute modules.  Each  module contains 4 compute nodes based
+on two Intel  Xeon E5-2670 processors operating at  2.6 GHz base clock
+speed, themselves  connected by a high speed  InfiniBand network based
+on Mellanox  SX6036 managed FDR switches,  with a 56  Gb/s speed. Each
+core has 32 KB  instruction and 32 KB data L1 caches  and 256 KB of L2
+cache.  All  the 8 core share  a 20 MB  L3 cache and the  platform has
+64GB of  DDR3 1600 MHz  RAM.  For our energy-to-solution  benchmark, a
+full  rack  of  Pilatus  constituted  of  42  standard  compute  nodes
+(pilatus[03-44]) was considered.
 
-\subsubsection{Tintorrum (UJI)}
-Tintorrum, which  is installed at the  Jaume I University  (UJI), is a
-heterogeneous  cluster   composed  of  28  compute   nodes.   For  our
-experiments only a subset of  these nodes were considered. This set is
-composed  of 16 nodes,  each of  which includes  two Intel  Xeon E5645
-hexa-core  processors  at  2.40   GHz  connected  via  Infiniband  QDR
+\paragraph{UJI - Tintorrum} 
+~\\
+\vskip -1em
+Tintorrum  is a heterogeneous  cluster composed  of 28  compute nodes.
+For our experiments only a subset of these nodes were considered. This
+set is  composed of 16  nodes, each of  which includes two  Intel Xeon
+E5645 hexa-core  processors at 2.40  GHz connected via  Infiniband QDR
 (Mellanox MTS3600 switch).  Each core  has 32 KB instruction and 32 KB
@@
 expected,  Xeon  E5 outperforms  Ivy  Bridge  EP,  being roughly  1.3x
-faster.   The  reason  for  that  is two-fold:  it  has  higher  clock
-frequency than  Ivy Bridge (2.6  GHz against 2.2  GHz) and it  aims at
-computing speed regardless to  energy consumption. In our experiments,
-Ivy Bridge EP showed  the best energy-to-solution, reducing the energy
-consumption of Xeon E5 by approximately $7\%$.
+faster.   The reason for  that is  two-fold: (1)  it has  higher clock
+frequency than Ivy  Bridge (2.6 GHz against 2.2 GHz),  and (2) it aims
+at  computing   speed  regardless   to  energy  consumption.   In  our
+experiments,  Ivy  Bridge   EP  showed  the  best  energy-to-solution,
+reducing the energy consumption of Xeon E5 by approximately $7\%$.
 
 \begin{figure}[htbf]
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART.eps}
+  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-0.eps}
   \caption{Time-to-solution and energy-to-solution comparison between
@@
 \end{figure}
+
+\begin{figure}[htbf]
+  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
+  \caption{Time-to-solution and energy-to-solution on Tintorrum -
+    Polling mode}
+  \label{fig:3}
+\end{figure}
+
+\begin{figure}[htbf]
+  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-2.eps}
+  \caption{Time-to-solution and energy-to-solution on Tintorrum -
+    Blocking mode}
+  \label{fig:3}
+\end{figure}

### 2014-04-23 paper_enahpc14/paper.tex
creates subsubsubsection; adds new figures

@@
 \DeclareRobustCommand\IPCClongname{}
+\setcounter{secnumdepth}{5}
 %
@@
 \begin{abstract}
-  In  this paper  we present  the energy-to-solution  of  COSMO-ART on
-  various platforms.   This model is  an extension of  the operational
-  weather  forecast  model  of   the  German  weather  service  (DWD),
-  developed for  the evaluation of the interactions  of reactive gases
-  and aerosol particles  with the state of atmosphere  at the regional
-  scale. It  includes secondary aerosols,  directly emitted components
-  like  soot,  mineral  dust,  sea  salt and  biological  material  as
-  pollen. Processes such  as emissions, coagulation, condensation, dry
-  deposition,  wet removal,  and sedimentation  of aerosols  are taken
-  into account.   The overall performance  of this application  on HPC
-  systems is analysed  by a profiling study to  determine hotspots and
-  identify  critical  CPU paths.   Moreover,  we describe  measurement
-  devices and energy-aware techniques  employed to evaluate the energy
-  footprint of the considered application and to get detailed insights
-  about power bottlenecks.  Our motivation is to improve corresponding
-  code    sections   to    sustain   performance    while   minimizing
-  energy-to-solution.   This  preliminary  work  sets  the  basis  for
-  subsequent studies to tackle  challenges related to energy efficient
-  high performance computing in the framework of the Exa2Green project
-  (\url{http://exa2green.eu/}).
+  In this  paper we investigate  the energy footprint  and performance
+  profiling of COSMO-ART  on various HPC platforms.  This  model is an
+  extension of  the operational weather  forecast model of  the German
+  weather  service   (DWD),  developed  for  the   evaluation  of  the
+  interactions of reactive gases  and aerosol particles with the state
+  of atmosphere at the  regional scale.  Different measurement devices
+  and energy-aware techniques are  described to evaluate both time and
+  energy  to  solution  of  the  considered application  and  to  gain
+  detailed  insights  into power  and  performance requirements.   Our
+  motivation  is to  improve  corresponding code  sections to  sustain
+  performance  while minimising energy-to-solution.   This preliminary
+  work  sets the  basis for  subsequent studies  to  tackle challenges
+  related  to  energy  efficient  high performance  computing  in  the
+  framework of the Exa2Green project (\url{http://exa2green.eu/}).
 

### 2014-04-24 paper_enahpc14/Sec4:Evaluation.tex
Manuel's changes

@@
 
-\begin{figure}[htbf]
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-2.eps}
-  \caption{Time-to-solution and energy-to-solution on Tintorrum -
-    Blocking mode}
+\begin{figure*}[htbf]
+  \centering
+  \includegraphics[width=0.78\textwidth]{Figs/23_24_blq1_stat1.eps}
+  \caption{blabla}
   \label{fig:3}
-\end{figure}
+\end{figure*}
+
+\begin{figure*}[htbf]
+  \centering
+  \scalebox{0.5}{\input{Figs/24hour.pstex_t}}
+  \caption{24 hours simulation trace using the MPI blocking mode on Tintorrum.}
+  \label{fig:3}
+\end{figure*}
+
+\begin{figure*}[htbf]
+  \centering
+  \scriptsize
+  MPI polling mode\\
+  \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}}\\
+  MPI blocking mode\\
+  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}
+  \caption{1 hour simulation (11h-12h) trace using the MPI blocking mode on Tintorrum.}
+  \label{fig:3}
+\end{figure*}
+
+\begin{figure*}[htbf]
+  \centering
+  \scriptsize
+  MPI polling mode\\
+  \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}}\\
+  MPI blocking mode\\
+  \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}
+  \caption{1 hour simulation (23h-24h) trace using the MPI blocking mode on Tintorrum.}
+  \label{fig:3}
+\end{figure*}

### 2014-04-24 paper_enahpc14/Sec3:Measurement-framework.tex
Manuel's changes

@@
 
-\subsubsection{UHAM - UJI}
+\subsubsection{Power-performance measurement framework at UJI}
+
 To assess the  performance and the energy efficiency  of COSMO-ART, we
-employ   a    version   of   the    integrated   framework   presented
-in~\cite{energy13} that works in  combination with Extrae and Paraver,
+employ   a    version   of   the \pmlib power measurement framework~\cite{energy13} that works in  combination with Extrae and Paraver~\ref{paraver},
 which are profiling/tracing and visualization tools, respectively.
 
-%The left part of the \vref{fig:Lustre} offers a graphical representation of
-%the Lustre architecture; the right depicts the tracing and profiling framework.
-To use our  approach, COSMO-ART is compiled using  the Extrae compiler
-wrappers,  which  automatically instrument  the  Fortran  code of  the
-model. Next, COSMO-ART  is run on the nodes,  thus dissipating certain
-amount  of power.   These  nodes are  connected  to power  measurement
-devices that account for the dissipated power/consumed energy and send
-the power  data to the  tracing server.  The client,  meanwhile, sends
-start/stop  primitives  in  order  to  gather  captured  data  by  the
-watt-meters onto the  tracing server, where an instance  of the \pmlib
-server is running.
+To use our approach, COSMO-ART is compiled using the Extrae library, 
+which  automatically instruments the Fortran  code of  the
+model in order to generate traces that record, e.g., MPI calls.
+Next, COSMO-ART  is run on the nodes,  thus dissipating certain
+amount  of power. These nodes are connected to power measurement devices that account
+for the dissipated power/consumed energy and send the power data to the tracing server. 
+The client,  meanwhile, sends start/stop  primitives  in  order  to  gather  captured  data  by  these PDUs onto the  tracing server, where an instance  of the \pmlib server is running.
 
-Once COSMO-ART run is finished, a file containing the power profile is
-created using the power data  received from the tracing server and the
-instrumentation post-processing generates the performance trace files.
-All  these files  are  combined then  in  Paraver which  allows us  to
-visualise the performance trace and the power profile of COSMO-ART all
-together.
+Once COSMO-ART run is finished, trace files containing the power profile 
+using the power data  received from the tracing server and the
+instrumentation trace files are generated.
+Afterwards, the combination of both trace files in Paraver allows us 
+to visualise and correlate the performance trace with the power profile of COSMO-ART.
 

### 2014-04-24 paper_enahpc14/Sec4:Evaluation.tex
Manuel's changes

--- 2014-04-23 paper_enahpc14/Sec4:Evaluation.tex
@@
 
-\paragraph{CSCS - Monch} 
-~\\
-\vskip -1em
-Monch is  a 10 rack  NEC-provided and dual-socket Intel  Ivy Bridge-EP
+\begin{itemize}
+
+\item \textsc{Monch} is a 10 rack NEC-provided and dual-socket Intel  Ivy Bridge-EP
 based  cluster, utilised  by  scientists  who are  part  of the  Swiss
@@
 
-\paragraph{CSCS - Pilatus} 
-~\\
-\vskip -1em
-Pilatus  is  a dual-socket  eight-core  Intel  Sandy  Bridge EP  based
+\item \textsc{Pilatus} is a dual-socket eight-core Intel Sandy Bridge EP based
 cluster used as Piz Daint pre-post processing cluster.  It is composed
@@
 
-\paragraph{UJI - Tintorrum} 
-~\\
-\vskip -1em
-Tintorrum  is a heterogeneous  cluster composed  of 28  compute nodes.
+\item \textsc{Tintorrum} is an heterogeneous  cluster composed  of 28  compute nodes.
 For our experiments only a subset of these nodes were considered. This
@@
 and the platform has 288 GB of DDR3 1333 MHz.
+
+\end{itemize}
 
@@
 
-\begin{figure}[htbf]
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-2.eps}
-  \caption{Time-to-solution and energy-to-solution on Tintorrum -
-    Blocking mode}
+\begin{figure*}[htbf]
+  \centering
+  \scriptsize
+  MPI polling mode\\
+  \includegraphics[width=0.78\textwidth]{Figs/23_24_blq1_stat.eps}\\
+  MPI blocking mode\\
+  \includegraphics[width=0.78\textwidth]{Figs/23_24_blq0_stat.eps}\\
+  \caption{blabla}
   \label{fig:3}
-\end{figure}
+\end{figure*}
+
+\begin{figure*}[htbf]
+  \centering\hspace{0.5cm}
+  \scalebox{0.5}{\input{Figs/24hour.pstex_t}}
+  \caption{24 hours simulation trace using the MPI blocking mode on Tintorrum.}
+  \label{fig:3}
+\end{figure*}
+
+\begin{figure*}[htbf]
+  \centering
+  \scriptsize
+  Time frame from 11\,h to 12\,h using the MPI polling mode\\
+  \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}}\\
+  Time frame from 11\,h to 12\,h using the MPI blocking mode
+  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}\\
+  Time frame from 23\,h to 24\,h using the MPI polling mode
+  \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}}\\
+  Time frame from 23\,h to 24\,h using the MPI blocking mode\\
+  \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}
+  \begin{tabular}{rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rl}
+& &  & &  & &  & &  \\[-0.15cm]
+Computation     & \multicolumn{1}{>{\columncolor[RGB]{  0,  , 255}}p{0.4cm}}{}  & & 
+Waiting a msg.  & \multicolumn{1}{>{\columncolor[RGB]{255,  0  ,0}}p{0.4cm}}{}  & & 
+Block. send     & \multicolumn{1}{>{\columncolor[RGB]{255,  0,174}}p{0.4cm}}{}  & & 
+Synchronization & \multicolumn{1}{>{\columncolor[RGB]{179,  0,  0}}p{0.4cm}}{}  & & 
+Wait/Wait all   & \multicolumn{1}{>{\columncolor[RGB]{235,  0,  0}}p{0.4cm}}{}  \\
+& &  & &  & &  & &  \\[-0.15cm]
+Immediate recv. & \multicolumn{1}{>{\columncolor[RGB]{100,100,177}}p{0.4cm}}{}  & &
+I/O             & \multicolumn{1}{>{\columncolor[RGB]{172,174, 41}}p{0.4cm}}{}  & &
+Group comm.     & \multicolumn{1}{>{\columncolor[RGB]{255,144, 26}}p{0.4cm}}{}  & &
+Others          & \multicolumn{1}{>{\columncolor[RGB]{192,224,  0}}p{0.4cm}}{}  & &
+Idle            & \multicolumn{1}{>{\columncolor[RGB]{117,195,255}}p{0.4cm}}{}  \\
+\end{tabular}
+  \caption{Traces for 1 hour simulation of COSMO-ART leveraging the polling and blocking MPI policies on Tintorrum.}
+  \label{fig:3}
+\end{figure*}

### 2014-04-24 paper_enahpc14/Sec3:Environment-setup.tex
changes structure of paper

--- 2014-04-23 trunk/paper_enahpc14/Sec4:Evaluation.tex
@@
-In  this section,  we  describe a  performance  and energy  efficiency
-evaluation  of  different  achitectures  when  running  the  COSMO-ART
-baseline.  We  start by  specifying our measurement  methodology along
-with the metrics used to  analyse the results on all platforms.  Then,
-we detail the environment  setup gathering the considered HPC systems,
-the  software  environment and  the  run  configuration.  Finally,  we
-discuss benchmark  results and  power-performance traces of  the model
-system.
+\subsection{Hardware platforms}
+\label{subsec:3.1}
 
-\subsection{Measurement methodology}
-\label{subsec:4.1}
-We  approach  the  assessment  of  the energy  footprint  and  overall
-performance    of    COSMO-ART    with    two    important    metrics:
-\textit{time-to-solution} (TTS) and \textit{energy-to-solution} (ETS).
-TTS refers to  the total wall clock time  of the application execution
-time. ETS  is the amount of  energy spent to  achieve results.  Energy
-consumption  is  assessed by  sampling  the  instantaneous peak  power
-during execution which  is then averaged and multiplied  by the TTS to
-determine  ETS.   Whenever   possible,  multiple  production  runs  of
-COSMO-ART  were performed  to  illustrate the  reproducibility of  the
-baseline,  and quantify  the  significant uncertainties  in the  power
-measurement, as dictated by the available technology.
-
-\subsection{Environment setup}
-\label{subsec:4.2}
 We have  chosen a state-of-the-art Intel's  third-generation Core (aka
@@
 
-\subsubsection{Sofware environment}
+\subsection{Software environment}
+\label{subsec:3.2}
+
 The COSMO-ART baseline  is a pure MPI based  Fortran~90 code currently
@@
 
-\subsection{Experimental results}
-\label{subsec:4.3}
+\subsection{Power-performance measurement framework}
+\label{subsec:3.3}
 
-\begin{figure}[htbf]
-  \begin{center}
-    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
-    \caption{Monch: Isola E1 Rack 2 Total Power}
-    \label{fig:1}
-  \end{center}
-\end{figure}
+We present  here two different  frameworks deployed on HPC  systems to
+measure  the  power  consumption   and  performance  of  the  baseline
+execution.
 
-\begin{figure}[htbf]
-  \begin{center}
-    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
-    \caption{Pilatus: Isola HD Total Power}
-    \label{fig:2}
-  \end{center}
-\end{figure}
+\subsubsection{CSCS - E3METER metering products}
+Supercomputer  clusters considered  for our  experiments at  the Swiss
+National Supercomputing Center (CSCS)  of ETH Zurich are equipped with
+E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
+high   quality   electricity  meters   released   by  Riedo   Networks
+(\url{http://riedonetworks.com/}),  that  enable  to monitor  and  log
+power  consumption of  the  IT infrastructure  as  well as  constantly
+analyze  line  voltage,  current,  power-factor,  frequency  with  1\%
+accuracy.   Using reliable  narrowband  powerline communication  (PLC)
+technology,  all metering  and power  quality data  from each  IPS are
+centrally collected by the E3METER Data Concentrator, via the existing
+power cables thus  avoiding the need for extra  cabling.  This data is
+made  available  via SNMP,  HTTP,  TELNET  through  the built-in  Fast
+Ethernet  port.   Time  synchronisation  is guaranteed  by  using  NTP
+servers.   Measured data  is accessed  through the  open  source Cacti
+software including  the E3METER  Cacti Plugin to  scan the  entire PLC
+network and monitor  in real-time the power usage  of individual rack,
+recorded in 5 minutes interval periods.
 
-Figures~\ref{fig:1} and  \ref{fig:2} account respectively  for Monch's
-Isola E1  Rack 2  and Pilatus' Isola  HD total power  measurements for
-1-day or 2-days simulations. On  the Intel Ivy Bridge EP based cluster
-(i.e. Monch), the 1-day simulation  was issued only twice due to usage
-restrictions. As time resolution was set to one update every 5 minutes
-for  power sampling,  the average  power consumption  was  computed by
-considering 6 values for each single  run.  On the Intel Xeon E5 based
-cluster (i.e.   Pilatus), the 1-day  simulation was issued  four times
-and a 2-days  run only once. Similarly, the  average power consumption
-was computed by  considering 4 values for each single  1-day run and 9
-values  for the  2-days  run. Corresponding  results  are gathered  in
-Table~\ref{tab:3}.
+\subsubsection{UHAM - UJI}
+To assess the  performance and the energy efficiency  of COSMO-ART, we
+employ   a    version   of   the    integrated   framework   presented
+in~\cite{energy13} that works in  combination with Extrae and Paraver,
+which are profiling/tracing and visualization tools, respectively.
 
-\begin{table}[htbf]
-  \begin{center}
-    \caption{Average power consumption (W) of the platforms}
-    \label{tab:3}
-    \begin{tabular}{ccc}
-      \hline\noalign{\smallskip}
-      \textbf{Simulation time} & \textbf{Xeon E5} & \textbf{Ivy Bridge EP} \\
-      \noalign{\smallskip}\hline\noalign{\smallskip}
-      \textbf{1 day} & 18122.01417 & 12658.52278 \\ 
-      & 17979.61083 & 12586.40833 \\
-      & 18065.45167 & - \\
-      & 17973.02833 & - \\
-      \noalign{\smallskip}\hline\noalign{\smallskip}
-      \textbf{2 days} & 17997.57815 & - \\
-      \noalign{\smallskip}\hline
-    \end{tabular}
-  \end{center}
-\end{table}
+%The left part of the \vref{fig:Lustre} offers a graphical representation of
+%the Lustre architecture; the right depicts the tracing and profiling framework.
+To use our  approach, COSMO-ART is compiled using  the Extrae compiler
+wrappers,  which  automatically instrument  the  Fortran  code of  the
+model. Next, COSMO-ART  is run on the nodes,  thus dissipating certain
+amount  of power.   These  nodes are  connected  to power  measurement
+devices that account for the dissipated power/consumed energy and send
+the power  data to the  tracing server.  The client,  meanwhile, sends
+start/stop  primitives  in  order  to  gather  captured  data  by  the
+watt-meters onto the  tracing server, where an instance  of the \pmlib
+server is running.
 
-In Figure~\ref{fig:3}, we compare both time-to-solution (right y-axis)
-and  energy-to-solution (left  y-axis) metrics  on both  platforms. As
-expected,  Xeon  E5 outperforms  Ivy  Bridge  EP,  being roughly  1.3x
-faster.   The reason for  that is  two-fold: (1)  it has  higher clock
-frequency than Ivy  Bridge (2.6 GHz against 2.2 GHz),  and (2) it aims
-at  computing   speed  regardless   to  energy  consumption.   In  our
-experiments,  Ivy  Bridge   EP  showed  the  best  energy-to-solution,
-reducing the energy consumption of Xeon E5 by approximately $7\%$.
+Once COSMO-ART run is finished, a file containing the power profile is
+created using the power data  received from the tracing server and the
+instrumentation post-processing generates the performance trace files.
+All  these files  are  combined then  in  Paraver which  allows us  to
+visualise the performance trace and the power profile of COSMO-ART all
+together.
 
-\begin{figure}[htbf]
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART.eps}
-  \caption{Time-to-solution and  energy-to-solution comparison between
-    Xeon E5 and Ivy Bridge-EP architectures}
-  \label{fig:3}
-\end{figure}
+%Once COSMO-ART run is finished, the VampirTrace \pmlib plugin receives
+%the  power   data  from  the  tracing   server.   The  instrumentation
+%post-processing generates  the performance trace files  and the \pmlib
+%plugin inserts the power data into them.
+
+%In  addition  to the  power  measurements,  we  also account  for  the
+%resource utilization values  of the nodes: CPU load,  memory usage and
+%storage device utilization. % and network utilization.
+
+%We  run special  \pmlib  server  instances on  the  server nodes  that
+%retrieve these  values from the \texttt{proc}  file system (leveraging
+%the  \texttt{psutil} Python library).   Thus, \pmlib  plugin instances
+%running  with the  instrumented  application connect  with the  \pmlib
+%servers.   Finally,   using  the   Vampir   visualization  tool,   the
+%power-performance traces  can be easily  analyzed through a  series of
+%plots and statistics.

### 2014-04-24 paper_enahpc14/Sec4:Experimental-results.tex
changes structure of paper

--- 2014-04-24 paper_enahpc14/Sec4:Evaluation.tex
@@
 \label{subsec:4.1}
+
 We  approach  the  assessment  of  the energy  footprint  and  overall
@@
 
-\subsection{Environment setup}
+\subsection{Time-energy analysis of COSMO-ART}
 \label{subsec:4.2}
-
-\subsubsection{HPC platforms}
-We have  chosen a state-of-the-art Intel's  third-generation Core (aka
-Ivy  Bridge) processing platform  at CSCS  (called ``Monch'')  for our
-power  measurements,  which  is  slated  to stay  in  service  without
-hardware  upgrade  for the  duration  of  the  Exa2Green project.   In
-principle at least,  this architecture could be recreated  or found in
-an identical configuration beyond  the lifetime of the project.  Since
-the baseline benchmark can  be reproduced within an expected variance,
-and  that the baseline  run configuration  can be  used in  all future
-versions  of the  code, a  fair comparison  will be  made  between the
-baseline and  the milestone versions  of COSMO-ART.  In this  study, a
-complementary  energy-to-solution benchmarking  comparison  is carried
-out   on  the   ``Pilatus''  cluster   at  CSCS,   based   on  Intel's
-previous-generation  Sandy  Bridge  processors,  conventional  in  HPC
-systems and known to be  more power consuming.  Finally a full tracing
-experiment is  conducted on the  ``Tintorrum'' cluster at the  Jaume I
-University (UJI) using an  integrated framework, to capture an overall
-power profile at a much finer resolution and to get more insights into
-the MPI blocking and polling influences on power savings.
-
-\paragraph{CSCS - Monch} 
-~\\
-\vskip -1em
-Monch is  a 10 rack  NEC-provided and dual-socket Intel  Ivy Bridge-EP
-based  cluster, utilised  by  scientists  who are  part  of the  Swiss
-Platform     for      Advanced     Scientific     Computing     (PASC,
-\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
-nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
-Each standard  compute node comprises  two Intel Ivy  Bridge Efficient
-Performance (EP)  E5-2660 v2 ten-core processors operating  at 2.2 GHz
-base  clock speed,  themselves connected  by a  high  speed InfiniBand
-network based on Mellanox SX6036  managed FDR switches, with a 56 Gb/s
-speed.  Each core  has 32 KB instruction and 32 KB  data L1 caches and
-256  KB of L2  cache. All  10 cores  share a  25 MB  L3 cache  and the
-platform has  32GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
-benchmark, a  full rack  of Monch constituted  of 52  standard compute
-nodes (monchc[029-080]) was considered.
-
-\paragraph{CSCS - Pilatus} 
-~\\
-\vskip -1em
-Pilatus  is  a dual-socket  eight-core  Intel  Sandy  Bridge EP  based
-cluster used as Piz Daint pre-post processing cluster.  It is composed
-of 42 compute  nodes and has 2 high-speed  interconnects based on FDR:
-the  first is  dedicated to  the  MPI traffic  and the  second to  the
-storage high  speed traffic.   The 2 login  nodes and the  42 computes
-nodes  consists in  11  twin-pair Intel  E5-Series  DALCO r2264i4t  2U
-scalable compute modules.  Each  module contains 4 compute nodes based
-on two Intel  Xeon E5-2670 processors operating at  2.6 GHz base clock
-speed, themselves  connected by a high speed  InfiniBand network based
-on Mellanox  SX6036 managed FDR switches,  with a 56  Gb/s speed. Each
-core has 32 KB  instruction and 32 KB data L1 caches  and 256 KB of L2
-cache.  All  the 8 core share  a 20 MB  L3 cache and the  platform has
-64GB of  DDR3 1600 MHz  RAM.  For our energy-to-solution  benchmark, a
-full  rack  of  Pilatus  constituted  of  42  standard  compute  nodes
-(pilatus[03-44]) was considered.
-
-\paragraph{UJI - Tintorrum} 
-~\\
-\vskip -1em
-Tintorrum  is a heterogeneous  cluster composed  of 28  compute nodes.
-For our experiments only a subset of these nodes were considered. This
-set is  composed of 16  nodes, each of  which includes two  Intel Xeon
-E5645 hexa-core  processors at 2.40  GHz connected via  Infiniband QDR
-(Mellanox MTS3600 switch).  Each core  has 32 KB instruction and 32 KB
-data L1 caches and 256 KB L2 cache. The 6 cores share a 12 MB L3 cache
-and the platform has 288 GB of DDR3 1333 MHz.
-
-\subsubsection{Sofware environment}
-The COSMO-ART baseline  is a pure MPI based  Fortran~90 code currently
-running on distributed multi-core systems only.  The software stack on
-both CSCS  platforms was controlled using the  modules framework which
-gives an easy and flexible mechanism  to access to all of the provided
-compilers, tools  and applications.  For our  initial benchmarking, we
-opted for the GNU compiler  (gcc/4.8.1 on Monch, gcc/4.8.2 on Pilatus)
-using the -O3  compiler flag in favor of  the intel compiler (14.0.1),
-delivering inferior performance.  In addition, we installed the MPICH2
-implementation of MPI (mvapich2/1.9) as well as the commonly used HDF5
-(1.8.12) and NetCDF (4.3.1)  libraries for the management of extremely
-large  and  complex data  collections.   All  computes  nodes have  an
-operating      system      based      on      GNU/Linux      featuring
-``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
-``3.0.101-0.15-default'' kernel in Pilatus.
-
-\subsubsection{Run configuration}
-A snapshot of the code, which includes, at least conceptually, all the
-information needed  to reproduce the  energy-to-solution benchmarks of
-COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
-node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
-Pilatus.   The  calculated  region  was mapped  to  the  participating
-processors using  a 2D-partitioning strategy.   The distribution along
-the $x$  and $y$ coordinates  was defined by setting:  $nprocx=40$ and
-$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
-usually  kept  bigger than  $nprocy$.   Besides,  as  this version  of
-COSMO-ART  doesn't  make  use  of  the traditional  GRIB  library,  we
-specified $nprocio=0$ for GRIB  I/O.  Hyperthreading is not considered
-in this study as previous attempts  of its use revealed that it always
-led to higher energy-to-solution.\\
-
-\subsection{Experimental results}
-\label{subsec:4.3}
 
@@
     Polling mode}
-  \label{fig:3}
+  \label{fig:4}
 \end{figure}
+
+\begin{figure}[htbf]
+  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-2.eps}
+  \caption{Time-to-solution and energy-to-solution on Tintorrum -
+    Polling mode}
+  \label{fig:5}
+\end{figure}
+
+\begin{figure}[htbf]
+  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-3.eps}
+  \caption{Time-to-solution and Power on Tintorrum - Polling mode}
+  \label{fig:6}
+\end{figure}
+
+\begin{figure}[htbf]
+  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-4.eps}
+  \caption{Time-to-solution and Power on Tintorrum - Blocking mode}
+  \label{fig:7}
+\end{figure}
+
+\subsection{Power-performance tracing of COSMO-ART}
+\label{subsec:4.3}
 
@@
   \caption{blabla}
-  \label{fig:3}
+  \label{fig:8}
 \end{figure*}
@@
   \caption{24 hours simulation trace using the MPI blocking mode on Tintorrum.}
-  \label{fig:3}
+  \label{fig:9}
 \end{figure*}
@@
   \caption{1 hour simulation (11h-12h) trace using the MPI blocking mode on Tintorrum.}
-  \label{fig:3}
+  \label{fig:10}
 \end{figure*}
@@
   \caption{1 hour simulation (23h-24h) trace using the MPI blocking mode on Tintorrum.}
-  \label{fig:3}
+  \label{fig:11}
 \end{figure*}

### 2014-04-24 trunk/paper_enahpc14/Introduction.tex
adds a few changes

@@
 GPUs  \citep{Gysi-2014, Lapillonne-2014} within  the framework  of the
-High Performance and High  Productivity Computing (HP2C) Initiative to
-optimise    it    for     computational    and    energy    efficiency
-(\url{http://www.hp2c.ch/}), significant investments  in ART are still
-required to take  it to a similar level.   The efficiency of COSMO-ART
-is     being    addressed    in     the    EU     Exa2Green    project
-(\url{http://exa2green.eu/}). The  ultimate goal of the  project is to
-deliver a  prototype code, which  provides an energy efficiency  of at
-least five times of the  baseline value.  Such an implementation would
-allow  the  community  to  investigate critical  questions  at  higher
-resolution  and   over  longer  periods,   at  reduced  cost   to  the
-environment.
+High    Performance   and    High   Productivity    Computing   (HP2C)
+Initiative \citep{HP2C},  to optimise it for  computational and energy
+efficiency, significant investments in  ART are still required to take
+it to a similar level.  The efficiency of COSMO-ART is being addressed
+in the  EU Exa2Green project \citep{EXA2GREEN} to  deliver a prototype
+code, which  provides an energy efficiency  of at least  five times of
+the baseline value.  Such  an implementation would allow the community
+to investigate critical questions at higher resolution and over longer
+periods, at reduced cost to the environment.
 
@@
 introduce  COSMO-ART and specify  its technical  setup related  to the
-investigated  performance  and  energy  evaluation  methods.   In  the
-subsequent Sec.~\ref{sec:3},  we describe different  power measurement
-equipments as well as energy-aware techniques employed to conduct this
-benchmarking  study.  Sec.~\ref{sec:4}  presents  the performance  and
-energy  requirements  of the  baseline  on  various architectures  and
-highlights  areas  where  improvements   will  be  necessary  for  the
-subsequent baseline refactoring.   Finally, we conclude with practical
-guidelines  to identifying  energy  saving opportunities  and give  an
-outlook to future research in Sec.~\ref{concl}.
+investigated   performance   and   energy  evaluation   methods.    In
+Sec.~\ref{sec:3}   we  describe   HPC  platforms,   power  measurement
+equipments   and  software  environment   employed  to   conduct  this
+benchmarking study.  Sec.~\ref{sec:4}  presents performance and energy
+requirements  of the  baseline on  these architectures  and highlights
+areas where improvements will be necessary for the subsequent baseline
+refactoring.   Finally,  we  conclude  with  practical  guidelines  to
+identifying energy saving opportunities  and give an outlook to future
+research in Sec.~\ref{concl}.

### 2014-04-24 trunk/paper_enahpc14/Sec3:Environment-setup.tex
adds a few changes

--- 2014-04-24 paper_enahpc14/Sec3:Environment-setup.tex
@@
 based  cluster, utilised  by  scientists  who are  part  of the  Swiss
-Platform     for      Advanced     Scientific     Computing     (PASC,
-\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
-nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
-Each standard  compute node comprises  two Intel Ivy  Bridge Efficient
-Performance (EP)  E5-2660 v2 ten-core processors operating  at 2.2 GHz
-base  clock speed,  themselves connected  by a  high  speed InfiniBand
-network based on Mellanox SX6036  managed FDR switches, with a 56 Gb/s
-speed.  Each core  has 32 KB instruction and 32 KB  data L1 caches and
-256  KB of L2  cache. All  10 cores  share a  25 MB  L3 cache  and the
-platform has  32GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
-benchmark, a  full rack  of Monch constituted  of 52  standard compute
-nodes (monchc[029-080]) was considered.
+Platform for Advanced Scientific  Computing (PASC) \citep{PASC}. It is
+composed of 312 standard  compute nodes, 24 large-memory compute nodes
+and  24  huge-memory  compute   nodes.   Each  standard  compute  node
+comprises two  Intel Ivy Bridge Efficient Performance  (EP) E5-2660 v2
+ten-core processors operating at  2.2 GHz base clock speed, themselves
+connected by a high speed  InfiniBand network based on Mellanox SX6036
+managed  FDR switches,  with a  56 Gb/s  speed.  Each  core has  32 KB
+instruction and 32  KB data L1 caches  and 256 KB of L2  cache. All 10
+cores share a  25 MB L3 cache  and the platform has 32GB  of DDR3 1600
+MHz RAM.  For our energy-to-solution  benchmark, a full rack  of Monch
+constituted  of  52   standard  compute  nodes  (monchc[029-080])  was
+considered.
 
@@
 
-\subsection{Software environment}
-\label{subsec:3.2}
-
-The COSMO-ART baseline  is a pure MPI based  Fortran~90 code currently
-running on distributed multi-core systems only.  The software stack on
-both CSCS  platforms was controlled using the  modules framework which
-gives an easy and flexible mechanism  to access to all of the provided
-compilers, tools  and applications.  For our  initial benchmarking, we
-opted for the GNU compiler  (gcc/4.8.1 on Monch, gcc/4.8.2 on Pilatus)
-using the -O3  compiler flag in favor of  the intel compiler (14.0.1),
-delivering inferior performance.  In addition, we installed the MPICH2
-implementation of MPI (mvapich2/1.9) as well as the commonly used HDF5
-(1.8.12) and NetCDF (4.3.1)  libraries for the management of extremely
-large  and  complex data  collections.   All  computes  nodes have  an
-operating      system      based      on      GNU/Linux      featuring
-``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
-``3.0.101-0.15-default'' kernel in Pilatus.
-
-\subsubsection{Run configuration}
-A snapshot of the code, which includes, at least conceptually, all the
-information needed  to reproduce the  energy-to-solution benchmarks of
-COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
-node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
-Pilatus.   The  calculated  region  was mapped  to  the  participating
-processors using  a 2D-partitioning strategy.   The distribution along
-the $x$  and $y$ coordinates  was defined by setting:  $nprocx=40$ and
-$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
-usually  kept  bigger than  $nprocy$.   Besides,  as  this version  of
-COSMO-ART  doesn't  make  use  of  the traditional  GRIB  library,  we
-specified $nprocio=0$ for GRIB  I/O.  Hyperthreading is not considered
-in this study as previous attempts  of its use revealed that it always
-led to higher energy-to-solution.\\
-
 \subsection{Power-performance measurement framework}
@@
 high   quality   electricity  meters   released   by  Riedo   Networks
-(\url{http://riedonetworks.com/}),  that  enable  to monitor  and  log
-power  consumption of  the  IT infrastructure  as  well as  constantly
-analyze  line  voltage,  current,  power-factor,  frequency  with  1\%
-accuracy.   Using reliable  narrowband  powerline communication  (PLC)
-technology,  all metering  and power  quality data  from each  IPS are
-centrally collected by the E3METER Data Concentrator, via the existing
-power cables thus  avoiding the need for extra  cabling.  This data is
-made  available  via SNMP,  HTTP,  TELNET  through  the built-in  Fast
-Ethernet  port.   Time  synchronisation  is guaranteed  by  using  NTP
-servers.   Measured data  is accessed  through the  open  source Cacti
-software including  the E3METER  Cacti Plugin to  scan the  entire PLC
-network and monitor  in real-time the power usage  of individual rack,
-recorded in 5 minutes interval periods.
+\citep{Riedonetworks},   that  enable   to  monitor   and   log  power
+consumption  of the IT  infrastructure as  well as  constantly analyze
+line  voltage,  current, power-factor,  frequency  with 1\%  accuracy.
+Using  reliable narrowband  powerline communication  (PLC) technology,
+all  metering and  power  quality  data from  each  IPS are  centrally
+collected  by the E3METER  Data Concentrator,  via the  existing power
+cables thus  avoiding the need for  extra cabling.  This  data is made
+available via  SNMP, HTTP, TELNET  through the built-in  Fast Ethernet
+port.   Time  synchronisation  is  guaranteed by  using  NTP  servers.
+Measured  data is  accessed  through the  open  source Cacti  software
+including the E3METER Cacti Plugin  to scan the entire PLC network and
+monitor in real-time the power usage of individual rack, recorded in 5
+minutes interval periods.
 
@@
 %plots and statistics.
+
+\subsection{Software environment}
+\label{subsec:3.2}
+
+The COSMO-ART baseline  is a pure MPI based  Fortran~90 code currently
+running on distributed multi-core systems only.  The software stack on
+both CSCS  platforms was controlled using the  modules framework which
+gives an easy and flexible mechanism  to access to all of the provided
+compilers, tools  and applications.  For our  initial benchmarking, we
+opted for the GNU compiler  (gcc/4.8.1 on Monch, gcc/4.8.2 on Pilatus)
+using the -O3  compiler flag in favor of  the intel compiler (14.0.1),
+delivering inferior performance.  In addition, we installed the MPICH2
+implementation of MPI (mvapich2/1.9) as well as the commonly used HDF5
+(1.8.12) and NetCDF (4.3.1)  libraries for the management of extremely
+large  and  complex data  collections.   All  computes  nodes have  an
+operating      system      based      on      GNU/Linux      featuring
+``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
+``3.0.101-0.15-default'' kernel in Pilatus.
+
+\subsubsection{Run configuration}
+A snapshot of the code, which includes, at least conceptually, all the
+information needed  to reproduce the  energy-to-solution benchmarks of
+COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
+node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
+Pilatus.   The  calculated  region  was mapped  to  the  participating
+processors using  a 2D-partitioning strategy.   The distribution along
+the $x$  and $y$ coordinates  was defined by setting:  $nprocx=40$ and
+$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
+usually  kept  bigger than  $nprocy$.   Besides,  as  this version  of
+COSMO-ART  doesn't  make  use  of  the traditional  GRIB  library,  we
+specified $nprocio=0$ for GRIB  I/O.  Hyperthreading is not considered
+in this study as previous attempts  of its use revealed that it always
+led to higher energy-to-solution.\\

### 2014-04-24 paper_enahpc14/Sec2:COSMO-ART.tex
adds a few changes

@@
 simulated gaseous and  aerosol concentrations.  This condition entails
-that at the  model initialization there is a  period of adjustment for
+that at the model initialization,  there is a period of adjustment for
 the model  to reach its equilibrium  state and minimize  the effect of

### 2014-04-24 paper_enahpc14/paper.tex
changes structure of paper
adds a few changes

@@
 \usepackage{color}
+\usepackage{colortbl}
 \usepackage{xspace}
@@
   related  to  energy  efficient  high performance  computing  in  the
-  framework of the Exa2Green project (\url{http://exa2green.eu/}).
+  framework of the Exa2Green project \citep{EXA2GREEN}.
 
@@
 
-\section{Related work \emph{(to be completed)}}
+\section{Related work}
 \label{sec:1}
@@
 
-\section{Power-performance measurement framework}
+\section{Environment setup}
 \label{sec:3}
-\input{Sec3:Measurement-framework.tex}
+\input{Sec3:Environment-setup.tex}
 
-\section{Evaluation}
+\section{Experimental results}
 \label{sec:4}
-\input{Sec4:Evaluation.tex}
+\input{Sec4:Experimental-results.tex}
 
-\section{Conclusion \emph{(to be completed)}}
+\section{Conclusion}
 \label{concl}
@@
 also gratefully acknowledge the High Performance and High Productivity
-Computing  Initiative  (\url{www.hp2c.ch}) for  results  that will  be
-leveraged in subsequent code refactoring.
+Computing Initiative  \citep{HP2C} for results that  will be leveraged
+in subsequent code refactoring.
 \end{acknowledgements}

### 2014-04-24 paper_enahpc14/Introduction.tex
gets rid of DOI information in paper.bib

@@
 concentrations impact the climate  system by scattering and absorption
-of   solar   radiation,    thereby   exerting   a   direct   radiative
-forcing  \citep{Liao-2005,  Bangert-2012,  Lundgren-2013}; or  by  the
-modification of  cloud properties, cloud fraction  and surface albedo,
-causing  a negative  indirect  radiative forcing  \citep{Haywood-2000,
-Lohmann-2005,     VandenHeever-2011,     Rosenfeld-2013}.      Despite
+of solar radiation, thereby exerting a direct radiative forcing; or by
+the  modification  of cloud  properties,  cloud  fraction and  surface
+albedo,  causing  a  negative  indirect  radiative  forcing.   Despite
 considerable  progress in  global aerosol  modelling \citep{Mann-2013}
-and    measurement-based    assessments   \citep{Myhre-2009},    large
-uncertainties  remain  in   current  estimates  of  aerosol  radiative
-forcing   \citep{Myhre-2013,    IPCC-2013,   Lee-2013,   Randles-2013,
-Rosenfeld-2013, Sherwood-2013, Stier-2013}.
+and  measurement-based  assessments,  large  uncertainties  remain  in
+current  estimates  of  aerosol radiative  forcing  \citep{Myhre-2013,
+IPCC-2013,  Lee-2013,   Randles-2013,  Rosenfeld-2013,  Sherwood-2013,
+Stier-2013}.
 
@@
 
-The extended  COSMO-ART model  is computationally much  more demanding
-than COSMO  since a large  number of additional tracers  and processes
-have to  be considered.  Thus the model  is currently severely-limited
-in  terms   of  applicability  and   expensive  in  terms   of  energy
-consumption.    Although   COSMO   has   recently   been   ported   to
+The extended atmospheric model  COSMO-ART is computationally much more
+demanding than  COSMO since a  large number of additional  tracers and
+processes  have  to  be  considered.   Thus  the  model  is  currently
+severely-limited in  terms of applicability and expensive  in terms of
+energy  consumption.   Although  COSMO  has recently  been  ported  to
 GPUs  \citep{Gysi-2014, Lapillonne-2014} within  the framework  of the
-High Performance and High  Productivity Computing (HP2C) Initiative to
-optimise    it    for     computational    and    energy    efficiency
-(\url{http://www.hp2c.ch/}), significant investments  in ART are still
-required to take  it to a similar level.   The efficiency of COSMO-ART
-is     being    addressed    in     the    EU     Exa2Green    project
-(\url{http://exa2green.eu/}). The  ultimate goal of the  project is to
-deliver a  prototype code, which  provides an energy efficiency  of at
-least five times of the  baseline value.  Such an implementation would
-allow  the  community  to  investigate critical  questions  at  higher
-resolution  and   over  longer  periods,   at  reduced  cost   to  the
-environment.
+High    Performance   and    High   Productivity    Computing   (HP2C)
+Initiative \citep{HP2C},  to optimise it for  computational and energy
+efficiency, significant investments in  ART are still required to take
+it to a similar level.  The efficiency of COSMO-ART is being addressed
+in the  EU Exa2Green project \citep{EXA2GREEN} to  deliver a prototype
+code, which  provides an energy efficiency  of at least  five times of
+the baseline value.  Such  an implementation would allow the community
+to investigate critical questions at higher resolution and over longer
+periods, at reduced cost to the environment.
 
-This  work is organized  as follows:  in Sec.~\ref{sec:1}  we describe
-related  work,  then in  Sec.~\ref{sec:2},  we  briefly introduce  the
-regional atmospheric  model COSMO-ART and specify  its technical setup
-related to the investigated performance and energy evaluation methods.
-Sec.~\ref{sec:5} describes the HPC  supercomputers used to compute the
-baseline.  Sec.~\ref{sec:3}   describes  different  power  measurement
-equipment  used for  conducting this  benchmarking study  and presents
-energy-aware techniques employed to  determine the energy footprint of
-the   considered  application.   Sec.~\ref{sec:4}   gathers  benchmark
-results  from  the  Exa2Green   project  and  highlights  areas  where
-improvements   will   be  necessary   for   the  subsequent   baseline
+This  work is  organized as  follows: in  Sec.~\ref{sec:1} we  give an
+overview  of  related  work,   then  in  Sec.~\ref{sec:2}  we  briefly
+introduce  COSMO-ART and specify  its technical  setup related  to the
+investigated   performance   and   energy  evaluation   methods.    In
+Sec.~\ref{sec:3}   we  describe   HPC  platforms,   power  measurement
+equipments   and  software  environment   employed  to   conduct  this
+benchmarking study.  Sec.~\ref{sec:4}  presents performance and energy
+requirements  of the  baseline on  these architectures  and highlights
+areas where improvements will be necessary for the subsequent baseline
 refactoring.   Finally,  we  conclude  with  practical  guidelines  to

### 2014-04-24 paper_enahpc14/Sec2:COSMO-ART.tex
reduces size of paper to 7 pages

@@
 \label{subsec:1.1}
+
 The  model  system COSMO-ART  described  in  \citep{Vogel-2009}, is  a
@@
 \label{subsec:1.2}
+
 Establishing effective energy performance benchmarking of a code under
@@
 Boundary data  for gas-phase species are taken  from IFS-MOZART output
-at  6h temporal resolution.  The model  setup incorporates  34~2-D and
-45~3-D  fields to  be written  out every  hour and  is devoid  of data
+at  6h temporal resolution.   The model  setup incorporates  34~2D and
+45~3D  fields to  be written  out  every hour  and is  devoid of  data
 assimilation methods.

### 2014-04-24 paper_enahpc14/Sec4:Experimental-results.tex
reduces size of paper to 7 pages

@@
-In  this section,  we  describe a  performance  and energy  efficiency
-evaluation  of  different  achitectures  when  running  the  COSMO-ART
-baseline.  We  start by  specifying our measurement  methodology along
-with the metrics used to  analyse the results on all platforms.  Then,
-we detail the environment  setup gathering the considered HPC systems,
-the  software  environment and  the  run  configuration.  Finally,  we
-discuss benchmark  results and  power-performance traces of  the model
-system.
+In this section, we present our measurement methodology along with the
+metrics considered and discuss energy efficiency and power-performance
+traces of the model system.
 
@@
 during execution which  is then averaged and multiplied  by the TTS to
-determine  ETS.   Whenever   possible,  multiple  production  runs  of
-COSMO-ART  were performed  to  illustrate the  reproducibility of  the
-baseline,  and quantify  the  significant uncertainties  in the  power
-measurement, as dictated by the available technology.
+determine  ETS.   Whenever  possible,  multiple production  runs  were
+performed  to  illustrate the  reproducibility  of  the baseline,  and
+quantify the  significant uncertainties  in the power  measurement, as
+dictated by the available technology.
 
@@
     \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
-    \caption{Monch: Isola E1 Rack 2 Total Power}
+    \caption{\textsc{Monch}: Isola E1 Rack 2 Total Power}
     \label{fig:1}
@@
     \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
-    \caption{Pilatus: Isola HD Total Power}
+    \caption{\textsc{Pilatus}: Isola HD Total Power}
     \label{fig:2}
@@
 
-Figures~\ref{fig:1} and  \ref{fig:2} account respectively  for Monch's
-Isola E1  Rack 2  and Pilatus' Isola  HD total power  measurements for
-1-day or 2-days simulations. On  the Intel Ivy Bridge EP based cluster
-(i.e. Monch), the 1-day simulation  was issued only twice due to usage
-restrictions. As time resolution was set to one update every 5 minutes
-for  power sampling,  the average  power consumption  was  computed by
-considering 6 values for each single  run.  On the Intel Xeon E5 based
-cluster (i.e.   Pilatus), the 1-day  simulation was issued  four times
-and a 2-days  run only once. Similarly, the  average power consumption
-was computed by  considering 4 values for each single  1-day run and 9
-values  for the  2-days  run. Corresponding  results  are gathered  in
-Table~\ref{tab:3}.
+Figures~\ref{fig:1}   and   \ref{fig:2}   account   respectively   for
+\textsc{Monch}'s Isola E1 Rack  2 and \textsc{Pilatus}' Isola HD total
+power measurements for  1-day or 2-days simulations. On  the Intel Ivy
+Bridge EP  based cluster (i.e.  \textsc{Monch}),  the 1-day simulation
+was issued  only twice due  to usage restrictions. As  time resolution
+was set to one update every  5 minutes for power sampling, the average
+power consumption was computed by considering 6 values for each single
+run.  On the Intel Xeon E5 based cluster (i.e.  \textsc{Pilatus}), the
+1-day simulation  was issued  four times and  a 2-days run  only once.
+Similarly, the average power consumption was computed by considering 4
+values  for  each  single  1-day  run  and 9  values  for  the  2-days
+run. Corresponding results are gathered in Table~\ref{tab:3}.
 
@@
 
-In Figure~\ref{fig:3}, we compare both time-to-solution (right y-axis)
-and  energy-to-solution (left  y-axis) metrics  on both  platforms. As
-expected,  Xeon  E5 outperforms  Ivy  Bridge  EP,  being roughly  1.3x
-faster.   The reason for  that is  two-fold: (1)  it has  higher clock
-frequency than Ivy  Bridge (2.6 GHz against 2.2 GHz),  and (2) it aims
-at  computing   speed  regardless   to  energy  consumption.   In  our
-experiments,  Ivy  Bridge   EP  showed  the  best  energy-to-solution,
-reducing the energy consumption of Xeon E5 by approximately $7\%$.
+In  Figure~\ref{fig:3}, we  compare both  TTS (right  y-axis)  and ETS
+(left  y-axis)  metrics  on  both  platforms.  As  expected,  Xeon  E5
+outperforms Ivy Bridge EP, being  roughly 1.3x faster.  The reason for
+that is  two-fold: (1) it has  higher clock frequency  than Ivy Bridge
+(2.6  GHz  against  2.2 GHz),  and  (2)  it  aims at  computing  speed
+regardless to  energy consumption.  In our experiments,  Ivy Bridge EP
+showed the best energy-to-solution, reducing the energy consumption of
+Xeon E5 by approximately $7\%$.
 
@@
   \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
-  \caption{Time-to-solution and energy-to-solution on Tintorrum -
-    Polling mode}
+  \caption{Time-to-solution and energy-to-solution on
+    \textsc{Tintorrum} - Polling mode}
   \label{fig:4}
@@
   \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-2.eps}
-  \caption{Time-to-solution and energy-to-solution on Tintorrum -
-    Polling mode}
+  \caption{Time-to-solution and energy-to-solution on
+    \textsc{Tintorrum} - Polling mode}
   \label{fig:5}
@@
   \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-3.eps}
-  \caption{Time-to-solution and Power on Tintorrum - Polling mode}
+  \caption{Time-to-solution and Power on \textsc{Tintorrum} - Polling
+    mode}
   \label{fig:6}
@@
   \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-4.eps}
-  \caption{Time-to-solution and Power on Tintorrum - Blocking mode}
+  \caption{Time-to-solution and Power on \textsc{Tintorrum} - Blocking
+    mode}
   \label{fig:7}
@@
   \scalebox{0.5}{\input{Figs/24hour.pstex_t}}
-  \caption{24 hours simulation trace using the MPI blocking mode on Tintorrum.}
+  \caption{24 hours simulation trace using the MPI blocking mode on
+    \textsc{Tintorrum}.}
   \label{fig:9}
@@
   \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}
-  \caption{1 hour simulation (11h-12h) trace using the MPI blocking mode on Tintorrum.}
+  \caption{1 hour simulation (11h-12h) trace using the MPI blocking
+    mode on \textsc{Tintorrum}.}
   \label{fig:10}
@@
   \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}
-  \caption{1 hour simulation (23h-24h) trace using the MPI blocking mode on Tintorrum.}
+  \caption{1 hour simulation (23h-24h) trace using the MPI blocking
+    mode on \textsc{Tintorrum}.}
   \label{fig:11}

### 2014-04-24 trunk/paper_enahpc14/Sec3:Environment-setup.tex
reduces size of paper to 7 pages

@@
 
-We have  chosen a state-of-the-art Intel's  third-generation Core (aka
-Ivy  Bridge) processing platform  at CSCS  (called ``Monch'')  for our
-power  measurements,  which  is  slated  to stay  in  service  without
-hardware  upgrade  for the  duration  of  the  Exa2Green project.   In
-principle at least,  this architecture could be recreated  or found in
-an identical configuration beyond  the lifetime of the project.  Since
-the baseline benchmark can  be reproduced within an expected variance,
-and  that the baseline  run configuration  can be  used in  all future
-versions  of the  code, a  fair comparison  will be  made  between the
-baseline and  the milestone versions  of COSMO-ART.  In this  study, a
-complementary  energy-to-solution benchmarking  comparison  is carried
-out   on  the   ``Pilatus''  cluster   at  CSCS,   based   on  Intel's
-previous-generation  Sandy  Bridge  processors,  conventional  in  HPC
-systems and known to be  more power consuming.  Finally a full tracing
-experiment is  conducted on the  ``Tintorrum'' cluster at the  Jaume I
-University (UJI) using an  integrated framework, to capture an overall
-power profile at a much finer resolution and to get more insights into
-the MPI blocking and polling influences on power savings.
+Benchmarks presented  in this work  were performed on  three different
+platforms  and can  be reproduced  within an  expected  variance, thus
+allowing a  fair comparison between the baseline  and future milestone
+versions of COSMO-ART.
 
-\subsubsection{CSCS - Monch}
-Monch is  a 10 rack  NEC-provided and dual-socket Intel  Ivy Bridge-EP
-based  cluster, utilised  by  scientists  who are  part  of the  Swiss
-Platform for Advanced Scientific  Computing (PASC) \citep{PASC}. It is
-composed of 312 standard  compute nodes, 24 large-memory compute nodes
-and  24  huge-memory  compute   nodes.   Each  standard  compute  node
-comprises two  Intel Ivy Bridge Efficient Performance  (EP) E5-2660 v2
-ten-core processors operating at  2.2 GHz base clock speed, themselves
-connected by a high speed  InfiniBand network based on Mellanox SX6036
-managed  FDR switches,  with a  56 Gb/s  speed.  Each  core has  32 KB
-instruction and 32  KB data L1 caches  and 256 KB of L2  cache. All 10
-cores share a  25 MB L3 cache  and the platform has 32GB  of DDR3 1600
-MHz RAM.  For our energy-to-solution  benchmark, a full rack  of Monch
-constituted  of  52   standard  compute  nodes  (monchc[029-080])  was
-considered.
+\begin{itemize}
+\item \textsc{Monch (CSCS)} is a 10 rack NEC-provided cluster composed
+  of 312 compute  nodes and utilised by members  of the Swiss Platform
+  for Advanced Scientific  Computing (PASC) project \citep{PASC}. Each
+  node comprises  two Intel Ivy Bridge EP  E5-2660 ten-core processors
+  operating at  2.2 GHz connected  via InfiniBand Mellanox  SX6036 and
+  FDR switches (56 Gb/s speed), and is equipped with 32GB of DDR3 1600
+  MHz  RAM.   For our  experiments,  a  full  rack constituted  of  52
+  standard compute  nodes was used.  \textsc{Monch} is  slated to stay
+  in  service  without  hardware  upgrade  for  the  duration  of  the
+  Exa2Green  project,  allowing   an  identical  configuration  to  be
+  recreated for future assessments of the baseline.\\
 
-\subsubsection{CSCS - Pilatus} 
-Pilatus  is  a dual-socket  eight-core  Intel  Sandy  Bridge EP  based
-cluster used as Piz Daint pre-post processing cluster.  It is composed
-of 42 compute  nodes and has 2 high-speed  interconnects based on FDR:
-the  first is  dedicated to  the  MPI traffic  and the  second to  the
-storage high  speed traffic.   The 2 login  nodes and the  42 computes
-nodes  consists in  11  twin-pair Intel  E5-Series  DALCO r2264i4t  2U
-scalable compute modules.  Each  module contains 4 compute nodes based
-on two Intel  Xeon E5-2670 processors operating at  2.6 GHz base clock
-speed, themselves  connected by a high speed  InfiniBand network based
-on Mellanox  SX6036 managed FDR switches,  with a 56  Gb/s speed. Each
-core has 32 KB  instruction and 32 KB data L1 caches  and 256 KB of L2
-cache.  All  the 8 core share  a 20 MB  L3 cache and the  platform has
-64GB of  DDR3 1600 MHz  RAM.  For our energy-to-solution  benchmark, a
-full  rack  of  Pilatus  constituted  of  42  standard  compute  nodes
-(pilatus[03-44]) was considered.
+\item \textsc{Pilatus (CSCS)} is  a dual-socket eight-core Intel Sandy
+  Bridge EP based cluster composed of 42 compute nodes and used as Piz
+  Daint pre-post  processing cluster at  CSCS.  The 2 login  nodes and
+  the 42 computes nodes consists in 11 twin-pair Intel E5-Series DALCO
+  r2264i4t 2U scalable compute modules. Each module contains 4 compute
+  nodes based  on two Intel  Xeon E5-2670 processors operating  at 2.6
+  GHz connected  via InfiniBand Mellanox  SX6036 and FDR  switches (56
+  Gb/s speed), and is equipped with 64GB of DDR3 1600 MHz RAM. For our
+  experiments, a  full rack constituted  of 42 standard  compute nodes
+  was  used. \textsc{Pilatus}  is based  on  Intel's second-generation
+  processors, conventional in  HPC systems and known to  be more power
+  consuming than Ivy Bridge successor.\\
 
-\subsubsection{UJI - Tintorrum} 
-Tintorrum  is a heterogeneous  cluster composed  of 28  compute nodes.
-For our experiments only a subset of these nodes were considered. This
-set is  composed of 16  nodes, each of  which includes two  Intel Xeon
-E5645 hexa-core  processors at 2.40  GHz connected via  Infiniband QDR
-(Mellanox MTS3600 switch).  Each core  has 32 KB instruction and 32 KB
-data L1 caches and 256 KB L2 cache. The 6 cores share a 12 MB L3 cache
-and the platform has 288 GB of DDR3 1333 MHz.
+\item \textsc{Tintorrum (UJI)} is  a heterogeneous cluster composed of
+  28 compute nodes.  For our  experiments only a subset of these nodes
+  were considered.   This set is composed  of 16 nodes,  each of which
+  includes  two Intel  Xeon  E5645 hexa-core  processors  at 2.40  GHz
+  connected via  Infiniband QDR (Mellanox MTS3600  switch).  Each core
+  has  32 KB  instruction  and 32  KB data  L1  caches and  256 KB  L2
+  cache. The 6 cores  share a 12 MB L3 cache and  the platform has 288
+  GB  of DDR3 1333  MHz.  A  full tracing  experiment is  conducted on
+  \textsc{Tintorrum}, to  capture an overall  power profile at  a much
+  finer resolution, and to get more insights into the MPI blocking and
+  polling influences on power savings.
+\end{itemize}
 
@@
 \subsubsection{CSCS - E3METER metering products}
+
 Supercomputer  clusters considered  for our  experiments at  the Swiss
@@
 \subsubsection{UHAM - UJI}
-To assess the  performance and the energy efficiency  of COSMO-ART, we
-employ   a    version   of   the    integrated   framework   presented
-in~\cite{energy13} that works in  combination with Extrae and Paraver,
-which are profiling/tracing and visualization tools, respectively.
+
+For our experiments at Jaume I University (UJI), we employed a version
+of the integrated framework presented in~\cite{energy13} that works in
+combination with  Extrae and Paraver, which  are profiling/tracing and
+visualization tools, respectively.
 
@@
 %the Lustre architecture; the right depicts the tracing and profiling framework.
-To use our  approach, COSMO-ART is compiled using  the Extrae compiler
-wrappers,  which  automatically instrument  the  Fortran  code of  the
-model. Next, COSMO-ART  is run on the nodes,  thus dissipating certain
-amount  of power.   These  nodes are  connected  to power  measurement
-devices that account for the dissipated power/consumed energy and send
-the power  data to the  tracing server.  The client,  meanwhile, sends
-start/stop  primitives  in  order  to  gather  captured  data  by  the
-watt-meters onto the  tracing server, where an instance  of the \pmlib
-server is running.
+COSMO-ART  is  compiled  using  the Extrae  compiler  wrappers,  which
+automatically  instrument  the  Fortran   code  of  the  model.  Next,
+COSMO-ART  is run  on the  nodes, thus  dissipating certain  amount of
+power.  These  nodes are connected  to power measurement  devices that
+account for  the dissipated power/consumed  energy and send  the power
+data to  the tracing server.  The client,  meanwhile, sends start/stop
+primitives in  order to gather  captured data by the  watt-meters onto
+the tracing server, where an instance of the \pmlib server is running.
 
-Once COSMO-ART run is finished, a file containing the power profile is
+Once  the run  is finished,  a file  containing the  power  profile is
 created using the power data  received from the tracing server and the
@@
 compilers, tools  and applications.  For our  initial benchmarking, we
-opted for the GNU compiler  (gcc/4.8.1 on Monch, gcc/4.8.2 on Pilatus)
-using the -O3  compiler flag in favor of  the intel compiler (14.0.1),
-delivering inferior performance.  In addition, we installed the MPICH2
-implementation of MPI (mvapich2/1.9) as well as the commonly used HDF5
-(1.8.12) and NetCDF (4.3.1)  libraries for the management of extremely
-large  and  complex data  collections.   All  computes  nodes have  an
-operating      system      based      on      GNU/Linux      featuring
-``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
-``3.0.101-0.15-default'' kernel in Pilatus.
+opted for the GNU  compiler (gcc/4.8.1 on \textsc{Monch}, gcc/4.8.2 on
+\textsc{Pilatus}) using  the -O3 compiler  flag in favor of  the intel
+compiler (14.0.1),  delivering inferior performance.   In addition, we
+installed the  MPICH2 implementation of MPI (mvapich2/1.9)  as well as
+the commonly used  HDF5 (1.8.12) and NetCDF (4.3.1)  libraries for the
+management  of  extremely large  and  complex  data collections.   All
+computes nodes  have an operating system based  on GNU/Linux featuring
+``2.6.32-358.11.1.el6.x86\_64''    kernel   in    \textsc{Monch}   and
+``3.0.101-0.15-default'' kernel in \textsc{Pilatus}.
 
-\subsubsection{Run configuration}
-A snapshot of the code, which includes, at least conceptually, all the
+A snapshot of the code,  which includes at least conceptually, all the
 information needed  to reproduce the  energy-to-solution benchmarks of
-COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
-node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
-Pilatus.   The  calculated  region  was mapped  to  the  participating
-processors using  a 2D-partitioning strategy.   The distribution along
-the $x$  and $y$ coordinates  was defined by setting:  $nprocx=40$ and
-$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
-usually  kept  bigger than  $nprocy$.   Besides,  as  this version  of
-COSMO-ART  doesn't  make  use  of  the traditional  GRIB  library,  we
-specified $nprocio=0$ for GRIB  I/O.  Hyperthreading is not considered
-in this study as previous attempts  of its use revealed that it always
-led to higher energy-to-solution.\\
+COSMO-ART, was produced and run on a 1040 cores using 20~MPI tasks per
+node on \textsc{Monch} and on a 1344 cores using 16~MPI tasks per node
+on  \textsc{Pilatus}.   The  calculated   region  was  mapped  to  the
+participating  processors  using   a  2D-partitioning  strategy.   The
+distribution along the $x$ and $y$ coordinates was defined by setting:
+$nprocx=40$  and $nprocy=26$  for \textsc{Monch}  and  $nprocx=28$ and
+$nprocy=24$  as  $nprocx$  is   usually  kept  bigger  than  $nprocy$.
+Besides,  as this  version doesn't  make use  of the  traditional GRIB
+library, we  specified $nprocio=0$  for GRIB I/O.   Hyperthreading was
+not considered in this study  as it previously revealed that it always
+led to higher energy-to-solution.

### 2014-04-25 trunk/paper_enahpc14/Sec2:COSMO-ART.tex
cancels previous commits of Cristiano as IBM Zurich is actually not going to contribute to this work
inserts related work

--- 2014-04-19 paper_enahpc14/Sec2:COSMO-ART.tex
@@
 \label{subsec:1.1}
-The  model  system COSMO-ART  described  in  \citep{Vogel-2009}, is  a
-regional  to  continental scale  model  coupled  online  to the  COSMO
-numerical weather  prediction and climate  model \citep{Baldauf-2011}.
-It  incorporates  sophisticated  modules  for  gaseous  chemistry  and
-aerosol dynamics  and allows the online calculation  of reactive trace
-substances and their interaction  with the atmosphere.  Detailed model
-description   can  be   found   in  \citep{Bangert-2012,   Knote-2011,
-  Knote-2013}.
+
+The model  system \textsc{Cosmo-art} described  in \citep{Vogel-2009},
+is  a  regional to  continental  scale  model  coupled online  to  the
+\textsc{Cosmo}   numerical  weather   prediction  and   climate  model
+\citep{Baldauf-2011}.   It   incorporates  sophisticated  modules  for
+gaseous  chemistry   and  aerosol  dynamics  and   allows  the  online
+calculation of  reactive trace  substances and their  interaction with
+the  atmosphere.    Detailed  model   description  can  be   found  in
+\citep{Bangert-2012, Knote-2011, Knote-2013}.
 
@@
 \label{subsec:1.2}
+
 Establishing effective energy performance benchmarking of a code under
-intense development such as COSMO-ART is a challenging task because of
-the  absolute necessity that  results must  be reproducible  within an
-expected variance for the duration of the Exa2Green project. To define
-a baseline,  it was necessary  to find a run-configuration  capable of
-being  recreated in  lal subsequent  versions.  Here  we give  a brief
-overview of the model setup  for our performance and energy efficiency
-evaluation.
+intense development  such as \textsc{Cosmo-art} is  a challenging task
+because of  the absolute necessity  that results must  be reproducible
+within  an  expected  variance  for  the  duration  of  the  Exa2Green
+project.  To   define  a  baseline,   it  was  necessary  to   find  a
+run-configuration  capable  of   being  recreated  in  all  subsequent
+versions.  Here  we give a brief  overview of the model  setup for our
+performance and energy efficiency evaluation.
 
@@
 Boundary data  for gas-phase species are taken  from IFS-MOZART output
-at 6h temporal resolution. The  model setup incorporates 34~2-D and
-45~3-D  fields  to be  written  out  every hour  and  is  devoid of  data
+at  6h temporal resolution.   The model  setup incorporates  34~2D and
+45~3D  fields to  be written  out  every hour  and is  devoid of  data
 assimilation methods.
 
-The  considered  COSMO-ART model  system  is  configured  with a  semi
-Lagrangian horizontal advection scheme with tricubic interpolation and
-selective filling  diffusion option in combination  with the dynamical
-core  using Runge-Kutta  time  stepping \citep{COSMO-PartI-2011}.   It
-also  makes use  of  the  Kinetic PreProcessor  solver  (KPP) for  the
-resolution  of atmospheric  chemistry ordinary  differential equations
-\citep{Damian-2002}.   Concerning the modelling  of wet  deposition in
-aerosols, the  baseline has only indirect cloud  feedbacks but does not
-include  in-cloud  scavenging  (rainout)  and  below-cloud  scavenging
-(washout)  yet.   Amongst  physical  parameterisations,  precipitation
-formation is performed by a two-moment cloud microphysics.
+The considered  \textsc{Cosmo-art} model  system is configured  with a
+semi-Lagrangian    horizontal   advection    scheme    with   tricubic
+interpolation  and selective filling  diffusion option  in combination
+with   the   dynamical    core   using   Runge-Kutta   time   stepping
+\citep{COSMO-PartI-2011}.    It  also   makes  use   of   the  Kinetic
+PreProcessor solver (KPP) for  the resolution of atmospheric chemistry
+ordinary  differential equations \citep{Damian-2002}.   Concerning the
+modelling  of  wet  deposition  in  aerosols, the  baseline  has  only
+indirect  cloud feedbacks  but  does not  include in-cloud  scavenging
+(rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
+parameterisations,   precipitation  formation   is   performed  by   a
+two-moment cloud microphysics.

### 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
inserts related work

--- 2014-04-24 paper_enahpc14/Sec4:Experimental-results.tex
@@
 We  approach  the  assessment  of  the energy  footprint  and  overall
-performance    of    COSMO-ART    with    two    important    metrics:
+performance   of  \textsc{Cosmo-art}   with  two   important  metrics:
 \textit{time-to-solution} (TTS) and \textit{energy-to-solution} (ETS).
@@
 
-\subsection{Time-energy analysis of COSMO-ART}
+\subsection{Time-energy analysis of \textsc{Cosmo-art}}
 \label{subsec:4.2}
@@
 
-\subsection{Power-performance tracing of COSMO-ART}
+\subsection{Power-performance tracing of \textsc{Cosmo-art}}
 \label{subsec:4.3}

### 2014-04-25 paper_enahpc14/Introduction.tex
inserts related work

@@
 scale model  system is coupled  with the operational  weather forecast
-model COSMO \citep{Baldauf-2011}, jointly developed by a consortium of
-European weather  centers and used in the  climate version (COSMO-CLM)
+model  \textsc{Cosmo}  \citep{Baldauf-2011},  jointly developed  by  a
+consortium of European weather centers and used in the climate version
 by a wide research community.
 
-The extended atmospheric model  COSMO-ART is computationally much more
-demanding than  COSMO since a  large number of additional  tracers and
-processes  have  to  be  considered.   Thus  the  model  is  currently
-severely-limited in  terms of applicability and expensive  in terms of
-energy  consumption.   Although  COSMO  has recently  been  ported  to
-GPUs  \citep{Gysi-2014, Lapillonne-2014} within  the framework  of the
-High    Performance   and    High   Productivity    Computing   (HP2C)
-Initiative \citep{HP2C},  to optimise it for  computational and energy
-efficiency, significant investments in  ART are still required to take
-it to a similar level.  The efficiency of COSMO-ART is being addressed
-in the  EU Exa2Green project \citep{EXA2GREEN} to  deliver a prototype
-code, which  provides an energy efficiency  of at least  five times of
-the baseline value.  Such  an implementation would allow the community
-to investigate critical questions at higher resolution and over longer
-periods, at reduced cost to the environment.
+The extended  atmospheric model \textsc{Cosmo-art}  is computationally
+much  more  demanding than  \textsc{Cosmo}  since  a  large number  of
+additional  tracers and  processes have  to be  considered.   Thus the
+model  is currently  severely-limited  in terms  of applicability  and
+expensive in terms of energy consumption.  Although \textsc{Cosmo} has
+recently been ported to GPUs \citep{Gysi-2014, Lapillonne-2014} within
+the framework of the  High Performance and High Productivity Computing
+(HP2C) Initiative  \citep{HP2C}, to optimise it  for computational and
+energy efficiency,  significant investments in ART  are still required
+to take it  to a similar level.  The  efficiency of \textsc{Cosmo-art}
+is being  addressed in the  EU Exa2Green project  \citep{EXA2GREEN} to
+deliver a  prototype code, which  provides an energy efficiency  of at
+least five times of the  baseline value.  Such an implementation would
+allow  the  community  to  investigate critical  questions  at  higher
+resolution  and   over  longer  periods,   at  reduced  cost   to  the
+environment.
 
@@
 overview  of  related  work,   then  in  Sec.~\ref{sec:2}  we  briefly
-introduce  COSMO-ART and specify  its technical  setup related  to the
-investigated   performance   and   energy  evaluation   methods.    In
+introduce \textsc{Cosmo-art}  and specify its  technical setup related
+to  the investigated  performance and  energy evaluation  methods.  In
 Sec.~\ref{sec:3}   we  describe   HPC  platforms,   power  measurement

### 2014-04-25 paper_enahpc14/Sec1:Related.tex
inserts related work

@@
-\begin{itemize}
-\item Application centric energy-efficiency study of distributed
-  multi-core and hybrid {CPU-GPU} systems \citep{Cumming-2014}
-\item Stella: A domain-specific language and tool for structured grid
-  methods \citep{Gysi-2014}
-\item Using compiler directives to port large scientific applications
-  to GPUs: An example from atmospherci science \citep{Lapillonne-2014}
-\item An analysis of energy-optimized lattice-Boltzmann CFD
-  simulations from the chip to the highly parallel level
-  \citep{Wittmann-2013}
-\item Energy efficiency vs. performance of the numerical solution of
-  PDEs: An application study on a low-power ARM-based cluster
-  \citep{Goddeke-2013}
-\item Analysis of Computing and Energy Performance of Multicore, NUMA,
-  and Manycore Platforms for an Irregular Application
-  \citep{Castro-2013}
-\item Exploring performance and power properties of modern multi-core
-  chips via simple machine models \citep{Hager-2013}
-\end{itemize}
+While the  TOP500 list was  introduced over 20  years ago to  rank the
+performance of HPC systems  worldwide, it's quite recently that energy
+efficiency  become  a  critical  constraint  in the  way  to  exascale
+computing.   Since 2007,  the Green500  power  measurement methodology
+encourages the design,  procurement and management of energy-efficient
+infrastuctures to contain performance in an affordable and competitive
+power  envelope.   However,  the state-of-the-art  research  assessing
+performance  and energy  efficiency  of applications  is still  pretty
+scarce.
+
+\emph{Padoin et al.}   \citep{Padoin-2013} investigate performance and
+power  consumption  of  an  agroforestry  application  and  show  that
+changing workload  can drastically improve energy efficiency  of CPU +
+GPU heterogeneous architectures.
+
+\emph{Ou  Pang et  al.}   \citep{Ou-2012} compare  ARM  and Intel  x86
+workstations  clusters  and   conclude  that  ARM-based  clusters  are
+advantageous  with   lightweight  applications  in   terms  of  energy
+efficiency.
+
+\emph{G\"oddeke et al.}  \citep{Goddeke-2013} evaluate weak and strong
+scalability of PDE solvers on a cluster of 96 ARM dual-core processors
+and demonstrate  that the ARM-based  cluster can be more  efficient in
+terms of energy-to-solution compared to x86-based cluster.
+
+\emph{Wittmann  et  al.}   \citep{Wittmann-2013}  perform  a  thorough
+analysis of  a lattice-Boltzmann method based CFD  simulation on Intel
+Sandy   Bridge  processors   and  show   extrapolated  results   on  a
+petascale-class machine.
+
+\emph{Cumming  et  al.}   \citep{Cumming-2014}  present a  simple  and
+practical  methodology  looking at  energy  minimisation  that can  be
+applied to various applications.

### 2014-04-25 paper_enahpc14/Sec3:Environment-setup.tex
inserts related work

@@
 
-We have  chosen a state-of-the-art Intel's  third-generation Core (aka
-Ivy  Bridge) processing platform  at CSCS  (called ``Monch'')  for our
-power  measurements,  which  is  slated  to stay  in  service  without
-hardware  upgrade  for the  duration  of  the  Exa2Green project.   In
-principle at least,  this architecture could be recreated  or found in
-an identical configuration beyond  the lifetime of the project.  Since
-the baseline benchmark can  be reproduced within an expected variance,
-and  that the baseline  run configuration  can be  used in  all future
-versions  of the  code, a  fair comparison  will be  made  between the
-baseline and  the milestone versions  of COSMO-ART.  In this  study, a
-complementary  energy-to-solution benchmarking  comparison  is carried
-out   on  the   ``Pilatus''  cluster   at  CSCS,   based   on  Intel's
-previous-generation  Sandy  Bridge  processors,  conventional  in  HPC
-systems and known to be  more power consuming.  Finally a full tracing
-experiment is  conducted on the  ``Tintorrum'' cluster at the  Jaume I
-University (UJI) using an  integrated framework, to capture an overall
-power profile at a much finer resolution and to get more insights into
-the MPI blocking and polling influences on power savings.
+Benchmarks presented  in this work  were performed on  three different
+platforms  and can  be reproduced  within an  expected  variance, thus
+allowing a  fair comparison between the baseline  and future milestone
+versions of \textsc{Cosmo-art}.
 
-\subsubsection{CSCS - Monch}
-Monch is  a 10 rack  NEC-provided and dual-socket Intel  Ivy Bridge-EP
-based  cluster, utilised  by  scientists  who are  part  of the  Swiss
-Platform     for      Advanced     Scientific     Computing     (PASC,
-\url{http://www.pasc-ch.org/}). It is composed of 312 standard compute
-nodes, 24 large-memory compute nodes and 24 huge-memory compute nodes.
-Each standard  compute node comprises  two Intel Ivy  Bridge Efficient
-Performance (EP)  E5-2660 v2 ten-core processors operating  at 2.2 GHz
-base  clock speed,  themselves connected  by a  high  speed InfiniBand
-network based on Mellanox SX6036  managed FDR switches, with a 56 Gb/s
-speed.  Each core  has 32 KB instruction and 32 KB  data L1 caches and
-256  KB of L2  cache. All  10 cores  share a  25 MB  L3 cache  and the
-platform has  32GB of  DDR3 1600 MHz  RAM. For  our energy-to-solution
-benchmark, a  full rack  of Monch constituted  of 52  standard compute
-nodes (monchc[029-080]) was considered.
+\begin{itemize}
+\item \textsc{Monch (CSCS)} is a 10 rack NEC-provided cluster composed
+  of 312 compute  nodes and utilised by members  of the Swiss Platform
+  for Advanced Scientific  Computing (PASC) project \citep{PASC}. Each
+  node comprises  two Intel Ivy Bridge EP  E5-2660 ten-core processors
+  operating at  2.2 GHz connected  via InfiniBand Mellanox  SX6036 and
+  FDR switches (56 Gb/s speed), and is equipped with 32GB of DDR3 1600
+  MHz  RAM.   For our  experiments,  a  full  rack constituted  of  52
+  standard compute  nodes was used.  \textsc{Monch} is  slated to stay
+  in  service  without  hardware  upgrade  for  the  duration  of  the
+  Exa2Green  project,  allowing   an  identical  configuration  to  be
+  recreated for future assessments of the baseline.\\
 
-\subsubsection{CSCS - Pilatus} 
-Pilatus  is  a dual-socket  eight-core  Intel  Sandy  Bridge EP  based
-cluster used as Piz Daint pre-post processing cluster.  It is composed
-of 42 compute  nodes and has 2 high-speed  interconnects based on FDR:
-the  first is  dedicated to  the  MPI traffic  and the  second to  the
-storage high  speed traffic.   The 2 login  nodes and the  42 computes
-nodes  consists in  11  twin-pair Intel  E5-Series  DALCO r2264i4t  2U
-scalable compute modules.  Each  module contains 4 compute nodes based
-on two Intel  Xeon E5-2670 processors operating at  2.6 GHz base clock
-speed, themselves  connected by a high speed  InfiniBand network based
-on Mellanox  SX6036 managed FDR switches,  with a 56  Gb/s speed. Each
-core has 32 KB  instruction and 32 KB data L1 caches  and 256 KB of L2
-cache.  All  the 8 core share  a 20 MB  L3 cache and the  platform has
-64GB of  DDR3 1600 MHz  RAM.  For our energy-to-solution  benchmark, a
-full  rack  of  Pilatus  constituted  of  42  standard  compute  nodes
-(pilatus[03-44]) was considered.
+\item \textsc{Pilatus (CSCS)} is  a dual-socket eight-core Intel Sandy
+  Bridge EP based cluster composed of 42 compute nodes and used as Piz
+  Daint pre-post  processing cluster at  CSCS.  The 2 login  nodes and
+  the 42 computes nodes consists in 11 twin-pair Intel E5-Series DALCO
+  r2264i4t 2U scalable compute modules. Each module contains 4 compute
+  nodes based  on two Intel  Xeon E5-2670 processors operating  at 2.6
+  GHz connected  via InfiniBand Mellanox  SX6036 and FDR  switches (56
+  Gb/s speed), and is equipped with 64GB of DDR3 1600 MHz RAM. For our
+  experiments, a  full rack constituted  of 42 standard  compute nodes
+  was  used. \textsc{Pilatus}  is based  on  Intel's second-generation
+  processors, conventional in  HPC systems and known to  be more power
+  consuming than Ivy Bridge successor.\\
 
-\subsubsection{UJI - Tintorrum} 
-Tintorrum  is a heterogeneous  cluster composed  of 28  compute nodes.
-For our experiments only a subset of these nodes were considered. This
-set is  composed of 16  nodes, each of  which includes two  Intel Xeon
-E5645 hexa-core  processors at 2.40  GHz connected via  Infiniband QDR
-(Mellanox MTS3600 switch).  Each core  has 32 KB instruction and 32 KB
-data L1 caches and 256 KB L2 cache. The 6 cores share a 12 MB L3 cache
-and the platform has 288 GB of DDR3 1333 MHz.
-
-\subsection{Software environment}
-\label{subsec:3.2}
-
-The COSMO-ART baseline  is a pure MPI based  Fortran~90 code currently
-running on distributed multi-core systems only.  The software stack on
-both CSCS  platforms was controlled using the  modules framework which
-gives an easy and flexible mechanism  to access to all of the provided
-compilers, tools  and applications.  For our  initial benchmarking, we
-opted for the GNU compiler  (gcc/4.8.1 on Monch, gcc/4.8.2 on Pilatus)
-using the -O3  compiler flag in favor of  the intel compiler (14.0.1),
-delivering inferior performance.  In addition, we installed the MPICH2
-implementation of MPI (mvapich2/1.9) as well as the commonly used HDF5
-(1.8.12) and NetCDF (4.3.1)  libraries for the management of extremely
-large  and  complex data  collections.   All  computes  nodes have  an
-operating      system      based      on      GNU/Linux      featuring
-``2.6.32-358.11.1.el6.x86\_64''      kernel      in     Monch      and
-``3.0.101-0.15-default'' kernel in Pilatus.
-
-\subsubsection{Run configuration}
-A snapshot of the code, which includes, at least conceptually, all the
-information needed  to reproduce the  energy-to-solution benchmarks of
-COSMO-ART, was produced and run on a 1040 cores using 20 MPI tasks per
-node on  Monch and  on a  1344 cores using  16 MPI  tasks per  node on
-Pilatus.   The  calculated  region  was mapped  to  the  participating
-processors using  a 2D-partitioning strategy.   The distribution along
-the $x$  and $y$ coordinates  was defined by setting:  $nprocx=40$ and
-$nprocy=26$ for  Monch and $nprocx=28$ and $nprocy=24$  as $nprocx$ is
-usually  kept  bigger than  $nprocy$.   Besides,  as  this version  of
-COSMO-ART  doesn't  make  use  of  the traditional  GRIB  library,  we
-specified $nprocio=0$ for GRIB  I/O.  Hyperthreading is not considered
-in this study as previous attempts  of its use revealed that it always
-led to higher energy-to-solution.\\
+\item \textsc{Tintorrum (UJI)} is  a heterogeneous cluster composed of
+  28 compute nodes.  For our  experiments only a subset of these nodes
+  were considered.   This set is composed  of 16 nodes,  each of which
+  includes  two Intel  Xeon  E5645 hexa-core  processors  at 2.40  GHz
+  connected via  Infiniband QDR (Mellanox MTS3600  switch).  Each core
+  has  32 KB  instruction  and 32  KB data  L1  caches and  256 KB  L2
+  cache. The 6 cores  share a 12 MB L3 cache and  the platform has 288
+  GB  of DDR3 1333  MHz.  A  full tracing  experiment is  conducted on
+  \textsc{Tintorrum}, to  capture an overall  power profile at  a much
+  finer resolution, and to get more insights into the MPI blocking and
+  polling influences on power savings.
+\end{itemize}
 
@@
 high   quality   electricity  meters   released   by  Riedo   Networks
-(\url{http://riedonetworks.com/}),  that  enable  to monitor  and  log
-power  consumption of  the  IT infrastructure  as  well as  constantly
-analyze  line  voltage,  current,  power-factor,  frequency  with  1\%
-accuracy.   Using reliable  narrowband  powerline communication  (PLC)
-technology,  all metering  and power  quality data  from each  IPS are
-centrally collected by the E3METER Data Concentrator, via the existing
-power cables thus  avoiding the need for extra  cabling.  This data is
-made  available  via SNMP,  HTTP,  TELNET  through  the built-in  Fast
-Ethernet  port.   Time  synchronisation  is guaranteed  by  using  NTP
-servers.   Measured data  is accessed  through the  open  source Cacti
-software including  the E3METER  Cacti Plugin to  scan the  entire PLC
-network and monitor  in real-time the power usage  of individual rack,
-recorded in 5 minutes interval periods.
+\citep{Riedonetworks},   that  enable   to  monitor   and   log  power
+consumption  of the IT  infrastructure as  well as  constantly analyze
+line  voltage,  current, power-factor,  frequency  with 1\%  accuracy.
+Using  reliable narrowband  powerline communication  (PLC) technology,
+all  metering and  power  quality  data from  each  IPS are  centrally
+collected  by the E3METER  Data Concentrator,  via the  existing power
+cables thus  avoiding the need for  extra cabling.  This  data is made
+available via  SNMP, HTTP, TELNET  through the built-in  Fast Ethernet
+port.   Time  synchronisation  is  guaranteed by  using  NTP  servers.
+Measured  data is  accessed  through the  open  source Cacti  software
+including the E3METER Cacti Plugin  to scan the entire PLC network and
+monitor in real-time the power usage of individual rack, recorded in 5
+minutes interval periods.
 
 \subsubsection{UHAM - UJI}
-To assess the  performance and the energy efficiency  of COSMO-ART, we
-employ   a    version   of   the    integrated   framework   presented
-in~\cite{energy13} that works in  combination with Extrae and Paraver,
-which are profiling/tracing and visualization tools, respectively.
+
+For our experiments at Jaume I University (UJI), we employed a version
+of the integrated framework presented in~\cite{energy13} that works in
+combination with  Extrae and Paraver, which  are profiling/tracing and
+visualization tools, respectively.
 
@@
 %the Lustre architecture; the right depicts the tracing and profiling framework.
-To use our  approach, COSMO-ART is compiled using  the Extrae compiler
-wrappers,  which  automatically instrument  the  Fortran  code of  the
-model. Next, COSMO-ART  is run on the nodes,  thus dissipating certain
+\textsc{Cosmo-art}  is compiled  using the  Extrae  compiler wrappers,
+which automatically  instrument the Fortran code of  the model.  Next,
+\textsc{Cosmo-art}  is  run on  the  nodes,  thus dissipating  certain
 amount  of power.   These  nodes are  connected  to power  measurement
@@
 
-Once COSMO-ART run is finished, a file containing the power profile is
+Once  the run  is finished,  a file  containing the  power  profile is
 created using the power data  received from the tracing server and the
@@
 All  these files  are  combined then  in  Paraver which  allows us  to
-visualise the performance trace and the power profile of COSMO-ART all
-together.
+visualise   the   performance  trace   and   the   power  profile   of
+\textsc{Cosmo-art} all together.
 
-%Once COSMO-ART run is finished, the VampirTrace \pmlib plugin receives
+%Once \textsc{Cosmo-art} run is finished, the VampirTrace \pmlib plugin receives
 %the  power   data  from  the  tracing   server.   The  instrumentation
@@
 %plots and statistics.
+
+\subsection{Software environment}
+\label{subsec:3.2}
+
+The \textsc{Cosmo-art}  baseline is a  pure MPI based  Fortran~90 code
+currently  running  on   distributed  multi-core  systems  only.   The
+software stack on both CSCS platforms was controlled using the modules
+framework which gives an easy  and flexible mechanism to access to all
+of the  provided compilers, tools  and applications.  For  our initial
+benchmarking,   we  opted   for   the  GNU   compiler  (gcc/4.8.1   on
+\textsc{Monch}, gcc/4.8.2 on  \textsc{Pilatus}) using the -O3 compiler
+flag  in favor  of the  intel compiler  (14.0.1),  delivering inferior
+performance.  In  addition, we installed the  MPICH2 implementation of
+MPI  (mvapich2/1.9) as  well as  the commonly  used HDF5  (1.8.12) and
+NetCDF  (4.3.1) libraries for  the management  of extremely  large and
+complex data collections.  All computes nodes have an operating system
+based on GNU/Linux featuring ``2.6.32-358.11.1.el6.x86\_64'' kernel in
+\textsc{Monch}     and      ``3.0.101-0.15-default''     kernel     in
+\textsc{Pilatus}.
+
+A snapshot of the code,  which includes at least conceptually, all the
+information needed  to reproduce the  energy-to-solution benchmarks of
+\textsc{Cosmo-art}, was produced and run  on a 1040 cores using 20~MPI
+tasks  per node on  \textsc{Monch} and  on a  1344 cores  using 16~MPI
+tasks per node on  \textsc{Pilatus}.  The calculated region was mapped
+to the participating processors using a 2D-partitioning strategy.  The
+distribution along the $x$ and $y$ coordinates was defined by setting:
+$nprocx=40$  and $nprocy=26$  for \textsc{Monch}  and  $nprocx=28$ and
+$nprocy=24$  as  $nprocx$  is   usually  kept  bigger  than  $nprocy$.
+Besides,  as this  version doesn't  make use  of the  traditional GRIB
+library, we  specified $nprocio=0$  for GRIB I/O.   Hyperthreading was
+not considered in this study  as it previously revealed that it always
+led to higher energy-to-solution.

### 2014-04-25 paper_enahpc14/paper.tex
inserts related work

@@
   In this  paper we investigate  the energy footprint  and performance
-  profiling of COSMO-ART  on various HPC platforms.  This  model is an
-  extension of  the operational weather  forecast model of  the German
-  weather  service   (DWD),  developed  for  the   evaluation  of  the
-  interactions of reactive gases  and aerosol particles with the state
-  of atmosphere at the  regional scale.  Different measurement devices
-  and energy-aware techniques are  described to evaluate both time and
-  energy  to  solution  of  the  considered application  and  to  gain
-  detailed  insights  into power  and  performance requirements.   Our
+  profiling  of  \textsc{COSMO-ART} on  various  HPC platforms.   This
+  model is an  extension of the operational weather  forecast model of
+  the German  weather service (DWD),  developed for the  evaluation of
+  the interactions  of reactive gases  and aerosol particles  with the
+  state of  atmosphere at  the regional scale.   Different measurement
+  devices and  energy-aware techniques are described  to evaluate both
+  time and  energy to  solution of the  considered application  and to
+  gain detailed insights into power and performance requirements.  Our
   motivation  is to  improve  corresponding code  sections to  sustain
@@
   chemistry  \and  Aerosols  modelling  \and  Profiling  methods  \and
-  Benchmark    analysis   \and    COSMO-ART    coupled   model    \and
+  Benchmark  analysis  \and   \textsc{COSMO-ART}  coupled  model  \and
   Time-to-solution \and Energy-to-solution \and Xeon processor}
@@
 
-\section{The COSMO-ART model system}
+\section{The \textsc{COSMO-ART} model system}
 \label{sec:2}

### 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
modifies ETS and TTS figures for MPI blocking and polling modes

@@
 \begin{figure}[htbf]
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
-  \caption{Time-to-solution and energy-to-solution on
-    \textsc{Tintorrum} - Polling mode}
+  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-5.eps}
+  \caption{Mean time-to-solution and energy-to-solution on
+    \textsc{Tintorrum} - MPI Polling and Blocking modes}
   \label{fig:4}
-\end{figure}
-
-\begin{figure}[htbf]
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-2.eps}
-  \caption{Time-to-solution and energy-to-solution on
-    \textsc{Tintorrum} - Polling mode}
-  \label{fig:5}
-\end{figure}
-
-\begin{figure}[htbf]
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-3.eps}
-  \caption{Time-to-solution and Power on \textsc{Tintorrum} - Polling
-    mode}
-  \label{fig:6}
-\end{figure}
-
-\begin{figure}[htbf]
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-4.eps}
-  \caption{Time-to-solution and Power on \textsc{Tintorrum} - Blocking
-    mode}
-  \label{fig:7}
 \end{figure}

### 2014-04-25 paper_enahpc14/Conclusion.tex
adds conclusion; modifies table

We  have   presented  a  methodology  for   comparing  performance  of
COSMO-ART,  a  regional  weather  forecast  model  augmented  for  the
interactions of  reactive gases and aerosol  particles.  The resulting
benchmarks illustrate  that the  best time-to-solution does  not imply
the best energy-to-solution: an Intel Sandybridge (2.6 GHz) system has
lower  time-to-solution but  higher energy-to-solution  than  an Intel
Ivybridge  (2.2  GHz) system,  although  the  two  metrics are  indeed
strongly  correlated.   A   tool  utilising  Paraver/Extrae  with  new
extensions  for power  consumption  was successfully  applied to  this
non-trivial application.  The  resulting profiles indicate that simple
changes, such as making use of an MPI version based on blocking rather
than polling, can reduce power consumption significantly.

This  reproducible  benchmark provides  a  baseline  for ongoing  work
package within  the EU-funded Exa2Green to  minimise power consumption
of COSMO-ART. Profiling  has given us insight into  the most expensive
code  components,  which are  now  being  altered  to utilise  revised
algorithms.  The  results of these  optimisations will be  reported in
future publications.

### 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
adds conclusion; modifies table

@@
     \label{tab:3}
-    \begin{tabular}{ccc}
+    \begin{tabular}{cccc}
       \hline\noalign{\smallskip}
-      \textbf{Simulation time} & \textbf{Xeon E5} & \textbf{Ivy Bridge EP} \\
+      \textbf{\scriptsize{Xeon E5}} & \textbf{\scriptsize{Ivy Bridge}} & \textbf{\scriptsize{Xeon  E5645}} & \textbf{\scriptsize{Xeon  E5645}}\\
+      & & \textbf{\scriptsize{(Polling)}} & \textbf{\scriptsize{(Blocking)}} \\
       \noalign{\smallskip}\hline\noalign{\smallskip}
-      \textbf{1 day} & 18122.01417 & 12658.52278 \\ 
-      & 17979.61083 & 12586.40833 \\
-      & 18065.45167 & - \\
-      & 17973.02833 & - \\
-      \noalign{\smallskip}\hline\noalign{\smallskip}
-      \textbf{2 days} & 17997.57815 & - \\
+      18035.0 & 12622.5 & 3713.6 & 3651.8 \\ 
       \noalign{\smallskip}\hline
@@
   \caption{Time-to-solution and energy-to-solution comparison between
-    Xeon E5 and Ivy Bridge-EP architectures}
+    Xeon E5 and Ivy Bridge-EP architectures for a 24h simulation}
   \label{fig:3}
@@
 \begin{figure}[htbf]
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
-  \caption{Time-to-solution and energy-to-solution on
-    \textsc{Tintorrum} - Polling mode}
+  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-5.eps}
+  \caption{Mean time-to-solution and energy-to-solution on
+    \textsc{Tintorrum} for a 24h simulation - MPI Polling and Blocking
+    modes}
   \label{fig:4}
@@
 
-\begin{figure}[htbf]
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-2.eps}
-  \caption{Time-to-solution and energy-to-solution on
-    \textsc{Tintorrum} - Polling mode}
-  \label{fig:5}
-\end{figure}
-
-\begin{figure}[htbf]
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-3.eps}
-  \caption{Time-to-solution and Power on \textsc{Tintorrum} - Polling
-    mode}
-  \label{fig:6}
-\end{figure}
-
-\begin{figure}[htbf]
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-4.eps}
-  \caption{Time-to-solution and Power on \textsc{Tintorrum} - Blocking
-    mode}
-  \label{fig:7}
-\end{figure}
-
-\subsection{Power-performance tracing of COSMO-ART}
+\subsection{Power-performance tracing of \textsc{Cosmo-art}}
 \label{subsec:4.3}

### 2014-04-25 trunk/paper_enahpc14/paper.tex
Manuel's changes
updates paper.tex

@@
 \title{Evaluating the Performance and Energy Efficiency of the
-  COSMO-ART model system}
+  \textsc{COSMO-ART} model system}
 %\thanks{}

### 2014-04-25 trunk/paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

@@
 allowing a  fair comparison between the baseline  and future milestone
-versions of COSMO-ART.
+versions of \textsc{Cosmo-art}.
 
 \begin{itemize}
-\item \textsc{Monch (CSCS)} is a 10 rack NEC-provided cluster composed
-  of 312 compute  nodes and utilised by members  of the Swiss Platform
-  for Advanced Scientific  Computing (PASC) project \citep{PASC}. Each
+\item \textsc{Monch} is a 10 rack NEC-provided cluster composed
+  of 312 compute  nodes and utilized by members  of the Swiss Platform
+  for Advanced Scientific  Computing (PASC) project~\citep{PASC} at CSCS. Each
   node comprises two Intel Ivy Bridge EP  E5-2660 ten-core processors
-  operating at  2.2 GHz connected  via InfiniBand Mellanox  SX6036 and
-  FDR switches (56 Gb/s speed), and is equipped with 32GB of DDR3 1600
-  MHz  RAM.   For our  experiments,  a  full  rack constituted  of  52
+  operating at  2.2 GHz, equipped with 32\,GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
+  a full rack constituted  of  52
   standard compute  nodes was used.  \textsc{Monch} is  slated to stay
@@
 
-\item \textsc{Pilatus (CSCS)} is  a dual-socket eight-core Intel Sandy
+\item \textsc{Pilatus} is a dual-socket eight-core Intel Sandy
   Bridge EP based cluster composed of 42 compute nodes and used as Piz
@@
   r2264i4t 2U scalable compute modules. Each module contains 4 compute
-  nodes based  on two Intel  Xeon E5-2670 processors operating  at 2.6
-  GHz connected  via InfiniBand Mellanox  SX6036 and FDR  switches (56
-  Gb/s speed), and is equipped with 64GB of DDR3 1600 MHz RAM. For our
+  nodes based  on two Intel Xeon E5-2670 processors operating  at 2.6\,GHz 
+  connected  via InfiniBand Mellanox SX6036 and FDR  switches (56
+  Gb/s speed), and is equipped with 64GB of DDR3 1600\,MHz RAM. For our
   experiments, a  full rack constituted  of 42 standard  compute nodes
@@
 
-\item \textsc{Tintorrum (UJI)} is  a heterogeneous cluster composed of
-  28 compute nodes.  For our  experiments only a subset of these nodes
-  were considered.   This set is composed  of 16 nodes,  each of which
-  includes  two Intel  Xeon  E5645 hexa-core  processors  at 2.40  GHz
-  connected via  Infiniband QDR (Mellanox MTS3600  switch).  Each core
-  has  32 KB  instruction  and 32  KB data  L1  caches and  256 KB  L2
-  cache. The 6 cores  share a 12 MB L3 cache and  the platform has 288
-  GB  of DDR3 1333  MHz.  A  full tracing  experiment is  conducted on
-  \textsc{Tintorrum}, to  capture an overall  power profile at  a much
-  finer resolution, and to get more insights into the MPI blocking and
-  polling influences on power savings.
+\item \textsc{Tintorrum} is  a heterogeneous cluster composed of
+  28 compute nodes at UJI. For our  experiments only a subset of 16 nodes was considered,  
+  Each of nodes of this subset includes a dual-socket hexa-core Intel Xeon E5645 processors 
+  running at 2.40\,GHz, equipped with 288\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS3600 switch. A  full tracing  experiment is  conducted on
+  \textsc{Tintorrum} in order to  capture an overall  power profile at a much
+  finer resolution. Also we analize the contribution of the MPI library
+  to the energy consumption and how the use of the blocking versus polling
+  message-passing policies can potentially render energy savings.
 \end{itemize}
 
-\subsection{Power-performance measurement framework}
+\subsection{Power measurement framework}
 \label{subsec:3.3}
 
-We present  here two different  frameworks deployed on HPC  systems to
-measure  the  power  consumption   and  performance  of  the  baseline
-execution.
+We present here the two different  frameworks deployed on the testbed HPC systems to
+measure  performance and  power  consumption of the baseline execution of COSMO-ART
 
-\subsubsection{CSCS - E3METER metering products}
+\subsubsection{E3METER metering framework}
 
-Supercomputer  clusters considered  for our  experiments at  the Swiss
-National Supercomputing Center (CSCS)  of ETH Zurich are equipped with
+Supercomputer  clusters considered  for our  experiments at CSCS of ETH Zurich are equipped with
 E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
@@
 consumption  of the IT  infrastructure as  well as  constantly analyze
-line  voltage,  current, power-factor,  frequency  with 1\%  accuracy.
+line  voltage,  current, power-factor,  frequency  with 1\,\%  accuracy.
 Using  reliable narrowband  powerline communication  (PLC) technology,
@@
 
-\subsubsection{UHAM - UJI}
+\subsubsection{Power-performance measurement framework}
 
 For our experiments at Jaume I University (UJI), we employed a version
-of the integrated framework presented in~\cite{energy13} that works in
+of the integrated framework presented in~\cite{energy13} that employs the \pmlib library in
 combination with Extrae and Paraver, which  are profiling/tracing and
@@
 
-%The left part of the \vref{fig:Lustre} offers a graphical representation of
-%the Lustre architecture; the right depicts the tracing and profiling framework.
-COSMO-ART  is  compiled  using  the Extrae  compiler  wrappers,  which
-automatically  instrument  the  Fortran   code  of  the  model.  Next,
-COSMO-ART  is run  on the  nodes, thus  dissipating certain  amount of
-power.  These  nodes are connected  to power measurement  devices that
-account for  the dissipated power/consumed  energy and send  the power
-data to  the tracing server.  The client,  meanwhile, sends start/stop
-primitives in  order to gather  captured data by the  watt-meters onto
-the tracing server, where an instance of the \pmlib server is running.
+\textsc{COSMO-ART}  is compiled  using the  Extrae  compiler wrappers,
+which automatically  instrument the Fortran code of  the model.  Next,
+\textsc{COSMO-ART}  is  run on  the  nodes,  thus dissipating  certain
+amount  of power.   These  nodes are  connected  to power  measurement
+devices that account for the dissipated power/consumed energy and send
+the power  data to the  tracing server.  The client,  meanwhile, sends
+start/stop  primitives  in  order  to  gather  captured  data  by  the
+watt-meters onto the  tracing server, where an instance  of the \pmlib
+server is running.
 
@@
 All  these files  are  combined then  in  Paraver which  allows us  to
-visualise the performance trace and the power profile of COSMO-ART all
-together.
-
-%Once COSMO-ART run is finished, the VampirTrace \pmlib plugin receives
-%the  power   data  from  the  tracing   server.   The  instrumentation
-%post-processing generates  the performance trace files  and the \pmlib
-%plugin inserts the power data into them.
-
-%In  addition  to the  power  measurements,  we  also account  for  the
-%resource utilization values  of the nodes: CPU load,  memory usage and
-%storage device utilization. % and network utilization.
-
-%We  run special  \pmlib  server  instances on  the  server nodes  that
-%retrieve these  values from the \texttt{proc}  file system (leveraging
-%the  \texttt{psutil} Python library).   Thus, \pmlib  plugin instances
-%running  with the  instrumented  application connect  with the  \pmlib
-%servers.   Finally,   using  the   Vampir   visualization  tool,   the
-%power-performance traces  can be easily  analyzed through a  series of
-%plots and statistics.
+visualise   the   performance  trace   and   the   power  profile   of
+\textsc{Cosmo-art} all together.
 
@@
 
-The COSMO-ART baseline  is a pure MPI based  Fortran~90 code currently
-running on distributed multi-core systems only.  The software stack on
-both CSCS  platforms was controlled using the  modules framework which
-gives an easy and flexible mechanism  to access to all of the provided
-compilers, tools  and applications.  For our  initial benchmarking, we
-opted for the GNU  compiler (gcc/4.8.1 on \textsc{Monch}, gcc/4.8.2 on
-\textsc{Pilatus}) using  the -O3 compiler  flag in favor of  the intel
-compiler (14.0.1),  delivering inferior performance.   In addition, we
-installed the  MPICH2 implementation of MPI (mvapich2/1.9)  as well as
-the commonly used  HDF5 (1.8.12) and NetCDF (4.3.1)  libraries for the
-management  of  extremely large  and  complex  data collections.   All
-computes nodes  have an operating system based  on GNU/Linux featuring
-``2.6.32-358.11.1.el6.x86\_64''    kernel   in    \textsc{Monch}   and
-``3.0.101-0.15-default'' kernel in \textsc{Pilatus}.
+The \textsc{Cosmo-art}  baseline is a  pure MPI based  Fortran~90 code
+currently  running  on   distributed  multi-core  systems  only.   The
+software stack on the platforms was controlled using the modules
+framework which gives an easy  and flexible mechanism to access to all
+of the  provided compilers, tools  and applications.  For  our initial
+benchmarking,   we  opted   for   the  GNU   compiler  gcc-4.8.1   on
+\textsc{Monch} and gcc-4.8.2 on  \textsc{Pilatus} and \textsc{Tintorrum}) using the -O3 compiler
+flag  in favor  of the  Intel compilers 14.0.1,  delivering inferior
+performance.
+In  addition, we installed the following implementations of MPI: MVAPICH2 1.9 on \textsc{Monch} and \textsc{Pilatus} and  OpenMPI 1.6.5 on \textsc{Tintorrum}.
+On the other hand, we  used HDF5 1.8.12 and NetCDF  (4.3.1) libraries for  the management  of extremely  large and
+complex data collections.  All computes nodes have an operating system
+based on GNU/Linux featuring ``2.6.32-358.11.1.el6.x86\_64'' kernel in
+\textsc{Monch}     and      ``3.0.101-0.15-default''     kernel     in
+\textsc{Pilatus}.
 
@@
 information needed  to reproduce the  energy-to-solution benchmarks of
-COSMO-ART, was produced and run on a 1040 cores using 20~MPI tasks per
-node on \textsc{Monch} and on a 1344 cores using 16~MPI tasks per node
-on  \textsc{Pilatus}.   The  calculated   region  was  mapped  to  the
-participating  processors  using   a  2D-partitioning  strategy.   The
+\textsc{COSMO-ART}, was produced and run  on a 1040 cores using 20~MPI
+tasks  per node on  \textsc{Monch} and  on a  1344 cores  using 16~MPI
+tasks per node on  \textsc{Pilatus}.  The calculated region was mapped
+to the participating processors using a 2D-partitioning strategy.  The
 distribution along the $x$ and $y$ coordinates was defined by setting:

### 2014-04-25 paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

@@
 allowing a  fair comparison between the baseline  and future milestone
-versions of \textsc{Cosmo-art}.
+versions of \textsc{COSMO-ART}.
 
 \begin{itemize}
-\item \textsc{Monch (CSCS)} is a 10 rack NEC-provided cluster composed
-  of 312 compute  nodes and utilised by members  of the Swiss Platform
-  for Advanced Scientific  Computing (PASC) project \citep{PASC}. Each
+\item \textsc{Monch} is a 10 rack NEC-provided cluster composed
+  of 312 compute  nodes %and utilized by members  of the Swiss Platform
+  %for Advanced Scientific  Computing (PASC) project~\citep{PASC}
+   at CSCS. Each
   node comprises two Intel Ivy Bridge EP  E5-2660 ten-core processors
-  operating at  2.2 GHz connected  via InfiniBand Mellanox  SX6036 and
-  FDR switches (56 Gb/s speed), and is equipped with 32GB of DDR3 1600
-  MHz  RAM.   For our  experiments,  a  full  rack constituted  of  52
-  standard compute  nodes was used.  \textsc{Monch} is  slated to stay
+  operating at  2.2 GHz, equipped with 32\,GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
+  a full rack constituted of 52 nodes was used. \textsc{Monch} is  slated to stay
   in  service  without  hardware  upgrade  for  the  duration  of  the
@@
 
-\item \textsc{Pilatus (CSCS)} is  a dual-socket eight-core Intel Sandy
-  Bridge EP based cluster composed of 42 compute nodes and used as Piz
-  Daint pre-post  processing cluster at  CSCS.  The 2 login  nodes and
+\item \textsc{Pilatus} is a dual-socket eight-core Intel Sandy
+  Bridge EP based cluster composed of 42 compute nodes %and used as Piz Daint pre-post  processing cluster 
+  at  CSCS.  The 2 login  nodes and
   the 42 computes nodes consists in 11 twin-pair Intel E5-Series DALCO
   r2264i4t 2U scalable compute modules. Each module contains 4 compute
-  nodes based  on two Intel  Xeon E5-2670 processors operating  at 2.6
-  GHz connected  via InfiniBand Mellanox  SX6036 and FDR  switches (56
-  Gb/s speed), and is equipped with 64GB of DDR3 1600 MHz RAM. For our
-  experiments, a  full rack constituted  of 42 standard  compute nodes
+  nodes based  on two Intel Xeon E5-2670 processors operating  at 2.6\,GHz equipped with 64GB of DDR3 1600\,MHz RAM and connected  via InfiniBand Mellanox SX6036 and FDR switches. 
+  For our  experiments, a  full rack constituted  of 42 standard  compute nodes
   was  used. \textsc{Pilatus}  is based  on  Intel's second-generation
@@
 
-\item \textsc{Tintorrum (UJI)} is  a heterogeneous cluster composed of
-  28 compute nodes.  For our  experiments only a subset of these nodes
-  were considered.   This set is composed  of 16 nodes,  each of which
-  includes  two Intel  Xeon  E5645 hexa-core  processors  at 2.40  GHz
-  connected via  Infiniband QDR (Mellanox MTS3600  switch).  Each core
-  has  32 KB  instruction  and 32  KB data  L1  caches and  256 KB  L2
-  cache. The 6 cores  share a 12 MB L3 cache and  the platform has 288
-  GB  of DDR3 1333  MHz.  A  full tracing  experiment is  conducted on
-  \textsc{Tintorrum}, to  capture an overall  power profile at  a much
-  finer resolution, and to get more insights into the MPI blocking and
-  polling influences on power savings.
+\item \textsc{Tintorrum} is  a heterogeneous cluster composed of
+  28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each of nodes of this subset includes a dual-socket hexa-core Intel Xeon E5645 processors 
+  running at 2.40\,GHz, equipped with 288\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS3600 switch. A  full tracing  experiment is  conducted on
+  \textsc{Tintorrum} in order to  capture an overall  power profile at a much
+  finer resolution. Also we analize the contribution of the MPI library
+  to the energy consumption and how the use of the blocking versus polling
+  message-passing policies can potentially render energy savings.
 \end{itemize}
 
-\subsection{Power-performance measurement framework}
+\subsection{Power measurement framework}
 \label{subsec:3.3}
 
-We present  here two different  frameworks deployed on HPC  systems to
-measure  the  power  consumption   and  performance  of  the  baseline
-execution.
+We present here the two different  frameworks deployed on the testbed HPC systems to
+measure  performance and  power  consumption of the baseline execution of COSMO-ART
 
-\subsubsection{CSCS - E3METER metering products}
+\subsubsection{E3METER metering framework}
 
-Supercomputer  clusters considered  for our  experiments at  the Swiss
-National Supercomputing Center (CSCS)  of ETH Zurich are equipped with
+Supercomputer  clusters considered  for our  experiments at CSCS of ETH Zurich are equipped with
 E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
@@
 
-%The left part of the \vref{fig:Lustre} offers a graphical representation of
-%the Lustre architecture; the right depicts the tracing and profiling framework.
-\textsc{Cosmo-art}  is compiled  using the  Extrae  compiler wrappers,
+\textsc{COSMO-ART}  is compiled  using the  Extrae  compiler wrappers,
 which automatically  instrument the Fortran code of  the model.  Next,
-\textsc{Cosmo-art}  is  run on  the  nodes,  thus dissipating  certain
+\textsc{COSMO-ART}  is  run on  the  nodes,  thus dissipating  certain
 amount  of power.   These  nodes are  connected  to power  measurement
@@
 
-%Once \textsc{Cosmo-art} run is finished, the VampirTrace \pmlib plugin receives
-%the  power   data  from  the  tracing   server.   The  instrumentation
-%post-processing generates  the performance trace files  and the \pmlib
-%plugin inserts the power data into them.
-
-%In  addition  to the  power  measurements,  we  also account  for  the
-%resource utilization values  of the nodes: CPU load,  memory usage and
-%storage device utilization. % and network utilization.
-
-%We  run special  \pmlib  server  instances on  the  server nodes  that
-%retrieve these  values from the \texttt{proc}  file system (leveraging
-%the  \texttt{psutil} Python library).   Thus, \pmlib  plugin instances
-%running  with the  instrumented  application connect  with the  \pmlib
-%servers.   Finally,   using  the   Vampir   visualization  tool,   the
-%power-performance traces  can be easily  analyzed through a  series of
-%plots and statistics.
-
 \subsection{Software environment}
@@
 currently  running  on   distributed  multi-core  systems  only.   The
-software stack on both CSCS platforms was controlled using the modules
+software stack on the platforms was controlled using the modules
 framework which gives an easy  and flexible mechanism to access to all
 of the  provided compilers, tools  and applications.  For  our initial
-benchmarking,   we  opted   for   the  GNU   compiler  (gcc/4.8.1   on
-\textsc{Monch}, gcc/4.8.2 on  \textsc{Pilatus}) using the -O3 compiler
-flag  in favor  of the  intel compiler  (14.0.1),  delivering inferior
-performance.  In  addition, we installed the  MPICH2 implementation of
-MPI  (mvapich2/1.9) as  well as  the commonly  used HDF5  (1.8.12) and
-NetCDF  (4.3.1) libraries for  the management  of extremely  large and
+benchmarking,   we  opted   for   the  GNU   compiler  gcc-4.8.1   on
+\textsc{Monch} and gcc-4.8.2 on  \textsc{Pilatus} and \textsc{Tintorrum}) using the -O3 compiler
+flag  in favor  of the  Intel compilers 14.0.1,  delivering inferior
+performance.
+In  addition, we installed the following implementations of MPI: MVAPICH2 1.9 on \textsc{Monch} and \textsc{Pilatus} and  OpenMPI 1.6.5 on \textsc{Tintorrum}.
+On the other hand, we  used HDF5 1.8.12 and NetCDF  (4.3.1) libraries for  the management  of extremely  large and
 complex data collections.  All computes nodes have an operating system
@@
 information needed  to reproduce the  energy-to-solution benchmarks of
-\textsc{Cosmo-art}, was produced and run  on a 1040 cores using 20~MPI
+\textsc{COSMO-ART}, was produced and run  on a 1040 cores using 20~MPI
 tasks  per node on  \textsc{Monch} and  on a  1344 cores  using 16~MPI

### 2014-04-25 paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

@@
 \item \textsc{Tintorrum} is  a heterogeneous cluster composed of
-  28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each of nodes of this subset includes a dual-socket hexa-core Intel Xeon E5645 processors 
+  28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each of them is composed dual-socket Intel Xeon E5645 hexa-core processors 
   running at 2.40\,GHz, equipped with 288\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS3600 switch. A  full tracing  experiment is  conducted on
   \textsc{Tintorrum} in order to  capture an overall  power profile at a much
-  finer resolution. Also we analize the contribution of the MPI library
+  finer resolution. In this sense, we also analyze the contribution of the MPI library
   to the energy consumption and how the use of the blocking versus polling

### 2014-04-25 paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

@@
 \textsc{COSMO-ART}  is  run on  the  nodes,  thus dissipating  certain
-amount  of power.   These  nodes are  connected  to power  measurement
-devices that account for the dissipated power/consumed energy and send
-the power  data to the  tracing server.  The client,  meanwhile, sends
-start/stop  primitives  in  order  to  gather  captured  data  by  the
-watt-meters onto the  tracing server, where an instance  of the \pmlib
-server is running.
+amount  of power.   These  nodes are connected to power distribution units (PDUs)
+APC8653 that account for the dissipated power of at 1\,Hz, these data is afterwards sent
+to the  tracing server.
+The client,  meanwhile, sends start/stop  primitives  in  order  to  gather  captured  data  by  the watt-meters onto the  tracing server, where an instance  of the \pmlib server is running.
 
@@
 visualise   the   performance  trace   and   the   power  profile   of
-\textsc{Cosmo-art} all together.
+\textsc{COSMO-ART} all together.
 
@@
 
-The \textsc{Cosmo-art}  baseline is a  pure MPI based  Fortran~90 code
+The \textsc{COSMO-ART}  baseline is a  pure MPI based  Fortran~90 code
 currently  running  on   distributed  multi-core  systems  only.   The

### 2014-04-25 trunk/paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

@@
 allowing a  fair comparison between the baseline  and future milestone
-versions of \textsc{Cosmo-art}.
+versions of \cosmoart.
 
 \begin{itemize}
-\item \textsc{Monch} is a 10 rack NEC-provided cluster composed
-  of 312 compute  nodes and utilized by members  of the Swiss Platform
-  for Advanced Scientific  Computing (PASC) project~\citep{PASC} at CSCS. Each
+\item \monch is a 10 rack NEC-provided cluster composed
+  of 312 compute  nodes %and utilized by members  of the Swiss Platform
+  %for Advanced Scientific  Computing (PASC) project~\citep{PASC}
+   at CSCS. Each
   node comprises two Intel Ivy Bridge EP  E5-2660 ten-core processors
   operating at  2.2 GHz, equipped with 32\,GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
-  a full rack constituted  of  52
-  standard compute  nodes was used.  \textsc{Monch} is  slated to stay
+  a full rack constituted of 52 nodes was used. \monch is  slated to stay
   in  service  without  hardware  upgrade  for  the  duration  of  the
@@
 
-\item \textsc{Pilatus} is a dual-socket eight-core Intel Sandy
-  Bridge EP based cluster composed of 42 compute nodes and used as Piz
-  Daint pre-post  processing cluster at  CSCS.  The 2 login  nodes and
+\item \pilat is a dual-socket eight-core Intel Sandy
+  Bridge EP based cluster composed of 42 compute nodes %and used as Piz Daint pre-post  processing cluster 
+  at  CSCS.  The 2 login  nodes and
   the 42 computes nodes consists in 11 twin-pair Intel E5-Series DALCO
   r2264i4t 2U scalable compute modules. Each module contains 4 compute
-  nodes based  on two Intel Xeon E5-2670 processors operating  at 2.6\,GHz 
-  connected  via InfiniBand Mellanox SX6036 and FDR  switches (56
-  Gb/s speed), and is equipped with 64GB of DDR3 1600\,MHz RAM. For our
-  experiments, a  full rack constituted  of 42 standard  compute nodes
-  was  used. \textsc{Pilatus}  is based  on  Intel's second-generation
+  nodes based  on two Intel Xeon E5-2670 processors operating  at 2.6\,GHz equipped with 64GB of DDR3 1600\,MHz RAM and connected  via InfiniBand Mellanox SX6036 and FDR switches. 
+  For our  experiments, a  full rack constituted  of 42 standard  compute nodes
+  was used. \pilat  is based  on  Intel's second-generation
   processors, conventional in  HPC systems and known to  be more power
@@
 
-\item \textsc{Tintorrum} is  a heterogeneous cluster composed of
-  28 compute nodes at UJI. For our  experiments only a subset of 16 nodes was considered,  
-  Each of nodes of this subset includes a dual-socket hexa-core Intel Xeon E5645 processors 
+\item \tinto is  a heterogeneous cluster composed of
+  28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each of them is composed dual-socket Intel Xeon E5645 hexa-core processors 
   running at 2.40\,GHz, equipped with 288\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS3600 switch. A  full tracing  experiment is  conducted on
-  \textsc{Tintorrum} in order to  capture an overall  power profile at a much
-  finer resolution. Also we analize the contribution of the MPI library
+  \tinto in order to  capture an overall  power profile at a much
+  finer resolution. In this sense, we also analyze the contribution of the MPI library
   to the energy consumption and how the use of the blocking versus polling
@@
 
-\subsubsection{Power-performance measurement framework}
+\subsubsection{Power-performance tracing framework}
 
@@
 
-\textsc{COSMO-ART}  is compiled  using the  Extrae  compiler wrappers,
+\cosmoart  is compiled  using the  Extrae  compiler wrappers,
 which automatically  instrument the Fortran code of  the model.  Next,
-\textsc{COSMO-ART}  is  run on  the  nodes,  thus dissipating  certain
-amount  of power.   These  nodes are  connected  to power  measurement
-devices that account for the dissipated power/consumed energy and send
-the power  data to the  tracing server.  The client,  meanwhile, sends
-start/stop  primitives  in  order  to  gather  captured  data  by  the
-watt-meters onto the  tracing server, where an instance  of the \pmlib
-server is running.
+\cosmoart  is  run on  the  nodes,  thus dissipating  certain
+amount  of power.   These  nodes are connected to power distribution units (PDUs)
+APC8653 that account for the dissipated power with a sampling frequency of 1\,Hz.
+The power data captured of these PDUs is afterwards collected by the tracing server.
+The client,  meanwhile, sends start/stop  primitives  in  order  to  gather  captured  data  by  the watt-meters onto the  tracing server, where an instance  of the \pmlib server is running.
 
@@
 visualise   the   performance  trace   and   the   power  profile   of
-\textsc{Cosmo-art} all together.
+\cosmoart all together.
 
@@
 
-The \textsc{Cosmo-art}  baseline is a  pure MPI based  Fortran~90 code
+The \cosmoart  baseline is a  pure MPI based  Fortran~90 code
 currently  running  on   distributed  multi-core  systems  only.   The
@@
 benchmarking,   we  opted   for   the  GNU   compiler  gcc-4.8.1   on
-\textsc{Monch} and gcc-4.8.2 on  \textsc{Pilatus} and \textsc{Tintorrum}) using the -O3 compiler
+\monch and gcc-4.8.2 on  \pilat and \tinto) using the -O3 compiler
 flag  in favor  of the  Intel compilers 14.0.1,  delivering inferior
 performance.
-In  addition, we installed the following implementations of MPI: MVAPICH2 1.9 on \textsc{Monch} and \textsc{Pilatus} and  OpenMPI 1.6.5 on \textsc{Tintorrum}.
+In  addition, we installed the following implementations of MPI: MVAPICH2 1.9 on \monch and \pilat and  OpenMPI 1.6.5 on \tinto.
 On the other hand, we  used HDF5 1.8.12 and NetCDF  (4.3.1) libraries for  the management  of extremely  large and
@@
 based on GNU/Linux featuring ``2.6.32-358.11.1.el6.x86\_64'' kernel in
-\textsc{Monch}     and      ``3.0.101-0.15-default''     kernel     in
-\textsc{Pilatus}.
+\monch, ``3.0.101-0.15-default'' in \pilat and ``2.6.18-238.19.1.el5.x86\_64'' in \tinto.
 
@@
 information needed  to reproduce the  energy-to-solution benchmarks of
-\textsc{COSMO-ART}, was produced and run  on a 1040 cores using 20~MPI
-tasks  per node on  \textsc{Monch} and  on a  1344 cores  using 16~MPI
-tasks per node on  \textsc{Pilatus}.  The calculated region was mapped
+\cosmoart, was produced and run  on a 1040 cores using 20~MPI
+tasks  per node on  \monch and  on a  1344 cores  using 16~MPI
+tasks per node on  \pilat.  The calculated region was mapped
 to the participating processors using a 2D-partitioning strategy.  The
 distribution along the $x$ and $y$ coordinates was defined by setting:
-$nprocx=40$  and $nprocy=26$  for \textsc{Monch}  and  $nprocx=28$ and
+$nprocx=40$  and $nprocy=26$  for \monch  and  $nprocx=28$ and
 $nprocy=24$  as  $nprocx$  is   usually  kept  bigger  than  $nprocy$.

### 2014-04-25 trunk/paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

@@
 We present here the two different  frameworks deployed on the testbed HPC systems to
-measure  performance and  power  consumption of the baseline execution of COSMO-ART
+measure  performance and  power  consumption of the baseline execution of \cosmoart
 
@@
 of the  provided compilers, tools  and applications.  For  our initial
-benchmarking,   we  opted   for   the  GNU   compiler  gcc-4.8.1   on
-\monch and gcc-4.8.2 on  \pilat and \tinto) using the -O3 compiler
+benchmarking,   we  opted   for   the  GNU   compiler  GCC 4.8.1   on
+\monch and GCC 4.8.2 on  \pilat and \tinto) using the -O3 compiler
 flag  in favor  of the  Intel compilers 14.0.1,  delivering inferior
@@
 complex data collections.  All computes nodes have an operating system
-based on GNU/Linux featuring ``2.6.32-358.11.1.el6.x86\_64'' kernel in
-\monch, ``3.0.101-0.15-default'' in \pilat and ``2.6.18-238.19.1.el5.x86\_64'' in \tinto.
+based on GNU/Linux featuring ``2.6.32-358.11.1.el6'' kernel in
+\monch, ``3.0.101-0.15-default'' in \pilat and ``2.6.18-238.19.1.el5'' in \tinto.
 

### 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
   \caption{Mean time-to-solution and energy-to-solution on
-    \textsc{Tintorrum} - MPI Polling and Blocking modes}
+    \textsc{Tintorrum} for a 24h simulation - MPI Polling and Blocking
+    modes}
   \label{fig:4}
@@
   MPI blocking mode\\
-  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}
-  \caption{1 hour simulation (11h-12h) trace using the MPI blocking
-    mode on \textsc{Tintorrum}.}
-  \label{fig:10}
-\end{figure*}
-
-\begin{figure*}[htbf]
-  \centering
-  \scriptsize
+  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}\\
   MPI polling mode\\
@@
   MPI blocking mode\\
-  \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}
+  \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}\\
   \caption{1 hour simulation (23h-24h) trace using the MPI blocking

### 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 We  approach  the  assessment  of  the energy  footprint  and  overall
-performance   of  \textsc{Cosmo-art}   with  two   important  metrics:
+performance   of   \cosmoart   with  two   important  metrics:
 \textit{time-to-solution} (TTS) and \textit{energy-to-solution} (ETS).
@@
 
-\subsection{Time-energy analysis of \textsc{Cosmo-art}}
+\subsection{Time-power-energy analysis of \textsc{Cosmo-art}}
 \label{subsec:4.2}

### 2014-04-25 paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

--- 2014-04-25 paper_enahpc14/Sec3:Environment-setup.tex
@@
 allowing a  fair comparison between the baseline  and future milestone
-versions of \textsc{COSMO-ART}.
+versions of \cosmoart.
 
 \begin{itemize}
-\item \textsc{Monch} is a 10 rack NEC-provided cluster composed
+\item \monch is a 10 rack NEC-provided cluster composed
   of 312 compute  nodes %and utilized by members  of the Swiss Platform
@@
   operating at  2.2 GHz, equipped with 32\,GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
-  a full rack constituted of 52 nodes was used. \textsc{Monch} is  slated to stay
+  a full rack constituted of 52 nodes was used. \monch is  slated to stay
   in  service  without  hardware  upgrade  for  the  duration  of  the
@@
 
-\item \textsc{Pilatus} is a dual-socket eight-core Intel Sandy
+\item \pilat is a dual-socket eight-core Intel Sandy
   Bridge EP based cluster composed of 42 compute nodes %and used as Piz Daint pre-post  processing cluster 
@@
   For our  experiments, a  full rack constituted  of 42 standard  compute nodes
-  was used. \textsc{Pilatus}  is based  on  Intel's second-generation
+  was used. \pilat  is based  on  Intel's second-generation
   processors, conventional in  HPC systems and known to  be more power
@@
 
-\item \textsc{Tintorrum} is  a heterogeneous cluster composed of
+\item \tinto is  a heterogeneous cluster composed of
   28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each of them is composed dual-socket Intel Xeon E5645 hexa-core processors 
   running at 2.40\,GHz, equipped with 288\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS3600 switch. A  full tracing  experiment is  conducted on
-  \textsc{Tintorrum} in order to  capture an overall  power profile at a much
+  \tinto in order to  capture an overall  power profile at a much
   finer resolution. In this sense, we also analyze the contribution of the MPI library
@@
 We present here the two different  frameworks deployed on the testbed HPC systems to
-measure  performance and  power  consumption of the baseline execution of COSMO-ART
+measure  performance and  power  consumption of the baseline execution of \cosmoart.
 
@@
 
-\textsc{COSMO-ART}  is compiled  using the  Extrae  compiler wrappers,
-which automatically  instrument the Fortran code of  the model.  Next,
-\textsc{COSMO-ART}  is  run on  the  nodes,  thus dissipating  certain
-amount  of power.   These  nodes are  connected  to power  measurement
-devices that account for the dissipated power/consumed energy and send
-the power  data to the  tracing server.  The client,  meanwhile, sends
-start/stop  primitives  in  order  to  gather  captured  data  by  the
-watt-meters onto the  tracing server, where an instance  of the \pmlib
-server is running.
+\cosmoart  is compiled and linked with the Extrae instrumentation library,
+which automatically  instrument the Fortran~90 code of the model, e.g., to record MPI calls.  Next, \cosmoart  is  run on  the  nodes,  thus dissipating  certain
+amount  of power.   These  nodes are connected to power distribution units (PDUs)
+APC 8653 that account for the dissipated power with a sampling frequency of 1\,Hz.
+The power data captured by these PDUs is collected into the tracing server.
+The client,  meanwhile, sends a start/stop  primitive  in  order  to  gather  captured  data  by  the PDUs.
 
-Once  the run  is finished,  a file  containing the  power  profile is
-created using the power data  received from the tracing server and the
-instrumentation post-processing generates the performance trace files.
-All  these files  are  combined then  in  Paraver which  allows us  to
-visualise   the   performance  trace   and   the   power  profile   of
-\textsc{Cosmo-art} all together.
+Once  the run  is finished,  a power-performance trace file containing the power 
+data received from the tracing server and the instrumentation post-process is generated.
+This file is finally loaded in Paraver in order to visualise the performance trace and to correlate it with the power profile generated by \cosmoart.
 
@@
 of the  provided compilers, tools  and applications.  For  our initial
-benchmarking,   we  opted   for   the  GNU   compiler  gcc-4.8.1   on
-\textsc{Monch} and gcc-4.8.2 on  \textsc{Pilatus} and \textsc{Tintorrum}) using the -O3 compiler
+benchmarking,   we  opted   for   the  GNU   compiler  GCC 4.8.1   on
+\monch and GCC 4.8.2 on  \pilat and \tinto) using the -O3 compiler
 flag  in favor  of the  Intel compilers 14.0.1,  delivering inferior
 performance.
-In  addition, we installed the following implementations of MPI: MVAPICH2 1.9 on \textsc{Monch} and \textsc{Pilatus} and  OpenMPI 1.6.5 on \textsc{Tintorrum}.
+In  addition, we installed the following implementations of MPI: MVAPICH2 1.9 on \monch and \pilat and  OpenMPI 1.6.5 on \tinto.
 On the other hand, we  used HDF5 1.8.12 and NetCDF  (4.3.1) libraries for  the management  of extremely  large and
 complex data collections.  All computes nodes have an operating system
-based on GNU/Linux featuring ``2.6.32-358.11.1.el6.x86\_64'' kernel in
-\textsc{Monch}     and      ``3.0.101-0.15-default''     kernel     in
-\textsc{Pilatus}.
+based on GNU/Linux featuring ``2.6.32-358.11.1.el6'' kernel in
+\monch, ``3.0.101-0.15-default'' in \pilat and ``2.6.18-238.19.1.el5'' in \tinto.
 

### 2014-04-25 trunk/paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

@@
 \begin{itemize}
-\item \monch is a 10 rack NEC-provided cluster composed
-  of 312 compute  nodes %and utilized by members  of the Swiss Platform
+\item \monch is a 10 rack NEC-provided cluste rat CSCS composed
+  of 312 compute  nodes. %and utilized by members  of the Swiss Platform
   %for Advanced Scientific  Computing (PASC) project~\citep{PASC}
-   at CSCS. Each
-  node comprises two Intel Ivy Bridge EP  E5-2660 ten-core processors
+  Each node comprises two Intel Ivy Bridge EP  E5-2660 ten-core processors
   operating at  2.2 GHz, equipped with 32\,GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
@@
 \item \pilat is a dual-socket eight-core Intel Sandy
-  Bridge EP based cluster composed of 42 compute nodes %and used as Piz Daint pre-post  processing cluster 
-  at  CSCS.  The 2 login  nodes and
+  Bridge EP based cluster at  CSCS composed of 42 compute nodes. %and used as Piz Daint pre-post  processing cluster 
+  The 2 login  nodes and
   the 42 computes nodes consists in 11 twin-pair Intel E5-Series DALCO
@@
 We present here the two different  frameworks deployed on the testbed HPC systems to
-measure  performance and  power  consumption of the baseline execution of \cosmoart
+measure  performance and  power  consumption of the baseline execution of \cosmoart.
 
@@
 
-\cosmoart  is compiled  using the  Extrae  compiler wrappers,
-which automatically  instrument the Fortran code of  the model.  Next,
-\cosmoart  is  run on  the  nodes,  thus dissipating  certain
+\cosmoart  is compiled and linked with the Extrae instrumentation library,
+which automatically  instrument the Fortran~90 code of the model, e.g., to record MPI calls.  Next, \cosmoart  is  run on  the  nodes,  thus dissipating  certain
 amount  of power.   These  nodes are connected to power distribution units (PDUs)
-APC8653 that account for the dissipated power with a sampling frequency of 1\,Hz.
-The power data captured of these PDUs is afterwards collected by the tracing server.
-The client,  meanwhile, sends start/stop  primitives  in  order  to  gather  captured  data  by  the watt-meters onto the  tracing server, where an instance  of the \pmlib server is running.
+APC 8653 that account for the dissipated power with a sampling frequency of 1\,Hz.
+The power data captured by these PDUs is collected into the tracing server.
+The client,  meanwhile, sends a start/stop  primitive  in  order  to  gather  captured  data  by  the PDUs.
 
-Once  the run  is finished,  a file  containing the  power  profile is
-created using the power data  received from the tracing server and the
-instrumentation post-processing generates the performance trace files.
-All  these files  are  combined then  in  Paraver which  allows us  to
-visualise   the   performance  trace   and   the   power  profile   of
-\cosmoart all together.
+Once  the run  is finished,  a power-performance trace file containing the power 
+data received from the tracing server and the instrumentation post-process is generated.
+This file is finally loaded in Paraver in order to visualise the performance trace and to correlate it with the power profile generated by \cosmoart.
 

### 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 
-\subsection{Time-energy analysis of \textsc{Cosmo-art}}
+\subsection{Time-power-energy analysis of \cosmoart}
 \label{subsec:4.2}

### 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
     \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
-    \caption{\textsc{Monch}: Isola E1 Rack 2 Total Power}
+    \caption{\textsc{Monch}: Isola E1 Rack 2 Total Power.}
     \label{fig:1}
@@
     \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
-    \caption{\textsc{Pilatus}: Isola HD Total Power}
+    \caption{\textsc{Pilatus}: Isola HD Total Power.}
     \label{fig:2}
@@
 
-In  Figure~\ref{fig:3}, we  compare both  TTS (right  y-axis)  and ETS
-(left  y-axis)  metrics  on  both  platforms.  As  expected,  Xeon  E5
+In  Figure~\ref{fig:3}, we  compare both  TTS (right  Y-axis)  and ETS
+(left  Y-axis)  metrics  on  both  platforms.  As  expected,  Xeon  E5
 outperforms Ivy Bridge EP, being  roughly 1.3x faster.  The reason for
-that is  two-fold: (1) it has  higher clock frequency  than Ivy Bridge
-(2.6  GHz  against  2.2 GHz),  and  (2)  it  aims at  computing  speed
+that is  two-fold: \emph{i}) it has  higher clock frequency  than Ivy Bridge
+(2.6  GHz  against  2.2 GHz),  and  \emph{ii})  it  aims at  computing  speed
 regardless to  energy consumption.  In our experiments,  Ivy Bridge EP
 showed the best energy-to-solution, reducing the energy consumption of
-Xeon E5 by approximately $7\%$.
+Xeon E5 by approximately 7\,\%.
 
@@
 
-\subsection{Power-performance tracing of \textsc{Cosmo-art}}
+\subsection{Power-performance tracing of \cosmoart}
 \label{subsec:4.3}
+
+OpenMPI 1.6.5, the communication library tested in \tinto, features two operation modes, 
+blocking and modes, which can be selected when the application is launched.
+In the polling configuration (the detault mode) ``MPI'' continually polls
+the netwrok interface to check for the completion of an event (e.g. send or receive).
+This usually yields low latency but high CPU utilization. Thus, can one expcet that this mode attanins the best performance, possibley at the cost a higher energy usage. In the blocking
+mode, the CPU is yield to other processes/threads if there are no incoming messages.
 
@@
   \scriptsize
-  MPI polling mode\\
+  Simulation of 1 hour during midday (from 11h to 12h) leveraging the polling MPI policy\\
   \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}}\\
-  MPI blocking mode\\
+  Simulation of 1 hour during midday (from 11h to 12h) leveraging the blocking MPI policy\\
   \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}\\
-  MPI polling mode\\
+  Simulation of 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
   \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}}\\
-  MPI blocking mode\\
+  Simulation of 1 hour during midnight (from 23h to 24h) leveraging the blocking MPI policy\\
   \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}\\

### 2014-04-25 trunk/paper_enahpc14/Introduction.tex
WS: proofread

@@
 While anthropogenic greenhouse gas emissions are driving unprecedented
-major  climate  changes  since   the  mid  20th  Century,  one  factor
+major  climate  changes  since   the  mid-20$^{th}$  Century,  one  factor
 overwhelmingly  affects  the  uncertainty  in  defining  human-induced
 radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
-Although they  are not considered  a heat-trapping greenhouse  gas and
-have  shorter  atmospheric lifetimes,  they  significantly modify  the
+Although they  are not considered  heat-trapping greenhouse  gases and
+have  shorter  atmospheric lifetimes,  aerosols  significantly modify  the
 global   radiation   budget   \citep{IPCC-2013}.   Enhanced   aerosols
@@
 of solar radiation, thereby exerting a direct radiative forcing; or by
-the  modification  of cloud  properties,  cloud  fraction and  surface
+modifying cloud  properties,  cloud  fraction and  surface
 albedo,  causing  a  negative  indirect  radiative  forcing.   Despite
 considerable  progress in  global aerosol  modelling \citep{Mann-2013}
-and    measurement-based    assessments   \citep{Myhre-2009},    large
-uncertainties  remain  in   current  estimates  of  aerosol  radiative
-forcing   \citep{Myhre-2013,    IPCC-2013,   Lee-2013,   Randles-2013,
-Rosenfeld-2013, Sherwood-2013, Stier-2013}.
+and  measurement-based  assessments,  large  uncertainties  remain  in
+current  estimates  of  aerosol radiative  forcing  \citep{Myhre-2013,
+IPCC-2013,  Lee-2013,   Randles-2013,  Rosenfeld-2013,  Sherwood-2013,
+Stier-2013}.
 
@@
 (KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
-scale model  system is coupled  with the operational  weather forecast
-model COSMO \citep{Baldauf-2011}, jointly developed by a consortium of
-European weather  centers and used in the  climate version (COSMO-CLM)
-by a wide research community.
+scale modelling system is coupled  with the operational  weather forecast
+model  \textsc{Cosmo}  \citep{Baldauf-2011},  jointly developed  by  a
+consortium of European weather centers and also used in a climate version
+by the wider research community.
 
-The extended atmospheric model  COSMO-ART is computationally much more
-demanding than  COSMO since a  large number of additional  tracers and
-processes  have  to  be  considered.   Thus  the  model  is  currently
-severely-limited in  terms of applicability and expensive  in terms of
-energy  consumption.   Although  COSMO  has recently  been  ported  to
-GPUs  \citep{Gysi-2014, Lapillonne-2014} within  the framework  of the
-High    Performance   and    High   Productivity    Computing   (HP2C)
-Initiative \citep{HP2C},  to optimise it for  computational and energy
-efficiency, significant investments in  ART are still required to take
-it to a similar level.  The efficiency of COSMO-ART is being addressed
-in the  EU Exa2Green project \citep{EXA2GREEN} to  deliver a prototype
-code, which  provides an energy efficiency  of at least  five times of
-the baseline value.  Such  an implementation would allow the community
-to investigate critical questions at higher resolution and over longer
-periods, at reduced cost to the environment.
+The extended  atmospheric model \textsc{Cosmo-art}  is com\-put\-ationally
+much  more  demanding than  \textsc{Cosmo}  since  a  large number  of
+additional  tracers and  processes have  to be  considered.   Thus the
+model  is currently  severely-limited  in terms  of applicability  and
+expensive in terms of energy consumption.  Although \textsc{Cosmo} has
+recently been ported to GPUs \citep{Gysi-2014, Lapillonne-2014} within
+the framework of the  High Performance and High Productivity Computing
+(HP2C) Initiative  \citep{HP2C} to optimise it  for computational and
+energy efficiency,  significant investments in ART  are still required
+to take it  to a similar level.  The  efficiency of \textsc{Cosmo-art}
+is being  addressed in the  EU Exa2Green project  \citep{EXA2GREEN} to
+deliver a  prototype code, which  provides an energy efficiency  of at
+least five times of the  baseline value.  Such an implementation would
+allow  the  community  to  investigate critical  questions  at  higher
+resolution  and   over  longer  periods,   at  reduced  cost   to  the
+environment.
 
@@
 areas where improvements will be necessary for the subsequent baseline
-refactoring.   Finally,  we  conclude  with  practical  guidelines  to
-identifying energy saving opportunities  and give an outlook to future
-research in Sec.~\ref{concl}.
+refactoring.   Finally,  we  conclude in Sec.~\ref{concl} with some 
+implications for the Exa2Green project and give an outlook for future
+research.

### 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
   \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}\\
-  \caption{1 hour simulation (23h-24h) trace using the MPI blocking
-    mode on \textsc{Tintorrum}.}
+  \begin{tabular}{rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rl}
+& &  & &  & &  & &  \\[-0.15cm]
+Computation     & \multicolumn{1}{>{\columncolor[RGB]{  0,  , 255}}p{0.4cm}}{}  & & 
+Waiting a msg.  & \multicolumn{1}{>{\columncolor[RGB]{255,  0  ,0}}p{0.4cm}}{}  & & 
+Block. send     & \multicolumn{1}{>{\columncolor[RGB]{255,  0,174}}p{0.4cm}}{}  & & 
+Synchronization & \multicolumn{1}{>{\columncolor[RGB]{179,  0,  0}}p{0.4cm}}{}  & & 
+Wait/Wait all   & \multicolumn{1}{>{\columncolor[RGB]{235,  0,  0}}p{0.4cm}}{}  \\
+& &  & &  & &  & &  \\[-0.15cm]
+Immediate recv. & \multicolumn{1}{>{\columncolor[RGB]{100,100,177}}p{0.4cm}}{}  & &
+I/O             & \multicolumn{1}{>{\columncolor[RGB]{172,174, 41}}p{0.4cm}}{}  & &
+Group comm.     & \multicolumn{1}{>{\columncolor[RGB]{255,144, 26}}p{0.4cm}}{}  & &
+Others          & \multicolumn{1}{>{\columncolor[RGB]{192,224,  0}}p{0.4cm}}{}  & &
+Idle            & \multicolumn{1}{>{\columncolor[RGB]{117,195,255}}p{0.4cm}}{}  \\
+\end{tabular}
+  \caption{Power-perfomance traces during midday and midnight leveraging the polling and blocking MPI policies on \tinto.}
   \label{fig:11}

### 2014-04-25 trunk/paper_enahpc14/Sec2:COSMO-ART.tex
WS: proofread

@@
 within  an  expected  variance  for  the  duration  of  the  Exa2Green
-project.  To   define  a  baseline,   it  was  necessary  to   find  a
+pro\-je\-ct.  To   define  a  baseline,   it  was  necessary  to   find  a
 run-configuration  capable  of   being  recreated  in  all  subsequent
@@
 Three-dimensional simulations are performed over large parts of Europe
-and the Mediterranean~Sea for April 13th 2010 in order to benefit from
-around  half  day  of  sun  exposure and  therefore  ensure  a  proper
+and the Mediterranean~Sea for April 13$^{th}$ 2010 in order to approximately
+have a half  day  of  sun  exposure and  therefore  ensure  a typical
 activation of the chemistry cycle.  The calculation domain corresponds
@@
 in  both directions  and 40~vertical  layers.  These  24-hour forecast
-simulations  are not  preceded  by  a spin-up  phase  to build-up  the
+simulations  are not  preceded  by  a spin-up  phase  to build up  the
 simulated gaseous and  aerosol concentrations.  This condition entails
-that at the model initialization,  there is a period of adjustment for
-the model  to reach its equilibrium  state and minimize  the effect of
+that at the model initialisation there is a period of adjustment for
+the model  to reach its equilibrium  state and minimise  the effect of
 initial conditions for gases and aerosols on the model predictions.
@@
 The considered  \textsc{Cosmo-art} model  system is configured  with a
-semi-Lagrangian    horizontal   advection    scheme    with   tricubic
+semi-Lagrangian    horizontal   advection    sche\-me    with   tricubic
 interpolation  and selective filling  diffusion option  in combination

### 2014-04-25 paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

@@
   28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each of them is composed dual-socket Intel Xeon E5645 hexa-core processors 
-  running at 2.40\,GHz, equipped with 288\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS3600 switch. A  full tracing  experiment is  conducted on
-  \tinto in order to  capture an overall  power profile at a much
-  finer resolution. In this sense, we also analyze the contribution of the MPI library
-  to the energy consumption and how the use of the blocking versus polling
-  message-passing policies can potentially render energy savings.
+  running at 2.40\,GHz, equipped with 288\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS3600 switch.
+
+  % A  full tracing  experiment is  conducted on
+  % \tinto in order to  capture an overall  power profile at a much
+  % finer resolution. In this sense, we also analyze the contribution of the MPI library
+  % to the energy consumption and how the use of the blocking versus polling
+  % message-passing policies can potentially render energy savings.
 \end{itemize}

### 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 
-OpenMPI 1.6.5, the communication library tested in \tinto, features two operation modes, 
-blocking and modes, which can be selected when the application is launched.
+A full tracing experiment is conducted on \tinto in order to capture an overall power
+profile at a much finer resolution.
+In this sense, we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
+
+Specifically, the OpenMPI 1.6.5 library leveraged in \tinto, features two operation modes, 
+blocking and modes, which can be selected before \cosmoart is launched.
 In the polling configuration (the detault mode) ``MPI'' continually polls
 the netwrok interface to check for the completion of an event (e.g. send or receive).
+
 This usually yields low latency but high CPU utilization. Thus, can one expcet that this mode attanins the best performance, possibley at the cost a higher energy usage. In the blocking

### 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 
-\subsection{Time-power-energy analysis of \textsc{Cosmo-art}}
+\subsection{Time-power-energy analysis of \cosmoart}
 \label{subsec:4.2}
@@
 
-\subsection{Power-performance tracing of \textsc{Cosmo-art}}
+\subsection{Power-performance tracing of \cosmoart}
 \label{subsec:4.3}
+
+A full tracing experiment is conducted on \tinto in order to capture an overall power
+profile at a much finer resolution. First we run \cosmoart for a full-day simulation
+and analyze the power consumption. Second we correlate performance with
+power traces of two 1-hor time frames during the day: midday (from 11h-12h) and
+midnight (23-24h).
+
+For all the experiments we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
+Specifically, the OpenMPI 1.6.5 library installed in \tinto, features two operation modes, blocking and modes, which can be selected before \cosmoart is launched.
+In the polling configuration (the detault mode) ``MPI'' continually polls
+the netwrok interface to check for the completion of an event (e.g. send or receive).
+This usually yields low latency but high CPU utilization. Thus, can one expcet that this mode attanins the best performance, possibley at the cost a higher energy usage. In the blocking
+mode, the CPU is yield to other processes/threads if there are no incoming messages.
 
@@
   \scriptsize
-  MPI polling mode\\
+  Simulation of 1 hour during midday (from 11h to 12h) leveraging the polling MPI policy\\
   \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}}\\
-  MPI blocking mode\\
+  Simulation of 1 hour during midday (from 11h to 12h) leveraging the blocking MPI policy\\
   \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}\\
-  MPI polling mode\\
+  Simulation of 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
   \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}}\\
-  MPI blocking mode\\
+  Simulation of 1 hour during midnight (from 23h to 24h) leveraging the blocking MPI policy\\
   \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}\\
-  \caption{1 hour simulation (23h-24h) trace using the MPI blocking
-    mode on \textsc{Tintorrum}.}
+  \begin{tabular}{rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rl}
+& &  & &  & &  & &  \\[-0.15cm]
+Computation     & \multicolumn{1}{>{\columncolor[RGB]{  0,  , 255}}p{0.4cm}}{}  & & 
+Waiting a msg.  & \multicolumn{1}{>{\columncolor[RGB]{255,  0  ,0}}p{0.4cm}}{}  & & 
+Block. send     & \multicolumn{1}{>{\columncolor[RGB]{255,  0,174}}p{0.4cm}}{}  & & 
+Synchronization & \multicolumn{1}{>{\columncolor[RGB]{179,  0,  0}}p{0.4cm}}{}  & & 
+Wait/Wait all   & \multicolumn{1}{>{\columncolor[RGB]{235,  0,  0}}p{0.4cm}}{}  \\
+& &  & &  & &  & &  \\[-0.15cm]
+Immediate recv. & \multicolumn{1}{>{\columncolor[RGB]{100,100,177}}p{0.4cm}}{}  & &
+I/O             & \multicolumn{1}{>{\columncolor[RGB]{172,174, 41}}p{0.4cm}}{}  & &
+Group comm.     & \multicolumn{1}{>{\columncolor[RGB]{255,144, 26}}p{0.4cm}}{}  & &
+Others          & \multicolumn{1}{>{\columncolor[RGB]{192,224,  0}}p{0.4cm}}{}  & &
+Idle            & \multicolumn{1}{>{\columncolor[RGB]{117,195,255}}p{0.4cm}}{}  \\
+\end{tabular}
+  \caption{Power-perfomance traces during midday and midnight leveraging the polling and blocking MPI policies on \tinto.}
   \label{fig:11}

### 2014-04-25 trunk/paper_enahpc14/Sec3:Environment-setup.tex
WS: proofread

@@
 \begin{itemize}
-\item \monch is a 10 rack NEC-provided cluste rat CSCS composed
-  of 312 compute  nodes. %and utilized by members  of the Swiss Platform
+\item \monch is a 10-rack NEC-provided clus\-ter at CSCS composed
+  of 312 compute  nodes. %and utilised by members  of the Swiss Platform
   %for Advanced Scientific  Computing (PASC) project~\citep{PASC}
-  Each node comprises two Intel Ivy Bridge EP  E5-2660 ten-core processors
+  Each node is comprised of two Intel Ivy Bridge EP  E5-2660 ten-core processors
   operating at  2.2 GHz, equipped with 32\,GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
@@
   The 2 login  nodes and
-  the 42 computes nodes consists in 11 twin-pair Intel E5-Series DALCO
-  r2264i4t 2U scalable compute modules. Each module contains 4 compute
+  the 42 computes nodes are consolidated into 11 twin-pair Intel E5-Series DALCO
+  r2264i4t 2U scalable compute modules, each containing 4 compute
   nodes based  on two Intel Xeon E5-2670 processors operating  at 2.6\,GHz equipped with 64GB of DDR3 1600\,MHz RAM and connected  via InfiniBand Mellanox SX6036 and FDR switches. 
@@
 \citep{Riedonetworks},   that  enable   to  monitor   and   log  power
-consumption  of the IT  infrastructure as  well as  constantly analyze
+consumption  of the IT  infrastructure as  well as  constantly analyse
 line  voltage,  current, power-factor,  frequency  with 1\,\%  accuracy.
@@
 combination with Extrae and Paraver, which  are profiling/tracing and
-visualization tools, respectively.
+visualisation tools, respectively.
 
 \cosmoart  is compiled and linked with the Extrae instrumentation library,
-which automatically  instrument the Fortran~90 code of the model, e.g., to record MPI calls.  Next, \cosmoart  is  run on  the  nodes,  thus dissipating  certain
+which automatically  instruments the Fortran~90 code of the model, e.g., to record MPI calls.  Next, \cosmoart  is  run on  the  nodes,  thus dissipating  certain
 amount  of power.   These  nodes are connected to power distribution units (PDUs)
@@
 \monch and GCC 4.8.2 on  \pilat and \tinto) using the -O3 compiler
-flag  in favor  of the  Intel compilers 14.0.1,  delivering inferior
+flag  in favor  of the  Intel compilers 14.0.1,  which delivered inferior
 performance.
@@
 
-A snapshot of the code,  which includes at least conceptually, all the
+A snapshot of the code,  which includes, at least conceptually, all the
 information needed  to reproduce the  energy-to-solution benchmarks of

### 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 
+In order to obtain an initial overview of the power dissipation of the system model, we run a
+full-day simulation of \cosmoart and capture the power profiles leveraging the two
+stated MPI policies, polling and blocking. Figure~\ref{fig:9} shows the power profiles
+of two \cosmoart runs featuring both MPI waiting modes. The first observation is that,
+for the polling MPI policy
+
+The first observation is that
+
 \begin{figure*}[htbf]

### 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

--- 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
@@
 and analyze the power consumption. Second we correlate performance with
-power traces of two 1-hor time frames during the day: midday (from 11h-12h) and
+power traces of two 1-hor time frames during the day: mid-day (from 11h-12h) and
 midnight (23-24h).
@@
 mode, the CPU is yield to other processes/threads if there are no incoming messages.
+
+In order to obtain an initial overview of the power dissipation of the system model, Figure~\ref{fig:9} depicts the power profiles of a full-day simulation of \cosmoart leveraging the  polling and blocking MPI policies on the top and bottom plots, respectively.
+The power profiles shown account for the sum of dissipated power of the 16 nodes used in \tinto captured with our PDUs.
+The first observation for the polling MPI policy is the quite plain power profile it generated.
+However the first and last hours of the day along a 24h simulation, dissipate slightly less power than the hours than the mid-day hours.
+We relate these variations due to the chemical reactions and aerosols, computed by ART, that are taking place during the daylight hours but not during night.
+In this case peak power dissipation is about 3821\,W.
+Looking at the power profile using the blocking MPI policy, we can easily observe
+that this pattern is repeated favoring the reduction of power during nighttime hours in a more noticeable way.
 

### 2014-04-25 trunk/paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

@@
   28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each of them is composed dual-socket Intel Xeon E5645 hexa-core processors 
-  running at 2.40\,GHz, equipped with 288\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS3600 switch.
+  running at 2.40\,GHz, equipped with 24\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS3600 switch.
 

### 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

--- 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
@@
 
-OpenMPI 1.6.5, the communication library tested in \tinto, features two operation modes, 
-blocking and modes, which can be selected when the application is launched.
+A full tracing experiment is conducted on \tinto in order to capture an overall power
+profile at a much finer resolution. First we run \cosmoart for a full-day simulation
+and analyze the power consumption. Second we correlate performance with
+power traces of two 1-hor time frames during the day: mid-day (from 11h-12h) and
+midnight (23-24h).
+
+For all the experiments we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
+Specifically, the OpenMPI 1.6.5 library installed in \tinto, features two operation modes, blocking and modes, which can be selected before \cosmoart is launched.
 In the polling configuration (the detault mode) ``MPI'' continually polls
@@
 mode, the CPU is yield to other processes/threads if there are no incoming messages.
+
+In order to obtain an initial overview of the power dissipation of the system model, Figure~\ref{fig:9} depicts the power profiles of a full-day simulation of \cosmoart leveraging the  polling and blocking MPI policies on the top and bottom plots, respectively.
+The power profiles shown account for the sum of dissipated power of the 16 nodes used in \tinto, captured with our PDUs.
+The first observation for the polling MPI policy is the quite plain power profile it generates.
+However, the first and last hours of the day along a 24h simulation, dissipate slightly less power than the hours than the mid-day hours.
+We relate these variations due to the chemical reactions and aerosols, computed by ART, that are taking place during the daylight hours but not during night.
+In this case peak power dissipation is about 3821\,W.
+Looking at the power profile using the blocking MPI policy, we can easily observe
+that the pattern is repeated favoring the reduction of power during nighttime hours in a more noticeable way.
 

### 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 
+The 24 hours simulation time-to-solution and energy-to-solution 
+results for Tintorrum are shown in Figure~\ref{fig:4}.  As expected, the mean energy
+consumed by COSMO-ART when using the MPI blocking mode is less than that
+consumed by the polling mode. This is due to the blocking-waits performed by the cores
+while using the blocking mode, in contrast of the polling-waits that take place when
+polling mode is used. From the point of view of time, the results are quite
+surprising since the blocking mode was expected to take less time than the polling
+one.
+
 \begin{figure}[htbf]
@@
 that the pattern is repeated favoring the reduction of power during nighttime hours in a more noticeable way.
+The peak power dissipation in this mode is 3768\,W, which means a reduction of 50\,W in total or, in other words, 1,\% of reductuon in the peak power.
+Also, we observe power drops in each hour simulated, basically related to periods in which the processes were performing blocking-waits instead of polling-waits, thus yielding the CPU to a low usage during this phases.
+Our final observation is the reduction in execution time when the blocking MPI policy is leveraged.
+One can expect that the costs of the interruptions used in the blocking can lead to an increase of the total execution time, however, applications with extensive computing may show better perforamnce with the blocking mode.
+This plays in favour for our results since the energy consumed is also less by just changing the behavior of
+the OpenMPI library from the polling to the blocking policy.
 

### 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
   \caption{Time-to-solution and energy-to-solution comparison between
-    Xeon E5 and Ivy Bridge-EP architectures for a 24h simulation}
+    Xeon E5 and Ivy Bridge-EP architectures for a 24h simulation.}
   \label{fig:3}
@@
   \caption{Mean time-to-solution and energy-to-solution on
-    \textsc{Tintorrum} for a 24h simulation - MPI Polling and Blocking
-    modes}
+    \textsc{Tintorrum} for a 24h simulation using the polling and blocking MPI policies.}
   \label{fig:4}
@@
 In order to obtain an initial overview of the power dissipation of the system model, Figure~\ref{fig:9} depicts the power profiles of a full-day simulation of \cosmoart leveraging the  polling and blocking MPI policies on the top and bottom plots, respectively.
-The power profiles shown account for the sum of dissipated power of the 16 nodes used in \tinto captured with our PDUs.
-The first observation for the polling MPI policy is the quite plain power profile it generated.
-However the first and last hours of the day along a 24h simulation, dissipate slightly less power than the hours than the mid-day hours.
+The power profiles shown account for the sum of dissipated power of the 16 nodes used in \tinto, captured with our PDUs.
+The first observation for the polling MPI policy is the quite plain power profile it generates.
+However, the first and last hours of the day along a 24h simulation, dissipate slightly less power than the hours than the mid-day hours.
 We relate these variations due to the chemical reactions and aerosols, computed by ART, that are taking place during the daylight hours but not during night.
@@
 Looking at the power profile using the blocking MPI policy, we can easily observe
-that this pattern is repeated favoring the reduction of power during nighttime hours in a more noticeable way.
+that the pattern is repeated favoring the reduction of power during nighttime hours in a more noticeable way.
+The peak power dissipation in this mode is 3768\,W, which means a reduction of 50\,W in total or, in other words, 1,\% of reductuon in the peak power.
+Also, we observe power drops in each hour simulated, basically related to periods in which the processes were performing blocking-waits instead of polling-waits, thus yielding the CPU to a low usage during this phases.
+Our final observation is the reduction in execution time when the blocking MPI policy is leveraged.
+One can expect that the costs of the interruptions used in the blocking can lead to an increase of the total execution time, however, applications with extensive computing may show better perforamnce with the blocking mode.
+This plays in favour for our results since the energy consumed is also less by just changing the behavior of
+the OpenMPI library from the polling to the blocking policy.
+
+To gain insights about
 

### 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 
-The 24 hours simulation time-to-solution and energy-to-solution 
-results for Tintorrum are shown in Figure~\ref{fig:4}.  As expected, the mean energy
-consumed by COSMO-ART when using the MPI blocking mode is less than that
-consumed by the polling mode. This is due to the blocking-waits performed by the cores
-while using the blocking mode, in contrast of the polling-waits that take place when
-polling mode is used. From the point of view of time, the results are quite
-surprising since the blocking mode was expected to take less time than the polling
-one.
+The TTS and ETS results for a full-day simulation on \tinto are shown in Figure~\ref{fig:4}.
+As expected, the mean energy consumed by COSMO-ART when using the blocking MPI policy is about 2\,W less than that consumed by the polling mode.
+This is due to the blocking-waits performed by the CPU cores while using the blocking mode, in contrast of the polling-waits that take place when polling mode is used.
+From the point of view of the execution time, there is a slight reduction when the blocking MPI policy is leveraged.
+One can expect that the costs of the interruptions used in the blocking MPI policy can lead to an increase of the total execution time, however, applications with extensive computing may show better perforamnce with the blocking modes.
+This plays in favour for our results since the energy consumed is also less by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
 
@@
 Also, we observe power drops in each hour simulated, basically related to periods in which the processes were performing blocking-waits instead of polling-waits, thus yielding the CPU to a low usage during this phases.
-Our final observation is the reduction in execution time when the blocking MPI policy is leveraged.
+Our final observation is the reduction in execution time when the blocking MPI policy is leveraged in contrast with the polling MPI.
 One can expect that the costs of the interruptions used in the blocking can lead to an increase of the total execution time, however, applications with extensive computing may show better perforamnce with the blocking mode.
-This plays in favour for our results since the energy consumed is also less by just changing the behavior of
-the OpenMPI library from the polling to the blocking policy.
+This plays in favour for our results since the energy consumed is also less by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
 

### 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 time. ETS  is the amount of  energy spent to  achieve results.  Energy
-consumption  is  assessed by  sampling  the  instantaneous peak  power
-during execution which  is then averaged and multiplied  by the TTS to
-determine  ETS.   Whenever  possible,  multiple production  runs  were
+consumption  is  assessed by  sampling  the  power during execution which is then averaged and multiplied  by the TTS to determine  ETS. Whenever  possible,  multiple production  runs  were
 performed  to  illustrate the  reproducibility  of  the baseline,  and
@@
 dictated by the available technology.
+
+For all the experiments we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
+Specifically, the OpenMPI 1.6.5 library installed in \tinto, features two operation modes, blocking and modes, which can be selected before \cosmoart is launched.
+In the polling configuration (the detault mode) ``MPI'' continually polls
+the netwrok interface to check for the completion of an event (e.g. send or receive).
+This usually yields low latency but high CPU utilization. Thus, can one expcet that this mode attanins the best performance, possibley at the cost a higher energy usage. In the blocking
+mode, the CPU is yield to other processes/threads if there are no incoming messages.
 
@@
     \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
-    \caption{\textsc{Monch}: Isola E1 Rack 2 Total Power.}
+    \caption{\monch: Isola E1 Rack 2 Total Power.}
     \label{fig:1}
@@
     \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
-    \caption{\textsc{Pilatus}: Isola HD Total Power.}
+    \caption{\pilat: Isola HD Total Power.}
     \label{fig:2}
@@
 Figures~\ref{fig:1}   and   \ref{fig:2}   account   respectively   for
-\textsc{Monch}'s Isola E1 Rack  2 and \textsc{Pilatus}' Isola HD total
+\monch's Isola E1 Rack  2 and \pilat' Isola HD total
 power measurements for  1-day or 2-days simulations. On  the Intel Ivy
-Bridge EP  based cluster (i.e.  \textsc{Monch}),  the 1-day simulation
+Bridge EP  based cluster (i.e.  \monch),  the 1-day simulation
 was issued  only twice due  to usage restrictions. As  time resolution
@@
 power consumption was computed by considering 6 values for each single
-run.  On the Intel Xeon E5 based cluster (i.e.  \textsc{Pilatus}), the
+run.  On the Intel Xeon E5 based cluster (i.e.  \pilat), the
 1-day simulation  was issued  four times and  a 2-days run  only once.
@@
 
-\subsection{Power-performance tracing of \cosmoart}
+\subsection{Power-performance tracing on \tintorrum}
 \label{subsec:4.3}
@@
 power traces of two 1-hor time frames during the day: mid-day (from 11h-12h) and
-midnight (23-24h).
+midnight (23-24h). All the experiments were performed using 192 \cosmoart processes
+mapping 12 processes in each of node.
 
@@
 
-In order to obtain an initial overview of the power dissipation of the system model, Figure~\ref{fig:9} depicts the power profiles of a full-day simulation of \cosmoart leveraging the  polling and blocking MPI policies on the top and bottom plots, respectively.
-The power profiles shown account for the sum of dissipated power of the 16 nodes used in \tinto, captured with our PDUs.
+In order to obtain an initial overview of the power dissipation of the system model, we have obtained the power profiles using our \pmlib framework of a full-day simulation of \cosmoart leveraging the polling and blocking MPI policies, these results are depicted in the top/bottom plots of Figure~\ref{fig:9}, respectvely.
+o, captured with our PDUs.
 The first observation for the polling MPI policy is the quite plain power profile it generates.
-However, the first and last hours of the day along a 24h simulation, dissipate slightly less power than the hours than the mid-day hours.
+However, the first and last hours of the full-day simulation dissipate slightly less power than the mid-day hours.
 We relate these variations due to the chemical reactions and aerosols, computed by ART, that are taking place during the daylight hours but not during night.
@@
 Also, we observe power drops in each hour simulated, basically related to periods in which the processes were performing blocking-waits instead of polling-waits, thus yielding the CPU to a low usage during this phases.
-Our final observation is the reduction in execution time when the blocking MPI policy is leveraged.
+Our final observation is the reduction in execution time when the blocking MPI policy is leveraged in contrast with the polling MPI.
 One can expect that the costs of the interruptions used in the blocking can lead to an increase of the total execution time, however, applications with extensive computing may show better perforamnce with the blocking mode.
-This plays in favour for our results since the energy consumed is also less by just changing the behavior of
-the OpenMPI library from the polling to the blocking policy.
+This plays in favour for our results since the energy consumed is also less by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
+
+To gain insights about
 

### 2014-04-25 trunk/paper_enahpc14/paper.tex
WS: proofread

@@
 \usepackage{xspace}
+\usepackage{epsfig}
 %
@@
 \newcommand{\pmlib}{\texttt{pmlib}\xspace}
+\newcommand{\monch}{\textsc{Monch}\xspace}
+\newcommand{\pilat}{\textsc{Pilatus}\xspace}
+\newcommand{\tinto}{\textsc{Tintorrum}\xspace}
+\newcommand{\cosmoart}{\textsc{Cosmo-Art}\xspace}
 %
@@
   Green computing  \and Numerical weather  prediction \and Atmospheric
-  chemistry  \and  Aerosols  modelling  \and  Profiling  methods  \and
+  chemistry  \and  Aerosol  modelling  \and  Profiling  methods  \and
   Benchmark  analysis  \and   \textsc{COSMO-ART}  coupled  model  \and

### 2014-04-25 paper_enahpc14/Introduction.tex
WS: proofread

@@
 While anthropogenic greenhouse gas emissions are driving unprecedented
-major  climate  changes  since   the  mid  20th  Century,  one  factor
-overwhelmingly  affects  the  uncertainty  in  defining  human-induced
+major  climate  changes  since   the  mid-20$^{th}$  Century,  one  factor
+overwhelmingly  affects  the  uncertainty  in  determining  human-induced
 radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
-Although they  are not considered  a heat-trapping greenhouse  gas and
-have  shorter  atmospheric lifetimes,  they  significantly modify  the
+Although they  are not considered  heat-trapping greenhouse  gases and
+have  shorter  atmospheric lifetimes,  aerosols  significantly modify  the
 global   radiation   budget   \citep{IPCC-2013}.   Enhanced   aerosols
@@
 of solar radiation, thereby exerting a direct radiative forcing; or by
-the  modification  of cloud  properties,  cloud  fraction and  surface
+modifying cloud  properties,  cloud  fraction and  surface
 albedo,  causing  a  negative  indirect  radiative  forcing.   Despite
@@
 (KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
-scale model  system is coupled  with the operational  weather forecast
-model  \textsc{Cosmo}  \citep{Baldauf-2011},  jointly developed  by  a
-consortium of European weather centers and used in the climate version
-by a wide research community.
+scale modelling system is coupled  with the operational  weather forecast
+model  \textsc{Cosmo}  \citep{Baldauf-2011},  jointly developed and used by  a
+consortium of European weather centers, as well as utilised in a climate version
+by the wider research community.
 
-The extended  atmospheric model \textsc{Cosmo-art}  is computationally
+The extended  atmospheric model \textsc{Cosmo-art}  is com\-put\-ationally
 much  more  demanding than  \textsc{Cosmo}  since  a  large number  of
 additional  tracers and  processes have  to be  considered.   Thus the
-model  is currently  severely-limited  in terms  of applicability  and
+model  is currently  severely limited  in terms  of applicability  and
 expensive in terms of energy consumption.  Although \textsc{Cosmo} has
@@
 the framework of the  High Performance and High Productivity Computing
-(HP2C) Initiative  \citep{HP2C}, to optimise it  for computational and
+(HP2C) Initiative  \citep{HP2C} to optimise it  for computational and
 energy efficiency,  significant investments in ART  are still required
@@
 deliver a  prototype code, which  provides an energy efficiency  of at
-least five times of the  baseline value.  Such an implementation would
+least five times the  baseline value.  Such an implementation would
 allow  the  community  to  investigate critical  questions  at  higher

### 2014-04-25 trunk/paper_enahpc14/Sec2:COSMO-ART.tex
WS: proofread

@@
 Three-dimensional simulations are performed over large parts of Europe
-and the Mediterranean~Sea for April 13$^{th}$ 2010 in order to approximately
+and the Mediterranean~Sea for April 13$^{th}$ 2010, near the equinox, in order to approximately
 have a half  day  of  sun  exposure and  therefore  ensure  a typical
@@
 simulations  are not  preceded  by  a spin-up  phase  to build up  the
-simulated gaseous and  aerosol concentrations.  This condition entails
-that at the model initialisation there is a period of adjustment for
-the model  to reach its equilibrium  state and minimise  the effect of
+simulated gaseous and  aerosol concentrations.  This condition means 
+that the model initialisation needs a period of adjustment to reach its equilibrium  state and minimise  the effect of
 initial conditions for gases and aerosols on the model predictions.
@@
 \citep{COSMO-PartI-2011}.    It  also   makes  use   of   the  Kinetic
-PreProcessor solver (KPP) for  the resolution of atmospheric chemistry
+PreProcessor solver (KPP) for  the solution of atmospheric chemistry
 ordinary  differential equations \citep{Damian-2002}.   Concerning the

### 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 power traces of two 1-hor time frames during the day: mid-day (from 11h-12h) and
-midnight (23-24h).
+midnight (23h-24h). All the experiments were performed using 192 \cosmoart processes
+mapping 12 processes in each of node.
 
-For all the experiments we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
-Specifically, the OpenMPI 1.6.5 library installed in \tinto, features two operation modes, blocking and modes, which can be selected before \cosmoart is launched.
-In the polling configuration (the detault mode) ``MPI'' continually polls
+For all the experiments we also analyze the contribution of the MPI library to the energy 
+consumption and how the use of the blocking versus polling message-passing policies can 
+potentially render energy savings. Specifically, the OpenMPI 1.6.5 library installed in \tinto, 
+features two operation modes, blocking and modes, which can be selected before \cosmoart is 
+launched. In the polling configuration (the detault mode) ``MPI'' continually polls
 the netwrok interface to check for the completion of an event (e.g. send or receive).
@@
 
-In order to obtain an initial overview of the power dissipation of the system model, Figure~\ref{fig:9} depicts the power profiles of a full-day simulation of \cosmoart leveraging the  polling and blocking MPI policies on the top and bottom plots, respectively.
-The power profiles shown account for the sum of dissipated power of the 16 nodes used in \tinto, captured with our PDUs.
-The first observation for the polling MPI policy is the quite plain power profile it generates.
-However, the first and last hours of the day along a 24h simulation, dissipate slightly less power than the hours than the mid-day hours.
-We relate these variations due to the chemical reactions and aerosols, computed by ART, that are taking place during the daylight hours but not during night.
-In this case peak power dissipation is about 3821\,W.
-Looking at the power profile using the blocking MPI policy, we can easily observe
-that the pattern is repeated favoring the reduction of power during nighttime hours in a more noticeable way.
-The peak power dissipation in this mode is 3768\,W, which means a reduction of 50\,W in total or, in other words, 1,\% of reductuon in the peak power.
-Also, we observe power drops in each hour simulated, basically related to periods in which the processes were performing blocking-waits instead of polling-waits, thus yielding the CPU to a low usage during this phases.
-Our final observation is the reduction in execution time when the blocking MPI policy is leveraged in contrast with the polling MPI.
-One can expect that the costs of the interruptions used in the blocking can lead to an increase of the total execution time, however, applications with extensive computing may show better perforamnce with the blocking mode.
-This plays in favour for our results since the energy consumed is also less by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
+In order to obtain an initial overview of the power dissipation of the system model, we have 
+obtained the power profiles using our \pmlib framework of a full-day simulation of \cosmoart 
+leveraging the polling and blocking MPI policies, these results are depicted in the top/bottom 
+plots of Figure~\ref{fig:9}, respectvely. o, captured with our PDUs. The first observation for 
+the polling MPI policy is the quite plain power profile it generates. However, the first and 
+last hours of the full-day simulation dissipate slightly less power than the mid-day hours. We 
+relate these variations due to the chemical reactions and aerosols, computed by ART, that are 
+taking place during the daylight hours but not during night. In this case peak power 
+dissipation is about 3821\,W. Looking at the power profile using the blocking MPI policy, we 
+can easily observe that the pattern is repeated favoring the reduction of power during 
+nighttime hours in a more  noticeable way. The peak power dissipation in this mode is 3768\,W, 
+which means a reduction of 50\,W in total or, in other words, 1,\% of reductuon in the peak 
+power. Also, we observe power  drops in each hour simulated, basically related to periods in 
+which the processes were performing blocking-waits instead of polling-waits, thus yielding the 
+CPU to a low usage during this phases. Our final observation is the reduction in execution time 
+when the blocking MPI policy is leveraged in contrast with the polling MPI. One can expect that 
+the costs of the interruptions used in the blocking can lead to an increase of the total 
+execution time, however, applications with extensive computing may show better perforamnce with 
+the blocking mode. This plays in favour for our results since the energy consumed is also less 
+by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
 
-To gain insights about
+To gain more insights about the power-performance behavior of \cosmoart, we also have conducted 
+experiments that show power-performance traces of 1 hour simulation. We limited the simulation 
+time to this bound since the pattern of the computation performed by \cosmoart is repeated each 
+hour, also because the weight of a full-day simulation trace can not be easily handled using 
+our machines and memory available (about 45\,GB per trace file). To soften the problem we 
+shrink the simulations to 1 hour during mid-day (11h-12h) and midnight (23h-24h), in which 
+\cosmoart perform different internal computations, as shown in Figure~\ref{fig:9}. At the same
+time we also leverage the two stated MPI policies: polling and waiting modes.
 

### 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 
-\subsection{Power-performance tracing on \tintorrum}
+\subsection{Power-performance tracing on \tinto}
 \label{subsec:4.3}
@@
 power traces of two 1-hor time frames during the day: mid-day (from 11h-12h) and
-midnight (23-24h). All the experiments were performed using 192 \cosmoart processes
+midnight (23h-24h). All the experiments were performed using 192 \cosmoart processes
 mapping 12 processes in each of node.
 
-For all the experiments we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
-Specifically, the OpenMPI 1.6.5 library installed in \tinto, features two operation modes, blocking and modes, which can be selected before \cosmoart is launched.
-In the polling configuration (the detault mode) ``MPI'' continually polls
+For all the experiments we also analyze the contribution of the MPI library to the energy 
+consumption and how the use of the blocking versus polling message-passing policies can 
+potentially render energy savings. Specifically, the OpenMPI 1.6.5 library installed in \tinto, 
+features two operation modes, blocking and modes, which can be selected before \cosmoart is 
+launched. In the polling configuration (the detault mode) ``MPI'' continually polls
 the netwrok interface to check for the completion of an event (e.g. send or receive).
@@
 
-In order to obtain an initial overview of the power dissipation of the system model, we have obtained the power profiles using our \pmlib framework of a full-day simulation of \cosmoart leveraging the polling and blocking MPI policies, these results are depicted in the top/bottom plots of Figure~\ref{fig:9}, respectvely.
-o, captured with our PDUs.
-The first observation for the polling MPI policy is the quite plain power profile it generates.
-However, the first and last hours of the full-day simulation dissipate slightly less power than the mid-day hours.
-We relate these variations due to the chemical reactions and aerosols, computed by ART, that are taking place during the daylight hours but not during night.
-In this case peak power dissipation is about 3821\,W.
-Looking at the power profile using the blocking MPI policy, we can easily observe
-that the pattern is repeated favoring the reduction of power during nighttime hours in a more noticeable way.
-The peak power dissipation in this mode is 3768\,W, which means a reduction of 50\,W in total or, in other words, 1,\% of reductuon in the peak power.
-Also, we observe power drops in each hour simulated, basically related to periods in which the processes were performing blocking-waits instead of polling-waits, thus yielding the CPU to a low usage during this phases.
-Our final observation is the reduction in execution time when the blocking MPI policy is leveraged in contrast with the polling MPI.
-One can expect that the costs of the interruptions used in the blocking can lead to an increase of the total execution time, however, applications with extensive computing may show better perforamnce with the blocking mode.
-This plays in favour for our results since the energy consumed is also less by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
+In order to obtain an initial overview of the power dissipation of the system model, we have 
+obtained the power profiles using our \pmlib framework of a full-day simulation of \cosmoart 
+leveraging the polling and blocking MPI policies, these results are depicted in the top/bottom 
+plots of Figure~\ref{fig:9}, respectvely. o, captured with our PDUs. The first observation for 
+the polling MPI policy is the quite plain power profile it generates. However, the first and 
+last hours of the full-day simulation dissipate slightly less power than the mid-day hours. We 
+relate these variations due to the chemical reactions and aerosols, computed by ART, that are 
+taking place during the daylight hours but not during night. In this case peak power 
+dissipation is about 3821\,W. Looking at the power profile using the blocking MPI policy, we 
+can easily observe that the pattern is repeated favoring the reduction of power during 
+nighttime hours in a more  noticeable way. The peak power dissipation in this mode is 3768\,W, 
+which means a reduction of 50\,W in total or, in other words, 1,\% of reductuon in the peak 
+power. Also, we observe power  drops in each hour simulated, basically related to periods in 
+which the processes were performing blocking-waits instead of polling-waits, thus yielding the 
+CPU to a low usage during this phases. Our final observation is the reduction in execution time 
+when the blocking MPI policy is leveraged in contrast with the polling MPI. One can expect that 
+the costs of the interruptions used in the blocking can lead to an increase of the total 
+execution time, however, applications with extensive computing may show better perforamnce with 
+the blocking mode. This plays in favour for our results since the energy consumed is also less 
+by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
 
-To gain insights about
+To gain more insights about the power-performance behavior of \cosmoart, we also have conducted 
+experiments that show power-performance traces of 1 hour simulation. We limited the simulation 
+time to this bound since the pattern of the computation performed by \cosmoart is repeated each 
+hour, also because the weight of a full-day simulation trace can not be easily handled using 
+our machines and memory available (about 45\,GB per trace file). To soften the problem we 
+shrink the simulations to 1 hour during mid-day (11h-12h) and midnight (23h-24h), in which 
+\cosmoart perform different internal computations, as shown in Figure~\ref{fig:9}. At the same
+time we also leverage the two stated MPI policies: polling and waiting modes.
 

### 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
   \scriptsize
-  Simulation of 1 hour during midday (from 11h to 12h) leveraging the polling MPI policy\\
-  \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}}\\
-  Simulation of 1 hour during midday (from 11h to 12h) leveraging the blocking MPI policy\\
-  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}\\
+  % Simulation of 1 hour during midday (from 11h to 12h) leveraging the polling MPI policy\\
+  % \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}}\\
+  % Simulation of 1 hour during midday (from 11h to 12h) leveraging the blocking MPI policy\\
+  % \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}\\
   Simulation of 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\

### 2014-04-25 paper_enahpc14/Sec3:Environment-setup.tex
WS: proofread

@@
 \begin{itemize}
-\item \monch is a 10 rack NEC-provided cluste rat CSCS composed
-  of 312 compute  nodes. %and utilized by members  of the Swiss Platform
+\item \monch is a 10-rack NEC-provided clus\-ter at CSCS composed
+  of 312 compute  nodes. %and utilised by members  of the Swiss Platform
   %for Advanced Scientific  Computing (PASC) project~\citep{PASC}
-  Each node comprises two Intel Ivy Bridge EP  E5-2660 ten-core processors
-  operating at  2.2 GHz, equipped with 32\,GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
+  Each node is comprised of two Intel Ivy Bridge EP  E5-2660 ten-core processors
+  operating at  2.2 GHz, equipped with 32GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
   a full rack constituted of 52 nodes was used. \monch is  slated to stay
@@
   Bridge EP based cluster at  CSCS composed of 42 compute nodes. %and used as Piz Daint pre-post  processing cluster 
-  The 2 login  nodes and
-  the 42 computes nodes consists in 11 twin-pair Intel E5-Series DALCO
-  r2264i4t 2U scalable compute modules. Each module contains 4 compute
+  The 2 login and
+  42 compute nodes are consolidated into 11 twin-pair Intel E5-Series DALCO
+  r2264i4t 2U scalable compute modules, each containing 4 compute
   nodes based  on two Intel Xeon E5-2670 processors operating  at 2.6\,GHz equipped with 64GB of DDR3 1600\,MHz RAM and connected  via InfiniBand Mellanox SX6036 and FDR switches. 
@@
   processors, conventional in  HPC systems and known to  be more power
-  consuming than Ivy Bridge successor.\\
+  consuming than its Ivy Bridge successor.\\
 
@@
   28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each of them is composed dual-socket Intel Xeon E5645 hexa-core processors 
-  running at 2.40\,GHz, equipped with 288\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS3600 switch.
+  running at 2.40\,GHz, equipped with 24\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS\-3600 switch.
 
@@
 cables thus  avoiding the need for  extra cabling.  This  data is made
-available via  SNMP, HTTP, TELNET  through the built-in  Fast Ethernet
+available via  SNMP, HTTP and TELNET  through the built-in  Fast Ethernet
 port.   Time  synchronisation  is  guaranteed by  using  NTP  servers.
@@
 
-The \cosmoart  baseline is a  pure MPI based  Fortran~90 code
+The \cosmoart  baseline is a  pure MPI-based  Fortran~90 code
 currently  running  on   distributed  multi-core  systems  only.   The

### 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 that is  two-fold: \emph{i}) it has  higher clock frequency  than Ivy Bridge
-(2.6  GHz  against  2.2 GHz),  and  \emph{ii})  it  aims at  computing  speed
+(2.6\,GHz  against  2.2\,GHz),  and  \emph{ii})  it  aims at  computing  speed
 regardless to  energy consumption.  In our experiments,  Ivy Bridge EP
@@
 The TTS and ETS results for a full-day simulation on \tinto are shown in Figure~\ref{fig:4}.  
-As expected, the mean energy consumed by COSMO-ART when using the blocking MPI policy is about 2\,W less than that consumed by the polling mode.
-This is due to the blocking-waits performed by the CPU cores while using the blocking mode, in contrast of the polling-waits that take place when polling mode is used.
-From the point of view of the execution time, there is a slight reduction when the blocking MPI policy is leveraged.
-One can expect that the costs of the interruptions used in the blocking MPI policy can lead to an increase of the total execution time, however, applications with extensive computing may show better perforamnce with the blocking modes.
-This plays in favour for our results since the energy consumed is also less by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
+As expected, the mean energy consumed by COSMO-ART when using the blocking MPI policy is about 
+2\,W less than that consumed by the polling mode. This is due to the blocking-waits performed 
+by the CPU cores while using the blocking mode, in contrast of the polling-waits that take 
+place when polling mode is used. From the point of view of the execution time, there is a 
+slight reduction when the blocking MPI policy is leveraged. One can expect that the costs of 
+the interruptions used in the blocking MPI policy can lead to an increase of the total 
+execution time, however, applications with extensive computing may show better perforamnce with 
+the blocking modes. This plays in favour for our results since the energy consumed is also less 
+by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
 
@@
 A full tracing experiment is conducted on \tinto in order to capture an overall power
-profile at a much finer resolution. First we run \cosmoart for a full-day simulation
-and analyze the power consumption. Second we correlate performance with
-power traces of two 1-hor time frames during the day: mid-day (from 11h-12h) and
+profile at a much finer resolution using the attached PDUs in each node. First we run \cosmoart 
+for a full-day simulation and analyze the power consumption. Second we correlate performance 
+with power traces of two 1-hor time frames during the day: mid-day (from 11h-12h) and
 midnight (23h-24h). All the experiments were performed using 192 \cosmoart processes
@@
 
-For all the experiments we also analyze the contribution of the MPI library to the energy 
-consumption and how the use of the blocking versus polling message-passing policies can 
-potentially render energy savings. Specifically, the OpenMPI 1.6.5 library installed in \tinto, 
-features two operation modes, blocking and modes, which can be selected before \cosmoart is 
-launched. In the polling configuration (the detault mode) ``MPI'' continually polls
-the netwrok interface to check for the completion of an event (e.g. send or receive).
-This usually yields low latency but high CPU utilization. Thus, can one expcet that this mode attanins the best performance, possibley at the cost a higher energy usage. In the blocking
-mode, the CPU is yield to other processes/threads if there are no incoming messages.
+% For all the experiments we also analyze the contribution of the MPI library to the energy 
+% consumption and how the use of the blocking versus polling message-passing policies can 
+% potentially render energy savings. Specifically, the OpenMPI 1.6.5 library installed in \tinto, 
+% features two operation modes, blocking and modes, which can be selected before \cosmoart is 
+% launched. In the polling configuration (the detault mode) ``MPI'' continually polls
+% the netwrok interface to check for the completion of an event (e.g. send or receive).
+% This usually yields low latency but high CPU utilization. Thus, can one expcet that this mode attanins the best performance, possibley at the cost a higher energy usage. In the blocking
+% mode, the CPU is yield to other processes/threads if there are no incoming messages.
 
@@
 leveraging the polling and blocking MPI policies, these results are depicted in the top/bottom 
-plots of Figure~\ref{fig:9}, respectvely. o, captured with our PDUs. The first observation for 
+plots of Figure~\ref{fig:9}, respectvely. The first observation for 
 the polling MPI policy is the quite plain power profile it generates. However, the first and 
@@
 
-\begin{figure*}[htbf]
-  \centering
-  \includegraphics[width=0.78\textwidth]{Figs/23_24_blq1_stat1.eps}
-  \caption{blabla}
-  \label{fig:8}
-\end{figure*}
+% \begin{figure*}[htbf]
+%   \centering
+%   Simulation of 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
+%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq0_stat.eps}\\
+%   Percentage of time for the 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
+%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq1_stat.eps}\\
+%   \caption{blabla}
+%   \label{fig:8}
+% \end{figure*}
 
@@
   \centering
+  \hspace{0.8cm}
   \scalebox{0.5}{\input{Figs/24hour.pstex_t}}
-  \caption{24 hours simulation trace using the MPI blocking mode on
-    \textsc{Tintorrum}.}
+  \caption{24 hours simulation trace using the MPI blocking mode on \tinto.}
   \label{fig:9}
@@
   \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}\\
-  Simulation of 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
+  1 hour simulation during midnight (from 23h to 24h) leveraging the polling MPI policy\\
   \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}}\\
-  Simulation of 1 hour during midnight (from 23h to 24h) leveraging the blocking MPI policy\\
+  1 hour simulation during midnight (from 23h to 24h) leveraging the blocking MPI policy\\
   \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}\\

### 2014-04-25 paper_enahpc14/Sec1:Related.tex
WS: revised
WS: proofread
Manuel's changes

@@
 While the  TOP500 list was  introduced over 20  years ago to  rank the
-performance of HPC systems  worldwide, it's quite recently that energy
+performance of HPC systems  worldwide, it is quite recently that energy
 efficiency  become  a  critical  constraint  in the  way  to  exascale
@@
 power  envelope.   However,  the state-of-the-art  research  assessing
-performance  and energy  efficiency  of applications  is still  pretty
-scarce.
+performance  and energy  efficiency  of applications  is still scarce.
 

### 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 
-For all the experiments we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
-Specifically, the OpenMPI 1.6.5 library installed in \tinto, features two operation modes, blocking and modes, which can be selected before \cosmoart is launched.
-In the polling configuration (the detault mode) ``MPI'' continually polls
-the netwrok interface to check for the completion of an event (e.g. send or receive).
-This usually yields low latency but high CPU utilization. Thus, can one expcet that this mode attanins the best performance, possibley at the cost a higher energy usage. In the blocking
+For the experiments with \tinto platform, we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
+Specifically, the OpenMPI 1.6.5 library,
+features two operation modes, blocking and polling, which can be selected before \cosmoart
+is launched. In the polling configuration (the detault mode) ``MPI'' continually polls
+the network interface to check for the completion of an event (e.g. send or receive).
+This usually yields low latency but high CPU utilization. Thus, one can expcet that this mode 
+attains the best performance, possibly at the cost a higher energy usage. In the blocking
 mode, the CPU is yield to other processes/threads if there are no incoming messages.
@@
 nighttime hours in a more  noticeable way. The peak power dissipation in this mode is 3768\,W, 
-which means a reduction of 50\,W in total or, in other words, 1,\% of reductuon in the peak 
+which means a reduction of 50\,W in total or, in other words, 1\,\% of reductuon in the peak 
 power. Also, we observe power  drops in each hour simulated, basically related to periods in 

### 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
WS: proofread

--- 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
@@
 
-For all the experiments we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
-Specifically, the OpenMPI 1.6.5 library installed in \tinto, features two operation modes, blocking and modes, which can be selected before \cosmoart is launched.
-In the polling configuration (the detault mode) ``MPI'' continually polls
-the netwrok interface to check for the completion of an event (e.g. send or receive).
-This usually yields low latency but high CPU utilization. Thus, can one expcet that this mode attanins the best performance, possibley at the cost a higher energy usage. In the blocking
+For the experiments with \tinto platform, we also analyse the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
+Specifically, the OpenMPI 1.6.5 library,
+features two operation modes, blocking and polling, which can be selected before \cosmoart
+is launched. In the polling configuration (the detault mode) ``MPI'' continually polls
+the network interface to check for the completion of an event (e.g. send or receive).
+This usually yields low latency but high CPU utilization. Thus, one can expcet that this mode 
+attains the best performance, possibly at the cost a higher energy usage. In the blocking
 mode, the CPU is yield to other processes/threads if there are no incoming messages.
@@
 \monch's Isola E1 Rack  2 and \pilat' Isola HD total
-power measurements for  1-day or 2-days simulations. On  the Intel Ivy
+power measurements for  1-day or 2-day simulations. On  the Intel Ivy
 Bridge EP  based cluster (i.e.  \monch),  the 1-day simulation
@@
 run.  On the Intel Xeon E5 based cluster (i.e.  \pilat), the
-1-day simulation  was issued  four times and  a 2-days run  only once.
+1-day simulation  was issued  four times and  a 2-day run  only once.
 Similarly, the average power consumption was computed by considering 4
@@
   \caption{Mean time-to-solution and energy-to-solution on
-    \textsc{Tintorrum} for a 24h simulation using the polling and blocking MPI policies.}
+    \textsc{Tintorrum} for a 24h simulation using the polling and blocking MPI policies. The 95\% 
+            confidence intervals of Student's $t$ distribution (10 samples) clearly indicates 
+            an improvement for blocking policy in both TTS and ETS.}
   \label{fig:4}
@@
 A full tracing experiment is conducted on \tinto in order to capture an overall power
-profile at a much finer resolution. First we run \cosmoart for a full-day simulation
-and analyze the power consumption. Second we correlate performance with
-power traces of two 1-hor time frames during the day: mid-day (from 11h-12h) and
+profile at a much finer resolution using the attached PDUs in each node. First we run \cosmoart 
+for a full-day simulation and analyse the power consumption. Second we correlate performance 
+with power traces of two 1-hor time frames during the day: mid-day (from 11h-12h) and
 midnight (23h-24h). All the experiments were performed using 192 \cosmoart processes
@@
 
-For all the experiments we also analyze the contribution of the MPI library to the energy 
-consumption and how the use of the blocking versus polling message-passing policies can 
-potentially render energy savings. Specifically, the OpenMPI 1.6.5 library installed in \tinto, 
-features two operation modes, blocking and modes, which can be selected before \cosmoart is 
-launched. In the polling configuration (the detault mode) ``MPI'' continually polls
-the netwrok interface to check for the completion of an event (e.g. send or receive).
-This usually yields low latency but high CPU utilization. Thus, can one expcet that this mode attanins the best performance, possibley at the cost a higher energy usage. In the blocking
-mode, the CPU is yield to other processes/threads if there are no incoming messages.
+% For all the experiments we also analyse the contribution of the MPI library to the energy 
+% consumption and how the use of the blocking versus polling message-passing policies can 
+% potentially render energy savings. Specifically, the OpenMPI 1.6.5 library installed in \tinto, 
+% features two operation modes, blocking and modes, which can be selected before \cosmoart is 
+% launched. In the polling configuration (the detault mode) ``MPI'' continually polls
+% the netwrok interface to check for the completion of an event (e.g. send or receive).
+% This usually yields low latency but high CPU utilization. Thus, can one expcet that this mode attanins the best performance, possibley at the cost a higher energy usage. In the blocking
+% mode, the CPU is yield to other processes/threads if there are no incoming messages.
 
@@
 relate these variations due to the chemical reactions and aerosols, computed by ART, that are 
-taking place during the daylight hours but not during night. In this case peak power 
-dissipation is about 3821\,W. Looking at the power profile using the blocking MPI policy, we 
+taking place during the daylight hours but not during night. In this case power consumption
+ is about 3821\,W. Looking at the power profile using the blocking MPI policy, we 
 can easily observe that the pattern is repeated favoring the reduction of power during 
 nighttime hours in a more  noticeable way. The peak power dissipation in this mode is 3768\,W, 
-which means a reduction of 50\,W in total or, in other words, 1,\% of reductuon in the peak 
+which means a reduction of 50\,W in total or, in other words, 1\,\% of reductuon in the peak 
 power. Also, we observe power  drops in each hour simulated, basically related to periods in 

### 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 
-For the experiments with \tinto platform, we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
+For the experiments with \tinto platform, we also analyse the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
 Specifically, the OpenMPI 1.6.5 library,
@@
 the interruptions used in the blocking MPI policy can lead to an increase of the total 
-execution time, however, applications with extensive computing may show better perforamnce with 
-the blocking modes. This plays in favour for our results since the energy consumed is also less 
-by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
+execution time, however, applications with extensive computing, e.g.
+\cosmoart, may show better perforamnce with the blocking modes This plays in favour for our results since the energy consumed is also less by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
 
@@
 profile at a much finer resolution using the attached PDUs in each node. First we run \cosmoart 
-for a full-day simulation and analyze the power consumption. Second we correlate performance 
-with power traces of two 1-hor time frames during the day: mid-day (from 11h-12h) and
-midnight (23h-24h). All the experiments were performed using 192 \cosmoart processes
+for a full-day simulation and analyse the power consumption. Second we correlate performance 
+with power traces of two 1-hor time frames during the day: mid-day (from 11h--12h) and
+midnight (23h--24h). All the experiments were performed using 192 \cosmoart processes
 mapping 12 processes in each of node.
 
-% For all the experiments we also analyze the contribution of the MPI library to the energy 
+% For all the experiments we also analyse the contribution of the MPI library to the energy 
 % consumption and how the use of the blocking versus polling message-passing policies can 
@@
 relate these variations due to the chemical reactions and aerosols, computed by ART, that are 
-taking place during the daylight hours but not during night. In this case peak power 
-dissipation is about 3821\,W. Looking at the power profile using the blocking MPI policy, we 
+taking place during the daylight hours but not during night. In this case power consumption
+ is about 3821\,W. Looking at the power profile using the blocking MPI policy, we 
 can easily observe that the pattern is repeated favoring the reduction of power during 
@@
 CPU to a low usage during this phases. Our final observation is the reduction in execution time 
-when the blocking MPI policy is leveraged in contrast with the polling MPI. One can expect that 
-the costs of the interruptions used in the blocking can lead to an increase of the total 
-execution time, however, applications with extensive computing may show better perforamnce with 
-the blocking mode. This plays in favour for our results since the energy consumed is also less 
-by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
+when the blocking MPI policy is leveraged in contrast with the polling MPI.
 
@@
 our machines and memory available (about 45\,GB per trace file). To soften the problem we 
-shrink the simulations to 1 hour during mid-day (11h-12h) and midnight (23h-24h), in which 
-\cosmoart perform different internal computations, as shown in Figure~\ref{fig:9}. At the same
-time we also leverage the two stated MPI policies: polling and waiting modes.
+shrink the simulations to 1 hour during mid-day (11h--12h) and midnight (23h--24h), in which 
+\cosmoart perform different internal computations due to the different chemical reactions that happen during the different time frames of the day.
+At the same time we also leverage the two stated MPI policies: polling and blocking modes.
+The traces are shown in Figure~\ref{fig:9}.
+As can be observed, the repetition
 

### 2014-04-25 paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

@@
   operating at  2.2 GHz, equipped with 32GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
-  a full rack constituted of 52 nodes was used. \monch is  slated to stay
-  in  service  without  hardware  upgrade  for  the  duration  of  the
-  Exa2Green  project,  allowing   an  identical  configuration  to  be
-  recreated for future assessments of the baseline.\\
+  a full rack constituted of 52 nodes was used. % \monch is  slated to stay
+  %in  service  without  hardware  upgrade  for  the  duration  of  the
+  %Exa2Green  project,  allowing   an  identical  configuration  to  be
+  %recreated for future assessments of the baseline.\\
 
@@
   Bridge EP based cluster at  CSCS composed of 42 compute nodes. %and used as Piz Daint pre-post  processing cluster 
-  The 2 login and
-  42 compute nodes are consolidated into 11 twin-pair Intel E5-Series DALCO
-  r2264i4t 2U scalable compute modules, each containing 4 compute
-  nodes based  on two Intel Xeon E5-2670 processors operating  at 2.6\,GHz equipped with 64GB of DDR3 1600\,MHz RAM and connected  via InfiniBand Mellanox SX6036 and FDR switches. 
-  For our  experiments, a  full rack constituted  of 42 standard  compute nodes
+  %The 2 login and 42 compute nodes %are consolidated into 11 twin-pair Intel E5-Series DALCO
+  %r2264i4t 2U scalable compute modules, each containing 4 compute
+  %nodes based 
+  %consisting of 
+  Each node is comprises two Intel Xeon E5-2670 processors operating  at 2.6\,GHz equipped with 64GB of DDR3 1600\,MHz RAM and connected  via InfiniBand Mellanox SX6036 and FDR switches. 
+  For our  experiments, a full rack constituted of 42 compute nodes
   was used. \pilat  is based  on  Intel's second-generation
@@
 \item \tinto is  a heterogeneous cluster composed of
-  28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each of them is composed dual-socket Intel Xeon E5645 hexa-core processors 
-  running at 2.40\,GHz, equipped with 24\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS\-3600 switch.
+  28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each node is composed of two Intel Xeon E5645 hexa-core processors 
+  running at 2.4\,GHz, equipped with 24\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS\-3600 switch.
 
@@
 We present here the two different  frameworks deployed on the testbed HPC systems to
-measure  performance and  power  consumption of the baseline execution of \cosmoart.
+measure  performance, power  dissipation and energy consumption of the baseline execution of \cosmoart.
 
@@
 
-For our experiments at Jaume I University (UJI), we employed a version
+For experiments at UJI on \tinto, we employed a version
 of the integrated framework presented in~\cite{energy13} that employs the \pmlib library in
@@
 
-\cosmoart  is compiled and linked with the Extrae instrumentation library,
-which automatically  instruments the Fortran~90 code of the model, e.g., to record MPI calls.  Next, \cosmoart  is  run on  the  nodes,  thus dissipating  certain
-amount  of power.   These  nodes are connected to power distribution units (PDUs)
-APC 8653 that account for the dissipated power with a sampling frequency of 1\,Hz.
-The power data captured by these PDUs is collected into the tracing server.
-The client,  meanwhile, sends a start/stop  primitive  in  order  to  gather  captured  data  by  the PDUs.
-
-Once  the run  is finished,  a power-performance trace file containing the power 
-data received from the tracing server and the instrumentation post-process is generated.
-This file is finally loaded in Paraver in order to visualise the performance trace and to correlate it with the power profile generated by \cosmoart.
+To employ our framework, \cosmoart is first compiled and linked with the Extrae instrumentation library, which automatically  instruments the Fortran~90 code of the model, e.g., to record MPI calls.
+Next, \cosmoart  is  run on  the  nodes,  thus dissipating  certain amount  of power.
+The  nodes are connected to power distribution units (PDUs)
+APC 8653 that account for the dissipated power with a sampling rate of 1\,Hz.
+The power data captured by these PDUs is collected by a \pmlib tracing server running on a separate machine.
+The client, running on the target nodes, sends start/stop  primitive in order to collect
+the power data during the \cosmoart run.
+Once  the run  is finished,  a power-performance trace file containing the power data is received from the tracing server and the instrumentation post-process is generated.
+A merged trace file is finally loaded in Paraver to visualise and correlate the performance trace and with the power profile generated by \cosmoart.
+This procedure allows to detect power bottlenecks, basically constituted by inefficient waiting methods.
 
@@
 information needed  to reproduce the  energy-to-solution benchmarks of
-\cosmoart, was produced and run  on a 1040 cores using 20~MPI
-tasks  per node on  \monch and  on a  1344 cores  using 16~MPI
-tasks per node on  \pilat.  The calculated region was mapped
+\cosmoart, was produced and run on 1040 cores using 20~MPI
+tasks  per node on  \monch, on 1344 cores  using 16~MPI
+tasks per node on  \pilat and on 192 cores  using 12~MPI
+tasks per node on \tinto.  The calculated region was mapped
 to the participating processors using a 2D-partitioning strategy.  The
-distribution along the $x$ and $y$ coordinates was defined by setting:
-$nprocx=40$  and $nprocy=26$  for \monch  and  $nprocx=28$ and
-$nprocy=24$  as  $nprocx$  is   usually  kept  bigger  than  $nprocy$.
-Besides,  as this  version doesn't  make use  of the  traditional GRIB
+distribution along the $x$ and $y$ coordinates was defined accordingly, keeping in mind
+% $by setting:
+%$nprocx=40$  and $nprocy=26$  for \monch  and  $nprocx=28$ and
+%$nprocy=24$  as  
+to keep $x$  bigger  than  $nprocy$.
+Besides,  as this  version does not  make use  of the  traditional GRIB
 library, we  specified $nprocio=0$  for GRIB I/O.   Hyperthreading was

### 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 In this section, we present our measurement methodology along with the
-metrics considered and discuss energy efficiency and power-performance
-traces of the model system.
+metrics considered and discuss about the performance-power-energy trade-off of the model system on the different platforms stated.
 
@@
 
-For the experiments with \tinto platform, we also analyse the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
+For the experiments with \tinto platform, we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
 Specifically, the OpenMPI 1.6.5 library,
 features two operation modes, blocking and polling, which can be selected before \cosmoart
-is launched. In the polling configuration (the detault mode) ``MPI'' continually polls
+is launched. In the polling configuration (the default mode) ``MPI'' continually polls
 the network interface to check for the completion of an event (e.g. send or receive).
-This usually yields low latency but high CPU utilization. Thus, one can expcet that this mode 
+This usually yields low latency but high CPU utilization. Thus, one can expect that this mode 
 attains the best performance, possibly at the cost a higher energy usage. In the blocking
@@
 (left  Y-axis)  metrics  on  both  platforms.  As  expected,  Xeon  E5
-outperforms Ivy Bridge EP, being  roughly 1.3x faster.  The reason for
+outperforms Ivy Bridge EP, being  roughly 130\,\% faster.  The reason for
 that is  two-fold: \emph{i}) it has  higher clock frequency  than Ivy Bridge
@@
 
-The TTS and ETS results for a full-day simulation on \tinto are shown in Figure~\ref{fig:4}.  
-As expected, the mean energy consumed by COSMO-ART when using the blocking MPI policy is about 
-2\,W less than that consumed by the polling mode. This is due to the blocking-waits performed 
-by the CPU cores while using the blocking mode, in contrast of the polling-waits that take 
-place when polling mode is used. From the point of view of the execution time, there is a 
-slight reduction when the blocking MPI policy is leveraged. One can expect that the costs of 
-the interruptions used in the blocking MPI policy can lead to an increase of the total 
-execution time, however, applications with extensive computing may show better perforamnce with 
-the blocking modes. This plays in favour for our results since the energy consumed is also less 
-by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
+Averaged TTS and ETS results over 10 repetitions of 24h simulation on \tinto are shown in Figure~\ref{fig:4}. Correponding results for average total power are shown in Table~\ref{tab:3}.  
+As expected, the average total power dissipated by \cosmoart while using the blocking MPI policy is about 2\,\% less than that consumed by the polling policy.
+This is due to the power-friendly (blocking) waits performed by the CPU cores while using the blocking MPI mode, in contrast to the polling waits that take place when the polling MPI mode is set.
+From the point of view of the execution time, there is a slight reduction for the blocking MPI mode.
+One can expect that the costs of the interruptions used in the blocking MPI policy can lead to an increase of the total execution time, however, applications with extensive computing, as e.g.
+\cosmoart, may show better performance with the blocking modes.
+This effect works in favor with our results since the energy consumed is also reduced (2\,\%) just by changing the behavior of the OpenMPI library to the blocking policy, i.e, with the \texttt{--mca mpi\_yield\_when\_idle} parameter set in the \texttt{mpirun} command.
 
@@
   \caption{Mean time-to-solution and energy-to-solution on
-    \textsc{Tintorrum} for a 24h simulation using the polling and blocking MPI policies. The 95\% 
-            confidence intervals of Student's $t$ distribution (10 samples) clearly indicates 
+    \tinto for a 24h simulation using the polling and blocking MPI policies. The 95\,\% confidence intervals of Student's $t$ distribution (10 samples) clearly indicates 
     an improvement for blocking policy in both TTS and ETS.}
@@
 profile at a much finer resolution using the attached PDUs in each node. First we run \cosmoart 
-for a full-day simulation and analyse the power consumption. Second we correlate performance 
-with power traces of two 1-hor time frames during the day: mid-day (from 11h-12h) and
-midnight (23h-24h). All the experiments were performed using 192 \cosmoart processes
-mapping 12 processes in each of node.
+for a full-day simulation and analyze the power consumption. Second we correlate performance 
+with power traces of two 1-hour time frames during the day: mid-day (11h--12h) and
+midnight (23h--24h). All the experiments were performed using 192 \cosmoart processes
+m 12 processes in each of node.
 
-% For all the experiments we also analyse the contribution of the MPI library to the energy 
+% For all the experiments we also analyze the contribution of the MPI library to the energy 
 % consumption and how the use of the blocking versus polling message-passing policies can 
@@
 % features two operation modes, blocking and modes, which can be selected before \cosmoart is 
-% launched. In the polling configuration (the detault mode) ``MPI'' continually polls
-% the netwrok interface to check for the completion of an event (e.g. send or receive).
-% This usually yields low latency but high CPU utilization. Thus, can one expcet that this mode attanins the best performance, possibley at the cost a higher energy usage. In the blocking
+% launched. In the polling configuration (the default mode) ``MPI'' continually polls
+% the network interface to check for the completion of an event (e.g. send or receive).
+% This usually yields low latency but high CPU utilization. Thus, can one expect that this mode attains the best performance, possibly at the cost a higher energy usage. In the blocking
 % mode, the CPU is yield to other processes/threads if there are no incoming messages.
 
-In order to obtain an initial overview of the power dissipation of the system model, we have 
-obtained the power profiles using our \pmlib framework of a full-day simulation of \cosmoart 
-leveraging the polling and blocking MPI policies, these results are depicted in the top/bottom 
-plots of Figure~\ref{fig:9}, respectvely. The first observation for 
-the polling MPI policy is the quite plain power profile it generates. However, the first and 
-last hours of the full-day simulation dissipate slightly less power than the mid-day hours. We 
-relate these variations due to the chemical reactions and aerosols, computed by ART, that are 
-taking place during the daylight hours but not during night. In this case power consumption
- is about 3821\,W. Looking at the power profile using the blocking MPI policy, we 
-can easily observe that the pattern is repeated favoring the reduction of power during 
-nighttime hours in a more  noticeable way. The peak power dissipation in this mode is 3768\,W, 
-which means a reduction of 50\,W in total or, in other words, 1\,\% of reductuon in the peak 
-power. Also, we observe power  drops in each hour simulated, basically related to periods in 
-which the processes were performing blocking-waits instead of polling-waits, thus yielding the 
-CPU to a low usage during this phases. Our final observation is the reduction in execution time 
-when the blocking MPI policy is leveraged in contrast with the polling MPI. One can expect that 
-the costs of the interruptions used in the blocking can lead to an increase of the total 
-execution time, however, applications with extensive computing may show better perforamnce with 
-the blocking mode. This plays in favour for our results since the energy consumed is also less 
-by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
+In order to obtain an initial overview of the power dissipation of the system model, we obtained power profiles leveraging our \pmlib power-measurement framework over a full-day simulation of \cosmoart.
+These profiles, obtained with both polling and blocking MPI policies, are depicted in  Figure~\ref{fig:9}.
+The first observation for the polling MPI policy is that the first and last hours of the day dissipate slightly less power than the mid-day hours.
+We relate these variations due to the chemical reactions and aerosols computed by ART, that are taking place during daylight hours but not during night.
+In this case peak power consumption is about 3821\,W.
+Looking at the power profile using the blocking MPI policy, we observe that this pattern is repeated, favoring the reduction of power during night hours in a more noticeable way than that for the polling case.The peak power dissipation during day hours in this mode is about 3768\,W, which means a reduction of around 50\,W with respect to the polling mode.
+Also, we notice systematic power drops in each hour simulated.
+We relate them to periods in which the processes were blocked waiting for incoming messages or synchronizations, thus yielding the CPU to low usage during.
+Our final observation is the reduction of the TTS in a factor of 2\,\% with the blocking MPI policy with respect to the polling one.
 
-To gain more insights about the power-performance behavior of \cosmoart, we also have conducted 
-experiments that show power-performance traces of 1 hour simulation. We limited the simulation 
-time to this bound since the pattern of the computation performed by \cosmoart is repeated each 
-hour, also because the weight of a full-day simulation trace can not be easily handled using 
-our machines and memory available (about 45\,GB per trace file). To soften the problem we 
-shrink the simulations to 1 hour during mid-day (11h-12h) and midnight (23h-24h), in which 
-\cosmoart perform different internal computations, as shown in Figure~\ref{fig:9}. At the same
-time we also leverage the two stated MPI policies: polling and waiting modes.
+To gain more insights about the power-performance behavior of \cosmoart, we obtained power-per\-for\-man\-ce traces so that they allow us to correlate the type of tasks being executed with the total power dissipated at each time (see Figure~\ref{fig:11}).
+For this experiment we limit the simulation time to 1 hour because \emph{i}) the pattern of the computation performed by \cosmoart is repeated hourly and, \emph{ii}) the weight of a full-day simulation trace can not be easily handled using our machines and memory available (each trace file was about 45\,GB).
+To face the problem we shrink the simulations to 1 hour and test a time frame during midday (from 11h to 12h) and during midnight (from 23h to 24h), in which \cosmoart perform different internal computations due to the different chemical reactions that happen during day and night.
+At the same time we also leverage the two already stated MPI policies: polling and blocking.
+
+All the traces in Figure~\ref{fig:11} are divided in different phases.
+In first phase all the processes perform group communications and execute MPI-receive primitives (e.g.
+\texttt{MPI\_Recv}).
+The second performs computation bursts, in between group communications, blocking sends, immediate receives and MPI-wait primitives are executed.
+Finally, group communications interlaced with computation and a significant synchronization phase is performed.
+From the performance point of view, one can notice a slight reduction of the TTS that blocking MPI mode leads over the polling MPI policy (around 2\,\%).
+A reduction of power of about  using the blocking MPI policy is produced only in routines that actually perfom synchronizations, group communications and MPI-receive/-wait primitives.
+Thus, the reduction of the pair power-energy directly depends on the percentage of time in which blocking MPI operations are executed.
+For our 1 hour simulation traces, an average of 50\,\% of the time the processes were blocked.
+Taking into account that  the total power reduction during blocking phases is about 5\,\%, one can compute the reduction of power, i.e., $50\,\% \times 5\,\% = 2.5\%$, which is consistent power reduction obtained for the full-day simulation (2\,\%).
+
+\begin{figure*}[ht]
+  \centering
+  \hspace{0.8cm}
+  \scalebox{0.55}{\input{Figs/24hour.pstex_t}}
+  \caption{24 hours simulation trace using the MPI blocking mode on \tinto.}
+  \label{fig:9}
+\end{figure*}
 
@@
 
-\begin{figure*}[htbf]
-  \centering
-  \hspace{0.8cm}
-  \scalebox{0.5}{\input{Figs/24hour.pstex_t}}
-  \caption{24 hours simulation trace using the MPI blocking mode on \tinto.}
-  \label{fig:9}
-\end{figure*}
-
-\begin{figure*}[htbf]
+\begin{figure*}[ht]
   \centering
   \scriptsize
-  Simulation of 1 hour during midday (from 11h to 12h) leveraging the polling MPI policy\\
+  Simulation of 1 hour during midday (11h--12h) leveraging the polling MPI policy\\
   \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}}\\
-  Simulation of 1 hour during midday (from 11h to 12h) leveraging the blocking MPI policy\\
+  Simulation of 1 hour during midday (11h--12h) leveraging the blocking MPI policy\\
   \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}\\
-  1 hour simulation during midnight (from 23h to 24h) leveraging the polling MPI policy\\
+  1 hour simulation during midnight (23h--24h) leveraging the polling MPI policy\\
   \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}}\\
-  1 hour simulation during midnight (from 23h to 24h) leveraging the blocking MPI policy\\
+  1 hour simulation during midnight (23h--24h) leveraging the blocking MPI policy\\
   \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}\\
-  \begin{tabular}{rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rl}
-& &  & &  & &  & &  \\[-0.15cm]
+  \begin{tabular}{rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rl}
+& &  & &  & &  \\[-0.15cm]
 Computation     & \multicolumn{1}{>{\columncolor[RGB]{  0,  , 255}}p{0.4cm}}{}  & & 
-Waiting a msg.  & \multicolumn{1}{>{\columncolor[RGB]{255,  0  ,0}}p{0.4cm}}{}  & & 
 Block. send     & \multicolumn{1}{>{\columncolor[RGB]{255,  0,174}}p{0.4cm}}{}  & & 
 Synchronization & \multicolumn{1}{>{\columncolor[RGB]{179,  0,  0}}p{0.4cm}}{}  & & 
-Wait/Wait all   & \multicolumn{1}{>{\columncolor[RGB]{235,  0,  0}}p{0.4cm}}{}  \\
-& &  & &  & &  & &  \\[-0.15cm]
+Wait/Wait all   & \multicolumn{1}{>{\columncolor[RGB]{235,255,  0}}p{0.4cm}}{}  \\
+& &  & &  & &  \\[-0.15cm]
 Immediate recv. & \multicolumn{1}{>{\columncolor[RGB]{100,100,177}}p{0.4cm}}{}  & &
-I/O             & \multicolumn{1}{>{\columncolor[RGB]{172,174, 41}}p{0.4cm}}{}  & &
+Waiting a msg.  & \multicolumn{1}{>{\columncolor[RGB]{255,  0  ,0}}p{0.4cm}}{}  & & 
 Group comm.     & \multicolumn{1}{>{\columncolor[RGB]{255,144, 26}}p{0.4cm}}{}  & &
-Others          & \multicolumn{1}{>{\columncolor[RGB]{192,224,  0}}p{0.4cm}}{}  & &
-Idle            & \multicolumn{1}{>{\columncolor[RGB]{117,195,255}}p{0.4cm}}{}  \\
+Others          & \multicolumn{1}{>{\columncolor[RGB]{192,224,  0}}p{0.4cm}}{}  \\
+%I/O             & \multicolumn{1}{>{\columncolor[RGB]{172,174, 41}}p{0.4cm}}{}  & &
+%
+%Idle            & \multicolumn{1}{>{\columncolor[RGB]{117,195,255}}p{0.4cm}}{}  \\
 \end{tabular}
-  \caption{Power-perfomance traces during midday and midnight leveraging the polling and blocking MPI policies on \tinto.}
+  \caption{Power-performance traces during midday and midnight leveraging the polling and blocking MPI policies on \tinto.}
   \label{fig:11}

### 2014-04-25 trunk/paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

@@
   Each node is comprised of two Intel Ivy Bridge EP  E5-2660 ten-core processors
-  operating at  2.2 GHz, equipped with 32\,GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
-  a full rack constituted of 52 nodes was used. \monch is  slated to stay
-  in  service  without  hardware  upgrade  for  the  duration  of  the
-  Exa2Green  project,  allowing   an  identical  configuration  to  be
-  recreated for future assessments of the baseline.\\
+  operating at  2.2 GHz, equipped with 32GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
+  a full rack constituted of 52 nodes was used. % \monch is  slated to stay
+  %in  service  without  hardware  upgrade  for  the  duration  of  the
+  %Exa2Green  project,  allowing   an  identical  configuration  to  be
+  %recreated for future assessments of the baseline.\\
 
@@
   Bridge EP based cluster at  CSCS composed of 42 compute nodes. %and used as Piz Daint pre-post  processing cluster 
-  The 2 login  nodes and
-  the 42 computes nodes are consolidated into 11 twin-pair Intel E5-Series DALCO
-  r2264i4t 2U scalable compute modules, each containing 4 compute
-  nodes based  on two Intel Xeon E5-2670 processors operating  at 2.6\,GHz equipped with 64GB of DDR3 1600\,MHz RAM and connected  via InfiniBand Mellanox SX6036 and FDR switches. 
-  For our  experiments, a  full rack constituted  of 42 standard  compute nodes
+  %The 2 login and 42 compute nodes %are consolidated into 11 twin-pair Intel E5-Series DALCO
+  %r2264i4t 2U scalable compute modules, each containing 4 compute
+  %nodes based 
+  %consisting of 
+  Each node is comprises two Intel Xeon E5-2670 processors operating  at 2.6\,GHz equipped with 64GB of DDR3 1600\,MHz RAM and connected  via InfiniBand Mellanox SX6036 and FDR switches. 
+  For our  experiments, a full rack constituted of 42 compute nodes
   was used. \pilat  is based  on  Intel's second-generation
   processors, conventional in  HPC systems and known to  be more power
-  consuming than Ivy Bridge successor.\\
+  consuming than its Ivy Bridge successor.\\
 
 \item \tinto is  a heterogeneous cluster composed of
-  28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each of them is composed dual-socket Intel Xeon E5645 hexa-core processors 
-  running at 2.40\,GHz, equipped with 24\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS3600 switch.
+  28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each node is composed of two Intel Xeon E5645 hexa-core processors 
+  running at 2.4\,GHz, equipped with 24\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS\-3600 switch.
 
@@
 \citep{Riedonetworks},   that  enable   to  monitor   and   log  power
-consumption  of the IT  infrastructure as  well as  constantly analyse
+dissipation  of the IT  infrastructure as  well as  constantly analyse
 line  voltage,  current, power-factor,  frequency  with 1\,\%  accuracy.

### 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

--- 2014-04-25 trunk/paper_enahpc14/Sec4:Experimental-results.tex
@@
 features two operation modes, blocking and polling, which can be selected before \cosmoart
-is launched. In the polling configuration (the detault mode) ``MPI'' continually polls
+is launched. In the polling configuration (the default mode) ``MPI'' continually polls
 the network interface to check for the completion of an event (e.g. send or receive).
-This usually yields low latency but high CPU utilization. Thus, one can expcet that this mode 
+This usually yields low latency but high CPU utilization. Thus, one can expect that this mode 
 attains the best performance, possibly at the cost a higher energy usage. In the blocking
@@
 \monch's Isola E1 Rack  2 and \pilat' Isola HD total
-power measurements for  1-day or 2-days simulations. On  the Intel Ivy
+power measurements for  1-day or 2-day simulations. On  the Intel Ivy
 Bridge EP  based cluster (i.e.  \monch),  the 1-day simulation
 was issued  only twice due  to usage restrictions. As  time resolution
-was set to one update every  5 minutes for power sampling, the average
-power consumption was computed by considering 6 values for each single
+was set to one update every  5 minutes for power sampling, the of average
+power dissipation was computed by considering 6 values for each single
 run.  On the Intel Xeon E5 based cluster (i.e.  \pilat), the
-1-day simulation  was issued  four times and  a 2-days run  only once.
-Similarly, the average power consumption was computed by considering 4
+1-day simulation  was issued  four times and  a 2-day run  only once.
+Similarly, the average of power dissipation was computed by considering 4
 values  for  each  single  1-day  run  and 9  values  for  the  2-days
@@
   \begin{center}
-    \caption{Average power consumption (W) of the platforms}
+    \caption{Average power dissipation (W) of the platforms}
     \label{tab:3}
@@
   \caption{Mean time-to-solution and energy-to-solution on
-    \textsc{Tintorrum} for a 24h simulation using the polling and blocking MPI policies.}
+    \tinto for a 24h simulation using the polling and blocking MPI policies. The 95\,\% confidence intervals of Student's $t$ distribution (10 samples) clearly indicates 
+    an improvement for blocking policy in both TTS and ETS.}
   \label{fig:4}
@@
 for a full-day simulation and analyze the power consumption. Second we correlate performance 
-with power traces of two 1-hor time frames during the day: mid-day (from 11h-12h) and
-midnight (23h-24h). All the experiments were performed using 192 \cosmoart processes
-mapping 12 processes in each of node.
+with power traces of two 1-hour time frames during the day: mid-day (11h--12h) and
+midnight (23h--24h). All the experiments were performed using 192 \cosmoart processes
+m 12 processes in each of node.
 
@@
 % features two operation modes, blocking and modes, which can be selected before \cosmoart is 
-% launched. In the polling configuration (the detault mode) ``MPI'' continually polls
-% the netwrok interface to check for the completion of an event (e.g. send or receive).
-% This usually yields low latency but high CPU utilization. Thus, can one expcet that this mode attanins the best performance, possibley at the cost a higher energy usage. In the blocking
+% launched. In the polling configuration (the default mode) ``MPI'' continually polls
+% the network interface to check for the completion of an event (e.g. send or receive).
+% This usually yields low latency but high CPU utilization. Thus, can one expect that this mode attains the best performance, possibly at the cost a higher energy usage. In the blocking
 % mode, the CPU is yield to other processes/threads if there are no incoming messages.
 
-In order to obtain an initial overview of the power dissipation of the system model, we have 
-obtained the power profiles using our \pmlib framework of a full-day simulation of \cosmoart 
-leveraging the polling and blocking MPI policies, these results are depicted in the top/bottom 
-plots of Figure~\ref{fig:9}, respectvely. The first observation for 
-the polling MPI policy is the quite plain power profile it generates. However, the first and 
-last hours of the full-day simulation dissipate slightly less power than the mid-day hours. We 
-relate these variations due to the chemical reactions and aerosols, computed by ART, that are 
-taking place during the daylight hours but not during night. In this case peak power 
-dissipation is about 3821\,W. Looking at the power profile using the blocking MPI policy, we 
-can easily observe that the pattern is repeated favoring the reduction of power during 
-nighttime hours in a more  noticeable way. The peak power dissipation in this mode is 3768\,W, 
-which means a reduction of 50\,W in total or, in other words, 1\,\% of reductuon in the peak 
-power. Also, we observe power  drops in each hour simulated, basically related to periods in 
-which the processes were performing blocking-waits instead of polling-waits, thus yielding the 
-CPU to a low usage during this phases. Our final observation is the reduction in execution time 
-when the blocking MPI policy is leveraged in contrast with the polling MPI. One can expect that 
-the costs of the interruptions used in the blocking can lead to an increase of the total 
-execution time, however, applications with extensive computing may show better perforamnce with 
-the blocking mode. This plays in favour for our results since the energy consumed is also less 
-by just changing the behavior of the OpenMPI library from the polling to the blocking policy.
+In order to obtain an initial overview of the power dissipation of the system model, we obtained power profiles leveraging our \pmlib power-measurement framework over a full-day simulation of \cosmoart.
+These profiles, obtained with both polling and blocking MPI policies, are depicted in  Figure~\ref{fig:9}.
+The first observation for the polling MPI policy is that the first and last hours of the day dissipate slightly less power than the mid-day hours.
+We relate these variations due to the chemical reactions and aerosols computed by ART, that are taking place during daylight hours but not during night.
+In this case peak power consumption is about 3821\,W.
+Looking at the power profile using the blocking MPI policy, we observe that this pattern is repeated, favoring the reduction of power during night hours in a more noticeable way than that for the polling case.The peak power dissipation during day hours in this mode is about 3768\,W, which means a reduction of around 50\,W with respect to the polling mode.
+Also, we notice systematic power drops in each hour simulated.
+We relate them to periods in which the processes were blocked waiting for incoming messages or synchronizations, thus yielding the CPU to low usage during.
+Our final observation is the reduction of the TTS in a factor of 2\,\% with the blocking MPI policy with respect to the polling one.
 
-To gain more insights about the power-performance behavior of \cosmoart, we also have conducted 
-experiments that show power-performance traces of 1 hour simulation. We limited the simulation 
-time to this bound since the pattern of the computation performed by \cosmoart is repeated each 
-hour, also because the weight of a full-day simulation trace can not be easily handled using 
-our machines and memory available (about 45\,GB per trace file). To soften the problem we 
-shrink the simulations to 1 hour during mid-day (11h-12h) and midnight (23h-24h), in which 
-\cosmoart perform different internal computations, as shown in Figure~\ref{fig:9}. At the same
-time we also leverage the two stated MPI policies: polling and waiting modes.
+To gain more insights about the power-performance behavior of \cosmoart, we obtained power-per\-for\-man\-ce traces so that they allow us to correlate the type of tasks being executed with the total power dissipated at each time (see Figure~\ref{fig:11}).
+For this experiment we limit the simulation time to 1 hour because \emph{i}) the pattern of the computation performed by \cosmoart is repeated hourly and, \emph{ii}) the weight of a full-day simulation trace can not be easily handled using our machines and memory available (each trace file was about 45\,GB).
+To face the problem we shrink the simulations to 1 hour and test a time frame during midday (from 11h to 12h) and during midnight (from 23h to 24h), in which \cosmoart perform different internal computations due to the different chemical reactions that happen during day and night.
+At the same time we also leverage the two already stated MPI policies: polling and blocking.
+
+All the traces in Figure~\ref{fig:11} are divided in different phases.
+In first phase all the processes perform group communications and execute MPI-receive primitives (e.g.
+\texttt{MPI\_Recv}).
+The second performs computation bursts, in between group communications, blocking sends, immediate receives and MPI-wait primitives are executed.
+Finally, group communications interlaced with computation and a significant synchronization phase is performed.
+From the performance point of view, one can notice a slight reduction of the TTS that blocking MPI mode leads over the polling MPI policy (around 2\,\%).
+A reduction of power of about  using the blocking MPI policy is produced only in routines that actually perfom synchronizations, group communications and MPI-receive/-wait primitives.
+Thus, the reduction of the pair power-energy directly depends on the percentage of time in which blocking MPI operations are executed.
+For our 1 hour simulation traces, an average of 50\,\% of the time the processes were blocked.
+Taking into account that  the total power reduction during blocking phases is about 5\,\%, one can compute the reduction of power, i.e., $50\,\% \times 5\,\% = 2.5\%$, which is consistent power reduction obtained for the full-day simulation (2\,\%).
+
+\begin{figure*}[ht]
+  \centering
+  \hspace{0.8cm}
+  \scalebox{0.55}{\input{Figs/24hour.pstex_t}}
+  \caption{24 hours simulation trace using the MPI blocking mode on \tinto.}
+  \label{fig:9}
+\end{figure*}
 

### 2014-04-25 trunk/paper_enahpc14/Conclusion.tex
Manuel's changes

--- 2014-04-25 paper_enahpc14/Conclusion.tex
@@
 Ivybridge  (2.2  GHz) system,  although  the  two  metrics are  indeed
-strongly  correlated.   A   tool  utilising  Paraver/Extrae  with  new
-extensions  for power  consumption  was successfully  applied to  this
-non-trivial application.  The  resulting profiles indicate that simple
-changes, such as making use of an MPI version based on blocking rather
-than polling, can reduce power consumption significantly.
+strongly  correlated. On the other hand, the usage of energy-friendly MPI waiting techniques analzyed with our power-performance framework over a series of runs of \cosmoart on \tinto and
+power traces demonstrate that it is possible to reduce both
+power dissipation and energy-to-solution while maintaining (or even decreasing) the time-to-solution. The  resulting profiles indicate that simple
+changes, such as making use of the blocking MPI policy rather
+than the polling is possible to slightly reduce both power dissipation and energy consumption (by 2\,\%).
 

### 2014-04-25 trunk/paper_enahpc14/Introduction.tex
Manuel's changes

@@
 major  climate  changes  since   the  mid-20$^{th}$  Century,  one  factor
-overwhelmingly  affects  the  uncertainty  in  defining  human-induced
+overwhelmingly  affects  the  uncertainty  in  determining  human-induced
 radiative  forcing:  the  effects   of  aerosols  in  the  atmosphere.
@@
 have  shorter  atmospheric lifetimes,  aerosols  significantly modify  the
-global   radiation   budget   \citep{IPCC-2013}.   Enhanced   aerosols
+global   radiation   budget   \cite{IPCC-2013}.   Enhanced   aerosols
 concentrations impact the climate  system by scattering and absorption
@@
 albedo,  causing  a  negative  indirect  radiative  forcing.   Despite
-considerable  progress in  global aerosol  modelling \citep{Mann-2013}
+considerable  progress in  global aerosol  modelling \cite{Mann-2013}
 and  measurement-based  assessments,  large  uncertainties  remain  in
-current  estimates  of  aerosol radiative  forcing  \citep{Myhre-2013,
-IPCC-2013,  Lee-2013,   Randles-2013,  Rosenfeld-2013,  Sherwood-2013,
-Stier-2013}.
+current  estimates  of  aerosol radiative  forcing  \cite{Myhre-2013,
+IPCC-2013, Lee-2013,  Sherwood-2013,  Stier-2013}.
 
@@
 gases),   developed   at  the   Karlsruhe   Institute  of   Technology
-(KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
+(KIT)  \cite{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
 scale modelling system is coupled  with the operational  weather forecast
-model  \textsc{Cosmo}  \citep{Baldauf-2011},  jointly developed  by  a
-consortium of European weather centers and also used in a climate version
+model  \textsc{Cosmo}  \cite{Baldauf-2011},  jointly developed and used by  a
+consortium of European weather centers, as well as utilized in a climate version
 by the wider research community.
 
-The extended  atmospheric model \textsc{Cosmo-art}  is com\-put\-ationally
+The extended  atmospheric model \cosmoart  is com\-put\-ationally
 much  more  demanding than  \textsc{Cosmo}  since  a  large number  of
 additional  tracers and  processes have  to be  considered.   Thus the
-model  is currently  severely-limited  in terms  of applicability  and
+model  is currently  severely limited  in terms  of applicability  and
 expensive in terms of energy consumption.  Although \textsc{Cosmo} has
-recently been ported to GPUs \citep{Gysi-2014, Lapillonne-2014} within
+recently been ported to GPUs \cite{Gysi-2014, Lapillonne-2014} 
+within
 the framework of the  High Performance and High Productivity Computing
-(HP2C) Initiative  \citep{HP2C} to optimise it  for computational and
+(HP2C) Initiative  \cite{HP2C} to optimize it  for computational and
 energy efficiency,  significant investments in ART  are still required
-to take it  to a similar level.  The  efficiency of \textsc{Cosmo-art}
-is being  addressed in the  EU Exa2Green project  \citep{EXA2GREEN} to
+to take it  to a similar level.  The  efficiency of \cosmoart
+is being  addressed in the  EU Exa2Green project  \cite{EXA2GREEN} to
 deliver a  prototype code, which  provides an energy efficiency  of at
-least five times of the  baseline value.  Such an implementation would
+least five times the  baseline value.  Such an implementation would
 allow  the  community  to  investigate critical  questions  at  higher
@@
 
-This  work is  organized as  follows: in  Sec.~\ref{sec:1} we  give an
-overview  of  related  work,   then  in  Sec.~\ref{sec:2}  we  briefly
-introduce \textsc{Cosmo-art}  and specify its  technical setup related
+This  work is  organized as  follows: in  Section~\ref{sec:1} we  give an
+overview  of  related  work,   then  in  Section~\ref{sec:2}  we  briefly
+introduce \cosmoart  and specify its  technical setup related
 to  the investigated  performance and  energy evaluation  methods.  In
-Sec.~\ref{sec:3}   we  describe   HPC  platforms,   power  measurement
+Section~\ref{sec:3}   we  describe   HPC  platforms,   power  measurement
 equipments   and  software  environment   employed  to   conduct  this
-benchmarking study.  Sec.~\ref{sec:4}  presents performance and energy
+benchmarking study.  Section~\ref{sec:4}  presents performance and energy
 requirements  of the  baseline on  these architectures  and highlights
 areas where improvements will be necessary for the subsequent baseline
-refactoring.   Finally,  we  conclude in Sec.~\ref{concl} with some 
+refactoring.   Finally,  we  conclude in Section~\ref{concl} with some 
 implications for the Exa2Green project and give an outlook for future

### 2014-04-25 trunk/paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes

@@
 \item \monch is a 10-rack NEC-provided clus\-ter at CSCS composed
-  of 312 compute  nodes. %and utilised by members  of the Swiss Platform
-  %for Advanced Scientific  Computing (PASC) project~\citep{PASC}
+  of 312 compute  nodes. %and utilized by members  of the Swiss Platform
+  %for Advanced Scientific  Computing (PASC) project~\cite{PASC}
   Each node is comprised of two Intel Ivy Bridge EP  E5-2660 ten-core processors
-  operating at  2.2 GHz, equipped with 32GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
+  operating at  2.2\,GHz, equipped with 32\,GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
   a full rack constituted of 52 nodes was used. % \monch is  slated to stay
@@
   %consisting of 
-  Each node is comprises two Intel Xeon E5-2670 processors operating  at 2.6\,GHz equipped with 64GB of DDR3 1600\,MHz RAM and connected  via InfiniBand Mellanox SX6036 and FDR switches. 
+  Each node is comprises two Intel Xeon E5-2670 processors operating  at 2.6\,GHz equipped with 64\,GB of DDR3 1600\,MHz RAM and connected  via InfiniBand Mellanox SX6036 and FDR switches. 
   For our  experiments, a full rack constituted of 42 compute nodes
@@
   processors, conventional in  HPC systems and known to  be more power
-  consuming than its Ivy Bridge successor.\\
+  consuming than its Ivy Bridge successor.
 
@@
   28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each node is composed of two Intel Xeon E5645 hexa-core processors 
-  running at 2.4\,GHz, equipped with 24\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS\-3600 switch.
+  running at 2.4\,GHz, equipped with 24\,GB  of DDR3 1333\,MHz and connected via InfiniBand QDR with a Mellanox MTS\-3600 switch.
 
@@
 E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
-high   quality   electricity  meters   released   by  Riedo   Networks
-\citep{Riedonetworks},   that  enable   to  monitor   and   log  power
-dissipation  of the IT  infrastructure as  well as  constantly analyse
+high   quality   electricity  meters   released   by  Riedo   Networks, 
+  that  enable   to  monitor   and   log  power
+dissipation  of the IT  infrastructure as  well as  constantly analyze
 line  voltage,  current, power-factor,  frequency  with 1\,\%  accuracy.
-Using  reliable narrowband  powerline communication  (PLC) technology,
+Using  reliable narrowband  power-line communication  (PLC) technology,
 all  metering and  power  quality  data from  each  IPS are  centrally
@@
 available via  SNMP, HTTP and TELNET  through the built-in  Fast Ethernet
-port.   Time  synchronisation  is  guaranteed by  using  NTP  servers.
+port.   Time  synchronization  is  guaranteed by  using  NTP  servers.
 Measured  data is  accessed  through the  open  source Cacti  software
@@
 combination with Extrae and Paraver, which  are profiling/tracing and
-visualisation tools, respectively.
+visualization tools, respectively.
 
@@
 Next, \cosmoart  is  run on  the  nodes,  thus dissipating  certain amount  of power.
-The  nodes are connected to power distribution units (PDUs)
-APC 8653 that account for the dissipated power with a sampling rate of 1\,Hz.
+The  nodes are connected to APC 8653 Power Distribution Units (PDUs)
+that account for the dissipated power with a sampling rate of 1\,Hz.
 The power data captured by these PDUs is collected by a \pmlib tracing server running on a separate machine.
-The client, running on the target nodes, sends start/stop  primitive in order to collect
+The client, running on the target nodes, sends start/stop  primitives in order to collect
 the power data during the \cosmoart run.
 Once  the run  is finished,  a power-performance trace file containing the power data is received from the tracing server and the instrumentation post-process is generated.
-A merged trace file is finally loaded in Paraver to visualise and correlate the performance trace and with the power profile generated by \cosmoart.
-This procedure allows to detect power bottlenecks, basically constituted by inefficient waiting methods.
+A merged trace file is finally loaded in Paraver to visualize and correlate the performance trace and with the power profile generated by \cosmoart.
+This procedure allows us to easily detect power bottlenecks, basically constituted by inefficient waiting methods.
 
@@
 performance.
-In  addition, we installed the following implementations of MPI: MVAPICH2 1.9 on \monch and \pilat and  OpenMPI 1.6.5 on \tinto.
+In  addition, we installed the following implementations of MPI: MVAPICH2 1.9 on \monch and \pilat, and  OpenMPI 1.6.5 on \tinto.
 On the other hand, we  used HDF5 1.8.12 and NetCDF  (4.3.1) libraries for  the management  of extremely  large and
@@
 to the participating processors using a 2D-partitioning strategy.  The
-distribution along the $x$ and $y$ coordinates was defined accordingly, keeping in mind
+distribution along the $x$ and $y$ coordinates was defined accordingly depeding on the selected platform.
 % $by setting:
@@
 %$nprocy=24$  as  
-to keep $x$  bigger  than  $nprocy$.
+%to keep $x$  bigger  than  $nprocy$.
 Besides,  as this  version does not  make use  of the  traditional GRIB

### 2014-04-25 trunk/paper_enahpc14/paper.tex
Manuel's changes

@@
 %
+\usepackage[numbers]{natbib}
+
+\usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{graphicx}
-\usepackage[T1]{fontenc}
-\usepackage{hyperref}
-\usepackage[round, sort&compress, numbers]{natbib}
 \usepackage{multirow}
-\usepackage{graphicx}
-\usepackage{units}
-\usepackage{color}
+\usepackage{rotating}
+\usepackage{algorithm}
+\usepackage{algorithmic}
 \usepackage{colortbl}
+\usepackage{url}
 \usepackage{xspace}
-\usepackage{epsfig}
+
 %
@@
 \title{Evaluating the Performance and Energy Efficiency of the
-  \textsc{COSMO-ART} model system}
+  \textsc{COSMO-ART} Model System}
 %\thanks{}
@@
   motivation  is to  improve  corresponding code  sections to  sustain
-  performance  while minimising energy-to-solution.   This preliminary
+  performance  while minimizing energy-to-solution.   This preliminary
   work  sets the  basis for  subsequent studies  to  tackle challenges
   related  to  energy  efficient  high performance  computing  in  the
-  framework of the Exa2Green project \citep{EXA2GREEN}.
+  framework of the Exa2Green project \cite{EXA2GREEN}.
 
-\keywords{High performance computing  \and Energy-aware computing \and
-  Green computing  \and Numerical weather  prediction \and Atmospheric
-  chemistry  \and  Aerosol  modelling  \and  Profiling  methods  \and
-  Benchmark  analysis  \and   \textsc{COSMO-ART}  coupled  model  \and
-  Time-to-solution \and Energy-to-solution \and Xeon processor}
+\keywords{High performance computing  \and Energy-aware computing
+  %Green computing  
+  \and Numerical weather  prediction \and Atmospheric
+  chemistry  \and  Aerosol  modelling \and
+  Benchmark  analysis  \and 
+  \cosmoart  coupled  model \and  
+  Power-performance profiling and tracing tools  \and
+  Energy-saving techniques}
 % \PACS{PACS code1 \and PACS code2 \and more} 
@@
 
-\section{Related work}
+\section{Related Work}
 \label{sec:1}
@@
 
-\section{The \textsc{COSMO-ART} model system}
+\section{The \textsc{COSMO-ART} Model System}
 \label{sec:2}
@@
 
-\section{Environment setup}
+\section{Environment Setup}
 \label{sec:3}
@@
 
-\section{Experimental results}
+\section{Experimental Results}
 \label{sec:4}
@@
 
-\section{Conclusion}
+\section{Concluding Remarks}
 \label{concl}
@@
 Programme Future and Emerging Technologies (FET) Proactive Initiative:
-Minimising Energy  Consumption of Computing to the  Limit (MINECC). We
+Minimizing Energy  Consumption of Computing to the  Limit (MINECC). We
 also gratefully acknowledge the High Performance and High Productivity
-Computing Initiative  \citep{HP2C} for results that  will be leveraged
+Computing Initiative  \cite{HP2C} for results that  will be leveraged
 in subsequent code refactoring.

### 2014-04-25 paper_enahpc14/Sec1:Related.tex
Manuel's changes

--- 2014-04-23 paper_enahpc14/Sec1:Related.tex
@@
-\begin{itemize}
-\item Application centric energy-efficiency study of distributed
-  multi-core and hybrid {CPU-GPU} systems \citep{Cumming-2014}
-\item Stella: A domain-specific language and tool for structured grid
-  methods \citep{Gysi-2014}
-\item Using compiler directives to port large scientific applications
-  to GPUs: An example from atmospherci science \citep{Lapillonne-2014}
-\item An analysis of energy-optimized lattice-Boltzmann CFD
-  simulations from the chip to the highly parallel level
-  \citep{Wittmann-2013}
-\item Energy efficiency vs. performance of the numerical solution of
-  PDEs: An application study on a low-power ARM-based cluster
-  \citep{Goddeke-2013}
-\item Analysis of Computing and Energy Performance of Multicore, NUMA,
-  and Manycore Platforms for an Irregular Application
-  \citep{Castro-2013}
-\item Exploring performance and power properties of modern multi-core
-  chips via simple machine models \citep{Hager-2013}
-\end{itemize}
+While the  TOP500 list was  introduced over 20  years ago to  rank the
+performance of HPC systems  worldwide, it is quite recently that energy
+efficiency  become  a  critical  constraint  in the  way  to  exascale
+computing.   Since 2007,  the Green500  power  measurement methodology
+encourages the design,  procurement and management of energy-efficient
+infrastructures to contain performance in an affordable and competitive
+power  envelope.   However,  the state-of-the-art  research  assessing
+performance  and energy  efficiency  of applications  is still scarce.
+
+\emph{Padoin et al.}   \cite{Padoin-2013} investigate performance and
+power  consumption  of  an  agroforestry  application  and  show  that
+changing workload  can drastically improve energy efficiency  of CPU +
+GPU heterogeneous architectures.
+
+\emph{Ou  Pang et  al.}   \cite{Ou-2012} compare  ARM  and Intel  x86
+workstations  clusters  and   conclude  that  ARM-based  clusters  are
+advantageous  with   lightweight  applications  in   terms  of  energy
+efficiency.
+
+\emph{G\"oddeke et al.}  \cite{Goddeke-2013} evaluate weak and strong
+scalability of PDE solvers on a cluster of 96 ARM dual-core processors
+and demonstrate  that the ARM-based  cluster can be more  efficient in
+terms of energy-to-solution compared to x86-based cluster.
+
+\emph{Wittmann  et  al.}   \cite{Wittmann-2013}  perform  a  thorough
+analysis of  a lattice-Boltzmann method based CFD  simulation on Intel
+Sandy Bridge  processors   and  show   extrapolated  results   on  a
+petascale-class machine.
+
+% \emph{Cumming  et  al.}   \cite{Cumming-2014}  present a  simple  and
+% practical  methodology  looking at  energy  minimization  that can  be
+% applied to various applications.

### 2014-04-25 paper_enahpc14/Sec2:COSMO-ART.tex
Manuel's changes

@@
 
-The  model  system COSMO-ART  described  in  \citep{Vogel-2009}, is  a
-regional  to  continental scale  model  coupled  online  to the  COSMO
-numerical weather  prediction and climate  model \citep{Baldauf-2011}.
-It  incorporates  sophisticated  modules  for  gaseous  chemistry  and
-aerosol dynamics  and allows the online calculation  of reactive trace
-substances and their interaction  with the atmosphere.  Detailed model
-description   can  be   found   in  \citep{Bangert-2012,   Knote-2011,
-  Knote-2013}.
+The model  system \cosmoart described  in \cite{Vogel-2009},
+is  a  regional to  continental  scale  model  coupled online  to  the
+\textsc{Cosmo}   numerical  weather   prediction  and   climate  model
+\cite{Baldauf-2011}.   It   incorporates  sophisticated  modules  for
+gaseous  chemistry   and  aerosol  dynamics  and   allows  the  online
+calculation of  reactive trace  substances and their  interaction with
+the  atmosphere.    Detailed  model   description  can  be   found  in
+\cite{Bangert-2012, Knote-2011, Knote-2013}.
 
@@
 Establishing effective energy performance benchmarking of a code under
-intense development such as COSMO-ART is a challenging task because of
-the  absolute necessity that  results must  be reproducible  within an
-expected variance for the duration of the Exa2Green project. To define
-a baseline,  it was necessary  to find a run-configuration  capable of
-being  recreated in  all subsequent  versions.  Here  we give  a brief
-overview of the model setup  for our performance and energy efficiency
-evaluation.
+intense development  such as \cosmoart is  a challenging task
+because of  the absolute necessity  that results must  be reproducible
+within  an  expected  variance  for  the  duration  of  the  Exa2Green
+pro\-je\-ct.  To   define  a  baseline,   it  was  necessary  to   find  a
+run-configuration  capable  of   being  recreated  in  all  subsequent
+versions.  Here  we give a brief  overview of the model  setup for our
+performance and energy efficiency evaluation.
 
 Three-dimensional simulations are performed over large parts of Europe
-and the Mediterranean~Sea for April 13th 2010 in order to benefit from
-around  half  day  of  sun  exposure and  therefore  ensure  a  proper
+and the Mediterranean~Sea for April 13$^{th}$ 2010, near the equinox, in order to approximately
+have a half  day  of  sun  exposure and  therefore  ensure  a typical
 activation of the chemistry cycle.  The calculation domain corresponds
@@
 in  both directions  and 40~vertical  layers.  These  24-hour forecast
-simulations  are not  preceded  by  a spin-up  phase  to build-up  the
-simulated gaseous and  aerosol concentrations.  This condition entails
-that at the model initialization,  there is a period of adjustment for
-the model  to reach its equilibrium  state and minimize  the effect of
+simulations  are not  preceded  by  a spin-up  phase  to build up  the
+simulated gaseous and  aerosol concentrations.  This condition means 
+that the model initialization needs a period of adjustment to reach its equilibrium  state and minimize  the effect of
 initial conditions for gases and aerosols on the model predictions.
@@
 
-The   considered  COSMO-ART   model  system   is  configured   with  a
-semi-Lagrangian    horizontal   advection    scheme    with   tricubic
+The considered  \cosmoart model  system is configured  with a
+semi-Lagrangian    horizontal   advection    sche\-me    with   tricubic
 interpolation  and selective filling  diffusion option  in combination
 with   the   dynamical    core   using   Runge-Kutta   time   stepping
-\citep{COSMO-PartI-2011}.    It  also   makes  use   of   the  Kinetic
-PreProcessor solver (KPP) for  the resolution of atmospheric chemistry
-ordinary  differential equations \citep{Damian-2002}.   Concerning the
+\cite{COSMO-PartI-2011}.    It  also   makes  use   of   the  Kinetic
+PreProcessor solver (KPP) for  the solution of atmospheric chemistry
+ordinary  differential equations \cite{Damian-2002}.   Concerning the
 modelling  of  wet  deposition  in  aerosols, the  baseline  has  only
@@
 (rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
-parameterisations,   precipitation  formation   is   performed  by   a
+parameterizations,   precipitation  formation   is   performed  by   a
 two-moment cloud microphysics.

### 2014-04-25 paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes

@@
 
-For the experiments with \tinto platform, we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
+For the experiments on \tinto platform, we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
 Specifically, the OpenMPI 1.6.5 library,
@@
 attains the best performance, possibly at the cost a higher energy usage. In the blocking
-mode, the CPU is yield to other processes/threads if there are no incoming messages.
+mode, the CPU is yield to other processes/threads if there are no incoming messages~\cite{PDP12}.
 
-\subsection{Time-power-energy analysis of \cosmoart}
+\subsection{Time-power-energy analysis}
 \label{subsec:4.2}
-
-\begin{figure}[htbf]
-  \begin{center}
-    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
-    \caption{\monch: Isola E1 Rack 2 Total Power.}
-    \label{fig:1}
-  \end{center}
-\end{figure}
-
-\begin{figure}[htbf]
-  \begin{center}
-    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
-    \caption{\pilat: Isola HD Total Power.}
-    \label{fig:2}
-  \end{center}
-\end{figure}
 
@@
 was issued  only twice due  to usage restrictions. As  time resolution
-was set to one update every  5 minutes for power sampling, the average
-power consumption was computed by considering 6 values for each single
+was set to one update every 5 minutes for power sampling, the averaged
+power dissipation was computed by considering 6 values for each single
 run.  On the Intel Xeon E5 based cluster (i.e.  \pilat), the
 1-day simulation  was issued  four times and  a 2-day run  only once.
-Similarly, the average power consumption was computed by considering 4
+Similarly, the averaged power dissipation was computed by considering 4
 values  for  each  single  1-day  run  and 9  values  for  the  2-days
@@
 
+\begin{figure}[htbf]
+  \centering
+    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
+    \caption{Isola E1 Rack 2 total power of \monch.}
+    \label{fig:1}
+\end{figure}
+
+\begin{figure}[htbf]
+  \centering
+    \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
+    \caption{Isola HD total power dissipation of \pilat.}
+    \label{fig:2}
+\end{figure}
+
 \begin{table}[htbf]
-  \begin{center}
-    \caption{Average power consumption (W) of the platforms}
+  \centering
+    \caption{Averaged power dissipation (W) of the platforms.}
     \label{tab:3}
@@
     \end{tabular}
-  \end{center}
   \end{table}
@@
 
-Averaged TTS and ETS results over 10 repetitions of 24h simulation on \tinto are shown in Figure~\ref{fig:4}. Correponding results for average total power are shown in Table~\ref{tab:3}.  
-As expected, the average total power dissipated by \cosmoart while using the blocking MPI policy is about 2\,\% less than that consumed by the polling policy.
+Averaged TTS and ETS results over 10 repetitions of 24h simulation on \tinto are shown in Figure~\ref{fig:4}.
+The corresponding results for the averaged total power are also shown in Table~\ref{tab:3}.
+As expected, the average total power dissipation by \cosmoart while using the blocking MPI policy is about 2\,\% less than that consumed by the polling policy.
 This is due to the power-friendly (blocking) waits performed by the CPU cores while using the blocking MPI mode, in contrast to the polling waits that take place when the polling MPI mode is set.
 From the point of view of the execution time, there is a slight reduction for the blocking MPI mode.
-One can expect that the costs of the interruptions used in the blocking MPI policy can lead to an increase of the total execution time, however, applications with extensive computing, as e.g.
+One can expect that the costs of the interruptions used in the blocking MPI policy can lead to an increase of the execution time, however, applications with extensive computing, as e.g.
 \cosmoart, may show better performance with the blocking modes.
@@
 \begin{figure}[htbf]
+  \centering
   \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-0.eps}
@@
 \begin{figure}[htbf]
+
+  \centering
   \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-5.eps}
+  \vspace{-1cm}
   \caption{Mean time-to-solution and energy-to-solution on
@@
 
-\subsection{Power-performance tracing on \tinto}
-\label{subsec:4.3}
-
-A full tracing experiment is conducted on \tinto in order to capture an overall power
-profile at a much finer resolution using the attached PDUs in each node. First we run \cosmoart 
-for a full-day simulation and analyze the power consumption. Second we correlate performance 
-with power traces of two 1-hour time frames during the day: mid-day (11h--12h) and
-midnight (23h--24h). All the experiments were performed using 192 \cosmoart processes
-m 12 processes in each of node.
-
-% For all the experiments we also analyze the contribution of the MPI library to the energy 
-% consumption and how the use of the blocking versus polling message-passing policies can 
-% potentially render energy savings. Specifically, the OpenMPI 1.6.5 library installed in \tinto, 
-% features two operation modes, blocking and modes, which can be selected before \cosmoart is 
-% launched. In the polling configuration (the default mode) ``MPI'' continually polls
-% the network interface to check for the completion of an event (e.g. send or receive).
-% This usually yields low latency but high CPU utilization. Thus, can one expect that this mode attains the best performance, possibly at the cost a higher energy usage. In the blocking
-% mode, the CPU is yield to other processes/threads if there are no incoming messages.
-
-In order to obtain an initial overview of the power dissipation of the system model, we obtained power profiles leveraging our \pmlib power-measurement framework over a full-day simulation of \cosmoart.
-These profiles, obtained with both polling and blocking MPI policies, are depicted in  Figure~\ref{fig:9}.
-The first observation for the polling MPI policy is that the first and last hours of the day dissipate slightly less power than the mid-day hours.
-We relate these variations due to the chemical reactions and aerosols computed by ART, that are taking place during daylight hours but not during night.
-In this case peak power consumption is about 3821\,W.
-Looking at the power profile using the blocking MPI policy, we observe that this pattern is repeated, favoring the reduction of power during night hours in a more noticeable way than that for the polling case.The peak power dissipation during day hours in this mode is about 3768\,W, which means a reduction of around 50\,W with respect to the polling mode.
-Also, we notice systematic power drops in each hour simulated.
-We relate them to periods in which the processes were blocked waiting for incoming messages or synchronizations, thus yielding the CPU to low usage during.
-Our final observation is the reduction of the TTS in a factor of 2\,\% with the blocking MPI policy with respect to the polling one.
-
-To gain more insights about the power-performance behavior of \cosmoart, we obtained power-per\-for\-man\-ce traces so that they allow us to correlate the type of tasks being executed with the total power dissipated at each time (see Figure~\ref{fig:11}).
-For this experiment we limit the simulation time to 1 hour because \emph{i}) the pattern of the computation performed by \cosmoart is repeated hourly and, \emph{ii}) the weight of a full-day simulation trace can not be easily handled using our machines and memory available (each trace file was about 45\,GB).
-To face the problem we shrink the simulations to 1 hour and test a time frame during midday (from 11h to 12h) and during midnight (from 23h to 24h), in which \cosmoart perform different internal computations due to the different chemical reactions that happen during day and night.
-At the same time we also leverage the two already stated MPI policies: polling and blocking.
-
-All the traces in Figure~\ref{fig:11} are divided in different phases.
-In first phase all the processes perform group communications and execute MPI-receive primitives (e.g.
-\texttt{MPI\_Recv}).
-The second performs computation bursts, in between group communications, blocking sends, immediate receives and MPI-wait primitives are executed.
-Finally, group communications interlaced with computation and a significant synchronization phase is performed.
-From the performance point of view, one can notice a slight reduction of the TTS that blocking MPI mode leads over the polling MPI policy (around 2\,\%).
-A reduction of power of about  using the blocking MPI policy is produced only in routines that actually perfom synchronizations, group communications and MPI-receive/-wait primitives.
-Thus, the reduction of the pair power-energy directly depends on the percentage of time in which blocking MPI operations are executed.
-For our 1 hour simulation traces, an average of 50\,\% of the time the processes were blocked.
-Taking into account that  the total power reduction during blocking phases is about 5\,\%, one can compute the reduction of power, i.e., $50\,\% \times 5\,\% = 2.5\%$, which is consistent power reduction obtained for the full-day simulation (2\,\%).
-
 \begin{figure*}[ht]
   \centering
-  \hspace{0.8cm}
-  \scalebox{0.55}{\input{Figs/24hour.pstex_t}}
-  \caption{24 hours simulation trace using the MPI blocking mode on \tinto.}
+  \hspace{0.6cm}
+  \scalebox{0.64}{\input{Figs/24hour.pstex_t}}
+  \caption{24 hours simulation power trace using the polling and blocking MPI policies on \tinto.}
   \label{fig:9}
 \end{figure*}
-
-% \begin{figure*}[htbf]
-%   \centering
-%   Simulation of 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
-%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq0_stat.eps}\\
-%   Percentage of time for the 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
-%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq1_stat.eps}\\
-%   \caption{blabla}
-%   \label{fig:8}
-% \end{figure*}
 
@@
 \end{figure*}
+
+\subsection{Power-performance profiling and tracing}
+\label{subsec:4.3}
+
+A full tracing experiment is conducted on \tinto in order to capture an overall power
+profile at a much finer resolution using the attached PDUs in each node. First we run \cosmoart 
+for a 1-day simulation and analyze the power consumption. Second we correlate performance 
+with power traces of two 1-hour time frames during the day: mid-day (11h--12h) and
+midnight (23h--24h). All the experiments were performed using 192 cores using 12 MPI processes
+per node on \tinto.
+
+% For all the experiments we also analyze the contribution of the MPI library to the energy 
+% consumption and how the use of the blocking versus polling message-passing policies can 
+% potentially render energy savings. Specifically, the OpenMPI 1.6.5 library installed in \tinto, 
+% features two operation modes, blocking and modes, which can be selected before \cosmoart is 
+% launched. In the polling configuration (the default mode) ``MPI'' continually polls
+% the network interface to check for the completion of an event (e.g. send or receive).
+% This usually yields low latency but high CPU utilization. Thus, can one expect that this mode attains the best performance, possibly at the cost a higher energy usage. In the blocking
+% mode, the CPU is yield to other processes/threads if there are no incoming messages.
+
+In order to obtain an initial overview of the power dissipation of the system model, we obtained power profiles leveraging our \pmlib power-measurement framework over a 1-day simulation of \cosmoart.
+These profiles, produced with both polling and blocking MPI policies, are depicted in  Figure~\ref{fig:9}.
+The first observation for the polling MPI policy is that the first and last hours of the day dissipate slightly less power than the mid-day hours.
+We relate these variations due to the different chemical reactions, aerosols, radiations and clouds formation processes, computed by ART, taking place during daylight hours but not during night.
+In this case the peak power consumption is about 3821\,W.
+Looking at the power profile using the blocking MPI policy, we observe that the 1-hour pattern is repeated along the 24-hours, favoring the reduction of power during night hours in a more noticeable way than that with the polling policy.
+The peak power dissipation during day hours of while the blocking MPI policy is leveraged is about 3768\,W, which means a reduction of around 50\,W with respect to the polling mode.
+Also, we notice systematic power drops for each simulated hour.
+We associate them to periods in which the processes were blocked waiting for incoming messages or synchronizations, thus yielding the CPU to low usage.
+Our final observation is the reduction of the TTS by 2\,\% with respect to the blocking MPI policy.
+
+To gain more insights about the power-performance behavior of \cosmoart, we obtained power-per\-for\-man\-ce traces to allow us correlate the type of tasks being executed with the total power dissipation along the time (see Figure~\ref{fig:11}).
+For this experiment we limit the simulation time to 1 hour because \emph{i}) the pattern of the computation performed by \cosmoart is repeated hourly and, \emph{ii}) the weight of a 1-day simulation trace can not be easily handled using our machines and memory available (each trace file was about 45\,GB).
+To face the problem we shrink the simulations to 1 hour and test a time frame during midday (from 11h to 12h) and  midnight (from 23h to 24h), in which \cosmoart perform different internal computations.
+At the same time we leverage the two already stated MPI policies: polling and blocking.
+
+All the traces in Figure~\ref{fig:11} are divided in different computation phases.
+In first phase all the processes perform group communications and execute MPI-receive primitives (e.g.
+\texttt{MPI\_Recv}).
+The second one performs computation bursts, combined in between with group communications, blocking sends, immediate receives and MPI-wait primitives are executed.
+Finally, group communications interlaced with computation and a significant synchronization phase are performed.
+From the performance point of view, one can notice a reduction of 2\,\% in the TTS that blocking MPI mode leads over the polling MPI policy.
+On the other hand an averaged reduction of power dissipation of 2\,\% is attained with the blocking MPI policy only in routines that actually perform synchronizations, group communications and MPI-receive/-wait primitives.
+Thus, the reduction of the power-energy pair directly depends on the percentage of time in which blocking MPI operations are executed.
+For the 1 hour simulation traces, an average of 50\,\% of the time the processes were blocked.
+Taking into account that  the total power reduction during blocking phases is about 5\,\%, one can compute an approximation of the power savings, i.e., $50\,\% \times 5\,\% = 2.5\%$, which is consistent power reduction obtained for the full-day simulation (around 2\,\%).
+However, we expect higher power-energy savings if future architectures tend proportional computing~\citep{Barroso:2007:CEC:1339817.1339894}.
+
+% \begin{figure*}[htbf]
+%   \centering
+%   Simulation of 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
+%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq0_stat.eps}\\
+%   Percentage of time for the 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
+%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq1_stat.eps}\\
+%   \caption{blabla}
+%   \label{fig:8}
+% \end{figure*}

### 2014-05-05 trunk/paper_enahpc14/Sec2:COSMO-ART.tex
cleans up latest version; improves format

@@
 
-The model  system \textsc{Cosmo-art} described  in \citep{Vogel-2009},
-is  a  regional to  continental  scale  model  coupled online  to  the
+The  model  system  \cosmoart  described in  \cite{Vogel-2009},  is  a
+regional   to  continental   scale   model  coupled   online  to   the
 \textsc{Cosmo}   numerical  weather   prediction  and   climate  model
-\citep{Baldauf-2011}.   It   incorporates  sophisticated  modules  for
+\cite{Baldauf-2011}.    It  incorporates  sophisticated   modules  for
 gaseous  chemistry   and  aerosol  dynamics  and   allows  the  online
@@
 the  atmosphere.    Detailed  model   description  can  be   found  in
-\citep{Bangert-2012, Knote-2011, Knote-2013}.
+\cite{Bangert-2012, Knote-2011, Knote-2013}.
 
@@
 Establishing effective energy performance benchmarking of a code under
-intense development  such as \textsc{Cosmo-art} is  a challenging task
-because of  the absolute necessity  that results must  be reproducible
-within  an  expected  variance  for  the  duration  of  the  Exa2Green
-pro\-je\-ct.  To   define  a  baseline,   it  was  necessary  to   find  a
-run-configuration  capable  of   being  recreated  in  all  subsequent
-versions.  Here  we give a brief  overview of the model  setup for our
-performance and energy efficiency evaluation.
+intense development such as \cosmoart is a challenging task because of
+the  absolute necessity that  results must  be reproducible  within an
+expected variance  for the duration of the  Exa2Green pro\-je\-ct.  To
+define  a  baseline, it  was  necessary  to  find a  run-configuration
+capable of being recreated in all subsequent versions.  Here we give a
+brief  overview of  the model  setup  for our  performance and  energy
+efficiency evaluation.
 
 Three-dimensional simulations are performed over large parts of Europe
-and the Mediterranean~Sea for April 13$^{th}$ 2010, near the equinox, in order to approximately
-have a half  day  of  sun  exposure and  therefore  ensure  a typical
-activation of the chemistry cycle.  The calculation domain corresponds
-to the CORDEX-EU-44 domain and is covered by a grid of $222\times 216$
-points with  a horizontal resolution of  $0.22\,^{\circ}$, i.e., 50~km
-in  both directions  and 40~vertical  layers.  These  24-hour forecast
-simulations  are not  preceded  by  a spin-up  phase  to build up  the
-simulated gaseous and  aerosol concentrations.  This condition means 
-that the model initialisation needs a period of adjustment to reach its equilibrium  state and minimise  the effect of
-initial conditions for gases and aerosols on the model predictions.
+and the Mediterranean~Sea for  April 13$^{th}$ 2010, near the equinox,
+in  order  to  approximately have  a  half  day  of sun  exposure  and
+therefore  ensure a typical  activation of  the chemistry  cycle.  The
+calculation  domain  corresponds to  the  CORDEX-EU-44  domain and  is
+covered  by  a  grid  of  $222\times 216$  points  with  a  horizontal
+resolution  of $0.22\,^{\circ}$,  i.e., 50~km  in both  directions and
+40~vertical  layers.   These  24-hour  forecast  simulations  are  not
+preceded  by a spin-up  phase to  build up  the simulated  gaseous and
+aerosol   concentrations.   This  condition   means  that   the  model
+initialization needs  a period of adjustment to  reach its equilibrium
+state  and minimize  the effect  of initial  conditions for  gases and
+aerosols on the model predictions.
 
@@
 
-The considered  \textsc{Cosmo-art} model  system is configured  with a
+The   considered  \cosmoart   model  system   is  configured   with  a
 semi-Lagrangian   horizontal    advection   sche\-me   with   tricubic
@@
 with   the   dynamical    core   using   Runge-Kutta   time   stepping
-\citep{COSMO-PartI-2011}.    It  also   makes  use   of   the  Kinetic
+\cite{COSMO-PartI-2011}.    It   also  makes   use   of  the   Kinetic
 PreProcessor solver  (KPP) for  the solution of  atmospheric chemistry
-ordinary  differential equations \citep{Damian-2002}.   Concerning the
+ordinary  differential equations  \cite{Damian-2002}.   Concerning the
 modelling  of  wet  deposition  in  aerosols, the  baseline  has  only

### 2014-05-05 trunk/paper_enahpc14/Sec3:Environment-setup.tex
cleans up latest version; improves format

@@
 \begin{itemize}
-\item \monch is a 10-rack NEC-provided clus\-ter at CSCS composed
-  of 312 compute  nodes. %and utilized by members  of the Swiss Platform
-  %for Advanced Scientific  Computing (PASC) project~\cite{PASC}
-  Each node is comprised of two Intel Ivy Bridge EP  E5-2660 ten-core processors
-  operating at  2.2\,GHz, equipped with 32\,GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
-  a full rack constituted of 52 nodes was used. % \monch is  slated to stay
-  %in  service  without  hardware  upgrade  for  the  duration  of  the
-  %Exa2Green  project,  allowing   an  identical  configuration  to  be
-  %recreated for future assessments of the baseline.\\
+\item \monch is  a 10-rack NEC-provided clus\-ter at  CSCS composed of
+  312 compute nodes. 
+  % and  utilized  by  members  of  the  Swiss  Platform  for  Advanced
+  % Scientific Computing (PASC) project~\cite{PASC}.
+  Each node is  comprised of two Intel Ivy  Bridge EP E5-2660 ten-core
+  processors  operating  at 2.2\,GHz,  equipped  with  32\,GB of  DDR3
+  1600\,MHz RAM  and connected via InfiniBand Mellanox  SX6036 and FDR
+  switches. For our  experiments, a full rack constituted  of 52 nodes
+  was used. 
+  % \monch is  slated to stay  in service without hardware  upgrade for
+  % the  duration  of  the  Exa2Green project,  allowing  an  identical
+  % configuration  to  be  recreated  for  future  assessments  of  the
+  % baseline.
 
-\item \pilat is a dual-socket eight-core Intel Sandy
-  Bridge EP based cluster at  CSCS composed of 42 compute nodes. %and used as Piz Daint pre-post  processing cluster 
-  %The 2 login and 42 compute nodes %are consolidated into 11 twin-pair Intel E5-Series DALCO
-  %r2264i4t 2U scalable compute modules, each containing 4 compute
-  %nodes based 
-  %consisting of 
-  Each node is comprises two Intel Xeon E5-2670 processors operating  at 2.6\,GHz equipped with 64\,GB of DDR3 1600\,MHz RAM and connected  via InfiniBand Mellanox SX6036 and FDR switches. 
-  For our  experiments, a full rack constituted of 42 compute nodes
-  was used. \pilat  is based  on  Intel's second-generation
-  processors, conventional in  HPC systems and known to  be more power
-  consuming than its Ivy Bridge successor.
+\item \pilat is  a dual-socket eight-core Intel Sandy  Bridge EP based
+  cluster at CSCS composed of 42 compute nodes.  
+  % and used as Piz Daint  pre-post processing cluster The 2 login and
+  % 42  compute  nodes  are   consolidated  into  11  twin-pair  Intel
+  % E5-Series  DALCO  r2264i4t   2U  scalable  compute  modules,  each
+  % containing 4 compute nodes
+  Each node  is comprises two Intel Xeon  E5-2670 processors operating
+  at 2.6\,GHz equipped with 64\,GB of DDR3 1600\,MHz RAM and connected
+  via  InfiniBand   Mellanox  SX6036   and  FDR  switches.    For  our
+  experiments, a full  rack constituted of 42 compute  nodes was used.
+  \pilat   is   based   on   Intel's   second-generation   processors,
+  conventional in  HPC systems  and known to  be more  power consuming
+  than its Ivy Bridge successor.
 
-\item \tinto is  a heterogeneous cluster composed of
-  28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each node is composed of two Intel Xeon E5645 hexa-core processors 
-  running at 2.4\,GHz, equipped with 24\,GB  of DDR3 1333\,MHz and connected via InfiniBand QDR with a Mellanox MTS\-3600 switch.
-
-  % A  full tracing  experiment is  conducted on
-  % \tinto in order to  capture an overall  power profile at a much
-  % finer resolution. In this sense, we also analyze the contribution of the MPI library
-  % to the energy consumption and how the use of the blocking versus polling
-  % message-passing policies can potentially render energy savings.
+\item \tinto is  a heterogeneous cluster composed of  28 compute nodes
+  at UJI.  For our experiments only  a subset of  16 homogeneous nodes
+  was  considered. Each  node  is  composed of  two  Intel Xeon  E5645
+  hexa-core processors  running at  2.4\,GHz, equipped with  24\,GB of
+  DDR3  1333\,MHz and  connected via  InfiniBand QDR  with  a Mellanox
+  MTS\-3600 switch.
 \end{itemize}
@@
 
-We present here the two different  frameworks deployed on the testbed HPC systems to
-measure  performance, power  dissipation and energy consumption of the baseline execution of \cosmoart.
+We present here  the two different frameworks deployed  on the testbed
+HPC  systems  to measure  performance,  power  dissipation and  energy
+consumption of the baseline execution of \cosmoart.
 
@@
 
-Supercomputer  clusters considered  for our  experiments at CSCS of ETH Zurich are equipped with
-E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
-high   quality   electricity  meters   released   by  Riedo   Networks, 
-  that  enable   to  monitor   and   log  power
-dissipation  of the IT  infrastructure as  well as  constantly analyze
-line  voltage,  current, power-factor,  frequency  with 1\,\%  accuracy.
-Using  reliable narrowband  power-line communication  (PLC) technology,
-all  metering and  power  quality  data from  each  IPS are  centrally
-collected  by the E3METER  Data Concentrator,  via the  existing power
-cables thus  avoiding the need for  extra cabling.  This  data is made
-available via  SNMP, HTTP and TELNET  through the built-in  Fast Ethernet
-port.   Time  synchronization  is  guaranteed by  using  NTP  servers.
-Measured  data is  accessed  through the  open  source Cacti  software
-including the E3METER Cacti Plugin  to scan the entire PLC network and
-monitor in real-time the power usage of individual rack, recorded in 5
-minutes interval periods.
+Supercomputer clusters  considered for our experiments at  CSCS of ETH
+Zurich are  equipped with E3METER  Intelligent Power Strips  (IPS) and
+Monitors (IPM)  which are high quality electricity  meters released by
+Riedo Networks,  that enable to  monitor and log power  dissipation of
+the  IT infrastructure  as well  as constantly  analyze  line voltage,
+current, power-factor, frequency  with 1\,\% accuracy.  Using reliable
+narrowband power-line communication (PLC) technology, all metering and
+power  quality data  from  each  IPS are  centrally  collected by  the
+E3METER Data Concentrator, via the existing power cables thus avoiding
+the need  for extra  cabling.  This data  is made available  via SNMP,
+HTTP  and  TELNET  through  the  built-in Fast  Ethernet  port.   Time
+synchronization is guaranteed by  using NTP servers.  Measured data is
+accessed through the open  source Cacti software including the E3METER
+Cacti Plugin to  scan the entire PLC network  and monitor in real-time
+the power  usage of  individual rack, recorded  in 5  minutes interval
+periods.
 
@@
 
-For experiments at UJI on \tinto, we employed a version
-of the integrated framework presented in~\cite{energy13} that employs the \pmlib library in
-combination with Extrae and Paraver, which  are profiling/tracing and
-visualization tools, respectively.
+For  experiments  at UJI  on  \tinto, we  employed  a  version of  the
+integrated  framework presented  in~\cite{energy13}  that employs  the
+\pmlib  library in  combination  with Extrae  and  Paraver, which  are
+profiling/tracing and visualization tools, respectively.
 
-To employ our framework, \cosmoart is first compiled and linked with the Extrae instrumentation library, which automatically  instruments the Fortran~90 code of the model, e.g., to record MPI calls.
-Next, \cosmoart  is  run on  the  nodes,  thus dissipating  certain amount  of power.
-The  nodes are connected to APC 8653 Power Distribution Units (PDUs)
-that account for the dissipated power with a sampling rate of 1\,Hz.
-The power data captured by these PDUs is collected by a \pmlib tracing server running on a separate machine.
-The client, running on the target nodes, sends start/stop  primitives in order to collect
-the power data during the \cosmoart run.
-Once  the run  is finished,  a power-performance trace file containing the power data is received from the tracing server and the instrumentation post-process is generated.
-A merged trace file is finally loaded in Paraver to visualize and correlate the performance trace and with the power profile generated by \cosmoart.
-This procedure allows us to easily detect power bottlenecks, basically constituted by inefficient waiting methods.
+To employ our  framework, \cosmoart is first compiled  and linked with
+the  Extrae instrumentation  library, which  automatically instruments
+the Fortran~90  code of the model,  e.g., to record  MPI calls.  Next,
+\cosmoart  is run  on the  nodes, thus  dissipating certain  amount of
+power. The  nodes are connected  to APC 8653 Power  Distribution Units
+(PDUs) that account  for the dissipated power with  a sampling rate of
+1\,Hz. The power data captured by  these PDUs is collected by a \pmlib
+tracing server running  on a separate machine. The  client, running on
+the target nodes, sends start/stop  primitives in order to collect the
+power  data during  the \cosmoart  run. Once  the run  is  finished, a
+power-performance  trace file  containing the  power data  is received
+from  the  tracing  server  and the  instrumentation  post-process  is
+generated.  A  merged trace  file  is  finally  loaded in  Paraver  to
+visualize  and correlate  the  performance trace  and  with the  power
+profile  generated by \cosmoart.  This procedure  allows us  to easily
+detect power bottlenecks, basically constituted by inefficient waiting
+methods.
 
@@
 
-The \cosmoart  baseline is a  pure MPI-based  Fortran~90 code
-currently  running  on   distributed  multi-core  systems  only.   The
-software stack on the platforms was controlled using the modules
-framework which gives an easy  and flexible mechanism to access to all
-of the  provided compilers, tools  and applications.  For  our initial
-benchmarking,   we  opted   for   the  GNU   compiler  GCC 4.8.1   on
-\monch and GCC 4.8.2 on  \pilat and \tinto) using the -O3 compiler
-flag  in favor  of the  Intel compilers 14.0.1,  which delivered inferior
-performance.
-In  addition, we installed the following implementations of MPI: MVAPICH2 1.9 on \monch and \pilat, and  OpenMPI 1.6.5 on \tinto.
-On the other hand, we  used HDF5 1.8.12 and NetCDF  (4.3.1) libraries for  the management  of extremely  large and
-complex data collections.  All computes nodes have an operating system
-based on GNU/Linux featuring ``2.6.32-358.11.1.el6'' kernel in
-\monch, ``3.0.101-0.15-default'' in \pilat and ``2.6.18-238.19.1.el5'' in \tinto.
+The \cosmoart  baseline is a pure MPI-based  Fortran~90 code currently
+running on distributed multi-core systems only.  The software stack on
+the platforms  was controlled using the modules  framework which gives
+an  easy and  flexible  mechanism to  access  to all  of the  provided
+compilers, tools  and applications.  For our  initial benchmarking, we
+opted for the GNU compiler GCC 4.8.1 on \monch and GCC 4.8.2 on \pilat
+and  \tinto)  using  the -O3  compiler  flag  in  favor of  the  Intel
+compilers 14.0.1, which  delivered inferior performance.  In addition,
+we  installed the following  implementations of  MPI: MVAPICH2  1.9 on
+\monch and \pilat, and OpenMPI 1.6.5  on \tinto. On the other hand, we
+used HDF5  1.8.12 and NetCDF  (4.3.1) libraries for the  management of
+extremely large and complex data collections.  All computes nodes have
+an     operating    system     based     on    GNU/Linux     featuring
+``2.6.32-358.11.1.el6'' kernel  in \monch, ``3.0.101-0.15-default'' in
+\pilat and ``2.6.18-238.19.1.el5'' in \tinto.
 
@@
 information needed  to reproduce the  energy-to-solution benchmarks of
-\cosmoart, was produced and run on 1040 cores using 20~MPI
-tasks  per node on  \monch, on 1344 cores  using 16~MPI
-tasks per node on  \pilat and on 192 cores  using 12~MPI
-tasks per node on \tinto.  The calculated region was mapped
-to the participating processors using a 2D-partitioning strategy.  The
-distribution along the $x$ and $y$ coordinates was defined accordingly depeding on the selected platform.
-% $by setting:
-%$nprocx=40$  and $nprocy=26$  for \monch  and  $nprocx=28$ and
-%$nprocy=24$  as  
-%to keep $x$  bigger  than  $nprocy$.
-Besides,  as this  version does not  make use  of the  traditional GRIB
-library, we  specified $nprocio=0$  for GRIB I/O.   Hyperthreading was
-not considered in this study  as it previously revealed that it always
-led to higher energy-to-solution.
+\cosmoart, was produced  and run on 1040 cores  using 20~MPI tasks per
+node on  \monch, on 1344 cores  using 16~MPI tasks per  node on \pilat
+and  on  192  cores  using  12~MPI  tasks per  node  on  \tinto.   The
+calculated region  was mapped to the participating  processors using a
+2D-partitioning  strategy.  The  distribution  along the  $x$ and  $y$
+coordinates  was   defined  accordingly  depending   on  the  selected
+platform.  Besides,  as  this  version   does  not  make  use  of  the
+traditional  GRIB  library, we  specified  $nprocio=0$  for GRIB  I/O.
+Hyperthreading  was not  considered  in this  study  as it  previously
+revealed that it always led to higher energy-to-solution.

### 2014-05-05 paper_enahpc14/Conclusion.tex
cleans up latest version; improves format

@@
 Ivybridge  (2.2  GHz) system,  although  the  two  metrics are  indeed
-strongly  correlated.   A   tool  utilising  Paraver/Extrae  with  new
-extensions  for power  consumption  was successfully  applied to  this
-non-trivial application.  The  resulting profiles indicate that simple
-changes, such as making use of an MPI version based on blocking rather
-than polling, can reduce power consumption significantly.
+strongly correlated.  On the other hand, the  usage of energy-friendly
+MPI waiting  techniques analzyed with  our power-performance framework
+over  a  series  of runs  of  \cosmoart  on  \tinto and  power  traces
+demonstrate that it  is possible to reduce both  power dissipation and
+energy-to-solution   while  maintaining   (or  even   decreasing)  the
+time-to-solution. The resulting profiles indicate that simple changes,
+such as making use of the  blocking MPI policy rather than the polling
+is  possible to  slightly  reduce both  power  dissipation and  energy
+consumption (by 2\,\%).
 

### 2014-05-05 paper_enahpc14/Introduction.tex
cleans up latest version; improves format

@@
 have shorter atmospheric  lifetimes, aerosols significantly modify the
-global   radiation   budget   \citep{IPCC-2013}.   Enhanced   aerosols
+global   radiation   budget   \cite{IPCC-2013}.    Enhanced   aerosols
 concentrations impact the climate  system by scattering and absorption
 of solar radiation, thereby exerting a direct radiative forcing; or by
-modifying cloud  properties,  cloud  fraction and  surface
-albedo,  causing  a  negative  indirect  radiative  forcing.   Despite
-considerable  progress in  global aerosol  modelling \citep{Mann-2013}
-and  measurement-based  assessments,  large  uncertainties  remain  in
-current  estimates  of  aerosol radiative  forcing  \citep{Myhre-2013,
-IPCC-2013,  Lee-2013,   Randles-2013,  Rosenfeld-2013,  Sherwood-2013,
-Stier-2013}.
+modifying cloud properties, cloud fraction and surface albedo, causing
+a negative indirect  radiative forcing.  Despite considerable progress
+in  global aerosol  modelling  \cite{Mann-2013} and  measurement-based
+assessments,  large  uncertainties  remain  in  current  estimates  of
+aerosol  radiative   forcing  \cite{Myhre-2013,  IPCC-2013,  Lee-2013,
+Sherwood-2013, Stier-2013}.
 
@@
 gases),   developed   at  the   Karlsruhe   Institute  of   Technology
-(KIT)  \citep{Vogel-2009,  Bangert-2011,  Knote-2013}.  This  regional
-scale modelling system is coupled  with the operational  weather forecast
-model  \textsc{Cosmo}  \citep{Baldauf-2011},  jointly developed and used by  a
-consortium of European weather centers, as well as utilised in a climate version
-by the wider research community.
+(KIT)  \cite{Vogel-2009,  Bangert-2011,  Knote-2013}.   This  regional
+scale  modelling  system  is  coupled  with  the  operational  weather
+forecast  model \textsc{Cosmo} \cite{Baldauf-2011},  jointly developed
+and  used by  a consortium  of European  weather centers,  as  well as
+utilized in a climate version by the wider research community.
 
-The extended  atmospheric model \textsc{Cosmo-art}  is com\-put\-ationally
-much  more  demanding than  \textsc{Cosmo}  since  a  large number  of
-additional  tracers and  processes have  to be  considered.   Thus the
-model  is currently  severely limited  in terms  of applicability  and
-expensive in terms of energy consumption.  Although \textsc{Cosmo} has
-recently been ported to GPUs \citep{Gysi-2014, Lapillonne-2014} within
-the framework of the  High Performance and High Productivity Computing
-(HP2C) Initiative  \citep{HP2C} to optimise it  for computational and
+The extended  atmospheric model \cosmoart  is com\-put\-ationally much
+more demanding than \textsc{Cosmo}  since a large number of additional
+tracers  and processes  have  to  be considered.   Thus  the model  is
+currently severely limited in  terms of applicability and expensive in
+terms  of energy  consumption.  Although  \textsc{Cosmo}  has recently
+been  ported  to  GPUs  \cite{Gysi-2014, Lapillonne-2014}  within  the
+framework  of the  High  Performance and  High Productivity  Computing
+(HP2C)  Initiative \cite{HP2C}  to optimize  it for  computational and
 energy efficiency,  significant investments in ART  are still required
-to take it  to a similar level.  The  efficiency of \textsc{Cosmo-art}
-is being  addressed in the  EU Exa2Green project  \citep{EXA2GREEN} to
-deliver a  prototype code, which  provides an energy efficiency  of at
-least five times the  baseline value.  Such an implementation would
-allow  the  community  to  investigate critical  questions  at  higher
-resolution  and   over  longer  periods,   at  reduced  cost   to  the
-environment.
+to take it  to a similar level.  The efficiency  of \cosmoart is being
+addressed in  the EU Exa2Green  project \cite{EXA2GREEN} to  deliver a
+prototype code, which  provides an energy efficiency of  at least five
+times  the baseline  value.  Such  an implementation  would  allow the
+community to  investigate critical questions at  higher resolution and
+over longer periods, at reduced cost to the environment.
 
-This  work is  organized as  follows: in  Sec.~\ref{sec:1} we  give an
-overview  of  related  work,   then  in  Sec.~\ref{sec:2}  we  briefly
-introduce \textsc{Cosmo-art}  and specify its  technical setup related
-to  the investigated  performance and  energy evaluation  methods.  In
-Sec.~\ref{sec:3}   we  describe   HPC  platforms,   power  measurement
+This work is  organized as follows: in Section~\ref{sec:1}  we give an
+overview  of  related work,  then  in  Section~\ref{sec:2} we  briefly
+introduce  \cosmoart and specify  its technical  setup related  to the
+investigated   performance   and   energy  evaluation   methods.    In
+Section~\ref{sec:3}  we  describe  HPC  platforms,  power  measurement
 equipments   and  software  environment   employed  to   conduct  this
-benchmarking study.  Sec.~\ref{sec:4}  presents performance and energy
-requirements  of the  baseline on  these architectures  and highlights
-areas where improvements will be necessary for the subsequent baseline
-refactoring.   Finally,  we  conclude in Sec.~\ref{concl} with some 
-implications for the Exa2Green project and give an outlook for future
-research.
+benchmarking  study.   Section~\ref{sec:4}  presents  performance  and
+energy  requirements  of  the  baseline  on  these  architectures  and
+highlights  areas  where  improvements   will  be  necessary  for  the
+subsequent   baseline   refactoring.     Finally,   we   conclude   in
+Section~\ref{concl} with  some implications for  the Exa2Green project
+and give an outlook for future research.

### 2014-05-05 paper_enahpc14/Sec4:Experimental-results.tex
Adding missing figure
cleans up latest version; improves format

@@
 In this section, we present our measurement methodology along with the
-metrics considered and discuss about the performance-power-energy trade-off of the model system on the different platforms stated.
+metrics  considered  and  discuss about  the  performance-power-energy
+trade-off of the model system on the different platforms stated.
 
@@
 time. ETS  is the amount of  energy spent to  achieve results.  Energy
-consumption  is  assessed by  sampling  the  power during execution which is then averaged and multiplied  by the TTS to determine  ETS. Whenever  possible,  multiple production  runs  were
-performed  to  illustrate the  reproducibility  of  the baseline,  and
-quantify the  significant uncertainties  in the power  measurement, as
-dictated by the available technology.
+consumption is  assessed by sampling the power  during execution which
+is then averaged and multiplied  by the TTS to determine ETS. Whenever
+possible, multiple  production runs  were performed to  illustrate the
+reproducibility  of   the  baseline,  and   quantify  the  significant
+uncertainties in  the power measurement, as dictated  by the available
+technology.
 
-For the experiments on \tinto platform, we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
-Specifically, the OpenMPI 1.6.5 library,
-features two operation modes, blocking and polling, which can be selected before \cosmoart
-is launched. In the polling configuration (the default mode) ``MPI'' continually polls
-the network interface to check for the completion of an event (e.g. send or receive).
-This usually yields low latency but high CPU utilization. Thus, one can expect that this mode 
-attains the best performance, possibly at the cost a higher energy usage. In the blocking
-mode, the CPU is yield to other processes/threads if there are no incoming messages~\cite{PDP12}.
+For  the   experiments  on  \tinto  platform,  we   also  analyze  the
+contribution of the MPI library  to the energy consumption and how the
+use  of  the  blocking  versus polling  message-passing  policies  can
+potentially  render energy  savings. Specifically,  the  OpenMPI 1.6.5
+library, features two operation modes, blocking and polling, which can
+be selected before \cosmoart is launched. In the polling configuration
+(the default mode) ``MPI''  continually polls the network interface to
+check for  the completion  of an event  (e.g. send or  receive).  This
+usually yields  low latency  but high CPU  utilization. Thus,  one can
+expect that  this mode attains  the best performance, possibly  at the
+cost a higher energy usage. In  the blocking mode, the CPU is yield to
+other     processes/threads    if     there     are    no     incoming
+messages~\cite{Castillo-2012}.
 
@@
 
-Figures~\ref{fig:1}   and   \ref{fig:2}   account   respectively   for
-\monch's Isola E1 Rack  2 and \pilat' Isola HD total
-power measurements for  1-day or 2-day simulations. On  the Intel Ivy
-Bridge EP  based cluster (i.e.  \monch),  the 1-day simulation
-was issued  only twice due  to usage restrictions. As  time resolution
-was set to one update every 5 minutes for power sampling, the averaged
-power dissipation was computed by considering 6 values for each single
-run.  On the Intel Xeon E5 based cluster (i.e.  \pilat), the
-1-day simulation  was issued  four times and  a 2-day run  only once.
-Similarly, the averaged power dissipation was computed by considering 4
-values  for  each  single  1-day  run  and 9  values  for  the  2-days
-run. Corresponding results are gathered in Table~\ref{tab:3}.
+Figures~\ref{fig:1} and \ref{fig:2}  account respectively for \monch's
+Isola  E1 Rack 2  and \pilat'  Isola HD  total power  measurements for
+1-day or 2-day  simulations. On the Intel Ivy  Bridge EP based cluster
+(i.e.   \monch), the  1-day simulation  was issued  only twice  due to
+usage restrictions. As  time resolution was set to  one update every 5
+minutes  for  power  sampling,  the  averaged  power  dissipation  was
+computed by  considering 6 values for  each single run.   On the Intel
+Xeon E5 based cluster (i.e.   \pilat), the 1-day simulation was issued
+four times and  a 2-day run only once.   Similarly, the averaged power
+dissipation was computed by considering 4 values for each single 1-day
+run  and  9 values  for  the  2-days  run. Corresponding  results  are
+gathered in Table~\ref{tab:1}.
 
@@
     \caption{Averaged power dissipation (W) of the platforms.}
-    \label{tab:3}
+    \label{tab:1}
     \begin{tabular}{cccc}
       \hline\noalign{\smallskip}
-      \textbf{\scriptsize{Xeon E5}} & \textbf{\scriptsize{Ivy Bridge}} & \textbf{\scriptsize{Xeon  E5645}} & \textbf{\scriptsize{Xeon  E5645}}\\
+      \textbf{\scriptsize{Xeon E5}} & \textbf{\scriptsize{Ivy Bridge}} & 
+      \textbf{\scriptsize{Xeon  E5645}} & \textbf{\scriptsize{Xeon  E5645}} \\
       & & \textbf{\scriptsize{(Polling)}} & \textbf{\scriptsize{(Blocking)}} \\
@@
 (left  Y-axis)  metrics  on  both  platforms.  As  expected,  Xeon  E5
-outperforms Ivy Bridge EP, being  roughly 130\,\% faster.  The reason for
-that is  two-fold: \emph{i}) it has  higher clock frequency  than Ivy Bridge
-(2.6\,GHz  against  2.2\,GHz),  and  \emph{ii})  it  aims at  computing  speed
-regardless to  energy consumption.  In our experiments,  Ivy Bridge EP
-showed the best energy-to-solution, reducing the energy consumption of
-Xeon E5 by approximately 7\,\%.
+outperforms Ivy  Bridge EP, being roughly 130\,\%  faster.  The reason
+for that is two-fold: \emph{i}) it has higher clock frequency than Ivy
+Bridge  (2.6\,GHz  against  2.2\,GHz),   and  \emph{ii})  it  aims  at
+computing speed regardless to energy consumption.  In our experiments,
+Ivy Bridge EP showed  the best energy-to-solution, reducing the energy
+consumption of Xeon E5 by approximately 7\,\%.
 
-Averaged TTS and ETS results over 10 repetitions of 24h simulation on \tinto are shown in Figure~\ref{fig:4}.
-The corresponding results for the averaged total power are also shown in Table~\ref{tab:3}.
-As expected, the average total power dissipation by \cosmoart while using the blocking MPI policy is about 2\,\% less than that consumed by the polling policy.
-This is due to the power-friendly (blocking) waits performed by the CPU cores while using the blocking MPI mode, in contrast to the polling waits that take place when the polling MPI mode is set.
-From the point of view of the execution time, there is a slight reduction for the blocking MPI mode.
-One can expect that the costs of the interruptions used in the blocking MPI policy can lead to an increase of the execution time, however, applications with extensive computing, as e.g.
-\cosmoart, may show better performance with the blocking modes.
-This effect works in favor with our results since the energy consumed is also reduced (2\,\%) just by changing the behavior of the OpenMPI library to the blocking policy, i.e, with the \texttt{--mca mpi\_yield\_when\_idle} parameter set in the \texttt{mpirun} command.
+Averaged TTS and ETS results  over 10 repetitions of 24h simulation on
+\tinto are shown in  Figure~\ref{fig:4}. The corresponding results for
+the  averaged total  power are  also shown  in  Table~\ref{tab:1}.  As
+expected, the average total power dissipation by \cosmoart while using
+the blocking MPI policy is about  2\,\% less than that consumed by the
+polling  policy. This is  due to  the power-friendly  (blocking) waits
+performed  by the  CPU cores  while using  the blocking  MPI  mode, in
+contrast to  the polling  waits that take  place when the  polling MPI
+mode is set. From the point of  view of the execution time, there is a
+slight reduction  for the blocking MPI  mode. One can  expect that the
+costs of the interruptions used in the blocking MPI policy can lead to
+an  increase  of  the   execution  time,  however,  applications  with
+extensive computing,  as e.g.  \cosmoart, may show  better performance
+with the blocking  modes. This effect works in  favor with our results
+since the energy consumed is also reduced (2\,\%) just by changing the
+behavior of the OpenMPI library  to the blocking policy, i.e, with the
+\texttt{--mca    mpi\_yield\_when\_idle}   parameter   set    in   the
+\texttt{mpirun} command.
 
@@
 \begin{figure}[htbf]
-
   \centering
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-5.eps}
+  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
   \vspace{-1cm}
-  \caption{Mean time-to-solution and energy-to-solution on
-    \tinto for a 24h simulation using the polling and blocking MPI policies. The 95\,\% confidence intervals of Student's $t$ distribution (10 samples) clearly indicates 
-    an improvement for blocking policy in both TTS and ETS.}
+  \caption{Mean time-to-solution and  energy-to-solution on \tinto for
+    a 24h simulation using the  polling and blocking MPI policies. The
+    95\,\%  confidence  intervals of  Student's  $t$ distribution  (10
+    samples) clearly  indicates an improvement for  blocking policy in
+    both TTS and ETS.}
   \label{fig:4}
@@
   \centering
-  \hspace{0.6cm}
-  \scalebox{0.64}{\input{Figs/24hour.pstex_t}}
-  \caption{24 hours simulation power trace using the polling and blocking MPI policies on \tinto.}
-  \label{fig:9}
+  \hspace{0.8cm} \scalebox{0.62}{\input{Figs/24hour.pstex_t}}
+  \caption{24  hours  simulation power  trace  using  the polling  and
+    blocking MPI policies on \tinto.}
+  \label{fig:5}
 \end{figure*}
@@
   \scriptsize
-  Simulation of 1 hour during midday (11h--12h) leveraging the polling MPI policy\\
-  \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}}\\
-  Simulation of 1 hour during midday (11h--12h) leveraging the blocking MPI policy\\
-  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}\\
-  1 hour simulation during midnight (23h--24h) leveraging the polling MPI policy\\
-  \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}}\\
-  1 hour simulation during midnight (23h--24h) leveraging the blocking MPI policy\\
-  \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}\\
+  Simulation of 1 hour during midday (11h--12h) leveraging the polling MPI policy \\
+  \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}} \\
+  Simulation of 1 hour during midday (11h--12h) leveraging the blocking MPI policy \\
+  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}} \\
+  1 hour simulation during midnight (23h--24h) leveraging the polling MPI policy \\
+  \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}} \\
+  1 hour simulation during midnight (23h--24h) leveraging the blocking MPI policy \\
+  \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}} \\
   \begin{tabular}{rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rl}
@@
 \end{tabular}
-  \caption{Power-performance traces during midday and midnight leveraging the polling and blocking MPI policies on \tinto.}
-  \label{fig:11}
+  \caption{Power-performance   traces  during   midday   and  midnight
+    leveraging the polling and blocking MPI policies on \tinto.}
+  \label{fig:6}
 \end{figure*}
@@
 
-A full tracing experiment is conducted on \tinto in order to capture an overall power
-profile at a much finer resolution using the attached PDUs in each node. First we run \cosmoart 
-for a 1-day simulation and analyze the power consumption. Second we correlate performance 
-with power traces of two 1-hour time frames during the day: mid-day (11h--12h) and
-midnight (23h--24h). All the experiments were performed using 192 cores using 12 MPI processes
-per node on \tinto.
+A full tracing  experiment is conducted on \tinto  in order to capture
+an overall power profile at a much finer resolution using the attached
+PDUs in each  node. First we run \cosmoart for  a 1-day simulation and
+analyze the  power consumption.  Second we correlate  performance with
+power  traces  of two  1-hour  time  frames  during the  day:  mid-day
+(11h--12h) and midnight (23h--24h). All the experiments were performed
+using 192 cores using 12 MPI processes per node on \tinto.
 
-% For all the experiments we also analyze the contribution of the MPI library to the energy 
-% consumption and how the use of the blocking versus polling message-passing policies can 
-% potentially render energy savings. Specifically, the OpenMPI 1.6.5 library installed in \tinto, 
-% features two operation modes, blocking and modes, which can be selected before \cosmoart is 
-% launched. In the polling configuration (the default mode) ``MPI'' continually polls
-% the network interface to check for the completion of an event (e.g. send or receive).
-% This usually yields low latency but high CPU utilization. Thus, can one expect that this mode attains the best performance, possibly at the cost a higher energy usage. In the blocking
-% mode, the CPU is yield to other processes/threads if there are no incoming messages.
+In order to obtain an initial overview of the power dissipation of the
+system  model,  we  obtained  power  profiles  leveraging  our  \pmlib
+power-measurement    framework   over    a    1-day   simulation    of
+\cosmoart. These profiles, produced with both polling and blocking MPI
+policies,  are depicted in  Figure~\ref{fig:5}. The  first observation
+for the polling MPI policy is that the first and last hours of the day
+dissipate slightly less power than  the mid-day hours. We relate these
+variations  due   to  the  different   chemical  reactions,  aerosols,
+radiations  and clouds  formation processes,  computed by  ART, taking
+place during  daylight hours  but not during  night. In this  case the
+peak power consumption is about  3821\,W. Looking at the power profile
+using the blocking  MPI policy, we observe that  the 1-hour pattern is
+repeated along  the 24-hours, favoring  the reduction of  power during
+night  hours in  a  more noticeable  way  than that  with the  polling
+policy.  The peak  power dissipation  during  day hours  of while  the
+blocking  MPI policy  is leveraged  is  about 3768\,W,  which means  a
+reduction of around  50\,W with respect to the  polling mode. Also, we
+notice systematic  power drops for  each simulated hour.  We associate
+them  to periods  in  which  the processes  were  blocked waiting  for
+incoming messages  or synchronizations, thus  yielding the CPU  to low
+usage. Our final observation is the reduction of the TTS by 2\,\% with
+respect to the blocking MPI policy.
 
-In order to obtain an initial overview of the power dissipation of the system model, we obtained power profiles leveraging our \pmlib power-measurement framework over a 1-day simulation of \cosmoart.
-These profiles, produced with both polling and blocking MPI policies, are depicted in  Figure~\ref{fig:9}.
-The first observation for the polling MPI policy is that the first and last hours of the day dissipate slightly less power than the mid-day hours.
-We relate these variations due to the different chemical reactions, aerosols, radiations and clouds formation processes, computed by ART, taking place during daylight hours but not during night.
-In this case the peak power consumption is about 3821\,W.
-Looking at the power profile using the blocking MPI policy, we observe that the 1-hour pattern is repeated along the 24-hours, favoring the reduction of power during night hours in a more noticeable way than that with the polling policy.
-The peak power dissipation during day hours of while the blocking MPI policy is leveraged is about 3768\,W, which means a reduction of around 50\,W with respect to the polling mode.
-Also, we notice systematic power drops for each simulated hour.
-We associate them to periods in which the processes were blocked waiting for incoming messages or synchronizations, thus yielding the CPU to low usage.
-Our final observation is the reduction of the TTS by 2\,\% with respect to the blocking MPI policy.
+To  gain  more  insights   about  the  power-performance  behavior  of
+\cosmoart,  we  obtained power-per\-for\-man\-ce  traces  to allow  us
+correlate  the type  of  tasks  being executed  with  the total  power
+dissipation  along  the   time  (see  Figure~\ref{fig:6}).   For  this
+experiment we  limit the simulation  time to 1 hour  because \emph{i})
+the  pattern of  the computation  performed by  \cosmoart  is repeated
+hourly and, \emph{ii}) the weight  of a 1-day simulation trace can not
+be easily handled using our  machines and memory available (each trace
+file was about 45\,GB). To  face the problem we shrink the simulations
+to 1 hour  and test a time  frame during midday (from 11h  to 12h) and
+midnight  (from 23h  to  24h), in  which  \cosmoart perform  different
+internal computations.  At  the same time we leverage  the two already
+stated MPI policies: polling and blocking.
 
-To gain more insights about the power-performance behavior of \cosmoart, we obtained power-per\-for\-man\-ce traces to allow us correlate the type of tasks being executed with the total power dissipation along the time (see Figure~\ref{fig:11}).
-For this experiment we limit the simulation time to 1 hour because \emph{i}) the pattern of the computation performed by \cosmoart is repeated hourly and, \emph{ii}) the weight of a 1-day simulation trace can not be easily handled using our machines and memory available (each trace file was about 45\,GB).
-To face the problem we shrink the simulations to 1 hour and test a time frame during midday (from 11h to 12h) and  midnight (from 23h to 24h), in which \cosmoart perform different internal computations.
-At the same time we leverage the two already stated MPI policies: polling and blocking.
-
-All the traces in Figure~\ref{fig:11} are divided in different computation phases.
-In first phase all the processes perform group communications and execute MPI-receive primitives (e.g.
-\texttt{MPI\_Recv}).
-The second one performs computation bursts, combined in between with group communications, blocking sends, immediate receives and MPI-wait primitives are executed.
-Finally, group communications interlaced with computation and a significant synchronization phase are performed.
-From the performance point of view, one can notice a reduction of 2\,\% in the TTS that blocking MPI mode leads over the polling MPI policy.
-On the other hand an averaged reduction of power dissipation of 2\,\% is attained with the blocking MPI policy only in routines that actually perform synchronizations, group communications and MPI-receive/-wait primitives.
-Thus, the reduction of the power-energy pair directly depends on the percentage of time in which blocking MPI operations are executed.
-For the 1 hour simulation traces, an average of 50\,\% of the time the processes were blocked.
-Taking into account that  the total power reduction during blocking phases is about 5\,\%, one can compute an approximation of the power savings, i.e., $50\,\% \times 5\,\% = 2.5\%$, which is consistent power reduction obtained for the full-day simulation (around 2\,\%).
-However, we expect higher power-energy savings if future architectures tend proportional computing~\citep{Barroso:2007:CEC:1339817.1339894}.
+All  the  traces  in   Figure~\ref{fig:6}  are  divided  in  different
+computation  phases. In first  phase all  the processes  perform group
+communications      and       execute      MPI-receive      primitives
+(e.g. \texttt{MPI\_Recv}). The second one performs computation bursts,
+combined  in  between   with  group  communications,  blocking  sends,
+immediate  receives and  MPI-wait primitives  are  executed.  Finally,
+group  communications interlaced  with computation  and  a significant
+synchronization  phase are  performed. From  the performance  point of
+view, one can notice a reduction of 2\,\% in the TTS that blocking MPI
+mode leads over the polling MPI  policy. On the other hand an averaged
+reduction of power dissipation of  2\,\% is attained with the blocking
+MPI policy  only in  routines that actually  perform synchronizations,
+group  communications  and  MPI-receive/-wait primitives.   Thus,  the
+reduction of the power-energy  pair directly depends on the percentage
+of time in which blocking MPI  operations are executed. For the 1 hour
+simulation traces, an average of 50\,\% of the time the processes were
+blocked.  Taking  into account that  the total power  reduction during
+blocking phases  is about 5\,\%,  one can compute an  approximation of
+the  power savings,  i.e., $50\,\%  \times  5\,\% =  2.5\%$, which  is
+consistent  power  reduction  obtained  for  the  full-day  simulation
+(around  2\,\%). However,  we  expect higher  power-energy savings  if
+future architectures tend proportional computing~\citep{Barroso-2007}.
 
@@
 %   \caption{blabla}
-%   \label{fig:8}
+%   \label{fig:7}
 % \end{figure*}

### 2014-05-05 trunk/paper_enahpc14/Sec3:Environment-setup.tex
updates working copy

@@
   312 compute nodes. 
-  % and  utilized  by  members  of  the  Swiss  Platform  for  Advanced
+  % and  utilised  by  members  of  the  Swiss  Platform  for  Advanced
   % Scientific Computing (PASC) project~\cite{PASC}.
@@
   % E5-Series  DALCO  r2264i4t   2U  scalable  compute  modules,  each
-  % containing 4 compute nodes
+  % containing 4 compute nodes based on two ...
   Each node  is comprises two Intel Xeon  E5-2670 processors operating
@@
 \pmlib  library in  combination  with Extrae  and  Paraver, which  are
-profiling/tracing and visualization tools, respectively.
+profiling/tracing and visualisation tools, respectively.
 
@@
 generated.  A  merged trace  file  is  finally  loaded in  Paraver  to
-visualize  and correlate  the  performance trace  and  with the  power
+visualise  and correlate  the  performance trace  and  with the  power
 profile  generated by \cosmoart.  This procedure  allows us  to easily
@@
 2D-partitioning  strategy.  The  distribution  along the  $x$ and  $y$
-coordinates  was   defined  accordingly  depending   on  the  selected
-platform.  Besides,  as  this  version   does  not  make  use  of  the
+coordinates was  defined by  setting: $nprocx=40$ and  $nprocy=26$ for
+\monch,  $nprocx=28$ and  $nprocy=24$ for  \pilat and  $nprocx=??$ and
+$nprocy=??$  for \tinto as  $nprocx$ is  usually kept  slightly bigger
+than  $nprocy$. Besides,  as this  version does  not make  use  of the
 traditional  GRIB  library, we  specified  $nprocio=0$  for GRIB  I/O.

### 2014-05-05 paper_enahpc14/Conclusion.tex
updates working copy

@@
 strongly correlated.  On the  other hand, the usage of energy-friendly
-MPI waiting  techniques analzyed with  our power-performance framework
+MPI waiting  techniques analysed with  our power-performance framework
 over  a  series  of runs  of  \cosmoart  on  \tinto and  power  traces

### 2014-05-05 paper_enahpc14/Introduction.tex
updates working copy

@@
 and  used by  a consortium  of European  weather centers,  as  well as
-utilized in a climate version by the wider research community.
+utilised in a climate version by the wider research community.
 
@@
 framework  of the  High  Performance and  High Productivity  Computing
-(HP2C)  Initiative \cite{HP2C}  to optimize  it for  computational and
+(HP2C)  Initiative \cite{HP2C}  to optimise  it for  computational and
 energy efficiency,  significant investments in ART  are still required
@@
 
-This work is  organized as follows: in Section~\ref{sec:1}  we give an
+This work is  organised as follows: in Section~\ref{sec:1}  we give an
 overview  of  related work,  then  in  Section~\ref{sec:2} we  briefly

### 2014-05-05 paper_enahpc14/Sec2:COSMO-ART.tex
updates working copy

@@
 
-The model  system \cosmoart described  in \cite{Vogel-2009},
-is  a  regional to  continental  scale  model  coupled online  to  the
+The  model  system  \cosmoart  described in  \cite{Vogel-2009},  is  a
+regional   to  continental   scale   model  coupled   online  to   the
 \textsc{Cosmo}   numerical  weather   prediction  and   climate  model
@@
 Establishing effective energy performance benchmarking of a code under
-intense development  such as \cosmoart is  a challenging task
-because of  the absolute necessity  that results must  be reproducible
-within  an  expected  variance  for  the  duration  of  the  Exa2Green
-pro\-je\-ct.  To   define  a  baseline,   it  was  necessary  to   find  a
-run-configuration  capable  of   being  recreated  in  all  subsequent
-versions.  Here  we give a brief  overview of the model  setup for our
-performance and energy efficiency evaluation.
+intense development such as \cosmoart is a challenging task because of
+the  absolute necessity that  results must  be reproducible  within an
+expected variance  for the duration of the  Exa2Green pro\-je\-ct.  To
+define  a  baseline, it  was  necessary  to  find a  run-configuration
+capable of being recreated in all subsequent versions.  Here we give a
+brief  overview of  the model  setup  for our  performance and  energy
+efficiency evaluation.
 
 Three-dimensional simulations are performed over large parts of Europe
-and the Mediterranean~Sea for April 13$^{th}$ 2010, near the equinox, in order to approximately
-have a half  day  of  sun  exposure and  therefore  ensure  a typical
-activation of the chemistry cycle.  The calculation domain corresponds
-to the CORDEX-EU-44 domain and is covered by a grid of $222\times 216$
-points with  a horizontal resolution of  $0.22\,^{\circ}$, i.e., 50~km
-in  both directions  and 40~vertical  layers.  These  24-hour forecast
-simulations  are not  preceded  by  a spin-up  phase  to build up  the
-simulated gaseous and  aerosol concentrations.  This condition means 
-that the model initialization needs a period of adjustment to reach its equilibrium  state and minimize  the effect of
-initial conditions for gases and aerosols on the model predictions.
+and the Mediterranean~Sea for  April 13$^{th}$ 2010, near the equinox,
+in  order  to  approximately have  a  half  day  of sun  exposure  and
+therefore  ensure a typical  activation of  the chemistry  cycle.  The
+calculation  domain  corresponds to  the  CORDEX-EU-44  domain and  is
+covered  by  a  grid  of  $222\times 216$  points  with  a  horizontal
+resolution  of $0.22\,^{\circ}$,  i.e., 50~km  in both  directions and
+40~vertical  layers.   These  24-hour  forecast  simulations  are  not
+preceded  by a spin-up  phase to  build up  the simulated  gaseous and
+aerosol   concentrations.   This  condition   means  that   the  model
+initialization needs  a period of adjustment to  reach its equilibrium
+state  and minimise  the effect  of initial  conditions for  gases and
+aerosols on the model predictions.
 
@@
 (rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
-parameterizations,   precipitation  formation   is   performed  by   a
+parameterisations,   precipitation  formation   is   performed  by   a
 two-moment cloud microphysics.

### 2014-05-05 paper_enahpc14/paper.tex
Manuel's changes
WS: epsfig
updates working copy

--- 2014-04-24 paper_enahpc14/paper.tex
@@
 %
+\usepackage[numbers]{natbib}
+\usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{graphicx}
-\usepackage[T1]{fontenc}
-\usepackage{hyperref}
-\usepackage[round, sort&compress, numbers]{natbib}
 \usepackage{multirow}
-\usepackage{graphicx}
-\usepackage{units}
-\usepackage{color}
+\usepackage{rotating}
+\usepackage{algorithm}
+\usepackage{algorithmic}
 \usepackage{colortbl}
+\usepackage{url}
 \usepackage{xspace}
@@
 \DeclareRobustCommand\IPCClongname{}
-\setcounter{secnumdepth}{5}
 %
@@
 \title{Evaluating  the  Performance   and  Energy  Efficiency  of  the
-  COSMO-ART model system}
+  \textsc{COSMO-ART} Model System}
 %\thanks{}
@@
   In this  paper we investigate  the energy footprint  and performance
-  profiling of COSMO-ART  on various HPC platforms.  This  model is an
+  profiling of \cosmoart  on various HPC platforms.  This  model is an
   extension of  the operational weather  forecast model of  the German
@@
   related  to  energy  efficient  high performance  computing  in  the
-  framework of the Exa2Green project \citep{EXA2GREEN}.
+  framework of the Exa2Green project \cite{EXA2GREEN}.
 
-\keywords{High performance computing  \and Energy-aware computing \and
-  Green computing  \and Numerical weather  prediction \and Atmospheric
-  chemistry  \and  Aerosols  modelling  \and  Profiling  methods  \and
-  Benchmark    analysis   \and    COSMO-ART    coupled   model    \and
-  Time-to-solution \and Energy-to-solution \and Xeon processor}
+\keywords{High performance computing  \and Energy-aware computing
+  %Green computing  
+  \and Numerical weather  prediction \and Atmospheric
+  chemistry  \and  Aerosol  modelling \and
+  Benchmark  analysis  \and 
+  \cosmoart  coupled  model \and  
+  Power-performance profiling and tracing tools  \and
+  Energy-saving techniques}
 % \PACS{PACS code1 \and PACS code2 \and more} 
@@
 
-\section{The COSMO-ART model system}
+\section{The \textsc{COSMO-ART} Model System}
 \label{sec:2}
@@
 also gratefully acknowledge the High Performance and High Productivity
-Computing Initiative  \citep{HP2C} for results that  will be leveraged
-in subsequent code refactoring.
+Computing Initiative \cite{HP2C} for results that will be leveraged in
+subsequent code refactoring.
 \end{acknowledgements}

### 2014-05-06 paper_enahpc14/Conclusion.tex
updates latest version

@@
 We  have   presented  a  methodology  for   comparing  performance  of
-COSMO-ART,  a  regional  weather  forecast  model  augmented  for  the
+\cosmoart,  a  regional  weather  forecast  model  augmented  for  the
 interactions of  reactive gases and aerosol  particles.  The resulting
@@
 package within  the EU-funded Exa2Green to  minimise power consumption
-of COSMO-ART. Profiling  has given us insight into  the most expensive
+of \cosmoart. Profiling  has given us insight into  the most expensive
 code  components,  which are  now  being  altered  to utilise  revised

### 2014-05-06 paper_enahpc14/Sec4:Experimental-results.tex
updates working copy
updates latest version

@@
   \begin{tabular}{cccc}
-      \hline\noalign{\smallskip}
-      \textbf{\scriptsize{Xeon E5}} & \textbf{\scriptsize{Ivy Bridge}} & 
-      \textbf{\scriptsize{Xeon  E5645}} & \textbf{\scriptsize{Xeon  E5645}} \\
-      & & \textbf{\scriptsize{(Polling)}} & \textbf{\scriptsize{(Blocking)}} \\
-      \noalign{\smallskip}\hline\noalign{\smallskip}
-      18035.0 & 12622.5 & 3713.6 & 3651.8 \\ 
-      \noalign{\smallskip}\hline
+    \hline\noalign{\smallskip}   \textbf{\scriptsize{Xeon   E5}}   &
+    \textbf{\scriptsize{Ivy   Bridge}}   &  \textbf{\scriptsize{Xeon
+        E5645}}   &   \textbf{\scriptsize{Xeon   E5645}}  \\   &   &
+    \textbf{\scriptsize{(Polling)}}                                 &
+    \textbf{\scriptsize{(Blocking)}}
+    \\   \noalign{\smallskip}\hline\noalign{\smallskip}   18035.0  &
+    12622.5 & 3713.6 & 3651.8 \\ \noalign{\smallskip}\hline
   \end{tabular}
@@
 analyze the  power consumption.  Second we  correlate performance with
-power  traces  of two  1-hour  time  frames  during the  day:  mid-day
+power  traces  of  two  1-hour  time frames  during  the  day:  midday
 (11h--12h) and midnight (23h--24h). All the experiments were performed
@@
 for the polling MPI policy is that the first and last hours of the day
-dissipate slightly less power than  the mid-day hours. We relate these
+dissipate slightly less  power than the midday hours.  We relate these
 variations  due   to  the  different   chemical  reactions,  aerosols,

### 2014-05-28 trunk/paper_enahpc14/RebuttalLetter/rebuttal.tex
Added Figures+RebuttalLeeter

\documentclass[a4paper,11pt]{article}

\title{Authors' Reply to Reviewers' Comments on\\ \itshape{``Evaluating the Performance and Energy Efficiency of the COSMO-ART Model System''}}

\author{Joseph~Charles, William~Sawyer, Manuel~F.~Dolz, Sandra~Catal\'an}

\usepackage[left=2.5cm,right=2.2cm,top=3cm,bottom=3cm]{geometry}
\usepackage[latin1]{inputenc}
\usepackage{color}
\usepackage{graphicx}
\usepackage{url}
\usepackage[urlcolor=blue,colorlinks=true]{hyperref}
\usepackage{ifthen}

%\newcommand{\xavi}[1]{\textcolor{blue}{#1}}
%\newcommand{\xavi}[1]{{#1}}
%\newcommand{\xaviNote}[1]{\footnote{\textcolor{blue}{#1}}}

\newcounter{question}
\setcounter{question}{1}

%\newcommand{\reviewerSay}{\vspace{0.2cm} {\bf \noindent \arabic{question}} \fbox{{\sc \textcolor{red}{Reviewer:}}} \medskip \rm}
\newcommand{\reviewerSay}{\vspace{0.2cm} {\bf \noindent } {\bf Reviewer:} \medskip \par}
\newcommand{\editorialSay}{\vspace{0.2cm} {\bf \noindent \arabic{question}.} {\bf Editorial:} \medskip \par}
%\newcommand{\weReply}{\vspace{0.25cm} {\bf \noindent \arabic{question}} \addtocounter{question}{1} \fbox{{\sc \textcolor{blue}{Authors:}}} \medskip \em\par}
\newcommand{\weReply}{\vspace{0.25cm} {\bf \noindent } \addtocounter{question}{1} \textbf{\emph{Authors:}} \medskip \em\par}
\newenvironment{weSay}{\bigskip \weReply \begin{it}}{\end{it}\begin{center} \samepage$\star \star \star$ \end{center}}
\newenvironment{weSayToEditorial}{\bigskip \weReply \begin{it}}{\end{it}\begin{center} \samepage$\star \star \star$ \end{center}}

%\newcommand{\reviewerSay}{\paragraph*{(\arabic{question}) Reviewer: }}
%\newcommand{\authorsReply}{\paragraph*{\emph{(\arabic{question}) \addtocounter{question}{1} Authors: }}}
%\newenvironment{weSay}{\vspace{-3ex} \authorsReply \begin{it}}{\end{it}\begin{center} \samepage$\star \star \star$ \end{center}}

\begin{document}

\maketitle

\noindent 
We would like to thank the Anonymous Reviewers
for providing us with a prompt and insightful review. 
We take this as an indication that there was genuine interest in the paper,
which is always a compliment regardless of the outcome of the review.

We addressed all the suggestions and recommendations from the reviewers to improve
the quality and flow of the paper. Below, we discuss the comments and suggestions
from the reviewers in detail.

\newpage
% ------------------------------------------------------------------------------------------------------------------------------------------------
\section*{Reviewer \#1}

\reviewerSay
The authors perform an energy performance and efficiency analysis of the COSMO-ART model system,
a complex weather prediction and climate model, focused on the study of aerosol concentrations in the
atmosphere. They present the model, the software and the hardware setup, including two metering
frameworks and perform out-of-band (i.e. measurement from external tools) power draw measurement
for three different setups. They use the experiments to perform two different analyses: energy 
efficiency for Sandy Bridge versus Ivy Bridge, and a power profile to compare polling and blocking 
MPI policies that also show the heavy energy consuming phases of the COSMO-ART model system.

The paper is well written and is easy to follow, and, in spite of the idea not being completely new, 
it offers a good insight about where are the most energy demanding phases and how the different 
policies on MPI affect them for the second experimentation part. The first experimentation part, 
besides being a common study, gives an interesting point of view about how energy efficient does 
not necessarily means time efficient.

%%% *** Manuel/Sandra ***
\begin{weSay}
The authors thank the reviewer for his/her general view of the paper.
\end{weSay}

Some suggestions regarding the paper:

- MONCH and PILATUS are presented as Ivy Bridge and Sandy Bridge clusters respectively in Section 4.1, 
and afterwards in section 5, they are referred as Ivy Bridge and Xeon E5 respectively. Apart from not 
being consistent, both processor families are Xeon E5 thus confusing a bit on what cluster is 
being referred. This also affects Table 1.

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

- It would be easier to compare values in Figure 4 if both energy ETSs and TTSs where together. 
Right now you show ETS TTS ETS TTS, and ETS ETS TTS TTS would be easier to understand.

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

- In multiple places: "power dissipation", "power draw". I assumed it this way, if it is not 
I suggest to explain it better.

%%% *** Manuel ***
\begin{weSay}
blabla
\end{weSay}

\newpage
\section*{Reviewer \#2}

\reviewerSay
The paper investigates the energy footprint of a weather forecasting model on various HPC platforms. 
As such the paper provides the overall energy footprint covering the overall run. In other words the 
granularity is very coarse and is at the application level. Although this will be interesting to the 
COSMO-ART community, the results include some good details on traces  on the
evaluation of polling vs blocking MPI strategies.

It could have been very beneficial to describe the nature of the application in detail instead of 
citing them. I assume that authors have resorted to that technique for the reasons of brevity  - yet 
I beleive that such a detailed description is essential to understand the behaviour of the application.

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

With missing crucial details of the application and in the abseence of any results towards various 
problem sizes or domains of cases, it is very difficult to do any post-analysis on the overall 
behaviour of the application - for example its runtime scalability (both weak and strong) vs energy 
scalability etc.

%%% *** Will ***
\begin{weSay}
blabla
\end{weSay}

In summary, although the paper is very interesting, the lack or depth of information in there severely 
limits the re-usable knowledge contributed by the paper.

%%% *** Will ***
\begin{weSay}
blabla
\end{weSay}

\newpage
\section*{Reviewer \#2}

The authors describe the extended atmospheric model COSMO-ART. They present
performance and energy measurements on a selection of HPC systems. There
is no new methodology presented in this paper.

%%% *** Will ***
\begin{weSay}
blabla
\end{weSay}

Related work is weak. I don't see how [20] and [11] are related to the content
of this paper (other algorithms, other architectures, no comparison of the
energy measurement infrastructure). I suggest to focus more on infrastructure
and tool support for system level power measurement in HPC environments in
the related work section. This would give you the opportunity to argue that
the 1 Hz resolution in one of your test systems is actually state-of-the-art.

%%% *** Will/Joseph ***
\begin{weSay}
blabla
\end{weSay}

The hardware description is a bit confusing.
- why does the NEC system have 10 racks for 312 nodes if one rack houses 52
     nodes?

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

- the "v2" is missing from the Ivy Bridge processor's name string

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

- sentence structure: "Each node is comprises two..."

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

- what do you mean by "second-generation processor" regarding Sandy Bridge?
     SB is sometimes referred to as 2nd gen Core-i* processor but this does not
     apply to the Xeon line

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

- the claim that SB is known to consume more power is wrong. The TDP classes
     are the same for IB and SB (e.g. E5-2670 and E5-2670v2 are both 115W parts)
     It presumably is less efficient, but even such a claim would need a reference

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

- It should be added if the Turbo Boost feature is enabled or disabled, this
     seems to be important for the discussion of the TTS results.

%%% *** Manuel/Sandra ***
\begin{weSay}
blabla
\end{weSay}

- It should be added which power saving mechanisms are used (cpufreq
     governor, available c-states)

%%% *** Manuel/Sandra ***
\begin{weSay}
blabla
\end{weSay}

- the second system is described as 16 cores x 42 nodes = 672 cores total. In
     4.3 it is stated that 1344 cores were used (Could be HyperThreading, however
     it is also stated that HT was not considered. Also, 16 MPI ranks per node
     indicate 84 nodes)

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

In 4.2.1
A 5 minute sampling interval seems useless to me. It would be nice to know if
this is actually an instantaneous value every 5 minutes (as I read it) or the
correct average of the past 5 minutes. Figure 5 suggests variation of the power
consumption in the 10\% range. Thus instantaneous values would have a high margin
for error and the 4,6, or 9 samples you use to calculate the average would be
too few to estimate a reliable average.

%%% *** Will/Joseph ***
\begin{weSay}
blabla
\end{weSay}

In 4.2.2
There is no information about the accuracy

%%% *** Manuel/Sandra ***
\begin{weSay}
blabla
\end{weSay}

In 4.3
Only -O3 is not appropriate for the Sandy bridge and Ivy Bridge system. At
least -mavx should be added. It looks like some low hanging fruit is reserved
to reach the targeted 5x improvement

%%% *** Will ***
\begin{weSay}
blabla
\end{weSay}

In 5.1
The blocking mode does not seem to be blocking. As there are no other
processes/threads available (as many MPI ranks as cores per node) the waiting
ranks will repeatedly call sched\_yield() only to be picked again by the OS
scheduler. This also explains the small benefit as no idle states are used.
You should check the CPU utilization. I guess it is 100\% on all cores in both
modes. If it is not, you should check with powertop which idle states are used.

%%% *** Manuel/Sandra ***
\begin{weSay}
blabla
\end{weSay}

In 5.2
What do you mean by "it aims at computing speed regardless to energy consumption"?
Does it mean it uses Turbo? Please clarify what exactly is different than on the
IB system.

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

It is confusing to compare "Ivy Bridge EP" against "Xeon E5" in the result
discussion, as the Ivy Bridge processor is also called Xeon E5 (v2) as well.

%%% *** Joseph ***
\begin{weSay}
blabla
\end{weSay}

Since the SB and IB measurements use different amounts of cores it is essential to
know if the application scales linearly with the core count. If the application
is scalable 1020 cores at 2.2 GHz should be faster than 672 cores at 2.6 GHz.

%%% *** Will ***
\begin{weSay}
blabla
\end{weSay}

In 5.3
Only 5\% energy reduction when blocked shows that the processors are not idle.
As mentioned earlier, you should add information if power management is
actually used while the processors are waiting.

%%% *** Manuel/Sandra ***
\begin{weSay}
blabla
\end{weSay}

\end{document}

### 2014-05-28 trunk/paper_enahpc14/Introduction.tex
updates a few things according to the reviewers' suggestions

@@
 of solar radiation, thereby exerting a direct radiative forcing; or by
-modifying cloud  properties,  cloud  fraction and  surface
-albedo,  causing  a  negative  indirect  radiative  forcing.   Despite
-considerable  progress in  global aerosol  modelling \cite{Mann-2013}
-and  measurement-based  assessments,  large  uncertainties  remain  in
-current  estimates  of  aerosol radiative  forcing  \cite{Myhre-2013,
-IPCC-2013, Lee-2013,  Sherwood-2013,  Stier-2013}.
+modifying cloud properties, cloud fraction and surface albedo, causing
+a negative indirect  radiative forcing.  Despite considerable progress
+in  global aerosol  modelling  \cite{Mann-2013} and  measurement-based
+assessments,  large  uncertainties  remain  in  current  estimates  of
+aerosol  radiative   forcing  \cite{IPCC-2013,  Lee-2013,  Myhre-2013,
+Randles-2013, Rosenfeld-2013, Sherwood-2013, Stier-2013}.
 
@@
 (KIT)  \cite{Vogel-2009,  Bangert-2011,  Knote-2013}.   This  regional
-scale modelling system is coupled  with the operational  weather forecast
-model  \textsc{Cosmo}  \cite{Baldauf-2011},  jointly developed and used by  a
-consortium of European weather centers, as well as utilized in a climate version
-by the wider research community.
+scale  modelling  system  is  coupled  with  the  operational  weather
+forecast  model \textsc{Cosmo} \cite{Baldauf-2011},  jointly developed
+and  used by  a consortium  of European  weather centers,  as  well as
+utilised in a climate version by the wider research community.
 
-The extended  atmospheric model \cosmoart  is com\-put\-ationally
-much  more  demanding than  \textsc{Cosmo}  since  a  large number  of
-additional  tracers and  processes have  to be  considered.   Thus the
-model  is currently  severely limited  in terms  of applicability  and
-expensive in terms of energy consumption.  Although \textsc{Cosmo} has
-recently been ported to GPUs \cite{Gysi-2014, Lapillonne-2014} 
-within
-the framework of the  High Performance and High Productivity Computing
-(HP2C) Initiative  \cite{HP2C} to optimize it  for computational and
+The extended  atmospheric model \cosmoart  is com\-put\-ationally much
+more demanding than \textsc{Cosmo}  since a large number of additional
+tracers  and processes  have  to  be considered.   Thus  the model  is
+currently severely limited in  terms of applicability and expensive in
+terms  of energy  consumption.  Although  \textsc{Cosmo}  has recently
+been  ported  to  GPUs  \cite{Gysi-2014, Lapillonne-2014}  within  the
+framework  of the  High  Performance and  High Productivity  Computing
+(HP2C)  Initiative \cite{HP2C}  to optimise  it for  computational and
 energy efficiency,  significant investments in ART  are still required
-to take it  to a similar level.  The  efficiency of \cosmoart
-is being  addressed in the  EU Exa2Green project  \cite{EXA2GREEN} to
-deliver a  prototype code, which  provides an energy efficiency  of at
-least five times the  baseline value.  Such an implementation would
-allow  the  community  to  investigate critical  questions  at  higher
-resolution  and   over  longer  periods,   at  reduced  cost   to  the
-environment.
+to take it  to a similar level.  The efficiency  of \cosmoart is being
+addressed in  the EU Exa2Green  project \cite{EXA2GREEN} to  deliver a
+prototype code, which  provides an energy efficiency of  at least five
+times  the baseline  value.  Such  an implementation  would  allow the
+community to  investigate critical questions at  higher resolution and
+over longer periods, at reduced cost to the environment.
 
-This  work is  organized as  follows: in  Section~\ref{sec:1} we  give an
+This work is  organised as follows: in Section~\ref{sec:1}  we give an
 overview  of  related work,  then  in  Section~\ref{sec:2} we  briefly
-introduce \cosmoart  and specify its  technical setup related
-to  the investigated  performance and  energy evaluation  methods.  In
+introduce  \cosmoart and specify  its technical  setup related  to the
+investigated   performance   and   energy  evaluation   methods.    In
 Section~\ref{sec:3}  we  describe  HPC  platforms,  power  measurement
 equipments   and  software  environment   employed  to   conduct  this
-benchmarking study.  Section~\ref{sec:4}  presents performance and energy
-requirements  of the  baseline on  these architectures  and highlights
-areas where improvements will be necessary for the subsequent baseline
-refactoring.   Finally,  we  conclude in Section~\ref{concl} with some 
-implications for the Exa2Green project and give an outlook for future
-research.
+benchmarking  study.   Section~\ref{sec:4}  presents  performance  and
+energy  requirements  of  the  baseline  on  these  architectures  and
+highlights  areas  where  improvements   will  be  necessary  for  the
+subsequent   baseline   refactoring.     Finally,   we   conclude   in
+Section~\ref{concl} with  some implications for  the Exa2Green project
+and give an outlook for future research.

### 2014-05-28 trunk/paper_enahpc14/Sec3:Environment-setup.tex
updates a few things according to the reviewers' suggestions

@@
 \begin{itemize}
-\item \monch is  a 10-rack NEC-provided clus\-ter at  CSCS composed of
-  312 compute nodes. 
+\item \monch is  a 10-rack NEC-provided clus\-ter at  CSCS composed by
+  376  compute  nodes  grouped  in  312  standard  compute  nodes,  24
+  large-memory compute nodes,  24 huge-memory nodes, 16 administration
+  nodes and 8 servers to provide I/O to Lustre.
   % and  utilised  by  members  of  the  Swiss  Platform  for  Advanced
-  % Scientific Computing (PASC) project~\cite{PASC}.
-  Each node is  comprised of two Intel Ivy  Bridge EP E5-2660 ten-core
-  processors  operating  at 2.2\,GHz,  equipped  with  32\,GB of  DDR3
-  1600\,MHz RAM  and connected via InfiniBand Mellanox  SX6036 and FDR
-  switches. For our  experiments, a full rack constituted  of 52 nodes
-  was used. 
+  % Scientific  Computing (PASC)  project~\cite{PASC}.   Each standard
+  compute  node is  comprised of  two  Intel Ivy  Bridge EP  E5-2660v2
+  ten-core processors  operating at 2.2\,GHz, equipped  with 32\,GB of
+  DDR3 1600\,MHz RAM and  connected via InfiniBand Mellanox SX6036 and
+  FDR  switches.   For  our   experiments,  a  full  rack  constituted
+  exclusively of 52 nodes was used.
   % \monch is  slated to stay  in service without hardware  upgrade for
@@
   cluster at CSCS composed of 42 compute nodes.  
-  % and used as Piz Daint  pre-post processing cluster The 2 login and
+  % and used as Piz Daint pre-post processing cluster. The 2 login and
   % 42  compute  nodes  are   consolidated  into  11  twin-pair  Intel
@@
   % containing 4 compute nodes based on two ...
-  Each node  is comprises two Intel Xeon  E5-2670 processors operating
+  Each node  is comprised two Intel Xeon  E5-2670 processors operating
   at 2.6\,GHz equipped with 64\,GB of DDR3 1600\,MHz RAM and connected

### 2014-05-28 paper_enahpc14/Conclusion.tex
updates a few things according to the reviewers' suggestions

@@
 the best energy-to-solution: an Intel Sandybridge (2.6 GHz) system has
-lower  time-to-solution but  higher energy-to-solution  than  an Intel
-Ivybridge  (2.2  GHz) system,  although  the  two  metrics are  indeed
-strongly correlated.  On the  other hand, the usage of energy-friendly
-MPI waiting  techniques analysed with  our power-performance framework
-over  a  series  of runs  of  \cosmoart  on  \tinto and  power  traces
-demonstrate that it  is possible to reduce both  power dissipation and
+lower time-to-solution but higher energy-to-solution than an Intel Ivy
+Bridge (2.2 GHz) system, although  the two metrics are indeed strongly
+correlated.   On the  other  hand, the  usage  of energy-friendly  MPI
+waiting techniques analysed  with our power-performance framework over
+a series of  runs of \cosmoart on \tinto  and power traces demonstrate
+that   it  is   possible  to   reduce  both   power   dissipation  and
 energy-to-solution   while  maintaining   (or  even   decreasing)  the

### 2014-05-28 paper_enahpc14/Sec2:COSMO-ART.tex
updates a few things according to the reviewers' suggestions

@@
 
-The  model  system  \cosmoart  described in  \cite{Vogel-2009},  is  a
-regional   to  continental   scale   model  coupled   online  to   the
-\textsc{Cosmo}   numerical  weather   prediction  and   climate  model
-\cite{Baldauf-2011}.    It  incorporates  sophisticated   modules  for
-gaseous  chemistry   and  aerosol  dynamics  and   allows  the  online
-calculation of  reactive trace  substances and their  interaction with
-the  atmosphere.    Detailed  model   description  can  be   found  in
-\cite{Bangert-2012, Knote-2011, Knote-2013}.
+The model system \cosmoart (ART stands for Aerosols and Reactive Trace
+gases) developed at KIT  Karlsruhe \cite{Vogel-2009}, is a regional to
+continental scale model coupled online to the \textsc{Cosmo} numerical
+weather  prediction (NWP) and  climate model  \cite{Baldauf-2011}.  It
+allows the  online calculation of reactive trace  substances and their
+interaction with  the atmosphere.  Physical  processes like transport,
+turbulent diffusion,  and dry and wet deposition  are treated together
+with photochemistry and aerosol dynamics using the modal approach.
+
+Aerosols dynamics  are simulated  with the modal  aerosol microphysics
+module  MADE \cite{Ackermann-1998},  expanded in  MADEsoot  to feature
+explicit  treatment of  soot aging  through condensation  of inorganic
+\cite{Riemer-2003} and organic  substances. MADE describes the aerosol
+population  through  five   modes  representing  sub-micron  particles
+consisting of sulphate, ammonium, nitrate, particulate organic matter,
+water and  soot \cite{Riemer-2004} in  a range of mixing  state. These
+modes are coupled  with the gas phase by  condensation and nucleation,
+and are  strongly influenced by  anthropogenic emissions of  gases and
+particles. MADEsoot  describes the sub-micrometer  aerosol population,
+composed  of  sea salt  \cite{Lundgren-2013},  mineral dust  particles
+\cite{Vogel-2006,   Stanelle-2010},  by   means  of   six  interacting
+log-normal modes.  For each  mode, mass contributions and total number
+concentration are prognostic  quantities, while the standard deviation
+is fixed.
+
+Specific  modules are included  to simulate  the dispersion  of pollen
+grains    \cite{Vogel-2008}    and    other   biological    particles.
+Meteorologically-influenced emissions  are also online  coupled within
+the  model  system.  The  biogenic  VOC  (volatile organic  compounds)
+emissions are  calculated as functions of  the land use  type based on
+the Global  Land Cover 2000  dataset and the modeled  temperatures and
+radiative fluxes \cite{Vogel-1995}.
+
+The gaseous chemistry in COSMO-ART  is solved by a modified version of
+the  Regional  Acid  Deposition  Model, Version  2  (RADM2)  mechanism
+\cite{Stockwell-1990}, which is extended to describe secondary organic
+aerosol  formation  based  on  a  volatile basis  set  (VBS)  approach
+\cite{Athanasopoulou-2013}  and  hydroxyl  radical  recycling  due  to
+isoprene chemistry  \cite{Geiger-2003}.  The thermodynamic equilibrium
+between the  gas and particulate  phases of the inorganic  material is
+achieved through the ISORROPIA II module \cite{Fountoukis-2007}.
+
+COSMO-ART  is  fully  online-coupled,  and  allows  for  feedbacks  of
+aerosols  on  temperature,  radiation  and cloud  condensation  nuclei
+(CCN).   Analytical  description  of  modules  incorporated  for  this
+purpose  exists  in  \cite{Vogel-2009, Bangert-2011}.   The  radiation
+scheme used  within the  model to calculate  the vertical  profiles of
+shortwave and longwave  radiative fluxes is GRAALS \cite{Ritter-1992}.
+In order to account for  the interaction of aerosol particles with the
+cloud microphysics and radiation,  COSMO-ART uses the two-moment cloud
+microphysics  scheme  of Seifert  and  Beheng \cite{Seifert-2006}  and
+comprehensive parameterizations for cloud condensation and ice crystal
+nucleation  \cite{Bangert-2011, Bangert-2012}.   The  system of  stiff
+ordinary differential equations described by chemical reactions in the
+aqueous-phase together with the transfer reactions is solved using the
+Kinetic PreProcessor (KPP, \cite{Damian-2002}).
+
+Further model description details  can be found in \cite{Bangert-2012,
+  Knote-2011, Knote-2013}.
 

### 2014-05-28 paper_enahpc14/Sec4:Experimental-results.tex
updates a few things according to the reviewers' suggestions

@@
 Isola  E1 Rack 2  and \pilat'  Isola HD  total power  measurements for
-1-day or 2-day  simulations. On the Intel Ivy  Bridge EP based cluster
-(i.e.   \monch), the  1-day simulation  was issued  only twice  due to
-usage restrictions. As  time resolution was set to  one update every 5
-minutes  for  power  sampling,  the  averaged  power  dissipation  was
-computed by  considering 6 values for  each single run.   On the Intel
-Xeon E5 based cluster (i.e.   \pilat), the 1-day simulation was issued
-four times and  a 2-day run only once.   Similarly, the averaged power
-dissipation was computed by considering 4 values for each single 1-day
-run  and  9 values  for  the  2-days  run. Corresponding  results  are
-gathered in Table~\ref{tab:1}.
+1-day or 2-day simulations. On \monch, the 1-day simulation was issued
+only twice due  to usage restrictions.  As time  resolution was set to
+one  update every  5 minutes  for power  sampling, the  averaged power
+dissipation was computed by considering  6 values for each single run.
+On \pilat, the 1-day simulation was  issued four times and a 2-day run
+only once.  Similarly, the  averaged power dissipation was computed by
+considering 4  values for each single  1-day run and 9  values for the
+2-days run. Corresponding results are gathered in Table~\ref{tab:1}.
 
@@
   \begin{tabular}{cccc}
-    \hline\noalign{\smallskip}   \textbf{\scriptsize{Xeon   E5}}   &
-    \textbf{\scriptsize{Ivy   Bridge}}   &  \textbf{\scriptsize{Xeon
-        E5645}}   &   \textbf{\scriptsize{Xeon   E5645}}  \\   &   &
-    \textbf{\scriptsize{(Polling)}}                                 &
-    \textbf{\scriptsize{(Blocking)}}
+    \hline\noalign{\smallskip}     \textbf{\scriptsize{PILATUS}}     &
+    \textbf{\scriptsize{MONCH}}  &  \textbf{\scriptsize{TINTORRUM}}  &
+    \textbf{\scriptsize{TINTORRUM}}           \\          &          &
+    \textbf{\scriptsize{(Polling)}} & \textbf{\scriptsize{(Blocking)}}
     \\    \noalign{\smallskip}\hline\noalign{\smallskip}   18035.0   &
@@
 Bridge  (2.6\,GHz  against  2.2\,GHz),   and  \emph{ii})  it  aims  at
-computing speed regardless to energy consumption.  In our experiments,
-Ivy Bridge EP showed  the best energy-to-solution, reducing the energy
-consumption of Xeon E5 by approximately 7\,\%.
+computing   speed  regardless   to  be   energy  efficient.    In  our
+experiments,  Ivy  Bridge   EP  showed  the  best  energy-to-solution,
+reducing the energy consumption of Xeon E5 by approximately 7\,\%.
 
@@
   \caption{Time-to-solution and  energy-to-solution comparison between
-    Xeon E5 and Ivy Bridge-EP architectures for a 24h simulation.}
+    \pilat and \monch for a 24h simulation.}
   \label{fig:3}

### 2014-05-28 trunk/paper_enahpc14/Sec3:Environment-setup.tex
updates Section 3 (detail)

@@
   % and  utilised  by  members  of  the  Swiss  Platform  for  Advanced
-  % Scientific  Computing (PASC)  project~\cite{PASC}.   Each standard
-  compute  node is  comprised of  two  Intel Ivy  Bridge EP  E5-2660v2
-  ten-core processors  operating at 2.2\,GHz, equipped  with 32\,GB of
-  DDR3 1600\,MHz RAM and  connected via InfiniBand Mellanox SX6036 and
-  FDR  switches.   For  our   experiments,  a  full  rack  constituted
+  % Scientific Computing (PASC) project~\cite{PASC}.  
+Each standard  compute node  is comprised of  two Intel Ivy  Bridge EP
+E5-2660v2  ten-core processors  operating at  2.2\,GHz,  equipped with
+32\,GB  of DDR3 1600\,MHz  RAM and  connected via  InfiniBand Mellanox
+SX6036 and FDR switches.  For our experiments, a full rack constituted
 exclusively of 52 nodes was used.

### 2014-05-29 paper_enahpc14/RebuttalLetter/rebuttal.tex
Manuel's changes + rebuttal letter

--- 2014-05-28 trunk/paper_enahpc14/RebuttalLetter/rebuttal.tex
@@
 \begin{weSay}
-The authors thank the reviewer for his/her general view of the paper.
+The authors thank the reviewer for his/her general view of the paper. We consider this work as
+a first step towards an energy-efficient version of COSMO-ART inside the Exa2Green project.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+We have fixed this issue along the manuscript. To do so we have explained in detail the specifications of the machines
+in the Environment setup section and afterwards we refer to the platforms' name or using the
+full processor's name and its microarchitecture.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+We have swapped the corresponding columns in the Figure 4, now ETS and TTS appear grouped.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+We have accordingly replaced ``power dissipation'' by ``power consumption'' along the manuscript.
 \end{weSay}
@@
 evaluation of polling vs blocking MPI strategies.
+
+%%% *** Manuel/Sandra ***
+\begin{weSay}
+The authors thank the reviewer for his/her general view of the paper.
+\end{weSay}
 
@@
 \newpage
-\section*{Reviewer \#2}
+\section*{Reviewer \#3}
 
@@
 \begin{weSay}
-blabla
+We have explained in detail the kind of nodes that this platform is comprised of.
+We only used a subset of 52 in our experiments.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+Fixed.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+We recognize the mistake and fix the sentence in the manuscript.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+We added information about the Turbo Boost feature for each platform \textbf{(MISSING FOR MONCH and PILATUS!!)}
 \end{weSay}
@@
 \begin{weSay}
-blabla
+We added information about the power saving mechanisms available on each platform \textbf{(MISSING FOR MONCH and PILATUS!!)}
 \end{weSay}
@@
 \begin{weSay}
-blabla
+We have fixed this mistake. HyperThreading was not considered in our experiments.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+\textbf{MISSING ANSWER!!}
+
+Figure 5 represents the power drawn of Tintorrum, for this case the PDU used has a sampling rate of 1 Hz instead of 5 minutes.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+We have added information about the accuracy of the PDU.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+We noticed the problem and fixed accordingly along the manuscript the misused blocking term.
+Now we refer to ``aggressive'' and ``degraded'' OpenMPI policies instead of ``polling''
+and ``blocking''. The reviewer is completely right on this issue, however even when using
+the ``degraded'' policy the CPU utilization is readuced to 50\,\%, thus saving energy.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+We have fixed the sentence accordingly so as not to be confusing.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+We have fixed this issue along the manuscript.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+The reviewer is right in this sense. The processors are not idle during blocking MPI routines, however a reduction of the CPU utilization is observed (50\,\%).
 \end{weSay}

### 2014-05-29 paper_enahpc14/Conclusion.tex
Manuel's changes + rebuttal letter

--- 2014-05-06 paper_enahpc14/Conclusion.tex
@@
 benchmarks illustrate  that the  best time-to-solution does  not imply
-the best energy-to-solution: an Intel Sandybridge (2.6 GHz) system has
-lower  time-to-solution but  higher energy-to-solution  than  an Intel
-Ivybridge  (2.2  GHz) system,  although  the  two  metrics are  indeed
-strongly correlated.  On the  other hand, the usage of energy-friendly
-MPI waiting  techniques analysed with  our power-performance framework
-over  a  series  of runs  of  \cosmoart  on  \tinto and  power  traces
-demonstrate that it  is possible to reduce both  power dissipation and
+the best energy-to-solution: an Intel Sandy Bridge (2.6 GHz) system has
+lower time-to-solution but higher energy-to-solution than an Intel Ivy
+Bridge (2.2 GHz) system, although  the two metrics are indeed strongly
+correlated.   On the  other  hand, the  usage  of energy-friendly  MPI
+waiting techniques analyzed  with our power-performance framework over
+a series of  runs of \cosmoart on \tinto  and power traces demonstrate
+that   it  is   possible  to   reduce  both   power   dissipation  and
 energy-to-solution   while  maintaining   (or  even   decreasing)  the
 time-to-solution. The resulting profiles indicate that simple changes,
-such as making use of the  blocking MPI policy rather than the polling
+such as making use of the  \emph{degraded} MPI policy rather than the \emph{aggressive}
 is  possible to  slightly  reduce both  power  dissipation and  energy
@@
 of \cosmoart. Profiling  has given us insight into  the most expensive
-code  components,  which are  now  being  altered  to utilise  revised
-algorithms.  The  results of these  optimisations will be  reported in
+code  components,  which are  now  being  altered  to utilize  revised
+algorithms.  The  results of these  optimizations will be  reported in
 future publications.

### 2014-05-29 paper_enahpc14/Introduction.tex
Manuel's changes + rebuttal letter

--- 2014-05-05 paper_enahpc14/Introduction.tex
@@
 a negative indirect  radiative forcing.  Despite considerable progress
-in  global aerosol  modelling  \cite{Mann-2013} and  measurement-based
+in  global aerosol  modeling  \cite{Mann-2013} and  measurement-based
 assessments,  large  uncertainties  remain  in  current  estimates  of
-aerosol  radiative   forcing  \cite{Myhre-2013,  IPCC-2013,  Lee-2013,
-Sherwood-2013, Stier-2013}.
+aerosol  radiative   forcing  \cite{IPCC-2013,  Lee-2013,  Myhre-2013,
+Randles-2013, Rosenfeld-2013, Sherwood-2013, Stier-2013}.
 
@@
 (KIT)  \cite{Vogel-2009,  Bangert-2011,  Knote-2013}.   This  regional
-scale  modelling  system  is  coupled  with  the  operational  weather
+scale  modeling  system  is  coupled  with  the  operational  weather
 forecast  model \textsc{Cosmo} \cite{Baldauf-2011},  jointly developed

### 2014-05-29 paper_enahpc14/Sec3:Environment-setup.tex
Manuel's changes + rebuttal letter

@@
 \begin{itemize}
-\item \monch is a 10-rack NEC-provided clus\-ter at CSCS composed
-  of 312 compute  nodes. %and utilised by members  of the Swiss Platform
-  %for Advanced Scientific  Computing (PASC) project~\citep{PASC}
-  Each node is comprised of two Intel Ivy Bridge EP  E5-2660 ten-core processors
-  operating at  2.2 GHz, equipped with 32GB of DDR3 1600\,MHz  RAM and connected  via InfiniBand Mellanox  SX6036 and FDR switches. For our  experiments, 
-  a full rack constituted of 52 nodes was used. % \monch is  slated to stay
-  %in  service  without  hardware  upgrade  for  the  duration  of  the
-  %Exa2Green  project,  allowing   an  identical  configuration  to  be
-  %recreated for future assessments of the baseline.\\
+\item \monch is  a 10-rack NEC-provided clus\-ter at  CSCS composed by
+  384   nodes  grouped  in  312  standard  compute  nodes,  24
+  large-memory compute nodes,  24 huge-memory nodes, 16 administration
+  nodes and 8 servers to provide I/O to Lustre.
+  % and  utilized  by  members  of  the  Swiss  Platform  for  Advanced
+  % Scientific Computing (PASC) project~\cite{PASC}.  
+Each standard  compute node  is comprised of  two Intel Xeon Ivy Bridge EP
+E5-2660v2  ten-core processors  operating at  2.2\,GHz,  equipped with
+32\,GB  of DDR3 1600\,MHz  RAM and  connected via  InfiniBand Mellanox
+SX6036 and FDR switches.  For our experiments, only a subset of 52 was used. %
+% only a full rack constituted exclusively of 52 nodes was used.
+  % \monch is  slated to stay  in service without hardware  upgrade for
+  % the  duration  of  the  Exa2Green project,  allowing  an  identical
+  % configuration  to  be  recreated  for  future  assessments  of  the
+  % baseline.
 
-\item \pilat is a dual-socket eight-core Intel Sandy
-  Bridge EP based cluster at  CSCS composed of 42 compute nodes. %and used as Piz Daint pre-post  processing cluster 
-  %The 2 login and 42 compute nodes %are consolidated into 11 twin-pair Intel E5-Series DALCO
-  %r2264i4t 2U scalable compute modules, each containing 4 compute
-  %nodes based 
-  %consisting of 
-  Each node is comprises two Intel Xeon E5-2670 processors operating  at 2.6\,GHz equipped with 64GB of DDR3 1600\,MHz RAM and connected  via InfiniBand Mellanox SX6036 and FDR switches. 
-  For our  experiments, a full rack constituted of 42 compute nodes
-  was used. \pilat  is based  on  Intel's second-generation
-  processors, conventional in  HPC systems and known to  be more power
-  consuming than its Ivy Bridge successor.\\
+\item \pilat is a cluster at CSCS composed of 42 compute nodes.  
+  % and used as Piz Daint pre-post processing cluster. The 2 login and
+  % 42  compute  nodes  are   consolidated  into  11  twin-pair  Intel
+  % E5-Series  DALCO  r2264i4t   2U  scalable  compute  modules,  each
+  % containing 4 compute nodes based on two ...
+  Each of them is comprised of two Intel Xeon Sandy Bridge EP E5-2670 eight-core processors operating
+  at 2.6\,GHz equipped with 64\,GB of DDR3 1600\,MHz RAM and connected
+  via  InfiniBand   Mellanox  SX6036   and  FDR  switches.    For  our
+  experiments, a subset constituted of 42 compute  nodes was used.
+  % \pilat   is   based   on   Intel's   second-generation   processors,
+  % conventional in  HPC systems  and known to  be more  power consuming
+  % than its Ivy Bridge successor.
 
-\item \tinto is  a heterogeneous cluster composed of
-  28 compute nodes at UJI. For our  experiments only a subset of 16 homogeneous nodes was considered. Each node is composed of two Intel Xeon E5645 hexa-core processors 
-  running at 2.4\,GHz, equipped with 24\,GB  of DDR3 1333\,MHz and connected via Infiniband QDR with a Mellanox MTS\-3600 switch.
-
-  % A  full tracing  experiment is  conducted on
-  % \tinto in order to  capture an overall  power profile at a much
-  % finer resolution. In this sense, we also analyze the contribution of the MPI library
-  % to the energy consumption and how the use of the blocking versus polling
-  % message-passing policies can potentially render energy savings.
+\item \tinto is  a heterogeneous cluster composed of  28 compute nodes
+  at UJI.  For our experiments only  a subset of  16 homogeneous nodes
+  was  considered. Each  node  is  comprised of  two  Intel Xeon Westmere EP E5645
+  hexa-core processors  running at  2.4\,GHz, equipped with  24\,GB of
+  DDR3  1333\,MHz and  connected via  InfiniBand QDR  with  a Mellanox
+  MTS\-3600 switch. Power saving mechanisms are enabled on the nodes are the C-states C1, C3 and C6, 
+  but no P-states (except P0) are used. TurboBoost feature is enabled for our experiments.
 \end{itemize}
@@
 
-We present here the two different  frameworks deployed on the testbed HPC systems to
-measure  performance, power  dissipation and energy consumption of the baseline execution of \cosmoart.
+We present here  the two different frameworks deployed  on the testbed
+HPC  systems  to measure  performance,  power and  energy
+consumption of the baseline execution of \cosmoart.
 
@@
 
-Supercomputer  clusters considered  for our  experiments at CSCS of ETH Zurich are equipped with
-E3METER Intelligent  Power Strips (IPS)  and Monitors (IPM)  which are
-high   quality   electricity  meters   released   by  Riedo   Networks
-\citep{Riedonetworks},   that  enable   to  monitor   and   log  power
-consumption  of the IT  infrastructure as  well as  constantly analyse
-line  voltage,  current, power-factor,  frequency  with 1\,\%  accuracy.
-Using  reliable narrowband  powerline communication  (PLC) technology,
-all  metering and  power  quality  data from  each  IPS are  centrally
-collected  by the E3METER  Data Concentrator,  via the  existing power
-cables thus  avoiding the need for  extra cabling.  This  data is made
-available via  SNMP, HTTP and TELNET  through the built-in  Fast Ethernet
-port.   Time  synchronisation  is  guaranteed by  using  NTP  servers.
-Measured  data is  accessed  through the  open  source Cacti  software
-including the E3METER Cacti Plugin  to scan the entire PLC network and
-monitor in real-time the power usage of individual rack, recorded in 5
-minutes interval periods.
+Supercomputer clusters  considered for our experiments at  CSCS of ETH
+Zurich are  equipped with E3METER  Intelligent Power Strips  (IPS) and
+Monitors (IPM)  which are high quality electricity  meters released by
+Riedo Networks,  that enable to  monitor and log power  consumption of
+the  IT infrastructure  as well  as constantly  analyze  line voltage,
+current, power-factor, frequency  with $\pm1\,\%$ accuracy.  Using reliable
+narrowband power-line communication (PLC) technology, all metering and
+power  quality data  from  each  IPS are  centrally  collected by  the
+E3METER Data Concentrator, via the existing power cables thus avoiding
+the need  for extra  cabling.  This data  is made available  via SNMP,
+HTTP  and  TELNET  through  the  built-in Fast  Ethernet  port.   Time
+synchronization is guaranteed by  using NTP servers.  Measured data is
+accessed through the open  source Cacti software including the E3METER
+Cacti Plugin to  scan the entire PLC network  and monitor in real-time
+the power  usage of  individual rack, recorded  in 5  minutes interval
+periods.
 
@@
 
-For experiments at UJI on \tinto, we employed a version
-of the integrated framework presented in~\cite{energy13} that employs the \pmlib library in
-combination with Extrae and Paraver, which  are profiling/tracing and
-visualisation tools, respectively.
+For  experiments  at UJI  on  \tinto, we  employed  a  version of  the
+integrated  framework presented  in~\cite{energy13}  that employs  the
+\pmlib  library in  combination  with Extrae  and  Paraver, which  are
+profiling/tracing and visualization tools, respectively.
 
-To employ our framework, \cosmoart is first compiled and linked with the Extrae instrumentation library, which automatically  instruments the Fortran~90 code of the model, e.g., to record MPI calls.
-Next, \cosmoart  is  run on  the  nodes,  thus dissipating  certain amount  of power.
-The  nodes are connected to power distribution units (PDUs)
-APC 8653 that account for the dissipated power with a sampling rate of 1\,Hz.
-The power data captured by these PDUs is collected by a \pmlib tracing server running on a separate machine.
-The client, running on the target nodes, sends start/stop  primitive in order to collect
-the power data during the \cosmoart run.
-Once  the run  is finished,  a power-performance trace file containing the power data is received from the tracing server and the instrumentation post-process is generated.
-A merged trace file is finally loaded in Paraver to visualise and correlate the performance trace and with the power profile generated by \cosmoart.
-This procedure allows to detect power bottlenecks, basically constituted by inefficient waiting methods.
+To employ our  framework, \cosmoart is first compiled  and linked with
+the  Extrae instrumentation  library, which  automatically instruments
+the Fortran~90  code of the model,  e.g., to record  MPI calls.  Next,
+\cosmoart  is run  on the  nodes, thus  consuming certain  amount of
+power. The  nodes are connected  to APC 8653 Power  Distribution Units
+(PDUs) that account  for the consumed   power with  a sampling rate of
+1\,Hz and $\pm3\,\%$ accuracy. The power data captured by  these PDUs is collected by a \pmlib
+tracing server running  on a separate machine. The  client, running on
+the target nodes, sends start/stop  primitives in order to collect the
+power  data during  the \cosmoart  run. Once  the run  is  finished, a
+power-performance  trace file  containing the  power data  is received
+from  the  tracing  server  and the  instrumentation  post-process  is
+generated.  A  merged trace  file  is  finally  loaded in  Paraver  to
+visualize  and correlate  the  performance trace  and  with the  power
+profile  generated by \cosmoart.  This procedure  allows us  to easily
+detect power bottlenecks, basically constituted by inefficient waiting
+methods.
 
@@
 
-The \cosmoart  baseline is a  pure MPI-based  Fortran~90 code
-currently  running  on   distributed  multi-core  systems  only.   The
-software stack on the platforms was controlled using the modules
-framework which gives an easy  and flexible mechanism to access to all
-of the  provided compilers, tools  and applications.  For  our initial
-benchmarking,   we  opted   for   the  GNU   compiler  GCC 4.8.1   on
-\monch and GCC 4.8.2 on  \pilat and \tinto) using the -O3 compiler
-flag  in favor  of the  Intel compilers 14.0.1,  which delivered inferior
-performance.
-In  addition, we installed the following implementations of MPI: MVAPICH2 1.9 on \monch and \pilat and  OpenMPI 1.6.5 on \tinto.
-On the other hand, we  used HDF5 1.8.12 and NetCDF  (4.3.1) libraries for  the management  of extremely  large and
-complex data collections.  All computes nodes have an operating system
-based on GNU/Linux featuring ``2.6.32-358.11.1.el6'' kernel in
-\monch, ``3.0.101-0.15-default'' in \pilat and ``2.6.18-238.19.1.el5'' in \tinto.
+The \cosmoart  baseline is a pure MPI-based  Fortran~90 code currently
+running on distributed multi-core systems only.  The software stack on
+the platforms  was controlled using the modules  framework which gives
+an  easy and  flexible  mechanism to  access  to all  of the  provided
+compilers, tools  and applications.  For our  initial benchmarking, we
+opted for the GNU compiler GCC 4.8.1 on \monch and GCC 4.8.2 on \pilat
+and  \tinto)  using  the \texttt{-O3}  compiler  flag  in  favor of  the  Intel
+compilers 14.0.1, which  delivered inferior performance.  In addition,
+we  installed the following  implementations of  MPI: MVAPICH2  1.9 on
+\monch and \pilat, and OpenMPI 1.6.5  on \tinto. On the other hand, we
+used HDF5  1.8.12 and NetCDF  (4.3.1) libraries for the  management of
+extremely large and complex data collections.  All computes nodes have
+an     operating    system     based     on    GNU/Linux     featuring
+``2.6.32-358.11.1.el6'' kernel  in \monch, ``3.0.101-0.15-default'' in
+\pilat and ``2.6.18-238.19.1.el5'' in \tinto.
 
@@
 information needed  to reproduce the  energy-to-solution benchmarks of
-\cosmoart, was produced and run on 1040 cores using 20~MPI
-tasks  per node on  \monch, on 1344 cores  using 16~MPI
-tasks per node on  \pilat and on 192 cores  using 12~MPI
-tasks per node on \tinto.  The calculated region was mapped
-to the participating processors using a 2D-partitioning strategy.  The
-distribution along the $x$ and $y$ coordinates was defined accordingly, keeping in mind
-% $by setting:
-%$nprocx=40$  and $nprocy=26$  for \monch  and  $nprocx=28$ and
-%$nprocy=24$  as  
-to keep $x$  bigger  than  $nprocy$.
-Besides,  as this  version does not  make use  of the  traditional GRIB
-library, we  specified $nprocio=0$  for GRIB I/O.   Hyperthreading was
-not considered in this study  as it previously revealed that it always
-led to higher energy-to-solution.
+\cosmoart, was produced  and run on 1040 cores  using 20~MPI tasks per
+node on  \monch, on 672 cores  using 16~MPI tasks per  node on \pilat
+and  on  192  cores  using  12~MPI  tasks per  node  on  \tinto.   The
+calculated region  was mapped to the participating  processors using a
+2D-partitioning  strategy.  The  distribution  along the  $x$ and  $y$
+coordinates was  defined by  setting: $nprocx=40$ and  $nprocy=26$ for
+\monch,  $nprocx=28$ and  $nprocy=24$ for  \pilat and  $nprocx=16$ and
+$nprocy=12$  for \tinto as  $nprocx$ is  usually kept  slightly bigger
+than  $nprocy$. Besides,  as this  version does  not make  use  of the
+traditional  GRIB  library, we  specified  $nprocio=0$  for GRIB  I/O.
+Hyperthreading  was not  considered  in this  study  as it  previously
+revealed that it always led to higher energy-to-solution.

### 2014-05-29 trunk/paper_enahpc14/Sec2:COSMO-ART.tex
Manuel's changes + rebuttal letter

@@
 ordinary  differential equations  \cite{Damian-2002}.   Concerning the
-modelling  of  wet  deposition  in  aerosols, the  baseline  has  only
+modeling  of  wet  deposition  in  aerosols, the  baseline  has  only
 indirect  cloud feedbacks  but  does not  include in-cloud  scavenging

### 2014-05-29 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes + rebuttal letter

@@
 time. ETS  is the amount of  energy spent to  achieve results.  Energy
-consumption  is  assessed by  sampling  the  power during execution which is then averaged and multiplied  by the TTS to determine  ETS. Whenever  possible,  multiple production  runs  were
-performed  to  illustrate the  reproducibility  of  the baseline,  and
-quantify the  significant uncertainties  in the power  measurement, as
-dictated by the available technology.
+consumption is  assessed by sampling the power  during execution which
+is then averaged and multiplied  by the TTS to determine ETS. Whenever
+possible, multiple  production runs  were performed to  illustrate the
+reproducibility  of   the  baseline,  and   quantify  the  significant
+uncertainties in  the power measurement, as dictated  by the available
+technology.
 
-For the experiments with \tinto platform, we also analyze the contribution of the MPI library to the energy consumption and how the use of the blocking versus polling message-passing policies can potentially render energy savings.
-Specifically, the OpenMPI 1.6.5 library,
-features two operation modes, blocking and polling, which can be selected before \cosmoart
-is launched. In the polling configuration (the default mode) ``MPI'' continually polls
-the network interface to check for the completion of an event (e.g. send or receive).
-This usually yields low latency but high CPU utilization. Thus, one can expect that this mode 
-attains the best performance, possibly at the cost a higher energy usage. In the blocking
-mode, the CPU is yield to other processes/threads if there are no incoming messages.
+For  the   experiments  on  \tinto  platform,  we   also  analyze  the
+contribution of the MPI library  to the energy consumption and how the
+use  of  the aggressive and degraded message-passing progression engine 
+policies can potentially  render energy  savings.
 
-\subsection{Time-power-energy analysis of \cosmoart}
+Specifically,   the  OpenMPI 1.6.5  library,  features  two  operation 
+policies, \emph{aggressive} and \emph{degraded}, which can be  selected 
+before \cosmoart is launched. In the \emph{aggressive} (default) policy,  
+normally used in exactly- or under-subscribed mode  (i.e., the  number 
+of running processes is equal to or less than the number of  available 
+cores), MPI processes will never voluntarily give up the processor  to 
+other processes. With some network transports, OpenMPI processes  will 
+spin  in  tight  loops  attempting  to  make message passing progress, 
+effectively causing other processes to not get  any  CPU  cycles  (and  
+therefore never make any  progress).  This  usually  yields  the  best 
+performance but high CPU utilization, possibly at the cost of a higher 
+energy usage.  On the other hand, the \emph{degraded} policy, suitable 
+for the oversubscribed mode (i.e., more  processes  are  running  than 
+total cores  available),  MPI  processes  will  frequently  yield  the 
+processor to its peers, thereby allowing all processes to make progress.
+
+In our experiments with \cosmoart only the exactly-subscribed mode  is
+used. Therefore, \cosmoart processes running  in  the  \emph{degraded} 
+policy are constantly checking for the completion  of  an  event (e.g. 
+send or receive) and repeatedly calling to  \texttt{sched\_yield()} to 
+be picked again by the Linux scheduler. Although one can expect no benefits 
+from   using   the  \emph{degraded}   policy   compared   with     the 
+\emph{aggressive} one, notable differences can be observed in the CPU utilization. 
+In a separated  test  with  the  \emph{aggressive}  policy 
+enabled, we noticed that a blocked MPI process waiting for an incoming 
+message has CPU utilization of 100\,\%,  however  when  the \emph{degraded} 
+mode  is  set,  the  CPU  utilization is only $50\,\%$, approximately. 
+Although neither of both policies avoid busy-waiting loops in blocking 
+routines, the use of the \emph{degraded} policy is as much as we can 
+do in OpenMPI to reduce the CPU utilization, and thus save energy.
+
+% polling configuration
+% (the default mode) ``MPI''  continually polls the network interface to
+% check for  the completion  of an event  (e.g. send or  receive).  This
+% usually yields  low latency  but high CPU  utilization. Thus,  one can
+% expect that  this mode attains  the best performance, possibly  at the
+% cost a higher energy usage. In  the blocking mode, the CPU is yield to
+% other     processes/threads    if     there     are    no     incoming
+% messages~\cite{Castillo-2012}.
+
+\subsection{Time-power-energy analysis}
 \label{subsec:4.2}
 
+Figures~\ref{fig:1} and \ref{fig:2}  account respectively for \monch's
+Isola  E1 Rack 2  and \pilat'  Isola HD  total power  measurements for
+1-day or 2-day simulations. On \monch, the 1-day simulation was issued
+only twice due  to usage restrictions.  As time  resolution was set to
+one  update every  5 minutes  for power  sampling, the  averaged power
+consumption was computed by considering  6 values for each single run.
+On \pilat, the 1-day simulation was  issued four times and a 2-day run
+only once.  Similarly, the  averaged power consumption was computed by
+considering 4  values for each single  1-day run and 9  values for the
+2-days run. Corresponding results are gathered in Table~\ref{tab:1}.
+
 \begin{figure}[htbf]
-  \begin{center}
+  \centering
   \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
-    \caption{\monch: Isola E1 Rack 2 Total Power.}
+  \caption{Isola E1 Rack 2 total power of \monch.}
   \label{fig:1}
-  \end{center}
 \end{figure}
@@
 \begin{figure}[htbf]
-  \begin{center}
+  \centering
   \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
-    \caption{\pilat: Isola HD Total Power.}
+  \caption{Isola HD total power consumption of \pilat.}
   \label{fig:2}
-  \end{center}
 \end{figure}
 
-Figures~\ref{fig:1}   and   \ref{fig:2}   account   respectively   for
-\monch's Isola E1 Rack  2 and \pilat' Isola HD total
-power measurements for  1-day or 2-day simulations. On  the Intel Ivy
-Bridge EP  based cluster (i.e.  \monch),  the 1-day simulation
-was issued  only twice due  to usage restrictions. As  time resolution
-was set to one update every  5 minutes for power sampling, the of average
-power dissipation was computed by considering 6 values for each single
-run.  On the Intel Xeon E5 based cluster (i.e.  \pilat), the
-1-day simulation  was issued  four times and  a 2-day run  only once.
-Similarly, the average of power dissipation was computed by considering 4
-values  for  each  single  1-day  run  and 9  values  for  the  2-days
-run. Corresponding results are gathered in Table~\ref{tab:3}.
-
 \begin{table}[htbf]
-  \begin{center}
-    \caption{Average power dissipation (W) of the platforms}
-    \label{tab:3}
+  \centering
+  \caption{Averaged power consumption (W) of the platforms.}
+  \label{tab:1}
   \begin{tabular}{cccc}
-      \hline\noalign{\smallskip}
-      \textbf{\scriptsize{Xeon E5}} & \textbf{\scriptsize{Ivy Bridge}} & \textbf{\scriptsize{Xeon  E5645}} & \textbf{\scriptsize{Xeon  E5645}}\\
-      & & \textbf{\scriptsize{(Polling)}} & \textbf{\scriptsize{(Blocking)}} \\
-      \noalign{\smallskip}\hline\noalign{\smallskip}
-      18035.0 & 12622.5 & 3713.6 & 3651.8 \\ 
-      \noalign{\smallskip}\hline
+    \hline\noalign{\smallskip}     \textbf{\scriptsize{PILATUS}}     &
+    \textbf{\scriptsize{MONCH}}  &  \textbf{\scriptsize{TINTORRUM}}  &
+    \textbf{\scriptsize{TINTORRUM}}           \\          &          &
+    \textbf{\scriptsize{(\emph{Aggressive})}} & \textbf{\scriptsize{(\emph{Degraded})}}
+    \\    \noalign{\smallskip}\hline\noalign{\smallskip}   18035.0   &
+    12622.5 & 3713.6 & 3651.8 \\ \noalign{\smallskip}\hline
   \end{tabular}
-  \end{center}
 \end{table}
@@
 In  Figure~\ref{fig:3}, we  compare both  TTS (right  Y-axis)  and ETS
-(left  Y-axis)  metrics  on  both  platforms.  As  expected,  Xeon  E5
-outperforms Ivy Bridge EP, being  roughly 130\,\% faster.  The reason for
-that is  two-fold: \emph{i}) it has  higher clock frequency  than Ivy Bridge
-(2.6\,GHz  against  2.2\,GHz),  and  \emph{ii})  it  aims at  computing  speed
-regardless to  energy consumption.  In our experiments,  Ivy Bridge EP
-showed the best energy-to-solution, reducing the energy consumption of
-Xeon E5 by approximately 7\,\%.
+(left  Y-axis)  metrics  on  both  platforms.  As  expected, Intel Sandy Bridge EP processors of \pilat
+outperform the Intel Ivy  Bridge EP processors of \monch, being roughly 130\,\%  faster.  The reason
+for that is two-fold: \emph{i}) they have higher clock frequency than Ivy
+Bridge  (2.6\,GHz  against  2.2\,GHz),   and  \emph{ii})  they  aim  at
+computing   speed  regardless   to  be   energy  efficient.    In  our
+experiments,  \monch  showed  the  best  energy-to-solution,
+reducing the energy consumption of \pilat by approximately 7\,\%.
 
-Averaged TTS and ETS results over 10 repetitions of 24h simulation on \tinto are shown in Figure~\ref{fig:4}. Correponding results for average total power are shown in Table~\ref{tab:3}.  
-As expected, the average total power dissipated by \cosmoart while using the blocking MPI policy is about 2\,\% less than that consumed by the polling policy.
-This is due to the power-friendly (blocking) waits performed by the CPU cores while using the blocking MPI mode, in contrast to the polling waits that take place when the polling MPI mode is set.
-From the point of view of the execution time, there is a slight reduction for the blocking MPI mode.
-One can expect that the costs of the interruptions used in the blocking MPI policy can lead to an increase of the total execution time, however, applications with extensive computing, as e.g.
-\cosmoart, may show better performance with the blocking modes.
-This effect works in favor with our results since the energy consumed is also reduced (2\,\%) just by changing the behavior of the OpenMPI library to the blocking policy, i.e, with the \texttt{--mca mpi\_yield\_when\_idle} parameter set in the \texttt{mpirun} command.
+Averaged TTS and ETS results  over 10 repetitions of 24h simulation on
+\tinto are shown in  Figure~\ref{fig:4}. The corresponding results for
+the  averaged total  power are  also shown  in  Table~\ref{tab:1}.  As
+expected, the average total power consumption by \cosmoart while using
+the \emph{degraded} MPI policy is about  2\,\% less than that consumed by the
+\emph{aggressive} one. This is  due to  the reduction of CPU usage (of about 50\,\%) while using  the \emph{degraded} MPI policy, in
+contrast with the \emph{aggressive} policy that always yields to 100\,\% of CPU utilization while waiting for an incoming message. From the point of  view of the execution time, there is a
+slight reduction  for the \emph{degraded} MPI  policy. One can  expect that the
+costs of the context switches produced in the \emph{degraded} MPI policy can lead to
+an  increase  of  the   execution  time,  however,  applications  with
+extensive computing,  as e.g.  \cosmoart, may show  better performance
+with the \emph{degraded} policy. This effect works in  favor with our results
+since the energy consumed is also reduced (2\,\%) just by changing the
+behavior of the OpenMPI library  to the \emph{degraded} policy, i.e, with the
+\texttt{--mca    mpi\_yield\_when\_idle 1}   parameter   set    in   the
+\texttt{mpirun} command.
 
 \begin{figure}[htbf]
+  \centering
   \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-0.eps}
   \caption{Time-to-solution and  energy-to-solution comparison between
-    Xeon E5 and Ivy Bridge-EP architectures for a 24h simulation.}
+    \pilat and \monch for a 24h simulation.}
   \label{fig:3}
@@
 \begin{figure}[htbf]
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-5.eps}
-  \caption{Mean time-to-solution and energy-to-solution on
-    \tinto for a 24h simulation using the polling and blocking MPI policies. The 95\,\% confidence intervals of Student's $t$ distribution (10 samples) clearly indicates 
-    an improvement for blocking policy in both TTS and ETS.}
+  \centering
+  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
+  \vspace{-1cm}
+  \caption{Mean time-to-solution and  energy-to-solution on \tinto for
+    a 24h simulation using the  \emph{aggressive} and \emph{degraded} MPI policies. The
+    95\,\%  confidence  intervals of  Student's  $t$ distribution  (10
+    samples) clearly  indicates an improvement for the \emph{degraded} policy in
+    both TTS and ETS.}
   \label{fig:4}
@@
 
-\subsection{Power-performance tracing on \tinto}
-\label{subsec:4.3}
-
-A full tracing experiment is conducted on \tinto in order to capture an overall power
-profile at a much finer resolution using the attached PDUs in each node. First we run \cosmoart 
-for a full-day simulation and analyze the power consumption. Second we correlate performance 
-with power traces of two 1-hour time frames during the day: mid-day (11h--12h) and
-midnight (23h--24h). All the experiments were performed using 192 \cosmoart processes
-m 12 processes in each of node.
-
-% For all the experiments we also analyze the contribution of the MPI library to the energy 
-% consumption and how the use of the blocking versus polling message-passing policies can 
-% potentially render energy savings. Specifically, the OpenMPI 1.6.5 library installed in \tinto, 
-% features two operation modes, blocking and modes, which can be selected before \cosmoart is 
-% launched. In the polling configuration (the default mode) ``MPI'' continually polls
-% the network interface to check for the completion of an event (e.g. send or receive).
-% This usually yields low latency but high CPU utilization. Thus, can one expect that this mode attains the best performance, possibly at the cost a higher energy usage. In the blocking
-% mode, the CPU is yield to other processes/threads if there are no incoming messages.
-
-In order to obtain an initial overview of the power dissipation of the system model, we obtained power profiles leveraging our \pmlib power-measurement framework over a full-day simulation of \cosmoart.
-These profiles, obtained with both polling and blocking MPI policies, are depicted in  Figure~\ref{fig:9}.
-The first observation for the polling MPI policy is that the first and last hours of the day dissipate slightly less power than the mid-day hours.
-We relate these variations due to the chemical reactions and aerosols computed by ART, that are taking place during daylight hours but not during night.
-In this case peak power consumption is about 3821\,W.
-Looking at the power profile using the blocking MPI policy, we observe that this pattern is repeated, favoring the reduction of power during night hours in a more noticeable way than that for the polling case.The peak power dissipation during day hours in this mode is about 3768\,W, which means a reduction of around 50\,W with respect to the polling mode.
-Also, we notice systematic power drops in each hour simulated.
-We relate them to periods in which the processes were blocked waiting for incoming messages or synchronizations, thus yielding the CPU to low usage during.
-Our final observation is the reduction of the TTS in a factor of 2\,\% with the blocking MPI policy with respect to the polling one.
-
-To gain more insights about the power-performance behavior of \cosmoart, we obtained power-per\-for\-man\-ce traces so that they allow us to correlate the type of tasks being executed with the total power dissipated at each time (see Figure~\ref{fig:11}).
-For this experiment we limit the simulation time to 1 hour because \emph{i}) the pattern of the computation performed by \cosmoart is repeated hourly and, \emph{ii}) the weight of a full-day simulation trace can not be easily handled using our machines and memory available (each trace file was about 45\,GB).
-To face the problem we shrink the simulations to 1 hour and test a time frame during midday (from 11h to 12h) and during midnight (from 23h to 24h), in which \cosmoart perform different internal computations due to the different chemical reactions that happen during day and night.
-At the same time we also leverage the two already stated MPI policies: polling and blocking.
-
-All the traces in Figure~\ref{fig:11} are divided in different phases.
-In first phase all the processes perform group communications and execute MPI-receive primitives (e.g.
-\texttt{MPI\_Recv}).
-The second performs computation bursts, in between group communications, blocking sends, immediate receives and MPI-wait primitives are executed.
-Finally, group communications interlaced with computation and a significant synchronization phase is performed.
-From the performance point of view, one can notice a slight reduction of the TTS that blocking MPI mode leads over the polling MPI policy (around 2\,\%).
-A reduction of power of about  using the blocking MPI policy is produced only in routines that actually perfom synchronizations, group communications and MPI-receive/-wait primitives.
-Thus, the reduction of the pair power-energy directly depends on the percentage of time in which blocking MPI operations are executed.
-For our 1 hour simulation traces, an average of 50\,\% of the time the processes were blocked.
-Taking into account that  the total power reduction during blocking phases is about 5\,\%, one can compute the reduction of power, i.e., $50\,\% \times 5\,\% = 2.5\%$, which is consistent power reduction obtained for the full-day simulation (2\,\%).
-
 \begin{figure*}[ht]
   \centering
-  \hspace{0.8cm}
-  \scalebox{0.55}{\input{Figs/24hour.pstex_t}}
-  \caption{24 hours simulation trace using the MPI blocking mode on \tinto.}
-  \label{fig:9}
+  \hspace{0.8cm} \scalebox{0.62}{\input{Figs/24hour.pstex_t}}
+  \caption{24  hours  simulation power  trace  using  the \emph{aggressive}  and
+    \emph{degraded} MPI policies on \tinto.}
+  \label{fig:5}
 \end{figure*}
-
-% \begin{figure*}[htbf]
-%   \centering
-%   Simulation of 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
-%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq0_stat.eps}\\
-%   Percentage of time for the 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
-%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq1_stat.eps}\\
-%   \caption{blabla}
-%   \label{fig:8}
-% \end{figure*}
 
@@
   \scriptsize
-  Simulation of 1 hour during midday (11h--12h) leveraging the polling MPI policy\\
-  \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}}\\
-  Simulation of 1 hour during midday (11h--12h) leveraging the blocking MPI policy\\
-  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}}\\
-  1 hour simulation during midnight (23h--24h) leveraging the polling MPI policy\\
-  \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}}\\
-  1 hour simulation during midnight (23h--24h) leveraging the blocking MPI policy\\
-  \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}}\\
+  Simulation of 1 hour during midday (11h--12h) using the \emph{aggressive} MPI policy \\
+  \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}} \\
+  Simulation of 1 hour during midday (11h--12h) using the \emph{degraded} MPI policy \\
+  \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}} \\
+  1 hour simulation during midnight (23h--24h) using the \emph{aggressive} MPI policy \\
+  \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}} \\
+  1 hour simulation during midnight (23h--24h) using the \emph{degraded} MPI policy \\
+  \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}} \\
   \begin{tabular}{rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rlp{-0.2cm}rl}
@@
   \end{tabular}
-  \caption{Power-performance traces during midday and midnight leveraging the polling and blocking MPI policies on \tinto.}
-  \label{fig:11}
+  \caption{Power-performance   traces  during   midday   and  midnight
+    leveraging the \emph{aggressive} and \emph{degraded} MPI policies on \tinto.}
+  \label{fig:6}
 \end{figure*}
+
+\subsection{Power-performance profiling and tracing}
+\label{subsec:4.3}
+
+A full tracing  experiment is conducted on \tinto  in order to capture
+an overall power profile at a much finer resolution using the attached
+PDUs in each  node. First we run \cosmoart for  a 1-day simulation and
+analyze the  power consumption.  Second we  correlate performance with
+power  traces  of  two  1-hour  time frames  during  the  day:  midday
+(11h--12h) and midnight (23h--24h). All the experiments on \tinto were performed
+using 192 cores with 12 MPI processes per node, i.e., using an exactly-subscribed MPI mode.
+
+In order to obtain an initial overview of the power consumption of the
+system  model,  we  obtained  power  profiles  leveraging  our  \pmlib
+power-measurement    framework   over    a    1-day   simulation    of
+\cosmoart. These profiles, produced with both \emph{aggressive} and \emph{degraded} MPI
+policies,  are depicted in  Figure~\ref{fig:5}. The  first observation
+for the \emph{aggressive} MPI policy is that the first and last hours of the day
+consume slightly less  power than the midday hours.  We relate these
+variations  due   to  the  different   chemical  reactions,  aerosols,
+radiations  and clouds  formation processes,  computed by  ART, taking
+place during  daylight hours  but not during  night. In this  case the
+peak power consumption is about  3821\,W. Looking at the power profile
+using the \emph{degraded}  MPI policy, we observe that  the 1-hour pattern is
+repeated along  the 24-hours, favoring  the reduction of  power during
+night  hours in  a  more noticeable  way  than that  with the   \emph{aggressive}
+policy.   The peak  power consumption  during day  hours  while the
+ \emph{degraded}  MPI policy  is leveraged  is  about 3768\,W,  which means  a
+reduction of around  50\,W with respect to the  \emph{aggressive} mode. Also, we
+notice systematic  power drops for each simulated  hour.  We associate
+them  to periods  in  which  the processes  were waiting  for
+incoming messages  or synchronizations, thus  yielding the CPU  to a lower
+usage. Our final observation is the reduction of the TTS by 2\,\% with
+respect to the  \emph{degraded} MPI policy.
+
+To  gain  more  insights   about  the  power-performance  behavior  of
+\cosmoart,  we  obtained power-per\-for\-man\-ce  traces  to allow  us
+correlate  the type  of  tasks  being executed  with  the total  power
+consumption  along  the   time  (see  Figure~\ref{fig:6}).   For  this
+experiment we  limit the simulation  time to 1 hour  because: \emph{i})
+the  pattern of  the computation  performed by  \cosmoart  is repeated
+hourly and, \emph{ii}) the weight  of a 1-day simulation trace can not
+be easily handled using our  machines and memory available (each trace
+file was about 45\,GB). Thus, we shrink the simulations
+to 1 hour during midday (from 11h  to 12h) and during
+midnight  (from 23h  to  24h), in  which  \cosmoart perform  different
+internal computations.  At  the same time we leverage  the two already
+stated MPI policies: \emph{aggressive} and \emph{degraded}.
+
+All  the  traces  in   Figure~\ref{fig:6}  are  divided  in  different
+computation  phases. In first  phase all  the processes  perform group
+communications      and       execute      MPI-receive      primitives
+(e.g. \texttt{MPI\_Recv}). The second one performs computation bursts,
+combined  in  between   with  group  communications,  blocking  sends,
+immediate  receives and  MPI-wait primitives.  Finally,
+group  communications interlaced  with computation  and  a significant
+synchronization  phase are  performed. From  the performance  point of
+view, one can notice a reduction of 2\,\% in the TTS that \emph{degraded}  MPI
+mode leads over the \emph{aggressive} MPI  policy. On the other hand an averaged
+reduction of power consumption of  2\,\% is attained with the \emph{degraded} 
+MPI policy  only in  routines that actually  perform synchronizations,
+group  communications  and  MPI-receive/-wait primitives.   Thus,  the
+reduction of the power-energy  pair directly depends on the percentage
+of time in which blocking MPI  operations are executed. For the 1 hour
+simulation traces, an average of 50\,\% of the time the processes were
+waiting for incoming messages.  Taking  into account that  the total power  reduction during
+blocking phases  is about 5\,\%,  one can compute an  approximation of
+the  power savings,  i.e., $50\,\%  \times  5\,\% =  2.5\%$, which  is
+consistent  power  reduction  obtained  for  the  full-day  simulation
+(around  2\,\%). However,  we  expect higher  power/energy savings  if
+upcoming versions of OpenMPI support full blocking policies as well as if 
+future architectures tend proportional computing~\citep{Barroso-2007}.
+
+% \begin{figure*}[htbf]
+%   \centering
+%   Simulation of 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
+%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq0_stat.eps}\\
+%   Percentage of time for the 1 hour during midnight (from 23h to 24h) leveraging the polling MPI policy\\
+%   \includegraphics[width=0.78\textwidth]{Figs/23_24_blq1_stat.eps}\\
+%   \caption{blabla}
+%   \label{fig:7}
+% \end{figure*}

### 2014-05-29 trunk/paper_enahpc14/paper.tex
cleans up latest version; improves format
Manuel's changes + rebuttal letter

@@
 \usepackage[numbers]{natbib}
-
 \usepackage{amsmath}
@@
 \usepackage{xspace}
-
 %
 \DeclareRobustCommand\IPCClongname{}
-\setcounter{secnumdepth}{5}
 %
@@
   In this  paper we investigate  the energy footprint  and performance
-  profiling  of  \textsc{COSMO-ART} on  various  HPC platforms.   This
-  model is an  extension of the operational weather  forecast model of
-  the German  weather service (DWD),  developed for the  evaluation of
-  the interactions  of reactive gases  and aerosol particles  with the
-  state of  atmosphere at  the regional scale.   Different measurement
-  devices and  energy-aware techniques are described  to evaluate both
-  time and  energy to  solution of the  considered application  and to
-  gain detailed insights into power and performance requirements.  Our
+  profiling of \cosmoart  on various HPC platforms.  This  model is an
+  extension of  the operational weather  forecast model of  the German
+  weather  service   (DWD),  developed  for  the   evaluation  of  the
+  interactions of reactive gases  and aerosol particles with the state
+  of atmosphere at the  regional scale.  Different measurement devices
+  and energy-aware techniques are  described to evaluate both time and
+  energy  to  solution  of  the  considered application  and  to  gain
+  detailed  insights  into power  and  performance requirements.   Our
   motivation  is to  improve  corresponding code  sections to  sustain
@@
   \and Numerical weather  prediction \and Atmospheric
-  chemistry  \and  Aerosol  modelling \and
+  chemistry  \and  Aerosol  modeling \and
   Benchmark  analysis  \and 
@@
 also gratefully acknowledge the High Performance and High Productivity
-Computing Initiative  \cite{HP2C} for results that  will be leveraged
-in subsequent code refactoring.
+Computing Initiative \cite{HP2C} for results that will be leveraged in
+subsequent code refactoring.
 \end{acknowledgements}

### 2014-05-29 paper_enahpc14/RebuttalLetter/rebuttal.tex
Manuel's changes + rebuttal letter

@@
 Now we refer to ``aggressive'' and ``degraded'' OpenMPI policies instead of ``polling''
-and ``blocking''. The reviewer is completely right on this issue, however even when using
+and ``blocking''. The reviewer is completely right on this issue, however even with 
 the ``degraded'' policy the CPU utilization is readuced to 50\,\%, thus saving energy.

### 2014-05-29 trunk/paper_enahpc14/Sec4:Experimental-results.tex
Manuel's changes + rebuttal letter

@@
   \scriptsize
-  Simulation of 1 hour during midday (11h--12h) using the \emph{aggressive} MPI policy \\
+  1 hour simulation during midday (11h--12h) using the \emph{aggressive} MPI policy \\
   \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}} \\
-  Simulation of 1 hour during midday (11h--12h) using the \emph{degraded} MPI policy \\
+  1 hour simulation during midday (11h--12h) using the \emph{degraded} MPI policy \\
   \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}} \\

### 2014-05-29 trunk/paper_enahpc14/Sec3:Environment-setup.tex

--- 2014-05-05 trunk/paper_enahpc14/Sec3:Environment-setup.tex
@@
 \begin{itemize}
-\item \monch is  a 10-rack NEC-provided clus\-ter at  CSCS composed of
-  312 compute nodes. 
-  % and  utilised  by  members  of  the  Swiss  Platform  for  Advanced
+\item \monch is  a 10-rack NEC-provided clus\-ter at  CSCS composed by
+  384   nodes  grouped  in  312  standard  compute  nodes,  24
+  large-memory compute nodes,  24 huge-memory nodes, 16 administration
+  nodes and 8 servers to provide I/O to Lustre.
+  % and  utilized  by  members  of  the  Swiss  Platform  for  Advanced
   % Scientific Computing (PASC) project~\cite{PASC}.  
-  Each node is  comprised of two Intel Ivy  Bridge EP E5-2660 ten-core
-  processors  operating  at 2.2\,GHz,  equipped  with  32\,GB of  DDR3
-  1600\,MHz RAM  and connected via InfiniBand Mellanox  SX6036 and FDR
-  switches. For our  experiments, a full rack constituted  of 52 nodes
-  was used. 
+Each standard  compute node  is comprised of  two Intel Xeon Ivy Bridge EP
+E5-2660v2  ten-core processors  operating at  2.2\,GHz,  equipped with
+32\,GB  of DDR3 1600\,MHz  RAM and  connected via  InfiniBand Mellanox
+SX6036 and FDR switches.  For our experiments, only a subset of 52 was used. %
+% only a full rack constituted exclusively of 52 nodes was used.
   % \monch is  slated to stay  in service without hardware  upgrade for
@@
 
-\item \pilat is  a dual-socket eight-core Intel Sandy  Bridge EP based
-  cluster at CSCS composed of 42 compute nodes.  
-  % and used as Piz Daint  pre-post processing cluster The 2 login and
+\item \pilat is a cluster at CSCS composed of 42 compute nodes.  
+  % and used as Piz Daint pre-post processing cluster. The 2 login and
   % 42  compute  nodes  are   consolidated  into  11  twin-pair  Intel
@@
   % containing 4 compute nodes based on two ...
-  Each node  is comprises two Intel Xeon  E5-2670 processors operating
+  Each of them is comprised of two Intel Xeon Sandy Bridge EP E5-2670 eight-core processors operating
   at 2.6\,GHz equipped with 64\,GB of DDR3 1600\,MHz RAM and connected
   via  InfiniBand   Mellanox  SX6036   and  FDR  switches.    For  our
-  experiments, a full  rack constituted of 42 compute  nodes was used.
-  \pilat   is   based   on   Intel's   second-generation   processors,
-  conventional in  HPC systems  and known to  be more  power consuming
-  than its Ivy Bridge successor.
+  experiments, a subset constituted of 42 compute  nodes was used.
+  % \pilat   is   based   on   Intel's   second-generation   processors,
+  % conventional in  HPC systems  and known to  be more  power consuming
+  % than its Ivy Bridge successor.
 
@@
   at UJI.  For our experiments only  a subset of  16 homogeneous nodes
-  was  considered. Each  node  is  composed of  two  Intel Xeon  E5645
+  was  considered. Each  node  is  comprised of  two  Intel Xeon Westmere EP E5645
   hexa-core processors  running at  2.4\,GHz, equipped with  24\,GB of
   DDR3  1333\,MHz and  connected via  InfiniBand QDR  with  a Mellanox
-  MTS\-3600 switch.
+  MTS\-3600 switch. Power saving mechanisms enabled on the nodes are the C-states C1, C3 and C6, 
+  but no P-states (except P0) are used. TurboBoost feature is enabled for our experiments.
 \end{itemize}
@@
 We present here  the two different frameworks deployed  on the testbed
-HPC  systems  to measure  performance,  power  dissipation and  energy
+HPC  systems  to measure  performance,  power and  energy
 consumption of the baseline execution of \cosmoart.
@@
 Monitors (IPM)  which are high quality electricity  meters released by
-Riedo Networks,  that enable to  monitor and log power  dissipation of
+Riedo Networks,  that enable to  monitor and log power  consumption of
 the  IT infrastructure  as well  as constantly  analyze  line voltage,
-current, power-factor, frequency  with 1\,\% accuracy.  Using reliable
+current, power-factor, frequency  with $\pm1\,\%$ accuracy.  Using reliable
 narrowband power-line communication (PLC) technology, all metering and
@@
 \pmlib  library in  combination  with Extrae  and  Paraver, which  are
-profiling/tracing and visualisation tools, respectively.
+profiling/tracing and visualization tools, respectively.
 
@@
 the Fortran~90  code of the model,  e.g., to record  MPI calls.  Next,
-\cosmoart  is run  on the  nodes, thus  dissipating certain  amount of
+\cosmoart  is run  on the  nodes, thus  consuming certain  amount of
 power. The  nodes are connected  to APC 8653 Power  Distribution Units
-(PDUs) that account  for the dissipated power with  a sampling rate of
-1\,Hz. The power data captured by  these PDUs is collected by a \pmlib
+(PDUs) that account  for the consumed   power with  a sampling rate of
+1\,Hz and $\pm3\,\%$ accuracy. The power data captured by  these PDUs is collected by a \pmlib
 tracing server running  on a separate machine. The  client, running on
@@
 generated.  A  merged trace  file  is  finally  loaded in  Paraver  to
-visualise  and correlate  the  performance trace  and  with the  power
+visualize  and correlate  the  performance trace  and  with the  power
 profile  generated by \cosmoart.  This procedure  allows us  to easily
@@
 opted for the GNU compiler GCC 4.8.1 on \monch and GCC 4.8.2 on \pilat
-and  \tinto)  using  the -O3  compiler  flag  in  favor of  the  Intel
+and  \tinto)  using  the \texttt{-O3}  compiler  flag  in  favor of  the  Intel
 compilers 14.0.1, which  delivered inferior performance.  In addition,
@@
 \cosmoart, was produced  and run on 1040 cores  using 20~MPI tasks per
-node on  \monch, on 1344 cores  using 16~MPI tasks per  node on \pilat
+node on  \monch, on 672 cores  using 16~MPI tasks per  node on \pilat
 and  on  192  cores  using  12~MPI  tasks per  node  on  \tinto.   The
@@
 coordinates was  defined by  setting: $nprocx=40$ and  $nprocy=26$ for
-\monch,  $nprocx=28$ and  $nprocy=24$ for  \pilat and  $nprocx=??$ and
-$nprocy=??$  for \tinto as  $nprocx$ is  usually kept  slightly bigger
+\monch,  $nprocx=28$ and  $nprocy=24$ for  \pilat and  $nprocx=16$ and
+$nprocy=12$  for \tinto as  $nprocx$ is  usually kept  slightly bigger
 than  $nprocy$. Besides,  as this  version does  not make  use  of the

### 2014-05-29 paper_enahpc14/Sec4:Experimental-results.tex

@@
 contribution of the MPI library  to the energy consumption and how the
-use  of  the  blocking  versus polling  message-passing  policies  can
-potentially  render energy  savings. Specifically,  the  OpenMPI 1.6.5
-library, features two operation modes, blocking and polling, which can
-be selected before \cosmoart is launched. In the polling configuration
-(the default mode) ``MPI''  continually polls the network interface to
-check for  the completion  of an event  (e.g. send or  receive).  This
-usually yields  low latency  but high CPU  utilization. Thus,  one can
-expect that  this mode attains  the best performance, possibly  at the
-cost a higher energy usage. In  the blocking mode, the CPU is yield to
-other     processes/threads    if     there     are    no     incoming
-messages~\cite{Castillo-2012}.
+use  of  the aggressive and degraded message-passing progression engine 
+policies can potentially  render energy  savings.
+
+Specifically,   the  OpenMPI 1.6.5  library,  features  two  operation 
+policies, \emph{aggressive} and \emph{degraded}, which can be  selected 
+before \cosmoart is launched. In the \emph{aggressive} (default) policy,  
+normally used in exactly- or under-subscribed mode  (i.e., the  number 
+of running processes is equal to or less than the number of  available 
+cores), MPI processes will never voluntarily give up the processor  to 
+other processes. With some network transports, OpenMPI processes  will 
+spin  in  tight  loops  attempting  to  make message passing progress, 
+effectively causing other processes to not get  any  CPU  cycles  (and  
+therefore never make any  progress).  This  usually  yields  the  best 
+performance but high CPU utilization, possibly at the cost of a higher 
+energy usage.  On the other hand, the \emph{degraded} policy, suitable 
+for the oversubscribed mode (i.e., more  processes  are  running  than 
+total cores  available),  MPI  processes  will  frequently  yield  the 
+processor to its peers, thereby allowing all processes to make progress.
+
+In our experiments with \cosmoart only the exactly-subscribed mode  is
+used. Therefore, \cosmoart processes running  in  the  \emph{degraded} 
+policy are constantly checking for the completion  of  an  event (e.g. 
+send or receive) and repeatedly calling to  \texttt{sched\_yield()} to 
+be picked again by the Linux scheduler. Although one can expect no benefits 
+from   using   the  \emph{degraded}   policy   compared   with     the 
+\emph{aggressive} one, notable differences can be observed in the CPU utilization. 
+In a separated  test  with  the  \emph{aggressive}  policy 
+enabled, we noticed that a blocked MPI process waiting for an incoming 
+message has CPU utilization of 100\,\%,  however  when  the \emph{degraded} 
+mode  is  set,  the  CPU  utilization is only $50\,\%$, approximately. 
+Although neither of both policies avoid busy-waiting loops in blocking 
+routines, the use of the \emph{degraded} policy is as much as we can 
+do in OpenMPI to reduce the CPU utilization, and thus save energy.
+
+% polling configuration
+% (the default mode) ``MPI''  continually polls the network interface to
+% check for  the completion  of an event  (e.g. send or  receive).  This
+% usually yields  low latency  but high CPU  utilization. Thus,  one can
+% expect that  this mode attains  the best performance, possibly  at the
+% cost a higher energy usage. In  the blocking mode, the CPU is yield to
+% other     processes/threads    if     there     are    no     incoming
+% messages~\cite{Castillo-2012}.
 
@@
 one  update every  5 minutes  for power  sampling, the  averaged power
-dissipation was computed by considering  6 values for each single run.
+consumption was computed by considering  6 values for each single run.
 On \pilat, the 1-day simulation was  issued four times and a 2-day run
-only once.  Similarly, the  averaged power dissipation was computed by
+only once.  Similarly, the  averaged power consumption was computed by
 considering 4  values for each single  1-day run and 9  values for the
@@
   \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
-  \caption{Isola E1 Rack 2 total power of \monch.}
+  \caption{Isola E1 Rack 2 total power consumption of \monch.}
   \label{fig:1}
@@
   \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
-  \caption{Isola HD total power dissipation of \pilat.}
+  \caption{Isola HD total power consumption of \pilat.}
   \label{fig:2}
@@
   \centering
-  \caption{Averaged power dissipation (W) of the platforms.}
+  \caption{Averaged power consumption (W) of the platforms.}
   \label{tab:1}
@@
     \textbf{\scriptsize{TINTORRUM}}           \\          &          &
-    \textbf{\scriptsize{(Polling)}} & \textbf{\scriptsize{(Blocking)}}
+    \textbf{\scriptsize{(\emph{Aggressive})}} & \textbf{\scriptsize{(\emph{Degraded})}}
     \\    \noalign{\smallskip}\hline\noalign{\smallskip}   18035.0   &
@@
 In  Figure~\ref{fig:3}, we  compare both  TTS (right  Y-axis)  and ETS
-(left  Y-axis)  metrics  on  both  platforms.  As  expected,  Xeon  E5
-outperforms Ivy  Bridge EP, being roughly 130\,\%  faster.  The reason
-for that is two-fold: \emph{i}) it has higher clock frequency than Ivy
-Bridge  (2.6\,GHz  against  2.2\,GHz),   and  \emph{ii})  it  aims  at
+(left  Y-axis)  metrics  on  both  platforms.  As  expected, Intel Sandy Bridge EP processors of \pilat
+outperform the Intel Ivy  Bridge EP processors of \monch, being roughly 130\,\%  faster.  The reason
+for that is two-fold: \emph{i}) they have higher clock frequency than Ivy
+Bridge  (2.6\,GHz  against  2.2\,GHz),   and  \emph{ii})  they  aim  at
 computing   speed  regardless   to  be   energy  efficient.    In  our
-experiments,  Ivy  Bridge   EP  showed  the  best  energy-to-solution,
-reducing the energy consumption of Xeon E5 by approximately 7\,\%.
+experiments,  \monch  showed  the  best  energy-to-solution,
+reducing the energy consumption of \pilat by approximately 7\,\%.
 
@@
 the  averaged total  power are  also shown  in  Table~\ref{tab:1}.  As
-expected, the average total power dissipation by \cosmoart while using
-the blocking MPI policy is about  2\,\% less than that consumed by the
-polling  policy. This is  due to  the power-friendly  (blocking) waits
-performed  by the  CPU cores  while using  the blocking  MPI  mode, in
-contrast to  the polling  waits that take  place when the  polling MPI
-mode is set. From the point of  view of the execution time, there is a
-slight reduction  for the blocking MPI  mode. One can  expect that the
-costs of the interruptions used in the blocking MPI policy can lead to
+expected, the average total power consumption by \cosmoart while using
+the \emph{degraded} MPI policy is about  2\,\% less than that consumed by the
+\emph{aggressive} one. This is  due to  the reduction of CPU usage (of about 50\,\%) while using  the \emph{degraded} MPI policy, in
+contrast with the \emph{aggressive} policy that always yields to 100\,\% of CPU utilization while waiting for an incoming message. From the point of  view of the execution time, there is a
+slight reduction  for the \emph{degraded} MPI  policy. One can  expect that the
+costs of the context switches produced in the \emph{degraded} MPI policy can lead to
 an  increase  of  the   execution  time,  however,  applications  with
 extensive computing,  as e.g.  \cosmoart, may show  better performance
-with the blocking  modes. This effect works in  favor with our results
+with the \emph{degraded} policy. This effect works in  favor with our results
 since the energy consumed is also reduced (2\,\%) just by changing the
-behavior of the OpenMPI library  to the blocking policy, i.e, with the
-\texttt{--mca    mpi\_yield\_when\_idle}   parameter   set    in   the
+behavior of the OpenMPI library  to the \emph{degraded} policy, i.e, with the
+\texttt{--mca    mpi\_yield\_when\_idle 1}   parameter   set    in   the
 \texttt{mpirun} command.
@@
   \caption{Mean time-to-solution and  energy-to-solution on \tinto for
-    a 24h simulation using the  polling and blocking MPI policies. The
+    a 24h simulation using the  \emph{aggressive} and \emph{degraded} MPI policies. The
     95\,\%  confidence  intervals of  Student's  $t$ distribution  (10
-    samples) clearly  indicates an improvement for  blocking policy in
+    samples) clearly  indicates an improvement for the \emph{degraded} policy in
     both TTS and ETS.}
@@
   \hspace{0.8cm} \scalebox{0.62}{\input{Figs/24hour.pstex_t}}
-  \caption{24  hours  simulation power  trace  using  the polling  and
-    blocking MPI policies on \tinto.}
+  \caption{24  hours  simulation power  trace  using  the \emph{aggressive}  and
+    \emph{degraded} MPI policies on \tinto.}
   \label{fig:5}
@@
   \scriptsize
-  Simulation of 1 hour during midday (11h--12h) leveraging the polling MPI policy \\
+  1 hour simulation during midday (11h--12h) using the \emph{aggressive} MPI policy \\
   \scalebox{0.5}{\input{Figs/11_12_blq0_figure.pstex_t}} \\
-  Simulation of 1 hour during midday (11h--12h) leveraging the blocking MPI policy \\
+  1 hour simulation during midday (11h--12h) using the \emph{degraded} MPI policy \\
   \scalebox{0.5}{\input{Figs/11_12_blq1_figure.pstex_t}} \\
-  1 hour simulation during midnight (23h--24h) leveraging the polling MPI policy \\
+  1 hour simulation during midnight (23h--24h) using the \emph{aggressive} MPI policy \\
   \scalebox{0.5}{\input{Figs/23_24_blq0_figure.pstex_t}} \\
-  1 hour simulation during midnight (23h--24h) leveraging the blocking MPI policy \\
+  1 hour simulation during midnight (23h--24h) using the \emph{degraded} MPI policy \\
   \scalebox{0.5}{\input{Figs/23_24_blq1_figure.pstex_t}} \\
@@
   \caption{Power-performance   traces  during   midday   and  midnight
-    leveraging the polling and blocking MPI policies on \tinto.}
+    leveraging the \emph{aggressive} and \emph{degraded} MPI policies on \tinto.}
   \label{fig:6}
@@
 power  traces  of  two  1-hour  time frames  during  the  day:  midday
-(11h--12h) and midnight (23h--24h). All the experiments were performed
-using 192 cores using 12 MPI processes per node on \tinto.
+(11h--12h) and midnight (23h--24h). All the experiments on \tinto were performed
+using 192 cores with 12 MPI processes per node, i.e., using an exactly-subscribed MPI mode.
 
-In order to obtain an initial overview of the power dissipation of the
+In order to obtain an initial overview of the power consumption of the
 system  model,  we  obtained  power  profiles  leveraging  our  \pmlib
 power-measurement    framework   over    a    1-day   simulation    of
-\cosmoart. These profiles, produced with both polling and blocking MPI
+\cosmoart. These profiles, produced with both \emph{aggressive} and \emph{degraded} MPI
 policies,  are depicted in  Figure~\ref{fig:5}. The  first observation
-for the polling MPI policy is that the first and last hours of the day
-dissipate slightly less  power than the midday hours.  We relate these
+for the \emph{aggressive} MPI policy is that the first and last hours of the day
+consume slightly less  power than the midday hours.  We relate these
 variations  due   to  the  different   chemical  reactions,  aerosols,
@@
 peak power consumption is about  3821\,W. Looking at the power profile
-using the blocking  MPI policy, we observe that  the 1-hour pattern is
+using the \emph{degraded}  MPI policy, we observe that  the 1-hour pattern is
 repeated along  the 24-hours, favoring  the reduction of  power during
-night  hours in  a  more noticeable  way  than that  with the  polling
-policy.   The peak  power dissipation  during day  hours of  while the
-blocking  MPI policy  is leveraged  is  about 3768\,W,  which means  a
-reduction of around  50\,W with respect to the  polling mode. Also, we
+night  hours in  a  more noticeable  way  than that  with the   \emph{aggressive}
+policy.   The peak  power consumption  during day  hours  while the
+ \emph{degraded}  MPI policy  is leveraged  is  about 3768\,W,  which means  a
+reduction of around  50\,W with respect to the  \emph{aggressive} mode. Also, we
 notice systematic  power drops for each simulated  hour.  We associate
-them  to periods  in  which  the processes  were  blocked waiting  for
-incoming messages  or synchronizations, thus  yielding the CPU  to low
+them  to periods  in  which  the processes  were waiting  for
+incoming messages  or synchronizations, thus  yielding the CPU  to a lower
 usage. Our final observation is the reduction of the TTS by 2\,\% with
-respect to the blocking MPI policy.
+respect to the  \emph{degraded} MPI policy.
 
@@
 correlate  the type  of  tasks  being executed  with  the total  power
-dissipation  along  the   time  (see  Figure~\ref{fig:6}).   For  this
-experiment we  limit the simulation  time to 1 hour  because \emph{i})
+consumption  along  the   time  (see  Figure~\ref{fig:6}).   For  this
+experiment we  limit the simulation  time to 1 hour  because: \emph{i})
 the  pattern of  the computation  performed by  \cosmoart  is repeated
@@
 be easily handled using our  machines and memory available (each trace
-file was about 45\,GB). To  face the problem we shrink the simulations
-to 1 hour  and test a time  frame during midday (from 11h  to 12h) and
+file was about 45\,GB). Thus, we shrink the simulations
+to 1 hour during midday (from 11h  to 12h) and during
 midnight  (from 23h  to  24h), in  which  \cosmoart perform  different
 internal computations.  At  the same time we leverage  the two already
-stated MPI policies: polling and blocking.
+stated MPI policies: \emph{aggressive} and \emph{degraded}.
 
@@
 combined  in  between   with  group  communications,  blocking  sends,
-immediate  receives and  MPI-wait primitives  are  executed.  Finally,
+immediate  receives and  MPI-wait primitives.  Finally,
 group  communications interlaced  with computation  and  a significant
 synchronization  phase are  performed. From  the performance  point of
-view, one can notice a reduction of 2\,\% in the TTS that blocking MPI
-mode leads over the polling MPI  policy. On the other hand an averaged
-reduction of power dissipation of  2\,\% is attained with the blocking
+view, one can notice a reduction of 2\,\% in the TTS that \emph{degraded}  MPI
+mode leads over the \emph{aggressive} MPI  policy. On the other hand an averaged
+reduction of power consumption of  2\,\% is attained with the \emph{degraded} 
 MPI policy  only in  routines that actually  perform synchronizations,
@@
 simulation traces, an average of 50\,\% of the time the processes were
-blocked.  Taking  into account that  the total power  reduction during
+waiting for incoming messages.  Taking  into account that  the total power  reduction during
 blocking phases  is about 5\,\%,  one can compute an  approximation of
@@
 consistent  power  reduction  obtained  for  the  full-day  simulation
-(around  2\,\%). However,  we  expect higher  power-energy savings  if
+(around  2\,\%). However,  we  expect higher  power/energy savings  if
+upcoming versions of OpenMPI support full blocking policies as well as if 
 future architectures tend proportional computing~\citep{Barroso-2007}.

### 2014-05-29 trunk/paper_enahpc14/Sec4:Experimental-results.tex

@@
 
-Averaged TTS and ETS results  over 10 repetitions of 24h simulation on
+Averaged TTS and ETS results  over 10 repetitions of 24-hours simulation on
 \tinto are shown in  Figure~\ref{fig:4}. The corresponding results for
@@
   \caption{Time-to-solution and  energy-to-solution comparison between
-    \pilat and \monch for a 24h simulation.}
+    \pilat and \monch for a 24-hours simulation.}
   \label{fig:3}
@@
   \caption{Mean time-to-solution and  energy-to-solution on \tinto for
-    a 24h simulation using the  \emph{aggressive} and \emph{degraded} MPI policies. The
+    a 24-hours simulation using the  \emph{aggressive} and \emph{degraded} MPI policies. The
     95\,\%  confidence  intervals of  Student's  $t$ distribution  (10
@@
   \hspace{0.8cm} \scalebox{0.62}{\input{Figs/24hour.pstex_t}}
-  \caption{24  hours  simulation power  trace  using  the \emph{aggressive}  and
+  \caption{24-hours simulation power  trace  using  the \emph{aggressive}  and
     \emph{degraded} MPI policies on \tinto.}

### 2014-05-29 trunk/paper_enahpc14/Sec4:Experimental-results.tex

@@
 
-Averaged TTS and ETS results  over 10 repetitions of 24-hours simulation on
+Averaged TTS and ETS results  over 10 repetitions of 24-hour simulation on
 \tinto are shown in  Figure~\ref{fig:4}. The corresponding results for
@@
   \caption{Time-to-solution and  energy-to-solution comparison between
-    \pilat and \monch for a 24-hours simulation.}
+    \pilat and \monch for a 24-hour simulation.}
   \label{fig:3}
@@
   \caption{Mean time-to-solution and  energy-to-solution on \tinto for
-    a 24-hours simulation using the  \emph{aggressive} and \emph{degraded} MPI policies. The
+    a 24-hour simulation using the  \emph{aggressive} and \emph{degraded} MPI policies. The
     95\,\%  confidence  intervals of  Student's  $t$ distribution  (10
@@
   \hspace{0.8cm} \scalebox{0.62}{\input{Figs/24hour.pstex_t}}
-  \caption{24-hours simulation power  trace  using  the \emph{aggressive}  and
+  \caption{24-hour simulation power  trace  using  the \emph{aggressive}  and
     \emph{degraded} MPI policies on \tinto.}
@@
 using the \emph{degraded}  MPI policy, we observe that  the 1-hour pattern is
-repeated along  the 24-hours, favoring  the reduction of  power during
+repeated along  the 24 hours, favoring  the reduction of  power during
 night  hours in  a  more noticeable  way  than that  with the   \emph{aggressive}

### 2014-05-30 paper_enahpc14/RebuttalLetter/rebuttal.tex
Added what I could.

@@
 \begin{weSay}
-blabla
+Indeed, there is a huge amount which could be said about COSMO, and COSMO-ART in particular, about strong and weak scalability, energy scalability, along with the underlying science.
+However, going into this detail would make this an entirely different paper, and one that has, at least for COSMO, already been published in several forums.
+Moreover, such a comprehensive analysis would have taken far more than eight pages.
+
+We had a modest goal in this paper of evaluating the COSMO-ART configuration, which is a benchmark for production at MeteoSwiss, in terms of energy-to-solution and time-to-solution.
+To this end we have used state-of-the-art profiling and energy measurement equipment.
+In subsequent papers, we plan to address a wider spectrum of energy issues.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+We would be happy to add specific technical details as requested by the reviewers, but the page limit constrains the amout of general detail we can add.
+Information about COSMO and COSMO-ART, as well as the tools employed, can be found in the appropriate citations in the paper.
+We have chosen to concentrate on documenting the aspects which are actually new to this paper, namely the application of the energy profiling tools to this non-trivial application, and the T2S and E2S evaluation of the COSMO-ART benchmark provided for MeteoSwiss.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+We do agree that this paper is not intended to develop a methodology for evaluating applications, but rather to use existing tools and methodology to evaluate a complex, real-world application, which brings with it a new spectrum of issues, like additional measurement complexity, large trace files, and the need for system-integrated power measurement tools.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+Indeed, there is only a tangential relationship between [11] and [20] and our paper, and we would be happy to remove these.
+There is in fact very little related work on energy evaluations of real applications; most work in the field limits itself to artificial toy examples, which is definitely not our topic of interest.
+We are trying to find more references on infrastructure and tool support, and hope to add these in the final version
+
+ [Jo/Manuel/Sandra: can you find anything appropriate?]
 \end{weSay}
@@
 \begin{weSay}
-blabla
+The configuration was taken from the MeteoSwiss benchmark, and was chosen by the developers.
+We deliberately chose to use their configuration directly to form the baseline, since this is how it would be run in production if running today.
+You are right that there is some low-hanging fruit, but optimizations like -mavx alone will not attain the 5x improvement.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+[Jo:  you have to try to answer this: there is information on the scalability of COSMO at least. ]
 \end{weSay}

### 2014-05-30 trunk/paper_enahpc14/RebuttalLetter/rebuttal.tex

@@
 
-\newpage
+%\newpage
 % ------------------------------------------------------------------------------------------------------------------------------------------------
@@
 \begin{weSay}
-blabla
+We have accordingly replaced ``power dissipation'' by ``power consumption'' along the manuscript.
 \end{weSay}
 
-\newpage
+%\newpage
 \section*{Reviewer \#2}
@@
 \begin{weSay}
-blabla
+We would be happy to add specific technical details as requested by the reviewers, but the page limit constrains the amout of general detail we can add.
+Information about COSMO and COSMO-ART, as well as the tools employed, can be found in the appropriate citations in the paper.
+We have chosen to concentrate on documenting the aspects which are actually new to this paper, namely the application of the energy profiling tools to this non-trivial application, and the T2S and E2S evaluation of the COSMO-ART benchmark provided for MeteoSwiss.
 \end{weSay}
 
-\newpage
-\section*{Reviewer \#2}
+%\newpage
+\section*{Reviewer \#3}
 
@@
 \begin{weSay}
-blabla
+We noticed the problem and fixed it accordingly along the manuscript. We misused blocking/polling terms.
+Now we have replaced the ``polling'' and ``blocking''` OpenMPI policies to ``aggressive'' and ``degraded'', 
+respectively. The reviewer is completely right on this issue, however even with 
+the ``degraded'' policy the CPU utilization is reduced to 50\,\% and thus allows us saving energy.
 \end{weSay}

### 2014-05-30 paper_enahpc14/RebuttalLetter/rebuttal.tex

@@
 
-\newpage
+%\newpage
 \section*{Reviewer \#2}
@@
 
-\newpage
+%\newpage
 \section*{Reviewer \#3}
@@
 \begin{weSay}
-We noticed the problem and fixed accordingly along the manuscript the misused blocking term.
-Now we refer to ``aggressive'' and ``degraded'' OpenMPI policies instead of ``polling''
-and ``blocking''. The reviewer is completely right on this issue, however even with 
-the ``degraded'' policy the CPU utilization is readuced to 50\,\%, thus saving energy.
+We noticed the problem and fixed it accordingly along the manuscript. We did a misuse of the blocking/polling terms.
+Now we have replaced them to ``aggressive'' and ``degraded'', since reflects exactly the experiments of our paper. 
+For the second comment, the reviewer is completely right on that issue, however even with  the ``degraded'' policy 
+the CPU utilization is reduced to 50\,\% and thus allows us to save energy.
 \end{weSay}

### 2014-05-30 trunk/paper_enahpc14/RebuttalLetter/rebuttal.tex

@@
 \begin{weSay}
-We noticed the problem and fixed it accordingly along the manuscript. We misused blocking/polling terms.
-Now we have replaced the ``polling'' and ``blocking''` OpenMPI policies to ``aggressive'' and ``degraded'', 
-respectively. The reviewer is completely right on this issue, however even with 
-the ``degraded'' policy the CPU utilization is reduced to 50\,\% and thus allows us saving energy.
+We noticed the problem and fixed it accordingly along the manuscript. We did a misuse of the blocking/polling terms.
+Now we have replaced them to ``aggressive'' and ``degraded'', since reflects exactly the experiments of our paper. 
+For the second comment, the reviewer is completely right on that issue, however even with  the ``degraded'' policy 
+the CPU utilization is reduced to 50\,\% and thus allows us to save energy.
 \end{weSay}
@@
 \begin{weSay}
-We have fixed the sentence accordingly so as not to be confusing.
+We have fixed the sentence accordingly to avoid any confusion.
 \end{weSay}
@@
 \begin{weSay}
-We have fixed this issue along the manuscript.
+We have fixed this issue along the manuscript. Now we employ the platform's name for comparing them.
 \end{weSay}

### 2014-05-30 trunk/paper_enahpc14/RebuttalLetter/rebuttal.tex

@@
 in the Environment setup section and afterwards we refer to the platforms' name or using the
-full processor's name and its microarchitecture.
+full processor's name plus its microarchitecture.
 \end{weSay}
@@
 \begin{weSay}
-We have accordingly replaced ``power dissipation'' by ``power consumption'' along the manuscript.
+We have accordingly replaced ``power dissipation'' by ``power consumption'' along the manuscript to avoid any confusion.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+We agree with the rewiever. Indeed we tried to shorten the COSMO-ART's explanation due to space restrictions in the last version of the paper. However
+with the page limit extension we added more details about the nature of the application to better understand its behaviour.
 \end{weSay}
@@
 \begin{weSay}
-Indeed, there is a huge amount which could be said about COSMO, and COSMO-ART in particular, about strong and weak scalability, energy scalability, along with the underlying science.
-However, going into this detail would make this an entirely different paper, and one that has, at least for COSMO, already been published in several forums.
-Moreover, such a comprehensive analysis would have taken far more than eight pages.
+Indeed, there is a huge amount which could be said about COSMO, and COSMO-ART in particular, 
+about strong and weak scalability, energy scalability, along with the underlying science.  
+However, going into this detail would make this an entirely different paper, and one that has, 
+at least for COSMO, already been published in several forums. %   Moreover, such a comprehensive 
+%analysis would have taken far more than eight pages.
 
-We had a modest goal in this paper of evaluating the COSMO-ART configuration, which is a benchmark for production at MeteoSwiss, in terms of energy-to-solution and time-to-solution.
-To this end we have used state-of-the-art profiling and energy measurement equipment.
-In subsequent papers, we plan to address a wider spectrum of energy issues.
+We had a modest goal in this paper of evaluating the COSMO-ART configuration, which is a benchmark 
+for production at MeteoSwiss, in terms of energy-to-solution and time-to-solution.  To this end we 
+have used state-of-the-art profiling and energy measurement equipment.   In subsequent papers, we 
+plan to address a wider spectrum of energy issues.    
 \end{weSay}
@@
 There is in fact very little related work on energy evaluations of real applications; most work in the field limits itself to artificial toy examples, which is definitely not our topic of interest.
-We are trying to find more references on infrastructure and tool support, and hope to add these in the final version
-
- [Jo/Manuel/Sandra: can you find anything appropriate?]
+Actually, we have added one reference that helps us to justify the use of the most appropriate wattmeters for an application like COSMO-ART.
+%We are trying to find more references on infrastructure and tool support, and hope to add these in the final version
+% [Jo/Manuel/Sandra: can you find anything appropriate?]
 \end{weSay}
@@
 \begin{weSay}
-We have explained in detail the kind of nodes that this platform is comprised of.
-We only used a subset of 52 in our experiments.
+We have described in detail the nodes' architectures that this platform is comprised of.
+We only used a subset of 52 in our experiments, that is why the reviwer was confused. We now rephrased the explanation in order to be more precise.
 \end{weSay}
@@
 \begin{weSay}
-blabla
+The reviwer is completely right. Thus, we have accordingly modified the text in order to avoid comparisons between SB and IB processors.
 \end{weSay}
@@
      seems to be important for the discussion of the TTS results.
-
-%%% *** Manuel/Sandra ***
-\begin{weSay}
-We added information about the Turbo Boost feature for each platform \textbf{(MISSING FOR MONCH and PILATUS!!)}
-\end{weSay}
-
 - It should be added which power saving mechanisms are used (cpufreq
@@
 \begin{weSay}
-We added information about the power saving mechanisms available on each platform \textbf{(MISSING FOR MONCH and PILATUS!!)}
+We have added the details about the Turbo Boost feature and power saving mechanisms for the used platforms.
 \end{weSay}
@@
 \begin{weSay}
-\textbf{MISSING ANSWER!!}
-
-Figure 5 represents the power drawn of Tintorrum, for this case the PDU used has a sampling rate of 1 Hz instead of 5 minutes.
+The E3METER metering framework records averaged power data every 5 minutes, this is as much as we can do in MONCH and PILATUS platforms to collect power consumption data.
+On the other hand, Figure 5 represents the power drawn of TINTORRUM, for this case a PDU with a sampling rate of 1 Hz was used instead.
 \end{weSay}
@@
 \begin{weSay}
-We have added information about the accuracy of the PDU.
+We have added information about the accuracy of the APC8653 PDU.
 \end{weSay}
@@
 We noticed the problem and fixed it accordingly along the manuscript. We did a misuse of the blocking/polling terms.
-Now we have replaced them to ``aggressive'' and ``degraded'', since reflects exactly the experiments of our paper. 
-For the second comment, the reviewer is completely right on that issue, however even with  the ``degraded'' policy 
-the CPU utilization is reduced to 50\,\% and thus allows us to save energy.
+We have replaced them to ``aggressive'' and ``degraded'', respectively, since exactly reflect the experiments of our paper. 
+For the second comment, the reviewer is completely right, however even with  the ``degraded'' policy 
+the CPU utilization is still reduced to 50\,\% and thus allows us to save energy.
 \end{weSay}
@@
 \begin{weSay}
-We have fixed the sentence accordingly to avoid any confusion.
+We have rephrased the paragraph accordingly to avoid any confusion.
 \end{weSay}
@@
 \begin{weSay}
-[Jo:  you have to try to answer this: there is information on the scalability of COSMO at least. ]
+We avoided to make comparisons between MONCH and PILATUS platforms - so far we have no more information about the scalability of COSMO-ART. We modified the text accordingly.
+%[Jo:  you have to try to answer this: there is information on the scalability of COSMO at least. ]
 \end{weSay}
@@
 \begin{weSay}
-The reviewer is right in this sense. The processors are not idle during blocking MPI routines, however a reduction of the CPU utilization is observed (50\,\%).
+The reviewer is right in this sense.
+The processors are not idle during blocking MPI routines, however a reduction of the CPU utilization is observed (50\,\%) even with the degraded policy that we used.
 \end{weSay}

### 2014-05-30 trunk/paper_enahpc14/Conclusion.tex

--- 2014-04-25 paper_enahpc14/Conclusion.tex
@@
 We  have   presented  a  methodology  for   comparing  performance  of
-COSMO-ART,  a  regional  weather  forecast  model  augmented  for  the
+\cosmoart,  a  regional  weather  forecast  model  augmented  for  the
 interactions of  reactive gases and aerosol  particles.  The resulting
 benchmarks illustrate  that the  best time-to-solution does  not imply
-the best energy-to-solution: an Intel Sandybridge (2.6 GHz) system has
-lower  time-to-solution but  higher energy-to-solution  than  an Intel
-Ivybridge  (2.2  GHz) system,  although  the  two  metrics are  indeed
-strongly  correlated.   A   tool  utilising  Paraver/Extrae  with  new
-extensions  for power  consumption  was successfully  applied to  this
-non-trivial application.  The  resulting profiles indicate that simple
-changes, such as making use of an MPI version based on blocking rather
-than polling, can reduce power consumption significantly.
+the best energy-to-solution: an Intel Sandy Bridge (2.6 GHz) system has
+lower time-to-solution but higher energy-to-solution than an Intel Ivy
+Bridge (2.2 GHz) system, although  the two metrics are indeed strongly
+correlated.   On the  other  hand, the  usage  of energy-friendly  MPI
+waiting techniques analyzed  with our power-per\-for\-man\-ce framework over
+a series of  runs of \cosmoart on \tinto  and power traces demonstrate
+that   it  is   possible  to   reduce  both   power   dissipation  and
+energy-to-solution   while  maintaining   (or  even   decreasing)  the
+time-to-solution. The resulting profiles indicate that simple changes,
+such as making use of the  \emph{degraded} MPI policy rather than the \emph{aggressive}
+is  possible to  slightly  reduce both  power  dissipation and  energy
+consumption (by 2\,\%).
 
 This  reproducible  benchmark provides  a  baseline  for ongoing  work
-package within  the EU-funded Exa2Green to  minimise power consumption
-of COSMO-ART. Profiling  has given us insight into  the most expensive
-code  components,  which are  now  being  altered  to utilise  revised
-algorithms.  The  results of these  optimisations will be  reported in
+package within  the EU-funded Exa2Green to  minimize power consumption
+of \cosmoart. Profiling  has given us insight into  the most expensive
+code  components,  which are  now  being  altered  to utilize  revised
+algorithms.  The  results of these  optimizations will be  reported in
 future publications.

### 2014-05-30 trunk/paper_enahpc14/Sec4:Experimental-results.tex

@@
 only twice due  to usage restrictions.  As time  resolution was set to
-one  update every  5 minutes  for power  sampling, the  averaged power
+one averaged update every  5 minutes  for power  sampling, the  averaged power
 consumption was computed by considering  6 values for each single run.
@@
 
-\begin{figure}[htbf]
+\begin{figure}[ht]
   \centering
-  \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Monch.eps}
+  \includegraphics[width=1\columnwidth]{Figs/NRJ_benchmark_Monch.eps}
   \caption{Isola E1 Rack 2 total power consumption of \monch.}
@@
 
-\begin{figure}[htbf]
+\begin{figure}[ht]
   \centering
-  \includegraphics[width=0.48\textwidth]{Figs/NRJ_benchmark_Pilatus.eps}
+  \includegraphics[width=1\columnwidth]{Figs/NRJ_benchmark_Pilatus.eps}
   \caption{Isola HD total power consumption of \pilat.}
@@
 
-\begin{table}[htbf]
+\begin{table}[ht]
   \centering
@@
 In  Figure~\ref{fig:3}, we  compare both  TTS (right  Y-axis)  and ETS
-(left  Y-axis)  metrics  on  both  platforms.  As  expected, Intel Sandy Bridge EP processors of \pilat
-outperform the Intel Ivy  Bridge EP processors of \monch, being roughly 130\,\%  faster.  The reason
-for that is two-fold: \emph{i}) they have higher clock frequency than Ivy
-Bridge  (2.6\,GHz  against  2.2\,GHz),   and  \emph{ii})  they  aim  at
-computing   speed  regardless   to  be   energy  efficient.    In  our
-experiments,  \monch  showed  the  best  energy-to-solution,
+(left  Y-axis)  metrics  on  both  platforms.  As can be observed, Intel Sandy Bridge EP processors of \pilat
+outperform the Intel Ivy  Bridge EP processors of \monch, being roughly 130\,\%  faster. 
+% The reason
+%for that is two-fold: \emph{i}) they have higher clock frequency than Ivy
+%Bridge  (2.6\,GHz  against  2.2\,GHz),   and  \emph{ii}) % they  aim  at
+% computing   speed  regardless   to  be   energy  efficient.
+In  our experiments,  \monch  showed  the  best  energy-to-solution,
 reducing the energy consumption of \pilat by approximately 7\,\%.
 
-Averaged TTS and ETS results  over 10 repetitions of 24-hour simulation on
+Averaged TTS and ETS results  over 10 repetitions of a 1-day simulation on
 \tinto are shown in  Figure~\ref{fig:4}. The corresponding results for
@@
 the \emph{degraded} MPI policy is about  2\,\% less than that consumed by the
-\emph{aggressive} one. This is  due to  the reduction of CPU usage (of about 50\,\%) while using  the \emph{degraded} MPI policy, in
-contrast with the \emph{aggressive} policy that always yields to 100\,\% of CPU utilization while waiting for an incoming message. From the point of  view of the execution time, there is a
+\emph{aggressive} one. This is  due to  the reduction of the CPU utilization (of about 50\,\%) while using  the \emph{degraded} MPI policy, in
+contrast with the \emph{aggressive} one, that always yields to 100\,\% of CPU load while waiting for incoming messages. From the point of  view of the execution time there is a
 slight reduction  for the \emph{degraded} MPI  policy. One can  expect that the
@@
 
-\begin{figure}[htbf]
+\begin{figure}[ht]
   \centering
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-0.eps}
+  \includegraphics[width=1\columnwidth]{Figs/Time_E2S_COSMO-ART-0.eps}
   \caption{Time-to-solution and  energy-to-solution comparison between
-    \pilat and \monch for a 24-hour simulation.}
+    \pilat and \monch for a 1-day simulation.}
   \label{fig:3}
@@
 
-\begin{figure}[htbf]
+\begin{figure}[ht]
   \centering
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
+  \includegraphics[width=1\columnwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
   \vspace{-1cm}
   \caption{Mean time-to-solution and  energy-to-solution on \tinto for
-    a 24-hour simulation using the  \emph{aggressive} and \emph{degraded} MPI policies. The
+    a 1-day simulation using the  \emph{aggressive} and \emph{degraded} MPI policies. The
     95\,\%  confidence  intervals of  Student's  $t$ distribution  (10
@@
   \hspace{0.8cm} \scalebox{0.62}{\input{Figs/24hour.pstex_t}}
-  \caption{24-hour simulation power  trace  using  the \emph{aggressive}  and
+  \caption{1-day simulation power  trace  using  the \emph{aggressive}  and
     \emph{degraded} MPI policies on \tinto.}
@@
 synchronization  phase are  performed. From  the performance  point of
-view, one can notice a reduction of 2\,\% in the TTS that \emph{degraded}  MPI
-mode leads over the \emph{aggressive} MPI  policy. On the other hand an averaged
-reduction of power consumption of  2\,\% is attained with the \emph{degraded} 
-MPI policy  only in  routines that actually  perform synchronizations,
-group  communications  and  MPI-receive/-wait primitives.   Thus,  the
+view, one can notice a reduction of 2\,\% in the TTS that the \emph{degraded}  MPI
+policy leads over the \emph{aggressive} one. On the other hand, a
+reduction of power consumption of  2\,\% in average is attained with the \emph{degraded} 
+policy only for routines that actually  perform synchronizations,
+group  communications  and MPI receive and wait primitives.   Thus,  the
 reduction of the power-energy  pair directly depends on the percentage
-of time in which blocking MPI  operations are executed. For the 1 hour
+of time in which blocking MPI  operations are executed. For the 1-hour
 simulation traces, an average of 50\,\% of the time the processes were
@@
 the  power savings,  i.e., $50\,\%  \times  5\,\% =  2.5\%$, which  is
-consistent  power  reduction  obtained  for  the  full-day  simulation
+consistent  power  reduction  obtained  for  the  1-day  simulation
 (around  2\,\%). However,  we  expect higher  power/energy savings  if
-upcoming versions of OpenMPI support full blocking policies as well as if 
+upcoming versions of OpenMPI support full blocking policies, as well as if 
 future architectures tend proportional computing~\citep{Barroso-2007}.

### 2014-05-30 paper_enahpc14/Sec1:Related.tex
cleans up latest version; improves format
updates a few things according to the reviewers' suggestions

@@
 encourages the design,  procurement and management of energy-efficient
-infrastructures to contain performance in an affordable and competitive
-power  envelope.   However,  the state-of-the-art  research  assessing
-performance  and energy  efficiency  of applications  is still scarce.
+infrastructures   to  contain   performance  in   an   affordable  and
+competitive  power envelope.   However, the  state-of-the-art research
+assessing performance  and energy efficiency of  applications is still
+scarce.
 
@@
 
-\emph{Ou  Pang et  al.}   \cite{Ou-2012} compare  ARM  and Intel  x86
-workstations  clusters  and   conclude  that  ARM-based  clusters  are
-advantageous  with   lightweight  applications  in   terms  of  energy
-efficiency.
+% \emph{Ou  Pang et  al.}   \cite{Ou-2012} compare  ARM  and Intel  x86
+% workstations  clusters  and   conclude  that  ARM-based  clusters  are
+% advantageous  with   lightweight  applications  in   terms  of  energy
+% efficiency.
 
-\emph{G\"oddeke et al.}  \cite{Goddeke-2013} evaluate weak and strong
-scalability of PDE solvers on a cluster of 96 ARM dual-core processors
-and demonstrate  that the ARM-based  cluster can be more  efficient in
-terms of energy-to-solution compared to x86-based cluster.
+% \emph{G\"oddeke et al.}  \cite{Goddeke-2013} evaluate weak and strong
+% scalability of PDE solvers on a cluster of 96 ARM dual-core processors
+% and demonstrate  that the ARM-based  cluster can be more  efficient in
+% terms of energy-to-solution compared to x86-based cluster.
 
@@
 
-% \emph{Cumming  et  al.}   \cite{Cumming-2014}  present a  simple  and
-% practical  methodology  looking at  energy  minimization  that can  be
-% applied to various applications.
+\emph{Cumming  et  al.}  \cite{Cumming-2014}  present  a simple  and
+practical  methodology  looking at  energy  minimization  that can  be
+applied to various applications.
+
+\emph{Diouri et al.} \cite{suscom} present a deep 
+study of different internal and external power monitoring devices that
+are widely used nowadays.

### 2014-05-30 paper_enahpc14/Sec3:Environment-setup.tex

@@
 allowing a  fair comparison between the baseline  and future milestone
-versions of \cosmoart.
+versions of \cosmoart:
 
@@
   DDR3  1333\,MHz and  connected via  InfiniBand QDR  with  a Mellanox
-  MTS\-3600 switch. Power saving mechanisms are enabled on the nodes are the C-states C1, C3 and C6, 
-  but no P-states (except P0) are used. TurboBoost feature is enabled for our experiments.
+  MTS\-3600 switch. 
 \end{itemize}
+
+Power saving mechanisms enabled on all the nodes' platforms are the C-states (C1, C3 and C6) 
+and P-states, but only P0 or the maximum frequency is used. TurboBoost feature is also enabled for our experiments.
 
@@
 HTTP  and  TELNET  through  the  built-in Fast  Ethernet  port.   Time
-synchronization is guaranteed by  using NTP servers.  Measured data is
+synchronization is guaranteed by  using NTP servers. Averaged measured data is
 accessed through the open  source Cacti software including the E3METER
@@
 generated.  A  merged trace  file  is  finally  loaded in  Paraver  to
-visualize  and correlate  the  performance trace  and  with the  power
-profile  generated by \cosmoart.  This procedure  allows us  to easily
+visualize  and correlate  the  tasks trace with the  power
+profile  produced by \cosmoart.  This procedure  allows us  to easily
 detect power bottlenecks, basically constituted by inefficient waiting

### 2014-05-30 paper_enahpc14/paper.tex

@@
   extension of  the operational weather forecast model of  the German
-  weather  service   (DWD),  developed  for  the   evaluation  of  the
+  wea\-ther  service   (DWD),  developed  for  the   evaluation  of  the
   interactions of reactive gases  and aerosol particles with the state
@@
   motivation  is to  improve  corresponding code  sections to  sustain
-  performance  while minimising energy-to-solution.   This preliminary
+  performance  while minimizing energy-to-solution.   This preliminary
   work  sets the  basis for  subsequent studies  to  tackle challenges
@@
 Programme Future and Emerging Technologies (FET) Proactive Initiative:
-Minimising Energy  Consumption of Computing to the  Limit (MINECC). We
+Minimizing Energy  Consumption of Computing to the  Limit (MINECC). We
 also gratefully acknowledge the High Performance and High Productivity

### 2014-05-31 trunk/paper_enahpc14/Sec3:Environment-setup.tex
Submitted version

@@
   DDR3  1333\,MHz and  connected via  InfiniBand QDR  with  a Mellanox
-  MTS\-3600 switch. Power saving mechanisms enabled on the nodes are the C-states C1, C3 and C6, 
-  but no P-states (except P0) are used. TurboBoost feature is enabled for our experiments.
+  MTS\-3600 switch. 
 \end{itemize}
+
+Power saving mechanisms enabled on all the nodes' platforms were the C-states (C1, C3 and C6) 
+and P-states, but only P0 or the maximum frequency was used. TurboBoost feature was also enabled for our experiments.
 

### 2014-05-31 paper_enahpc14/Conclusion.tex
Submitted version

@@
 benchmarks illustrate  that the  best time-to-solution does  not imply
-the best energy-to-solution: an Intel Sandy Bridge (2.6 GHz) system has
-lower time-to-solution but higher energy-to-solution than an Intel Ivy
-Bridge (2.2 GHz) system, although  the two metrics are indeed strongly
+the best energy-to-solution: a system based on Intel Xeon Sandy Bridge (2.6 GHz) has
+lower time-to-solution but higher energy-to-solution than on a platform based on Intel Xeon Ivy
+Bridge (2.2 GHz), although  the two metrics are indeed strongly
 correlated.   On the  other  hand, the  usage  of energy-friendly  MPI
-waiting techniques analyzed  with our power-performance framework over
-a series of  runs of \cosmoart on \tinto  and power traces demonstrate
-that   it  is   possible  to   reduce  both   power   dissipation  and
-energy-to-solution   while  maintaining   (or  even   decreasing)  the
-time-to-solution. The resulting profiles indicate that simple changes,
-such as making use of the  \emph{degraded} MPI policy rather than the \emph{aggressive}
-is  possible to  slightly  reduce both  power  dissipation and  energy
-consumption (by 2\,\%).
+waiting techniques analyzed  with our power-per\-for\-man\-ce framework demonstrate
+that   it  is   possible  to   reduce the energy-to-solution   while  maintaining   (or  even   decreasing)  the
+time-to-so\-lu\-tion. The resulting power profiles of \cosmoart indicate that simple changes,
+such as making use of the  \emph{degraded} MPI policy rather than the \emph{aggressive} (by default) is  possible to  slightly  reduce both  power and  energy consumption (by 2\,\%).
 
 This  reproducible  benchmark provides  a  baseline  for ongoing  work
-package within  the EU-funded Exa2Green to  minimise power consumption
+package within  the EU-funded Exa2Green to  minimize power consumption
 of \cosmoart. Profiling  has given us insight into  the most expensive

### 2014-05-31 paper_enahpc14/Sec4:Experimental-results.tex
Submitted version

@@
 contribution of the MPI library  to the energy consumption and how the
-use  of  the aggressive and degraded message-passing progression engine 
-policies can potentially  render energy  savings.
-
-Specifically,   the  OpenMPI 1.6.5  library,  features  two  operation 
+use  of different message-passing progression engine 
+modes can potentially  render energy  savings. Specifically,   the  OpenMPI 1.6.5  library,  features  two  operation 
 policies, \emph{aggressive} and \emph{degraded}, which can be  selected 
-before \cosmoart is launched. In the \emph{aggressive} (default) policy,  
-normally used in exactly- or under-subscribed mode  (i.e., the  number 
+before \cosmoart is launched. In the \emph{aggressive} (by default) policy,  
+normally used in exactly- and under-subscribed modes  (i.e., the  number 
 of running processes is equal to or less than the number of  available 
@@
 therefore never make any  progress).  This  usually  yields  the  best 
-performance but high CPU utilization, possibly at the cost of a higher 
+performance but a higher CPU utilization, possibly at the cost of a higher 
 energy usage.  On the other hand, the \emph{degraded} policy, suitable 
@@
 enabled, we noticed that a blocked MPI process waiting for an incoming 
-message has CPU utilization of 100\,\%,  however  when  the \emph{degraded} 
-mode  is  set,  the  CPU  utilization is only $50\,\%$, approximately. 
+message utilized 100\,\% of CPU,  however  when  the \emph{degraded} 
+mode  was  set,  the  CPU  utilization was only $50\,\%$, approximately. 
 Although neither of both policies avoid busy-waiting loops in blocking 
@@
 In  Figure~\ref{fig:3}, we  compare both  TTS (right  Y-axis)  and ETS
-(left  Y-axis)  metrics  on  both  platforms.  As  expected, Intel Sandy Bridge EP processors of \pilat
-outperform the Intel Ivy  Bridge EP processors of \monch, being roughly 130\,\%  faster.  The reason
-for that is two-fold: \emph{i}) they have higher clock frequency than Ivy
-Bridge  (2.6\,GHz  against  2.2\,GHz),   and  \emph{ii})  they  aim  at
-computing   speed  regardless   to  be   energy  efficient.    In  our
-experiments,  \monch  showed  the  best  energy-to-solution,
+(left  Y-axis)  metrics  on  both  platforms.  As can be observed, Intel Sandy Bridge EP processors of \pilat
+outperform the Intel Ivy  Bridge EP processors of \monch, being roughly 130\,\%  faster. 
+% The reason
+%for that is two-fold: \emph{i}) they have higher clock frequency than Ivy
+%Bridge  (2.6\,GHz  against  2.2\,GHz),   and  \emph{ii}) % they  aim  at
+% computing   speed  regardless   to  be   energy  efficient.
+In  our experiments,  \monch  showed  the  best  energy-to-solution,
 reducing the energy consumption of \pilat by approximately 7\,\%.
 
-Averaged TTS and ETS results  over 10 repetitions of 24h simulation on
+Averaged TTS and ETS results  over 10 repetitions of a 1-day simulation on
 \tinto are shown in  Figure~\ref{fig:4}. The corresponding results for
@@
 extensive computing,  as e.g.  \cosmoart, may show  better performance
-with the \emph{degraded} policy. This effect works in  favor with our results
+with the \emph{degraded} policy. This effect works in  favor with our results,
 since the energy consumed is also reduced (2\,\%) just by changing the
@@
 
-\begin{figure}[htbf]
+\begin{figure}[ht]
   \centering
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-0.eps}
+  \includegraphics[width=1\columnwidth]{Figs/Time_E2S_COSMO-ART-0.eps}
   \caption{Time-to-solution and  energy-to-solution comparison between
-    \pilat and \monch for a 24h simulation.}
+    \pilat and \monch for a 1-day simulation.}
   \label{fig:3}
@@
 
-\begin{figure}[htbf]
+\begin{figure}[ht]
   \centering
-  \includegraphics[width=0.5\textwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
+  \includegraphics[width=1\columnwidth]{Figs/Time_E2S_COSMO-ART-1.eps}
   \vspace{-1cm}
   \caption{Mean time-to-solution and  energy-to-solution on \tinto for
-    a 24h simulation using the  \emph{aggressive} and \emph{degraded} MPI policies. The
+    a 1-day simulation using the  \emph{aggressive} and \emph{degraded} MPI policies. The
     95\,\%  confidence  intervals of  Student's  $t$ distribution  (10
@@
   \hspace{0.8cm} \scalebox{0.62}{\input{Figs/24hour.pstex_t}}
-  \caption{24  hours  simulation power  trace  using  the \emph{aggressive}  and
+  \caption{1-day simulation power  trace  using  the \emph{aggressive}  and
     \emph{degraded} MPI policies on \tinto.}
@@
 incoming messages  or synchronizations, thus  yielding the CPU  to a lower
-usage. Our final observation is the reduction of the TTS by 2\,\% with
-respect to the  \emph{degraded} MPI policy.
+usage. Our final observation is the reduction of 2\,\% in the TTS when the \emph{degraded} policy is used.
 
@@
 synchronization  phase are  performed. From  the performance  point of
-view, one can notice a reduction of 2\,\% in the TTS that \emph{degraded}  MPI
-mode leads over the \emph{aggressive} MPI  policy. On the other hand an averaged
-reduction of power consumption of  2\,\% is attained with the \emph{degraded} 
-MPI policy  only in  routines that actually  perform synchronizations,
-group  communications  and  MPI-receive/-wait primitives.   Thus,  the
+view, one can notice a reduction of 2\,\% in the TTS that the \emph{degraded}  MPI
+policy leads over the \emph{aggressive} one. On the other hand, a
+reduction of power consumption of  2\,\% in average is attained with the \emph{degraded} 
+policy, only for routines that actually  perform synchronizations,
+group  communications  and MPI receive and wait primitives.   Thus,  the
 reduction of the power-energy  pair directly depends on the percentage
-of time in which blocking MPI  operations are executed. For the 1 hour
+of time in which blocking MPI  operations are executed. For the 1-hour
 simulation traces, an average of 50\,\% of the time the processes were

### 2014-06-16 paper_enahpc14/Sec2:COSMO-ART.tex
updates paper according to page proofs

@@
 aqueous-phase together with the transfer reactions is solved using the
-Kinetic PreProcessor (KPP, \cite{Damian-2002}).
-
-Further model description details  can be found in \cite{Bangert-2012,
-  Knote-2011, Knote-2013}.
+Kinetic   PreProcessor   (KPP,   \cite{Damian-2002}).  Further   model
+description  details can be  found in  \cite{Bangert-2012, Knote-2011,
+  Knote-2013}.
 
@@
 initialization needs  a period of adjustment to  reach its equilibrium
-state  and minimise  the effect  of initial  conditions for  gases and
+state  and minimize  the effect  of initial  conditions for  gases and
 aerosols on the model predictions.
@@
 The meteorological initial and boundary conditions are obtained by the
-the ECMWF  global spectral model IFS  with an update  frequency of 3h.
+ECMWF  global spectral  model  IFS  with an  update  frequency of  3h.
 Boundary data  for gas-phase species are taken  from IFS-MOZART output
@@
 ordinary  differential equations  \cite{Damian-2002}.   Concerning the
-modelling  of  wet  deposition  in  aerosols, the  baseline  has  only
+modeling  of  wet  deposition  in  aerosols, the  baseline  has  only
 indirect  cloud feedbacks  but  does not  include in-cloud  scavenging
 (rainout) and below-cloud  scavenging (washout) yet.  Amongst physical
-parameterisations,   precipitation  formation   is   performed  by   a
+parameterizations,   precipitation  formation   is   performed  by   a
 two-moment cloud microphysics.

### 2014-06-16 paper_enahpc14/Sec4:Experimental-results.tex
updates paper according to page proofs

@@
 contribution of the MPI library  to the energy consumption and how the
-use  of different message-passing progression engine 
-modes can potentially  render energy  savings. Specifically,   the  OpenMPI 1.6.5  library,  features  two  operation 
-policies, \emph{aggressive} and \emph{degraded}, which can be  selected 
-before \cosmoart is launched. In the \emph{aggressive} (by default) policy,  
-normally used in exactly- and under-subscribed modes  (i.e., the  number 
-of running processes is equal to or less than the number of  available 
-cores), MPI processes will never voluntarily give up the processor  to 
-other processes. With some network transports, OpenMPI processes  will 
-spin  in  tight  loops  attempting  to  make message passing progress, 
+use  of   different  message-passing  progression   engine  modes  can
+potentially  render energy  savings. Specifically,  the  OpenMPI 1.6.5
+library,  features  two   operation  policies,  \emph{aggressive}  and
+\emph{degraded},   which   can  be   selected   before  \cosmoart   is
+launched. In the \emph{aggressive}  (by default) policy, normally used
+in exactly-  and under-subscribed modes  (i.e., the number  of running
+processes is equal to or less than the number of available cores), MPI
+processes  will  never voluntarily  give  up  the  processor to  other
+processes. With  some network transports, OpenMPI  processes will spin
+in  tight   loops  attempting   to  make  message   passing  progress,
 effectively causing  other processes  to not get  any CPU  cycles (and
 therefore  never make  any progress).   This usually  yields  the best
-performance but a higher CPU utilization, possibly at the cost of a higher 
-energy usage.  On the other hand, the \emph{degraded} policy, suitable 
-for the oversubscribed mode (i.e., more  processes  are  running  than 
-total cores  available),  MPI  processes  will  frequently  yield  the 
-processor to its peers, thereby allowing all processes to make progress.
+performance but  a higher CPU utilization,  possibly at the  cost of a
+higher energy  usage.  On the other hand,  the \emph{degraded} policy,
+suitable for the oversubscribed mode (i.e., more processes are running
+than total  cores available), MPI processes will  frequently yield the
+processor  to  its  peers,  thereby  allowing all  processes  to  make
+progress.
 
@@
 send or receive) and  repeatedly calling to \texttt{sched\_yield()} to
-be picked again by the Linux scheduler. Although one can expect no benefits 
-from   using   the  \emph{degraded}   policy   compared   with     the 
-\emph{aggressive} one, notable differences can be observed in the CPU utilization. 
-In a separated  test  with  the  \emph{aggressive}  policy 
+be picked  again by  the Linux scheduler.  Although one can  expect no
+benefits  from  using the  \emph{degraded}  policy  compared with  the
+\emph{aggressive} one, notable differences  can be observed in the CPU
+utilization.  In  a separated  test with the  \emph{aggressive} policy
 enabled, we noticed that a blocked MPI process waiting for an incoming
-message utilized 100\,\% of CPU,  however  when  the \emph{degraded} 
-mode  was  set,  the  CPU  utilization was only $50\,\%$, approximately. 
+message utilized 100\,\% of CPU, however when the \emph{degraded} mode
+was  set,  the  CPU  utilization  was  only  $50\,\%$,  approximately.
 Although neither of both policies avoid busy-waiting loops in blocking
@@
 only twice due  to usage restrictions.  As time  resolution was set to
-one averaged update every  5 minutes  for power  sampling, the  averaged power
-consumption was computed by considering  6 values for each single run.
-On \pilat, the 1-day simulation was  issued four times and a 2-day run
-only once.  Similarly, the  averaged power consumption was computed by
-considering 4  values for each single  1-day run and 9  values for the
-2-days run. Corresponding results are gathered in Table~\ref{tab:1}.
+one averaged update  every 5 minutes for power  sampling, the averaged
+power consumption was computed by considering 6 values for each single
+run.   On \pilat, the  1-day simulation  was issued  four times  and a
+2-day run  only once.  Similarly,  the averaged power  consumption was
+computed  by considering  4 values  for each  single 1-day  run  and 9
+values  for the  2-days  run. Corresponding  results  are gathered  in
+Table~\ref{tab:1}.
 
@@
 In  Figure~\ref{fig:3}, we  compare both  TTS (right  Y-axis)  and ETS
-(left  Y-axis)  metrics  on  both  platforms.  As can be observed, Intel Sandy Bridge EP processors of \pilat
-outperform the Intel Ivy  Bridge EP processors of \monch, being roughly 130\,\%  faster. 
+(left Y-axis)  metrics on both  platforms.  As can be  observed, Intel
+Sandy Bridge EP  processors of \pilat outperform the  Intel Ivy Bridge
+EP processors of \monch, being roughly 130\,\% faster.
 % The reason
@@
 
-Averaged TTS and ETS results  over 10 repetitions of a 1-day simulation on
-\tinto are shown in  Figure~\ref{fig:4}. The corresponding results for
-the  averaged total  power are  also shown  in  Table~\ref{tab:1}.  As
+Averaged TTS and ETS results over 10 repetitions of a 1-day simulation
+on \tinto  are shown in Figure~\ref{fig:4}.  The corresponding results
+for the averaged total power  are also shown in Table~\ref{tab:1}.  As
 expected, the average total power consumption by \cosmoart while using
-the \emph{degraded} MPI policy is about  2\,\% less than that consumed by the
-\emph{aggressive} one. This is  due to  the reduction of the CPU utilization (of about 50\,\%) while using  the \emph{degraded} MPI policy, in
-contrast with the \emph{aggressive} one, that always yields to 100\,\% of CPU load while waiting for incoming messages. From the point of  view of the execution time there is a
-slight reduction  for the \emph{degraded} MPI  policy. One can  expect that the
-costs of the context switches produced in the \emph{degraded} MPI policy can lead to
-an  increase  of  the   execution  time,  however,  applications  with
+the \emph{degraded} MPI policy is  about 2\,\% less than that consumed
+by the \emph{aggressive} one. This is  due to the reduction of the CPU
+utilization  (of about  50\,\%)  while using  the \emph{degraded}  MPI
+policy, in contrast with the \emph{aggressive} one, that always yields
+to 100\,\% of  CPU load while waiting for  incoming messages. From the
+point of  view of the execution  time there is a  slight reduction for
+the \emph{degraded} MPI  policy. One can expect that  the costs of the
+context switches  produced in the \emph{degraded} MPI  policy can lead
+to  an increase  of  the execution  time,  however, applications  with
 extensive computing,  as e.g.  \cosmoart, may  show better performance
-with the \emph{degraded} policy. This effect works in  favor with our results,
-since the energy consumed is also reduced (2\,\%) just by changing the
-behavior of the OpenMPI library  to the \emph{degraded} policy, i.e, with the
-\texttt{--mca    mpi\_yield\_when\_idle 1}   parameter   set    in   the
-\texttt{mpirun} command.
+with the \emph{degraded}  policy. This effect works in  favor with our
+results, since  the energy  consumed is also  reduced (2\,\%)  just by
+changing the  behavior of the  OpenMPI library to  the \emph{degraded}
+policy,   i.e,  with   the  \texttt{--mca   mpi\_yield\_when\_idle  1}
+parameter set in the \texttt{mpirun} command.
 
@@
   \caption{Mean time-to-solution and  energy-to-solution on \tinto for
-    a 1-day simulation using the  \emph{aggressive} and \emph{degraded} MPI policies. The
-    95\,\%  confidence  intervals of  Student's  $t$ distribution  (10
-    samples) clearly  indicates an improvement for the \emph{degraded} policy in
-    both TTS and ETS.}
+    a 1-day simulation using the \emph{aggressive} and \emph{degraded}
+    MPI  policies. The  95\,\% confidence  intervals of  Student's $t$
+    distribution (10 samples) clearly indicates an improvement for the
+    \emph{degraded} policy in both TTS and ETS.}
   \label{fig:4}
@@
   \hspace{0.8cm} \scalebox{0.62}{\input{Figs/24hour.pstex_t}}
-  \caption{1-day simulation power  trace  using  the \emph{aggressive}  and
-    \emph{degraded} MPI policies on \tinto.}
+  \caption{1-day  simulation power  trace using  the \emph{aggressive}
+    and \emph{degraded} MPI policies on \tinto.}
   \label{fig:5}
@@
 power  traces  of  two  1-hour  time frames  during  the  day:  midday
-(11h--12h) and midnight (23h--24h). All the experiments on \tinto were performed
-using 192 cores with 12 MPI processes per node, i.e., using an exactly-subscribed MPI mode.
+(11h--12h) and midnight (23h--24h). All the experiments on \tinto were
+performed using 192 cores with  12 MPI processes per node, i.e., using
+an exactly-subscribed MPI mode.
 
@@
 power-measurement    framework   over    a    1-day   simulation    of
-\cosmoart. These profiles, produced with both \emph{aggressive} and \emph{degraded} MPI
-policies,  are depicted in  Figure~\ref{fig:5}. The  first observation
-for the \emph{aggressive} MPI policy is that the first and last hours of the day
-consume slightly less  power than the midday hours.  We relate these
-variations  due   to  the  different   chemical  reactions,  aerosols,
-radiations  and clouds  formation processes,  computed by  ART, taking
-place during  daylight hours  but not during  night. In this  case the
-peak power consumption is about  3821\,W. Looking at the power profile
-using the \emph{degraded}  MPI policy, we observe that  the 1-hour pattern is
-repeated along  the 24 hours, favoring  the reduction of  power during
-night  hours in  a  more noticeable  way  than that  with the   \emph{aggressive}
-policy.   The peak  power consumption  during day  hours  while the
- \emph{degraded}  MPI policy  is leveraged  is  about 3768\,W,  which means  a
-reduction of around  50\,W with respect to the  \emph{aggressive} mode. Also, we
+\cosmoart.  These profiles, produced  with both  \emph{aggressive} and
+\emph{degraded} MPI policies,  are depicted in Figure~\ref{fig:5}. The
+first  observation for the  \emph{aggressive} MPI  policy is  that the
+first and last  hours of the day consume slightly  less power than the
+midday  hours.   We  relate  these  variations due  to  the  different
+chemical   reactions,  aerosols,   radiations  and   clouds  formation
+processes, computed by ART, taking place during daylight hours but not
+during  night.  In this  case  the  peak  power consumption  is  about
+3821\,W. Looking  at the power  profile using the  \emph{degraded} MPI
+policy, we  observe that the 1-hour  pattern is repeated  along the 24
+hours, favoring  the reduction of power  during night hours  in a more
+noticeable way than that  with the \emph{aggressive} policy.  The peak
+power  consumption  during day  hours  while  the \emph{degraded}  MPI
+policy  is leveraged  is about  3768\,W,  which means  a reduction  of
+around  50\,W with  respect to  the \emph{aggressive}  mode.  Also, we
 notice systematic  power drops for each simulated  hour.  We associate
-them  to periods  in  which  the processes  were waiting  for
-incoming messages  or synchronizations, thus  yielding the CPU  to a lower
-usage. Our final observation is the reduction of 2\,\% in the TTS when the \emph{degraded} policy is used.
+them  to periods  in which  the  processes were  waiting for  incoming
+messages  or  synchronizations,  thus  yielding  the CPU  to  a  lower
+usage. Our final observation is the reduction of 2\,\% in the TTS when
+the \emph{degraded} policy is used.
 
 To  gain  more  insights   about  the  power-performance  behavior  of
-\cosmoart,  we  obtained power-per\-for\-man\-ce  traces  to allow  us
+\cosmoart, we  obtained power-per\-for\-man\-ce traces to  allow us to
 correlate  the type  of  tasks  being executed  with  the total  power
@@
 be easily handled using our  machines and memory available (each trace
-file was about 45\,GB). Thus, we shrink the simulations
-to 1 hour during midday (from 11h  to 12h) and during
-midnight  (from 23h  to  24h), in  which  \cosmoart perform  different
-internal computations.  At  the same time we leverage  the two already
-stated MPI policies: \emph{aggressive} and \emph{degraded}.
+file was  about 45\,GB).   Thus, we shrink  the simulations to  1 hour
+during midday (from 11h to 12h) and during midnight (from 23h to 24h),
+in which  \cosmoart perform  different internal computations.   At the
+same  time   we  leverage  the   two  already  stated   MPI  policies:
+\emph{aggressive} and \emph{degraded}.
 
@@
 combined  in  between   with  group  communications,  blocking  sends,
-immediate  receives and  MPI-wait primitives.  Finally,
-group  communications interlaced  with computation  and  a significant
+immediate   receives   and   MPI-wait  primitives.    Finally,   group
+communications   interlaced  with   computation   and  a   significant
 synchronization  phase are  performed. From  the performance  point of
-view, one can notice a reduction of 2\,\% in the TTS that the \emph{degraded}  MPI
-policy leads over the \emph{aggressive} one. On the other hand, a
-reduction of power consumption of  2\,\% in average is attained with the \emph{degraded} 
-policy, only for routines that actually  perform synchronizations,
-group  communications  and MPI receive and wait primitives.   Thus,  the
-reduction of the power-energy  pair directly depends on the percentage
-of time in which blocking MPI  operations are executed. For the 1-hour
-simulation traces, an average of 50\,\% of the time the processes were
-waiting for incoming messages.  Taking  into account that  the total power  reduction during
+view,  one  can notice  a  reduction  of 2\,\%  in  the  TTS that  the
+\emph{degraded} MPI  policy leads  over the \emph{aggressive}  one. On
+the other hand,  a reduction of power consumption  of 2\,\% in average
+is attained  with the \emph{degraded}  policy, only for  routines that
+actually  perform  synchronizations,   group  communications  and  MPI
+receive and wait primitives.   Thus, the reduction of the power-energy
+pair directly depends on the  percentage of time in which blocking MPI
+operations are executed. For  the 1-hour simulation traces, an average
+of  50\,\%  of  the  time  the processes  were  waiting  for  incoming
+messages.  Taking  into account that the total  power reduction during
 blocking phases  is about 5\,\%,  one can compute an  approximation of
 the  power savings,  i.e., $50\,\%  \times  5\,\% =  2.5\%$, which  is
-consistent  power  reduction  obtained  for  the  1-day  simulation
-(around  2\,\%). However,  we  expect higher  power/energy savings  if
-upcoming versions of OpenMPI support full blocking policies, as well as if 
+consistent power  reduction obtained for the  1-day simulation (around
+2\,\%).  However, we  expect higher  power/energy savings  if upcoming
+versions  of OpenMPI  support full  blocking policies,  as well  as if
 future architectures tend proportional computing~\citep{Barroso-2007}.

### 2014-06-16 trunk/paper_enahpc14/Conclusion.tex
updates paper according to page proofs

@@
 benchmarks illustrate  that the  best time-to-solution does  not imply
-the best energy-to-solution: an Intel Sandy Bridge (2.6 GHz) system has
-lower time-to-solution but higher energy-to-solution than an Intel Ivy
-Bridge (2.2 GHz) system, although  the two metrics are indeed strongly
-correlated.   On the  other  hand, the  usage  of energy-friendly  MPI
-waiting techniques analyzed  with our power-per\-for\-man\-ce framework over
-a series of  runs of \cosmoart on \tinto  and power traces demonstrate
-that   it  is   possible  to   reduce  both   power   dissipation  and
-energy-to-solution   while  maintaining   (or  even   decreasing)  the
-time-to-solution. The resulting profiles indicate that simple changes,
-such as making use of the  \emph{degraded} MPI policy rather than the \emph{aggressive}
-is  possible to  slightly  reduce both  power  dissipation and  energy
+the best energy-to-solution: a system based on Intel Xeon Sandy Bridge
+(2.6  GHz) has  lower time-to-solution  but  higher energy-to-solution
+than on a platform based on  Intel Xeon Ivy Bridge (2.2 GHz), although
+the two  metrics are indeed  strongly correlated.  On the  other hand,
+the usage of energy-friendly  MPI waiting techniques analyzed with our
+power-per\-for\-man\-ce framework  demonstrate that it  is possible to
+reduce the  energy-to-solution while maintaining  (or even decreasing)
+the  time-to-so\-lu\-tion. The resulting  power profiles  of \cosmoart
+indicate   that    simple   changes,    such   as   making    use   of
+the \emph{degraded}  MPI policy rather than  the \emph{aggressive} (by
+default)  is  possible  to  slightly  reduce  both  power  and  energy
 consumption (by 2\,\%).

### 2014-06-16 trunk/paper_enahpc14/Sec3:Environment-setup.tex
updates paper according to page proofs

@@
 32\,GB  of DDR3 1600\,MHz  RAM and  connected via  InfiniBand Mellanox
-SX6036 and FDR switches.  For our experiments, only a subset of 52 was used. %
+SX6036 and FDR switches.  For our experiments, a full rack  of 52 standard compute nodes was used. %
 % only a full rack constituted exclusively of 52 nodes was used.
@@
   via  InfiniBand   Mellanox  SX6036   and  FDR  switches.    For  our
-  experiments, a subset constituted of 42 compute  nodes was used.
+  experiments, a full rack constituted of 42 compute  nodes was used.
   % \pilat   is   based   on   Intel's   second-generation   processors,
@@
   at UJI.  For  our experiments only a subset  of 16 homogeneous nodes
-  was  considered. Each  node  is  comprised of  two  Intel Xeon Westmere EP E5645
-  hexa-core processors  running at  2.4\,GHz, equipped with  24\,GB of
-  DDR3  1333\,MHz and  connected via  InfiniBand QDR  with  a Mellanox
+  was considered. Each node is comprised of two Intel Xeon Westmere EP
+  E5645 hexa-core processors running at 2.4\,GHz, equipped with 24\,GB
+  of DDR3 1333\,MHz  and connected via InfiniBand QDR  with a Mellanox
   MTS\-3600 switch.
@@
 
-Power saving mechanisms enabled on all the nodes' platforms were the C-states (C1, C3 and C6) 
-and P-states, but only P0 or the maximum frequency was used. TurboBoost feature was also enabled for our experiments.
+Power saving mechanisms  enabled on all the nodes'  platforms were the
+C-states (C1,  C3 and  C6) and  P-states, but only  P0 or  the maximum
+frequency  was  used. TurboBoost  feature  was  also  enabled for  our
+experiments.
 
@@
 the  IT infrastructure  as well  as constantly  analyze  line voltage,
-current, power-factor, frequency  with $\pm1\,\%$ accuracy.  Using reliable
-narrowband power-line communication (PLC) technology, all metering and
-power  quality data  from  each  IPS are  centrally  collected by  the
-E3METER Data Concentrator, via the existing power cables thus avoiding
-the need  for extra  cabling.  This data  is made available  via SNMP,
-HTTP  and  TELNET  through  the  built-in Fast  Ethernet  port.   Time
-synchronization is guaranteed by  using NTP servers. Averaged measured data is
-accessed through the open  source Cacti software including the E3METER
-Cacti Plugin to  scan the entire PLC network  and monitor in real-time
-the power  usage of  individual rack, recorded  in 5  minutes interval
-periods.
+current,  power-factor,  frequency  with $\pm1\,\%$  accuracy.   Using
+reliable  narrowband power-line  communication  (PLC) technology,  all
+metering and power quality data  from each IPS are centrally collected
+by the E3METER  Data Concentrator, via the existing  power cables thus
+avoiding the need for extra  cabling.  This data is made available via
+SNMP, HTTP and  TELNET through the built-in Fast  Ethernet port.  Time
+synchronization is guaranteed by  using NTP servers. Averaged measured
+data is accessed through the  open source Cacti software including the
+E3METER Cacti  Plugin to  scan the entire  PLC network and  monitor in
+real-time the  power usage of  individual rack, recorded in  5 minutes
+interval periods.
 
@@
 (PDUs) that  account for  the consumed power  with a sampling  rate of
-1\,Hz and $\pm3\,\%$ accuracy. The power data captured by  these PDUs is collected by a \pmlib
-tracing server running  on a separate machine. The  client, running on
-the target nodes, sends start/stop  primitives in order to collect the
-power  data during  the \cosmoart  run. Once  the run  is  finished, a
-power-performance  trace file  containing the  power data  is received
-from  the  tracing  server  and the  instrumentation  post-process  is
-generated.  A  merged trace  file  is  finally  loaded in  Paraver  to
-visualize  and correlate  the  tasks trace with the  power
-profile  produced by \cosmoart.  This procedure  allows us  to easily
-detect power bottlenecks, basically constituted by inefficient waiting
-methods.
+1\,Hz and $\pm3\,\%$  accuracy. The power data captured  by these PDUs
+is  collected  by  a  \pmlib  tracing server  running  on  a  separate
+machine.  The client, running  on the  target nodes,  sends start/stop
+primitives in  order to  collect the power  data during  the \cosmoart
+run.  Once  the  run  is  finished,  a  power-performance  trace  file
+containing the power data is  received from the tracing server and the
+instrumentation  post-process is  generated.  A  merged trace  file is
+finally loaded in  Paraver to visualize and correlate  the tasks trace
+with the  power profile produced by \cosmoart.   This procedure allows
+us  to  easily  detect  power bottlenecks,  basically  constituted  by
+inefficient waiting methods.
 
@@
 \monch and \pilat, and OpenMPI 1.6.5  on \tinto. On the other hand, we
-used HDF5  1.8.12 and NetCDF  (4.3.1) libraries for the  management of
+used  HDF5 1.8.12  and NetCDF  4.3.1 libraries  for the  management of
 extremely large and complex data collections.  All computes nodes have
